### 2015-07-16 paper/paper.tex
initial commit

\documentclass[a4paper]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{minted}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

\usepackage{hyperref}
\usepackage{url}

% Additional packages 
\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{float}
\usepackage{enumitem}

\title{Notes on Design and Implementation of Anglican}

\begin{document}

\maketitle

\begin{abstract}
\end{abstract}

\section{Introduction}

\section{Design Outline}

\section{Macro-based Compilation}

\section{Inference Algorithms}

\section{Runtime Library}

\bibliographystyle{alpha}
\bibliography{refs}

\end{document}

### 2015-07-16 paper/paper.tex
introduction drafted

@@
 \usepackage{graphicx}
-\usepackage{minted}
+\usepackage{listings}
 
@@
 
+Anglican is a probabilistic programming language integrated with
+Clojure.  There are several ways to build a programming language
+on top or besides another language.  The easiest to grasp is an
+interpreter --- a program that reads a program, in its entirety
+or line-by-line, and executes it by applying operational
+semantics of a certain kind to the language. \textsc{Basic} is
+famous for line-by-line interpreted implementations.
+
+Another approach is to write a compiler, either to a virtual
+architecture, so called p-code or byte-code, or to real
+hardware. Here, the whole program is translated from the
+`higher-level' source language to a `lower-level' object
+language, which can be directly executed, either by hardware or
+by an interpreter --- but the latter interpreter can be made
+simpler and more efficient  than an interpreter for the source
+language.
+
+On top of these two approaches are methods in which a new
+language is implemented `inside' another language of the same
+level of abstraction. Different languages provide different
+means for this; Lisp is notorious for the macro facility
+that allows to extend the language almost without
+restriction --- by writing \textit{macros}, one adds new
+constructs to the existing language. There are several uses of
+macros --- one is to extend the language \textit{syntax}, for
+example, by adding new control structures; another is to keep
+the existing syntax but alter the operational semantic --- the
+way programs are executed and compute their outputs.
+
+Anglican is implemented in just this way --- a macro facility
+provided by Clojure, a Lisp dialect, is used both to add to
+Clojure constructs that delimit probabilistic, rather than
+deterministic, code, and to alter the operational semantics of
+Clojure constructs inside probabilistic code fragments. Anglican
+claims its right to count as a separate language because of the
+ubiquitous probabilistic execution semantics rather than because
+of a different syntax, which is actually an advantage rather
+than a drawback --- Clojure programmers only need to know how to
+specify the boundaries of Anglican programs, but can use
+familiar Clojure syntax to write probabilistic code.
+
+An implementation of Anglican must therefore address three issues:
+\begin{itemize}
+	\item the Clojure syntax to introduce probabilistic Anglican
+		code inside Clojure modules;
+	\item source-to-source transformation of Anglican programs
+		into Clojure, so that probabilistic execution becomes
+		possible;
+	\item algorithms which run Clojure code, obtained by
+		transforming Anglican programs, according to the
+		probabilistic operational semantics.
+\end{itemize}
+The following sections explain the way Anglican is implemented,
+from source-to-source transformation and syntactic wrappers to
+inference algorithms which accept Clojure functions built from
+anglican code as a parameter, and produce probabilistic outputs.
+
 \section{Design Outline}
@@
 
-\section{Runtime Library}
+\section{Definitions and Runtime Library}
 

### 2015-07-17 paper/paper.tex
in progress

@@
 
+\usepackage{enumitem}
 \usepackage{amsmath}
 \usepackage{amssymb}
+\usepackage{amsfonts}
 \usepackage{graphicx}
-\usepackage{listings}
 
-% For algorithms
 \usepackage{algorithm}
 \usepackage{algorithmic}
+\usepackage{listings}
 
@@
 \usepackage{url}
-
-% Additional packages 
-\usepackage{amsfonts}
-%\usepackage{amsthm}
-\usepackage{float}
-\usepackage{enumitem}
 
@@
 \begin{abstract}
+	In these notes, are explain the rationale behind the
+	macro-compiled implementation of Anglican, outline
+	design choices, and describe important implementation
+	aspects.
 \end{abstract}
@@
 deterministic, code, and to alter the operational semantics of
-Clojure constructs inside probabilistic code fragments. Anglican
+Clojure expressions inside probabilistic code fragments. Anglican
 claims its right to count as a separate language because of the
@@
 inference algorithms which accept Clojure functions built from
-anglican code as a parameter, and produce probabilistic outputs.
+Anglican code as a parameter, and produce probabilistic outputs.
 

### 2015-07-17 paper/paper.tex
design outline in progress

@@
 
+A probabilistic program, or query, mostly runs deterministic
+code, except for certain checkpoints, in which probabilities are
+involved, and normal, linear execution of the program is
+disrupted. In Anglican and similar languages there are two
+types of such checkpoints:
+\begin{itemize}
+	\item drawing a value from a random source (\texttt{sample});
+	\item conditioning the posterior distribution by
+		conditioning a computed value on a random source
+		(\texttt{observe}).
+\end{itemize}
+
+Anglican can be mostly implemented as a regular programming
+language, except for the handling of these checkpoints.
+Depending on the \textit{inference algorithm}, \texttt{sample}
+and \texttt{observe} may result in implicit input/output
+operations and control changes. For example, \texttt{observe} in
+particle filtering inference algorithms is a non-deterministic
+control statement at which a particle can be either replicated
+or terminated. Similarly, in Metropolis-Hastings,
+\texttt{sample} is both an input and a non-deterministic control
+statement (with delayed effect), eventually affecting acceptance
+or rejection of a sample.
+
+Because of the checkpoints, Anglican programs must allow the
+inference algorithm to intervene, recording information and
+affecting control flow. This can be implemented through coroutines/
+cooperative multitasking, parallel execution/preemptive
+multitasking and shared memory, as well as through explicit
+maintenance of program continuations at checkpoints. Clojure is
+a functional language, and continuation-passing style (CPS)
+transformation is a well-developed technique in the area of
+functional languages. Implementing a variant of CPS
+transformation seemed to be the most flexible and lightweight
+option --- any other form of concurrency would put a higher burden
+on the underlying runtime (JVM) and the operating system.
+Consequently, Anglican has been implemented as a CPS-transformed
+computation with access to continuations in probabilistic
+checkpoints. Anglican `compiler', represented by a set of
+functions in the \texttt{anglican.trap} namespace, accepts a
+Clojure subset and transforms it into a variant of CPS
+representation, which allows inference algorithms to intervene
+in the execution flow at probabilistic checkpoints.
+
+Anglican is intended to co-exist with Clojure and be a part of
+the source of a Clojure program. To facilitate this, Anglican
+programs, or queries, are wrapped by macros (defined in the
+\texttt{anglican.emit} namespace), which call the CPS
+transformations and define Clojure objects suitable for passing
+as arguments to inference algorithms (\texttt{defquery},
+\texttt{query}). In addition to defining entire queries,
+Anglican promotes modularization of inference algorithm through
+definition of \textit{probabilistic functions} using
+\texttt{defm} and \texttt{fm} (Anglican counterparts of Clojure
+\texttt{defn} and \texttt{fn}), which are written in Anglican,
+may include probabilistic forms \texttt{sample} and
+\texttt{observe} (as well \texttt{predict} for the output), and
+can be seamlessly called from inside Anglican queries, just like
+functions locally defined within the same query.
+
 \section{Macro-based Compilation}

### 2015-07-17 paper/paper.tex
design outline in progress
editorial

@@
 Because of the checkpoints, Anglican programs must allow the
-inference algorithm to intervene, recording information and
+inference algorithm to step in, recording information and
 affecting control flow. This can be implemented through coroutines/
@@
 
+Operational semantic of Anglican queries is different from that
+of Clojure code, therefore queries must be called through
+inference algorithms, rather than `directly'.
+
 \section{Macro-based Compilation}

### 2015-07-18 paper/paper.tex
in progress

@@
 of Clojure code, therefore queries must be called through
-inference algorithms, rather than `directly'.
+inference algorithms, rather than `directly'.  The
+\texttt{anglican.inference} namespace supplies the
+\texttt{infer} multimethod, which accepts an Anglican query and
+returns a lazy sequence of weighted samples from the
+distribution defined by the query. Inference algorithms must
+provide an implementation for \texttt{infer}, as well as
+override some of the methods of the \textit{checkpoint}
+multimethod, called to handle \texttt{sample} and
+\texttt{observe} in an algorithmic-specific manner, as well
+as on termination of a probabilistic program.
 

### 2015-07-18 paper/paper.tex
design outline drafted

@@
 returns a lazy sequence of weighted samples from the
-distribution defined by the query. Inference algorithms must
+distribution defined by the query.  When inference
+is performed on an Anglican query, the query is run by a
+particular inference algorithm. Inference algorithms must
 provide an implementation for \texttt{infer}, as well as
@@
 as on termination of a probabilistic program.
+
+Finally, Anglican queries use `primitive', or commonly known
+and used, distributions, to draw random samples and condition
+observations. Many primitive distributions are provided by the
+\texttt{anglican.runtime} namespace, and additional
+distributions can be defined by the user by implementing the
+\texttt{distribution} protocol. The \texttt{defdist} macro
+provides a convenient syntax for defining primitive distributions.
+
+Compilation, invocation, and runtime support of anglican queries
+are discussed in detail in the following sections.
 

### 2015-07-19 paper/paper.tex
better english

@@
 Anglican is implemented in just this way --- a macro facility
-provided by Clojure, a Lisp dialect, is used both to add to
-Clojure constructs that delimit probabilistic, rather than
-deterministic, code, and to alter the operational semantics of
-Clojure expressions inside probabilistic code fragments. Anglican
-claims its right to count as a separate language because of the
-ubiquitous probabilistic execution semantics rather than because
-of a different syntax, which is actually an advantage rather
-than a drawback --- Clojure programmers only need to know how to
-specify the boundaries of Anglican programs, but can use
-familiar Clojure syntax to write probabilistic code.
+provided by Clojure, a Lisp dialect, is used both to extend
+Clojure with constructs that delimit probabilistic code, and to
+alter the operational semantics of Clojure expressions inside
+probabilistic code fragments. Anglican claims its right to count
+as a separate language because of the ubiquitous probabilistic
+execution semantics rather than because of a different syntax,
+which is actually an advantage rather than a drawback ---
+Clojure programmers only need to know how to specify the
+boundaries of Anglican programs, but can use familiar Clojure
+syntax to write probabilistic code.
 

### 2015-07-19 paper/paper.tex
better english

@@
 as a separate language because of the ubiquitous probabilistic
-execution semantics rather than because of a different syntax,
+execution semantics rather than a different syntax,
 which is actually an advantage rather than a drawback ---

### 2015-07-19 paper/paper.tex
better english

@@
 inference algorithm to step in, recording information and
-affecting control flow. This can be implemented through coroutines/
-cooperative multitasking, parallel execution/preemptive
+affecting control flow. This can be implemented through
+coroutines/cooperative multitasking, parallel execution/preemptive
 multitasking and shared memory, as well as through explicit
@@
 
-Compilation, invocation, and runtime support of anglican queries
-are discussed in detail in the following sections.
+Compilation, invocation, and runtime support of Anglican queries
+are discussed in detail in further sections.
 

### 2015-07-19 paper/paper.tex
better english

@@
 \texttt{query}). In addition to defining entire queries,
-Anglican promotes modularization of inference algorithm through
+Anglican promotes modularization of inference algorithms through
 definition of \textit{probabilistic functions} using

### 2015-07-19 paper/paper.tex
formatting

@@
 \texttt{defm} and \texttt{fm} (Anglican counterparts of Clojure
-\texttt{defn} and \texttt{fn}), which are written in Anglican,
-may include probabilistic forms \texttt{sample} and
-\texttt{observe} (as well \texttt{predict} for the output), and
-can be seamlessly called from inside Anglican queries, just like
-functions locally defined within the same query.
+\texttt{defn} and \texttt{fn}). Probabilistic functions are
+written in Anglican, may include probabilistic forms
+\texttt{sample} and \texttt{observe} (as well \texttt{predict}
+for the output), and can be seamlessly called from inside
+Anglican queries, just like functions locally defined within the
+same query.
 
@@
 provide an implementation for \texttt{infer}, as well as
-override some of the methods of the \textit{checkpoint}
+override some of the methods of the \texttt{checkpoint}
 multimethod, called to handle \texttt{sample} and

### 2015-07-19 paper/paper.tex
proofread

@@
 example, by adding new control structures; another is to keep
-the existing syntax but alter the operational semantic --- the
-way programs are executed and compute their outputs.
+the existing syntax but alter the \textit{operational semantics}
+--- the way programs are executed and compute their outputs.
 
@@
 
-Operational semantic of Anglican queries is different from that
+Operational semantics of Anglican queries is different from that
 of Clojure code, therefore queries must be called through

### 2015-07-24 paper/paper.tex
white paper in progress --- transformation

@@
 
+Compilation of Anglican into Clojure relies on the Clojure
+\textit{macro} facility. However, the compilation algorithm is
+implemented as a library of \textit{functions} in namespace
+\texttt{anglican.trap}, which are invoked by macros. Namespace
+\texttt{anglican.trap-test} contains many transformation tests,
+which both help assure proper functioning of the transformation
+functions, and serve as illustrative examples of individual
+transformations, such as of functions, flow control, and
+probabilistic contstructs.
+
+The CPS transformation is organized in top-down manner. The
+top-level function is  \texttt{cps-of-expression}, which
+receives an expression and a continuation, and returns the
+expression in the CPS form, with the computed result passed to
+the continuation. A continuation accepts \texttt{two arguments}:
+\begin{itemize}
+	\item the computed value;
+	\item the internal state, bound to the local variable
+		\texttt{\$state} in every lexical scope.
+\end{itemize}
+The state (\texttt{\$state}) is threaded through the computation
+and contains data used by inference algorithm. \texttt{\$state}
+is a Clojure hash map, and the map entries are
+algorithm-dependent. Except for transformation of
+\texttt{mem}, the memoization form, CPS transformation routines
+are not aware of contents of \texttt{\$state}, do not access or
+modify it directly, but rather just thread the state unmodified
+through the computation. Algorithm-specific handlers of
+checkpoints corresponding to the probabilistic forms
+(\texttt{sample}, \texttt{observe}) modify the
+state and reinject a new state into the computation.
+
+\subsection{Expression Kinds}
+
+There are three different kinds of inputs to CPS transformation:
+\begin{itemize}
+	\item literals, which are passed as an argument to the
+		continuation unmodified;
+	\item value expressions (e.g. the \texttt{fn} form) 
+		(called \textit{opaque}
+		expressions in the code) which must be transformed to
+		CPS, but the transformed object is passed to the
+		continuation as a whole, opaquely;
+	\item general expressions (let's call them
+		\textit{transparent}), through which the continuation is
+		threaded in an expression-specific way, and can be
+		called in multiple locations of the CPS-transformed
+		code, such as in all branches of an \texttt{if}
+		statement.
+\end{itemize}
+
+\subsection{Literals}
+
+Literals are the same in Anglican and Clojure. They are 
+left unmodified. Literals are a subset of opaque expressions,
+and are identified by test \texttt{simple?} called from
+\texttt{opaque?}.  However, The Clojure syntax has a peculiarity
+of using the syntax of compound literals (vectors, hash maps,
+and sets) for data constructors. Hence, compound literals must
+be traversed recursively, and if there is a nested non-literal
+component, transformed into a call to the corresponding data
+constructor. Functions \texttt{cps-of-vector},
+\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
+\texttt{cps-of-expression}, transform Clojure constructor syntax
+(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
+corresponding calls.
+
+\subsection{Opaque Expressions}
+
+Opaque, or value, expressions, have a different shape in the
+original and the CPS form. However, their CPS form follows the
+pattern \texttt{(continuation transformed-expression)}, and thus
+the transformation does not depend on the continuation, and
+can be accomplished without passing the continuation as a
+transformation argument. Primitive (non-CPS) procedures used in
+Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
+forms are opaque and transformed by
+\texttt{primitive-procedure-cps}, \texttt{fn-cps}, and
+\texttt{mem-cps}, correspondingly.
+
+\subsection{General Expressions}
+
 \section{Inference Algorithms}

### 2015-07-26 paper/paper.tex
general expressions in progress

@@
 
+The most general form of CPS transformation receives an
+expression and a continuation, and returns the expression
+in CPS form with the continuation potentially called in multiple
+tail positions. General expressions can be somewhat voluntarily
+divided into several groups:
+\begin{itemize}
+	\item binding forms --- \texttt{let} and
+		\texttt{loop/recur};
+	\item flow control --- \texttt{if}, \texttt{when},
+		\texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
+	\item function applications and \texttt{apply};
+	\item probabilistic forms --- \texttt{predict},
+		\texttt{observe}, \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
+\end{itemize}
+
 \section{Inference Algorithms}

### 2015-07-28 paper/paper.tex
implementation highlights in progress
continuations

@@
 \usepackage{url}
+
+\input{anglican}
 
@@
 
-\subsection{Literals}
+\subsubsection{Literals}
 
@@
 
-\subsection{Opaque Expressions}
+\subsubsection{Opaque Expressions}
 
@@
 
-\subsection{General Expressions}
+\subsubsection{General Expressions}
 
@@
 \end{itemize}
+Functions that transform general expressions accept the
+expression and the continuation as parameters, and are
+consistently named \texttt{cps-of-}\textit{form}, for example,
+\texttt{cps-of-do}, \texttt{cps-of-store}.
+
+\subsection{Implementation Highlights}
+
+\subsubsection{Continuations}
+
+Continuations are functions that are called in tail positions
+with the computed value and state as their arguments --- in CPS
+there is always a function call in every tail position and never
+a value. Continuations are passed to CPS transformers, and when
+transformers are called recursively, the continuations are
+generated on the fly.
+
+There are two critical issues related to generation of
+continuations:
+\begin{itemize}
+	\item unbounded \textit{stack growth} in recursive code;
+	\item code size \textit{explosion} when a non-atomic
+		continuation is symbolically substituted in multiple
+		locations.
+\end{itemize}
+
+In implementations of functional programming languages stack
+growth is avoided through \textit{tail call optimization} (TCO).
+However, Clojure does not support a general form of TCO, and
+CPS-transformed code that creates deeply nested calls will
+easily exhaust the stack. Anglican employs a workaround called
+\textit{trampolining} --- instead of inserting a continuation
+call directly, the transformer always wraps the call into a
+\textit{thunk}, or parameterless function. The thunk is returned
+and called by the trampoline (Clojure provides function
+\texttt{trampoline} for this purpose) --- this way the
+computation continues but the stack is collapsed on every
+continuation call. Function \texttt{continue} implements the
+wrapping.
+
+To realize the potential danger of code size explosion, consider
+CPS transformation of code
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+(if (adult? person)
+  (if (male? person)
+	 (choose-beer)
+	 (choose_wine))
+  (choose-juice))
+\end{lstlisting}
+with continuation 
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+(fn [choice _]
+  (case (kind choice)
+    :beer (beer-jar choice)
+    :wine (wine-glass choice)
+    :juice (juice-bottle choice)))
+\end{lstlisting}
+The code of continuation, represented by an \texttt{fn} form, will be
+repeated three times. In general, CPS code can grow extremely
+large if symbolic continuations are inserted repeatedly.
+
+To circumvent this inefficiency, CPS transformers for
+expressions with multiple continuation points (\texttt{if} and
+derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
+the continuation to a fresh symbol if it is not yet a symbol.
+Macro \texttt{defn-with-named-cont} establishes the binding
+automatically.
+
+\subsubsection{Primitive Procedures}
+
+\subsubsection{Probabilistic forms}
+
+\subsubsection{Stochastic memoization}
 

### 2015-07-28 paper/anglican.tex
added anglican for listings

% !TEX root =  mdpp.tex

% listings definition for Anglican programming language

\lstdefinelanguage{Anglican}%
{morekeywords={*,*1,*2,*3,*agent*,*allow-unresolved-vars*,*assert*,*clojure-version*,*command-line-args*,%
		*compile-files*,*compile-path*,*e,*err*,*file*,*flush-on-newline*,*in*,*macro-meta*,%
		*math-context*,*ns*,*out*,*print-dup*,*print-length*,*print-level*,*print-meta*,*print-readably*,%
		*read-eval*,*source-path*,*use-context-classloader*,*warn-on-reflection*,+,-,->,->>,..,/,:else,%
		<,<=,=,==,>,>=,@,accessor,aclone,add-classpath,add-watch,agent,agent-errors,aget,alength,alias,%
		all-ns,alter,alter-meta!,alter-var-root,amap,ancestors,and,apply,areduce,array-map,aset,%
		aset-boolean,aset-byte,aset-char,aset-double,aset-float,aset-int,aset-long,aset-short,assert,%
		assoc,assoc!,assoc-in,associative?,atom,await,await-for,await1,bases,bean,bigdec,bigint,binding,%
		bit-and,bit-and-not,bit-clear,bit-flip,bit-not,bit-or,bit-set,bit-shift-left,bit-shift-right,%
		bit-test,bit-xor,boolean,boolean-array,booleans,bound-fn,bound-fn*,butlast,byte,byte-array,%
		bytes,case,cast,char,char-array,char-escape-string,char-name-string,char?,chars,chunk,chunk-append,%
		chunk-buffer,chunk-cons,chunk-first,chunk-next,chunk-rest,chunked-seq?,class,class?,%
		clear-agent-errors,clojure-version,coll?,comment,commute,comp,comparator,compare,compare-and-set!,%
		compile,complement,concat,cond,condp,conj,conj!,cons,constantly,construct-proxy,contains?,count,%
		counted?,create-ns,create-struct,cycle,dec,decimal?,declare,def,defdist,definline,defm,defmacro,defmethod,%
		defmulti,defn,defn-,defonce,defquery,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%
		descendants,destructure,disj,disj!,dissoc,dissoc!,distinct,distinct?,do,do-template,doall,doc,%
		dorun,doseq,dosync,dotimes,doto,double,double-array,doubles,drop,drop-last,drop-while,empty,empty?,%
		ensure,enumeration-seq,eval,even?,every?,false,false?,ffirst,file-seq,filter,finally,find,find-doc,%
		find-ns,find-var,first,float,float-array,float?,floats,flush,fm,fn,fn?,fnext,for,force,format,future,%
		future-call,future-cancel,future-cancelled?,future-done?,future?,gen-class,gen-interface,gensym,%
		get,get-in,get-method,get-proxy-class,get-thread-bindings,get-validator,hash,hash-map,hash-set,%
		identical?,identity,if,if-let,if-not,ifn?,import,in-ns,inc,init-proxy,instance?,int,int-array,%
		integer?,interleave,intern,interpose,into,into-array,ints,io!,isa?,iterate,iterator-seq,juxt,%
		key,keys,keyword,keyword?,last,lazy-cat,lazy-seq,let,letfn,line-seq,list,list*,list?,load,load-file,%
		load-reader,load-string,loaded-libs,locking,long,long-array,longs,loop,macroexpand,macroexpand-1,%
		make-array,make-hierarchy,map,map?,mapcat,max,max-key,memfn,memoize,merge,merge-with,meta,%
		method-sig,methods,min,min-key,mod,monitor-enter,monitor-exit,name,namespace,neg?,new,newline,%
		next,nfirst,nil,nil?,nnext,not,not-any?,not-empty,not-every?,not=,ns,ns-aliases,ns-imports,%
		ns-interns,ns-map,ns-name,ns-publics,ns-refers,ns-resolve,ns-unalias,ns-unmap,nth,nthnext,num,%
		number?,observe,odd?,or,parents,partial,partition,pcalls,peek,persistent!,pmap,pop,pop!,pop-thread-bindings,%
		pos?,pr,pr-str,predict,prefer-method,prefers,primitives-classnames,print,print-ctor,print-doc,print-dup,%
		print-method,print-namespace-doc,print-simple,print-special-doc,print-str,printf,println,println-str,%
		prn,prn-str,promise,proxy,proxy-call-with-super,proxy-mappings,proxy-name,proxy-super,%
		push-thread-bindings,pvalues,query,quot,rand,rand-int,range,ratio?,rational?,rationalize,re-find,%
		re-groups,re-matcher,re-matches,re-pattern,re-seq,read,read-line,read-string,recur,reduce,ref,%
		ref-history-count,ref-max-history,ref-min-history,ref-set,refer,refer-clojure,reify,%
		release-pending-sends,rem,remove,remove-method,remove-ns,remove-watch,repeat,repeatedly,%
		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,reverse,reversible?,%
		rseq,rsubseq,sample,second,select-keys,send,send-off,seq,seq?,seque,sequence,sequential?,set,set!,%
		set-validator!,set?,short,short-array,shorts,shutdown-agents,slurp,some,sort,sort-by,sorted-map,%
		sorted-map-by,sorted-set,sorted-set-by,sorted?,special-form-anchor,special-symbol?,split-at,%
		split-with,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%
		sync,syntax-symbol-anchor,take,take-last,take-nth,take-while,test,the-ns,throw,time,to-array,%
		to-array-2d,trampoline,transient,tree-seq,true,true?,try,type,unchecked-add,unchecked-dec,%
		unchecked-divide,unchecked-inc,unchecked-multiply,unchecked-negate,unchecked-remainder,%
		unchecked-subtract,underive,unquote,unquote-splicing,update-in,update-proxy,use,val,vals,%
		var,var-get,var-set,var?,vary-meta,vec,vector,vector?,when,when-first,when-let,when-not,%
		while,with-bindings,with-bindings*,with-in-str,with-loading-context,with-local-vars,%
		with-meta,with-open,with-out-str,with-precision,xml-seq,zero?,zipmap
	},%
	   sensitive,% ???
	      alsodigit=-,%
		     morecomment=[l];,%
			    morestring=[b]"%
				  }[keywords,comments,strings]%

### 2015-07-28 paper/paper.tex
typo
author

@@
 \title{Notes on Design and Implementation of Anglican}
+\author{David Tolpin\\dtolpin@robots.ox.ac.uk}
 
@@
 \begin{abstract}
-	In these notes, are explain the rationale behind the
+	In these notes, I explain the rationale behind the
 	macro-compiled implementation of Anglican, outline
@@
 \texttt{trampoline} for this purpose) --- this way the
-computation continues but the stack is collapsed on every
+computation continues, but the stack is collapsed on every
 continuation call. Function \texttt{continue} implements the
@@
 
-To realize the potential danger of code size explosion, consider
+To realize potential danger of code size explosion, consider
 CPS transformation of code

### 2015-07-29 paper/paper.tex
managing primitive procedures

@@
 
+When an Anglican function is transformed into a Clojure function
+by \texttt{fn-cps}, two auxiliary parameters are added to the
+beginning of the parameter list --- continuation and state.
+Correspondingly, when a function \textit{call} is transformed
+(by \texttt{cps-of-application} or \texttt{cps-of-apply}), the
+current continuation and the state are passed to the called
+function. Anglican can also call Clojure functions; however
+Clojure functions do not expect these auxiliary parameters.
+To allow the mixing of Anglican (CPS-transformed) and Clojure
+function calls in Anglican code, the Anglican compiler must be
+able to recognize `primitive' (that is, implemented in Clojure
+rather than in Anglican) functions.
+
+Providing an explicit syntax for differentiating between
+Anglican and Clojure function calls would be cumbersome. Another
+option would be to use meta-data to identify Anglican function
+calls at runtime, however this would impact performance, and
+a good runtime performance is critical for probabilistic
+programs. The approach taken by Anglican is to maintain a list
+of unqualified names of primitive functions, as well of
+namespaces in which all functions are primitive, and recognize
+primitive functions by name --- if a function name is not in the
+list, the function is a Clojure function. Global dynamically
+bound variables \texttt{*primitive-procedures*} and
+\texttt{*primitive-namespace*} contain the initial lists of
+names and namespaces, correspondingly. Of course, local bindings
+can shade global primitive function names. For example,
+\texttt{first} is a Clojure function inside the let block in the
+following example:
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+(let [first (fn [[x & y]] x)]
+  (first '[1 2 3]))
+\end{lstlisting}
+The Anglican compiler takes care of the shading by rebinding
+\texttt{*primitive-procedures*} in every lexical scope
+(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
+\texttt{shading-primitive-procedures} automates the shading.
+
 \subsubsection{Probabilistic forms}

### 2015-07-29 paper/paper.tex
stochastic memoization

@@
 
-\subsubsection{Stochastic memoization}
+There are two proper probabilistic forms turning Anglican into a
+probabilistic programming language --- \texttt{sample} and
+\texttt{observe}. Their purpose is to interrupt deterministic
+computation and transfer control to the inference algorithm.
+Practically, this is achieved through returning
+\textit{checkpoints} --- Clojure records of the corresponding
+types (\texttt{anglican.trap.sample} or
+\texttt{anglican.trap.observe}). The records contain fields
+specific to each form, as well as the continuation; calling the
+continuation resumes the computation. Checkpoints expose the
+program state to the inference algorithm, and the updated state
+is re-injected into the computation when the continuation is
+called.
+
+In addition to checkpoints, there are a few other special forms
+--- \texttt{predict}, \texttt{store}, \texttt{retrieve},
+\texttt{mem} --- which modify program state. These forms are
+translated into expressions involving calls of functions from the
+\texttt{anglican.state} namespace. The \texttt{mem} form, which
+implements stochastic memoization, deserves a more detailed
+explanation.
+
+\textit{Memoization} is often implemented on top of a mutable
+dictionary, where the key is the argument list and the value is
+the return value. There are no mutable data structures in a
+probabilistic program, and \texttt{mem}'s memory is stored in a
+nested dictionary in the program state (function
+\texttt{mem-cps}).  Each memoized function receives a unique
+automatically generated identifier. Each time a memoized
+function is called, one of two continuations is chosen,
+depending on whether the same function (a function with the same
+identifier) was previously called in the same run of the
+probabilistic program with the same arguments. If the memoized
+result is available, the continuation of the memoized function
+call is immediately called with the stored result. Otherwise,
+the argument of \texttt{mem} is called with a continuation that
+first creates an updated state with the memoized result, and
+then calls the `outer' continuation with the result and the
+updated state.
+
+Memoized results are not shared among multiple runs of a
+probabilistic program, which is intended. Otherwise, it would be
+impossible to memoize functions with random results.
 

### 2015-07-29 paper/paper.tex
typos

@@
 dictionary, where the key is the argument list and the value is
-the return value. There are no mutable data structures in a
-probabilistic program, and \texttt{mem}'s memory is stored in a
-nested dictionary in the program state (function
-\texttt{mem-cps}).  Each memoized function receives a unique
-automatically generated identifier. Each time a memoized
+the returned value. However, there are no mutable data
+structures in a probabilistic program, hence \texttt{mem}'s
+memory is stored as a nested dictionary in the program state
+(function \texttt{mem-cps}).  Each memoized function receives a
+unique automatically generated identifier. Each time a memoized
 function is called, one of two continuations is chosen,

### 2015-07-29 paper/paper.tex
typos
another typo fixed

@@
 \end{lstlisting}
-The code of continuation, represented by an \texttt{fn} form, will be
-repeated three times. In general, CPS code can grow extremely
-large if symbolic continuations are inserted repeatedly.
+The code of the continuation, represented by an \texttt{fn}
+form, will be repeated three times. In general, CPS code can
+grow extremely large if symbolic continuations are inserted
+repeatedly.
 
@@
 memory is stored as a nested dictionary in the program state
-(function \texttt{mem-cps}).  Each memoized function receives a
+(function \texttt{mem-cps}).  Every memoized function gets a
 unique automatically generated identifier. Each time a memoized

### 2015-08-19 paper/paper.tex
inference algorithms in progress

@@
 \usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{graphicx}
+\usepackage{amssymb} \usepackage{amsfonts} \usepackage{graphicx}
 
@@
 
+An inference algorithm is an implementation of \texttt{infer}
+multimethod from \texttt{anglican.inference} namespace. The
+multimethod accepts an algorithm identifier, a query --- the
+probabilistic program in which to perform the inference, an
+initial value for the query, and optional algorithm parameters.
+
+The sole purpose of the algorithm identifier is method dispatch
+--- conventionally, the identifier is a keyword related to the
+algorithm name (\texttt{:lmh} for Lightweight
+Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
+etc.). The second parameter is a query as defined by
+\texttt{query} or \texttt{defquery} forms. If needed, the query
+can be defined anonymously, right in the argument list of a call
+to \texttt{infer}:
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+    (let [x 1]
+      (infer :pgibbs (query (predict x)) nil))
+\end{lstlisting}
+A query is executed by calling the initial continuation of the
+query, which accepts a value and a state. The state is supplied
+by the inference algorithm, however the value is provided as a
+parameter of \texttt{infer}. A query does not have to have any
+parameters, in which case the value can be simply \texttt{nil}.
+When a query is defined with a binding for the initial value,
+the value becomes available inside the query. The value is
+destructured in case of a structured binding, for example:
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+    (defquery my-query [mean sd]
+       (predict (sample (normal mean sd))))
+
+    (define samples (infer :lmh my-query [1.0 3.0]))
+\end{lstlisting}
+Finally, an indefinite number of auxiliary arguments can be
+passed to \texttt{infer}. By convention, the arguments should be
+keyword arguments, and are interpreted in the algorithm-specific
+manner.
+
 \section{Definitions and Runtime Library}

### 2015-08-19 paper/paper.tex
inference algorithms

@@
 \usepackage{enumitem}
+\usepackage[lighttt]{lmodern}
 \usepackage{amsmath}
@@
 CPS transformation of code
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
 (if (adult? person)
@@
      (choose-beer)
-     (choose_wine))
+     (choose-wine))
   (choose-juice))
@@
 with continuation 
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
 (fn [choice _]
@@
 following example:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
 (let [first (fn [[x & y]] x)]
@@
 
+\subsection{The \texttt{infer} multimethod}
+
 The sole purpose of the algorithm identifier is method dispatch
@@
 to \texttt{infer}:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
 (let [x 1]
@@
 destructured in case of a structured binding, for example:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\scriptsize\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
 (defquery my-query [mean sd]
@@
 
-    (define samples (infer :lmh my-query [1.0 3.0]))
+(def samples (infer :lmh my-query [1.0 3.0]))
 \end{lstlisting}
@@
 manner.
+
+\subsection{Internals of an inference algorithm}
+
+The simplest inference algorithm is \textit{importance
+sampling}:
+\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
+(derive ::algorithm :anglican.inference/algorithm)
+
+(defmethod infer :importance [_ prog value & {}]
+  (letfn [(sample-seq []
+            (lazy-seq
+              (cons
+                (:state (exec ::algorithm prog value
+                              initial-state))
+                (sample-seq))))]
+    (sample-seq)))
+\end{lstlisting}
+\texttt{anglican.importance/infer} just calls
+\texttt{anglican.inference/exec} and relies on default
+implementations of checkpoint handlers. A different inference
+algorithm would provide its own implementations of
+\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
+both, as well as invoke \texttt{exec} from an elaborated
+conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
+(\texttt{anglican.smc}) are examples of inference algorithms
+where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
+handler is overridden. In addition, SMC runs multiple particles
+(program instances) simultaneously, while LMH runs re-runs
+programs from an intermediate continuation rather than from the
+beginning.
 

### 2015-08-19 paper/paper.tex
first finished draft

@@
 
+A Clojure namespace that includes a definition of an Anglican
+program imports (`requires') two essential namespaces:
+\texttt{anglican.emit} and \texttt{anglican.runtime}. The former
+provides macros for defining Anglican programs
+(\texttt{defquery}, \texttt{query}) and functions
+(\texttt{defm}, \texttt{fm}, \texttt{mem}), as well as Anglican
+bootstrap definitions that must be included with every program
+--- first of all, CPS implementations of higher-order functions.
+\texttt{anglican.emit} can be viewed as the Anglican
+\textit{compiler tool}, which helps transform Anglican code into
+Clojure before any inference is performed.
+
+\texttt{anglican.runtime} is the Anglican runtime library. For
+convenience, it exposes common mathematical functions
+(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
+\texttt{exp}, etc.), but most importantly, it provides
+definitions of common distributions. Each distribution object
+implements the \texttt{anglican.runtime/distribution} protocol,
+with two methods: \texttt{sample} and \texttt{observe}. The
+names of the methods coincide with Anglican special forms
+\texttt{sample} and \texttt{observe}, but they are by no means
+the same. The \texttt{sample} \textit{method} returns a random sample,
+the \texttt{observe} \textit{method} returns the log probability
+of the value. The methods can be, and sometimes are called from
+handlers of the corresponding checkpoints, but do not have to
+be. For example, in LMH either the \texttt{sample} or
+the \texttt{observe} \textit{method} is called for a
+\texttt{sample} checkpoint, depending on whether the value is
+drawn or re-used.
+
+Macro \texttt{defdist} should be used to define distributions.
+\texttt{defdist} takes care of defining a separate record type
+for every distribution so that multimethods can be dispatched on
+distribution types when needed, e.g. for custom proposals.
+
+Similar to distributions, \textit{random processes}, related to
+so called `exchangeable random procedures', are defined using
+macro \texttt{defproc} and implement the
+\texttt{anglican.runtime/random-process} protocol. The protocol
+has twho methods --- \texttt{produce}, which returns a
+distribution object, and \texttt{absorb}, which returns a new
+random process object obtained by absorbing a value drawn from
+the process. While distributions are random and non-functional,
+random processes are deterministic and functional, hence
+\texttt{produce} and \texttt{absorb} are called directly and do
+not have corresponding special forms in Anglican.
+
 \bibliographystyle{alpha}

### 2016-07-04 paper/paper.tex
layout

@@
-\documentclass[a4paper]{article}
+\documentclass[preprint]{sigplanconf}
 
@@
 \usepackage{amsmath}
-\usepackage{amssymb} \usepackage{amsfonts} \usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{amsfonts}
+\usepackage{graphicx}
 
@@
 
-\title{Notes on Design and Implementation of Anglican}
-\author{David Tolpin\\dtolpin@robots.ox.ac.uk}
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\title{Design and Implementation of Probabilistic Programming Language Anglican}
+\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
 
@@
 forms are opaque and transformed by
-\texttt{primitive-procedure-cps}, \texttt{fn-cps}, and
+\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps}, \texttt{fn-cps}, and
 \texttt{mem-cps}, correspondingly.
@@
 CPS transformation of code
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
 (if (adult? person)
@@
 with continuation 
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
 (fn [choice _]
@@
 following example:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
 (let [first (fn [[x & y]] x)]
@@
 to \texttt{infer}:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
 (let [x 1]
@@
 destructured in case of a structured binding, for example:
-\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
+\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
 (defquery my-query [mean sd]
@@
 sampling}:
-\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
-(derive ::algorithm :anglican.inference/algorithm)
-
+\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+(derive ::algorithm
+        :anglican.inference/algorithm)
 (defmethod infer :importance [_ prog value & {}]
-  (letfn [(sample-seq []
+  (letfn
+    [(sample-seq []
        (lazy-seq
@@
 \end{lstlisting}
-\texttt{anglican.importance/infer} just calls
-\texttt{anglican.inference/exec} and relies on default
+\texttt{anglican.importance/{\linebreak[0]}infer} just calls
+\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
 implementations of checkpoint handlers. A different inference
@@
 definitions of common distributions. Each distribution object
-implements the \texttt{anglican.runtime/distribution} protocol,
+implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
 with two methods: \texttt{sample} and \texttt{observe}. The
@@
 macro \texttt{defproc} and implement the
-\texttt{anglican.runtime/random-process} protocol. The protocol
+\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
 has twho methods --- \texttt{produce}, which returns a
@@
 
-\bibliographystyle{alpha}
+\acks
+
+\bibliographystyle{abbrvnat}
 \bibliography{refs}

### 2016-07-04 paper/paper.tex
in progress

@@
 \begin{abstract}
+    Anglican is a probabilistic programming system designed to interoperate
+    with Clojure and other JVM languages. We describe the implementation of
+    Anglican and illustrate how its design facilitates both explorative and
+    industrial use of probabilistic programming.
+
     In these notes, I explain the rationale behind the
@@
 \section{Introduction}
+For data science practitioners, statistical inference is typically
+but one step in a more elaborate analysis workflow. The first stage
+of this work involves data acquisition, pre-processing and cleaning.
+This is often followed by several iterations of exploratory model
+design and testing of inference algorithms. Once a sufficiently
+robust statistical model and corresponding inference algorithm have
+been identified, analysis results must be post-processed, visualized,
+and in some cases integrated into a wider production system.
+
+Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
+represent generative models as programs written in a specialized
+language that provides syntax for the definition and conditioning of
+random variables. The code for such models is generally concise,
+modular, and easy to modify or extend. Typically inference can be
+performed for any probabilistic program using one or more generic
+inference techniques provided by the system backend, such as
+Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
+\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
+of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
+Although these generic techniques are not always as statistically
+efficient as techniques that take advantage of model-specific
+optimizations, probabilistic programming makes it easier to optimize
+models for a specific application in a manner that is efficient in
+terms of the dimensionality of its latent variables.
+
+While probabilistic programming systems shorten the iteration cycle in
+exploratory model design, they typically lack basic functionality needed for
+data I/O, pre-processing, and analysis and visualization of inference results.
+In this demonstration, we describe the implementation of
+Anglican~(\url{http://www.robots.ox.ac.uk/~fwood/anglican/}), a probabilistic
+programming language that tightly integrates with
+Clojure~(\url{http://clojure.org/}), a general-purpose programming language
+that runs on the Java Virtual Machine (JVM). Both languages share a common
+syntax, and can be invoked from each other. This allows Anglican programs to
+make use of a rich set of libraries written in both Clojure and Java.
+Conversely Anglican allows intuitive and compact specification of models for
+which inference may be performed as part of a larger Clojure project.
 
@@
 \section{Design Outline}
+
+An Anglican program, or \textit{query}, is compiled into a Clojure
+function. When inference is performed with a provided algorithm, this
+produces a sequence of return values, or \textit{predicts}. Anglican
+shares a common syntax with Clojure; Clojure functions can be called
+from Anglican code and vice versa. A simple program in Anglican can
+look like the following code:
+
+%\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+%(defquery models
+%  "chooses a distribution which describes the data"
+%  (let [;; Model -- randomly choose a distribution and parameters
+%        dist (sample (categorical [[uniform-discrete 1]
+%                                   [uniform-continuous 1]
+%                                   [normal 1] 
+%                                   [gamma 1]]))
+%        a (sample (gamma 1 1)) b (sample (gamma 1 1))
+%        d (dist a b)]
+%    ;; Data  --- samples from the unknown distribution
+%    (observe d 1) (observe d 2) (observe d 4) (observe d 7)
+%    ;; Output --- predicted distribution type and parameters
+%    (predict :d (type d))
+%    (predict :a a) (predict :b b)))
+%\end{lstlisting}
+
+Internally, an Anglican query is represented by a computation in
+\textit{continuation passing style} (CPS), and inference algorithms
+exploit the CPS structure of the code to intercept probabilistic
+operations in an algorithm-specific way. Among the available
+inference algorithms there are Particle Cascade~\cite{PWD+14},
+Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
+Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
+Inference on Anglican queries generates a lazy sequence of
+samples, which can be processed asynchronously in Clojure
+code for analysis, integration, and decision making.
+
+Clojure (and Anglican) run on the JVM and get access to a wide choice of Java
+libraries for data processing, networking, presentation, and imaging.
+Conversely, Anglican queries can be called from Java and other JVM languages.
+Programs involving Anglican queries can be deployed as JVM
+\textit{jars}, and run without modification on any platform for 
+which JVM is available.
 

### 2016-07-04 paper/paper.tex
formatting issues mostly resolved, now writing

@@
 \usepackage{algorithmic}
+\usepackage[lighttt]{lmodern}
+\ttfamily
+\DeclareFontShape{OT1}{lmtt}{m}{it}
+     {<->sub*lmtt/m/sl}{}
 	 \usepackage{listings}
@@
 look like the following code:
-
-%\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
-%(defquery models
-%  "chooses a distribution which describes the data"
-%  (let [;; Model -- randomly choose a distribution and parameters
-%        dist (sample (categorical [[uniform-discrete 1]
-%                                   [uniform-continuous 1]
-%                                   [normal 1] 
-%                                   [gamma 1]]))
-%        a (sample (gamma 1 1)) b (sample (gamma 1 1))
-%        d (dist a b)]
-%    ;; Data  --- samples from the unknown distribution
-%    (observe d 1) (observe d 2) (observe d 4) (observe d 7)
-%    ;; Output --- predicted distribution type and parameters
-%    (predict :d (type d))
-%    (predict :a a) (predict :b b)))
-%\end{lstlisting}
+\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+(defquery models
+  "chooses a distribution
+   which describes the data"
+  (let [;; Model - randomly choose distribution
+	    dist (sample (categorical
+                       [[uniform-continuous 1]
+                        [normal 1] 
+                        [gamma 1]]))
+        a (sample (gamma 1 1))
+        b (sample (gamma 1 1))
+        d (dist a b)]
+    ;; Data - samples from the distribution
+    (observe d 1)
+    (observe d 2)
+    (observe d 4)
+    (observe d 7)
+    ;; Output --- predicted distribution
+    (predict :d (type d))
+    (predict :a a) (predict :b b)))
+\end{lstlisting}
 

### 2016-07-04 paper/paper.tex
style for listings

@@
 \usepackage{algorithmic}
+\usepackage{listings}
 \usepackage[lighttt]{lmodern}
@@
      {<->sub*lmtt/m/sl}{}
-	 \usepackage{listings}
+\lstdefinestyle{default}{frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}
 
@@
 look like the following code:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (defquery models
@@
 CPS transformation of code
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (if (adult? person)
@@
 with continuation 
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (fn [choice _]
@@
 following example:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (let [first (fn [[x & y]] x)]
@@
 to \texttt{infer}:
-\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (let [x 1]
@@
 destructured in case of a structured binding, for example:
-\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (defquery my-query [mean sd]
@@
 sampling}:
-\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
+\begin{lstlisting}[style=default]
 (derive ::algorithm

### 2016-07-05 paper/paper.tex
typo fixed

@@
 primitive functions by name --- if a function name is not in the
-list, the function is a Clojure function. Global dynamically
+list, the function is an Anglican function. Global dynamically
 bound variables \texttt{*primitive-procedures*} and

### 2016-07-05 paper/paper.tex
some citations

@@
 In this demonstration, we describe the implementation of
-Anglican~(\url{http://www.robots.ox.ac.uk/~fwood/anglican/}), a probabilistic
+Anglican~(\cite{TMW15}), a probabilistic
 programming language that tightly integrates with
@@
 
-Anglican is a probabilistic programming language integrated with
-Clojure.  There are several ways to build a programming language
+There are several ways to build a programming language
 on top or besides another language.  The easiest to grasp is an
@@
 \end{itemize}
-The following sections explain the way Anglican is implemented,
-from source-to-source transformation and syntactic wrappers to
-inference algorithms which accept Clojure functions built from
-Anglican code as a parameter, and produce probabilistic outputs.
+
+Anglican language syntax, compilation, invocation, and runtime
+support of Anglican queries are discussed in detail in further
+sections.
 
@@
 function. When inference is performed with a provided algorithm, this
-produces a sequence of return values, or \textit{predicts}. Anglican
+produces a sequence of output values, or \textit{predicts}. Anglican
 shares a common syntax with Clojure; Clojure functions can be called
@@
 \begin{lstlisting}[style=default]
-(defquery models
+(defquery models data
   "chooses a distribution
    which describes the data"
-  (let [;; Model - randomly choose distribution
+  (let [;; Guessing a distribution
         dist (sample (categorical
-                       [[uniform-continuous 1]
-                        [normal 1] 
-                        [gamma 1]]))
+                       [[uniform-continuous 0 1]
+                        [normal 0.5 1] 
+                        [gamma 1 1]]))
         a (sample (gamma 1 1))
@@
         d (dist a b)]
-    ;; Data - samples from the distribution
-    (observe d 1)
-    (observe d 2)
-    (observe d 4)
-    (observe d 7)
-    ;; Output --- predicted distribution
+
+    ;; Observing samples from the distribution
+    (loop [data data]
+      (when (seq data)
+        (let [[x & data] data]
+          (observe d x))
+        (recur data)))
+
+    ;; Predicting the distribution
     (predict :d (type d))
@@
 exploit the CPS structure of the code to intercept probabilistic
-operations in an algorithm-specific way. Among the available
+operations in an algorithm-specific way\footnote{\cite{GS15}
+also describe a CPS-based implementation of a probabilistic
+programming language.}. Among the available
 inference algorithms there are Particle Cascade~\cite{PWD+14},
@@
 operations and control changes. For example, \texttt{observe} in
-particle filtering inference algorithms is a non-deterministic
+particle filtering inference algorithms \cite{WVM14} a non-deterministic
 control statement at which a particle can be either replicated
-or terminated. Similarly, in Metropolis-Hastings,
+or terminated. Similarly, in Metropolis-Hastings \cite{WSG11},
 \texttt{sample} is both an input and a non-deterministic control
@@
 provides a convenient syntax for defining primitive distributions.
-
-Compilation, invocation, and runtime support of Anglican queries
-are discussed in detail in further sections.
 
@@
 can shade global primitive function names. For example,
-\texttt{first} is a Clojure function inside the let block in the
+\texttt{first} is an Anglican function inside the let block in the
 following example:

### 2016-07-05 paper/paper.tex
more citations

@@
 In this demonstration, we describe the implementation of
-Anglican~(\cite{TMW15}), a probabilistic
+Anglican~(\cite{TMW15, Anglican}), a probabilistic
 programming language that tightly integrates with
-Clojure~(\url{http://clojure.org/}), a general-purpose programming language
+Clojure~\cite{H08,Clojure}, a general-purpose programming language
 that runs on the Java Virtual Machine (JVM). Both languages share a common
@@
 provides a convenient syntax for defining primitive distributions.
+
+\section{Anglican Syntax}
+
+\cite{SDF+10, PC94, H08}
 

### 2016-07-05 paper/paper.tex
minor changes before specifying syntax

@@
 Similar to distributions, \textit{random processes}, related to
-so called `exchangeable random procedures', are defined using
+so called `exchangeable random procedures'\footnote{However, random sequences
+generated by Anglican random processes are not required to be
+exchangeable.}, are defined using
 macro \texttt{defproc} and implement the
@@
 
+This work was partially supported under DARPA PPAML through the
+U.S. AFRL under Cooperative Agreement number FA8750- 14-2-0006,
+Sub Award number 61160290-111668.
+
 \bibliographystyle{abbrvnat}

### 2016-07-07 paper/paper.tex
abstract and summary

@@
     Anglican is a probabilistic programming system designed to interoperate
-    with Clojure and other JVM languages. We describe the implementation of
-    Anglican and illustrate how its design facilitates both explorative and
-    industrial use of probabilistic programming.
+    with Clojure and other JVM languages. We introduce the
+	programming language Anglican, outline our design choices,
+	and discuss in depth the implementation of the Anglican language
+	and runtime, including macro-based compilation, extended
+	CPS-based evaluation model, and functional representations for
+	probabilistic paradigms, such as distribution,
+	random processes, and inference algorithms.
 
-    In these notes, I explain the rationale behind the
-    macro-compiled implementation of Anglican, outline
-    design choices, and describe important implementation
-    aspects.
+	We show that a probabilistic functional language can be
+	implemented efficiently and integrated tightly with a
+	conventional functional language with only moderate
+	computational overhead. We also demonstrate how advanced
+	probabilistic modeling concepts are mapped naturally
+	to a functional representation.
 \end{abstract}
@@
 
+\section{Summary}
+
+In this paper, we presented design and implementation internals
+of the Probabilistic Programming System Anglican. Implementing
+a language is an interesting endeavour, in particular when the
+language implements a new paradigm, in this case probabilistic
+programming. Functional programming is a natural complement of
+probabilistic programming --- the latter allows both concise and
+expressive specification of probabilistic generative models and
+efficient implementation of inference algorithm.
+
+Implementing a probabilistic language on top and in tight
+integration with a functional language, Clojure, both helped
+us to accomplish an ambitious goal in a short time span, and
+provided important insights on structure and semantics of
+probabilistic concepts incorporated in Anglican. Efficiency
+and expressive power of Anglican owe to adherence to the
+functional approach as much as to rich inference opportunities
+by the Anglican environment.
+
 \acks

### 2016-07-07 paper/paper.tex
typos fixed

@@
 	CPS-based evaluation model, and functional representations for
-	probabilistic paradigms, such as distribution,
-	random processes, and inference algorithms.
+	probabilistic paradigms, such as a distribution,
+	a random process, and an inference algorithm.
 
@@
 	probabilistic modeling concepts are mapped naturally
-	to a functional representation.
+	to the functional foundation.
 \end{abstract}
@@
 Internally, an Anglican query is represented by a computation in
-\textit{continuation passing style} (CPS), and inference algorithms
+\textit{continuation passing style} (CPS)~\cite{AJ89}, and inference algorithms
 exploit the CPS structure of the code to intercept probabilistic
@@
 involved, and normal, linear execution of the program is
-disrupted. In Anglican and similar languages there are two
-types of such checkpoints:
-\begin{itemize}
-    \item drawing a value from a random source (\texttt{sample});
-    \item conditioning the posterior distribution by
-        conditioning a computed value on a random source
+disrupted. In Anglican and similar languages there are two types of such checkpoints:
+\begin{itemize} \item drawing a value from a random source (\texttt{sample});
+    \item conditioning a computed value on a random source
         (\texttt{observe}).
@@
 transformations, such as of functions, flow control, and
-probabilistic contstructs.
+probabilistic constructs.
 
@@
 expression in the CPS form, with the computed result passed to
-the continuation. A continuation accepts \texttt{two arguments}:
+the continuation. A continuation accepts \textit{two} arguments:
 \begin{itemize}
@@
 \texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
-has twho methods --- \texttt{produce}, which returns a
+has two methods --- \texttt{produce}, which returns a
 distribution object, and \texttt{absorb}, which returns a new

### 2016-07-07 paper/paper.tex
language definition

@@
 
-\section{Anglican Syntax}
+\section{Language}
 
-\cite{SDF+10, PC94, H08}
+The Anglican language is a subset of Clojure\footnote{It would be
+possible to support almost full Clojure by expanding all macros
+in the Anglican source code, however in Clojure, unlike in
+Scheme~\cite{SDF+10}, or Common Lisp~\cite{PC94}, the result
+of macro-expansion of derived special forms is not well
+specified and implementation specific.}.  Within \texttt{defquery},
+\texttt{let}, \texttt{if}, \texttt{when}, \texttt{cond},
+\texttt{case}, \texttt{and}, \texttt{or}, \texttt{fn} forms are
+supported (others may be in the future but are not now). 
+In \texttt{let} bindings and \texttt{fn} argument lists,
+vector destructuring (but not hash map destructuring) is
+supported. Compound literals for vectors, hash maps, and
+sets are supported just like in Clojure.
+
+\subsection{\texttt{recur}}
+
+Anglican is stackless, therefore \texttt{recur} is
+unnecessary, no recursive call can lead to stack overflow;
+Recursive calls to functions should be used instead. However,
+\texttt{loop}/\texttt{recur} is provided for convenience as a way to express
+loops. \texttt{recur} outside of \texttt{loop} will lead to unpredictable
+behaviour and hard-to-catch errors.
+
+\subsection{Core library}
+
+All of Clojure's core library except for higher-order functions
+(functions that accept other functions as arguments) is
+available in Anglican. Higher order functions cannot be reused
+from Clojure, they have to be re-implemented to accept
+functional arguments in CPS form. The following higher-order
+functions are implemented: \texttt{map}, \texttt{reduce},
+\texttt{filter}, \texttt{some}, \texttt{repeatedly},
+\texttt{comp}, \texttt{partial}.
 
@@
 and are identified by test \texttt{simple?} called from
-\texttt{opaque?}.  However, The Clojure syntax has a peculiarity
+\texttt{opaque?}.  However, the Clojure syntax has a peculiarity
 of using the syntax of compound literals (vectors, hash maps,

### 2016-07-07 paper/paper.tex
simplifications and removal of non-essential details

@@
 (functions that accept other functions as arguments) is
-available in Anglican. Higher order functions cannot be reused
+available in Anglican. Higher-order functions cannot be reused
 from Clojure, they have to be re-implemented to accept
@@
 implemented as a library of \textit{functions} in namespace
-\texttt{anglican.trap}, which are invoked by macros. Namespace
-\texttt{anglican.trap-test} contains many transformation tests,
-which both help assure proper functioning of the transformation
-functions, and serve as illustrative examples of individual
-transformations, such as of functions, flow control, and
-probabilistic constructs.
-
+\texttt{anglican.trap}, which are invoked by macros.
 The CPS transformation is organized in top-down manner. The
@@
 Literals are the same in Anglican and Clojure. They are 
-left unmodified. Literals are a subset of opaque expressions,
-and are identified by test \texttt{simple?} called from
-\texttt{opaque?}.  However, the Clojure syntax has a peculiarity
+left unmodified; literals are a subset of opaque expressions.
+However, the Clojure syntax has a peculiarity
 of using the syntax of compound literals (vectors, hash maps,
@@
 implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
-with two methods: \texttt{sample} and \texttt{observe}. The
-names of the methods coincide with Anglican special forms
-\texttt{sample} and \texttt{observe}, but they are by no means
-the same. The \texttt{sample} \textit{method} returns a random sample,
-the \texttt{observe} \textit{method} returns the log probability
-of the value. The methods can be, and sometimes are called from
+with two methods: \texttt{sample} and \texttt{log-prob}.
+The \texttt{sample} \textit{method} returns a random sample and
+roughly corresponds to the \texttt{sample} checkpoint,
+the \texttt{log-prob} \textit{method} returns the log probability
+of the value and roughly corresponds to the \texttt{observe}
+checkpoint. The methods can be, and sometimes are called from
 handlers of the corresponding checkpoints, but do not have to
 be. For example, in LMH either the \texttt{sample} or
-the \texttt{observe} \textit{method} is called for a
+the \texttt{log-prob} \textit{method} is called for a
 \texttt{sample} checkpoint, depending on whether the value is

### 2016-07-31 paper/paper.tex
reformatted
in progress

@@
 \title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
+\authorinfo{David Tolpin \and Jan Willem van de Meent \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and fwood@robots.ox.ac.uk}
 
@@
 involved, and normal, linear execution of the program is
-disrupted. In Anglican and similar languages there are two types of such checkpoints:
-\begin{itemize} \item drawing a value from a random source (\texttt{sample});
+disrupted. In Anglican and similar languages there are two types
+of such checkpoints:
+
+\begin{itemize}
+	\item drawing a value from a random source (\texttt{sample});
     \item conditioning a computed value on a random source
@@
 \end{itemize}
+The CPS transformation of constant \texttt{1} with continutation
+\texttt{cont} thus takes the following form:
+\begin{lstlisting}[style=default]
+=> (cps-of-expression 1 'cont)
+(cont 1 $state)
+\end{lstlisting}
+
 The state (\texttt{\$state}) is threaded through the computation
 and contains data used by inference algorithm. \texttt{\$state}
-is a Clojure hash map, and the map entries are
-algorithm-dependent. Except for transformation of
+is a Clojure hash map:
+\begin{lstlisting}[style=default]
+(def initial-state
+  "initial program state"
+  {:log-weight 0.0
+   :predicts []
+   :result nil
+   ::mem {}
+   ::store nil
+   ... })
+\end{lstlisting}
+The full list of map entries depends on the inference
+algorithm. Except for transformation of
 \texttt{mem}, the memoization form, CPS transformation routines
@@
 (\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
-corresponding calls.
+corresponding calls:
+\begin{lstlisting}[style=default]
+=> (cps-of-vector [0 1 2] 'cont)
+(cont (vector 0 1 2) $state)
+=> (cps-of-hash-map {:a 1 :b 2} 'cont)
+(cont (hash-map :a 1, :b 2) $state)
+=> (cps-of-set #{0 1} 'cont)
+(cont (set (list 0 1)) $state)
+\end{lstlisting}
 
@@
 forms are opaque and transformed by
-\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps}, \texttt{fn-cps}, and
-\texttt{mem-cps}, correspondingly.
+\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps},
+\texttt{fn-cps}, and \texttt{mem-cps}, correspondingly: a
+slightly simplified CPS form of expression
+\begin{lstlisting}[style=default]
+(fn [x y]
+  (+ x y))
+\end{lstlisting}
+would be
+\begin{lstlisting}[style=default]
+(fn [cont $state x y] 
+  (cont (+ x y) $state))
+\end{lstlisting}
+In the actual code an automatically generated fresh symbol is
+used instead of \texttt{cont}.
 
@@
 
+\paragraph{Trampolining}
+
 In implementations of functional programming languages stack
@@
 continuation call. Function \texttt{continue} implements the
-wrapping.
+wrapping and is invoked on every continuation call:
+\begin{lstlisting}[style=default]
+=> (continue 'cont 'value 'state)
+(fn [] (cont value state))
+\end{lstlisting}
+Correspondingly, the full, wrapped CPS form of
+\begin{lstlisting}[style=default]
+(fn [x y] (+ x y))
+\end{lstlisting}
+becomes
+\begin{lstlisting}[style=default]
+(fn [cont $state x y] 
+  (fn [] 
+    (cont (+ x y) $state)))
+\end{lstlisting}
+When the CPS-transformed function is called, it returns a
+\textit{thunk} (a parameterless function) which is then
+re-invoked through the trampoline, with the stack collapsed.
 

### 2016-07-31 paper/paper.tex
example of factored continuation

@@
 
-\usepackage{algorithm}
-\usepackage{algorithmic}
 \usepackage{listings}
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}
 
@@
 
-\paragraph{Trampolining}
+\paragraph{Managing stack size}
 
@@
 
+\paragraph{Avoiding exponential code growth}
+
 To realize potential danger of code size explosion, consider
@@
 \end{lstlisting}
-The code of the continuation, represented by an \texttt{fn}
+The code of the continuation, represented by the \texttt{fn}
 form, will be repeated three times. In general, CPS code can
@@
 the continuation to a fresh symbol if it is not yet a symbol.
-Macro \texttt{defn-with-named-cont} establishes the binding
-automatically.
+Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
+establishes the binding automatically:
+\begin{lstlisting}[style=default]
+=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
+(let [cont (fn [x] (* 2 x))]
+  (if c
+      (fn [] (cont t $state))
+      (fn [] (cont f $state))))
+\end{lstlisting}
 

### 2016-07-31 paper/paper.tex
in progress

@@
 Practically, this is achieved through returning
-\textit{checkpoints} --- Clojure records of the corresponding
+\textbf{checkpoints} --- Clojure records of the corresponding
 types (\texttt{anglican.trap.sample} or
@@
 is re-injected into the computation when the continuation is
-called.
+called: 
+\begin{lstlisting}[style=default]
+=> (cps-of-expression '(sample dist) 'cont)
+(->sample dist cont $state)
+=> (cps-of-expression '(observe dist val) 'cont)
+(->observe dist val cont $state)
+\end{lstlisting}
 
@@
 
-\textit{Memoization} is often implemented on top of a mutable
+\textbf{Memoization} is often implemented on top of a mutable
 dictionary, where the key is the argument list and the value is
@@
 call is immediately called with the stored result. Otherwise,
-the argument of \texttt{mem} is called with a continuation that
+the argument of \texttt{mem} is called with a continuation which
 first creates an updated state with the memoized result, and
 then calls the `outer' continuation with the result and the
-updated state.
+updated state:
+\begin{lstlisting}[style=default]
+=> (mem-cps '(foo))
+(let [M (gensym "M")]
+  (fn mem [C $state & P]
+    (if (in-mem? $state M P)
+      ;; previously memoized result
+      (fn []
+        (C (get-mem $state M P) $state))
+      ;; new computation
+      (clojure.core/apply
+        foo
+        ;; memoize result in continuation
+        (fn [V $state]
+          (fn [] (C V (set-mem $state M P V))))
+        $state
+        P)))))
+\end{lstlisting}
 
@@
 implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
-with two methods: \texttt{sample} and \texttt{log-prob}.
-The \texttt{sample} \textit{method} returns a random sample and
+with two methods: \texttt{sample*} and \texttt{observe*}.
+The \texttt{sample*} \textit{method} returns a random sample and
 roughly corresponds to the \texttt{sample} checkpoint,
-the \texttt{log-prob} \textit{method} returns the log probability
+the \texttt{observe*} \textit{method} returns the log probability
 of the value and roughly corresponds to the \texttt{observe}
@@
 handlers of the corresponding checkpoints, but do not have to
-be. For example, in LMH either the \texttt{sample} or
-the \texttt{log-prob} \textit{method} is called for a
+be. For example, in LMH either the \texttt{sample*} or
+the \texttt{observe*} \textit{method} is called for a
 \texttt{sample} checkpoint, depending on whether the value is

### 2016-07-31 paper/paper.tex
draft ready

@@
 (let [M (gensym "M")]
-  (fn mem [C $state & P]
+  (fn mem23623 [C $state & P]
     (if (in-mem? $state M P)
@@
 definitions of common distributions. Each distribution object
-implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
+implements the \texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution} protocol,
 with two methods: \texttt{sample*} and \texttt{observe*}.
@@
 for every distribution so that multimethods can be dispatched on
-distribution types when needed, e.g. for custom proposals.
+distribution types when needed, e.g. for custom proposals. The
+Bernoulli distribution could be defined as follows:
+\begin{lstlisting}[style=default]
+(defdist bernoulli
+  "Bernoulli distribution"
+  [p]
+  (sample* [this] (if (< (rand) p) 1 0))
+  (observe* [this value]
+    (Math/log (case value
+                1 p
+                0 (- 1. p)
+                0.))))
+\end{lstlisting}
 
@@
 random process object obtained by absorbing a value drawn from
-the process. While distributions are random and non-functional,
+the process. Here is a definition of the Poisson process:
+\begin{lstlisting}[style=default]
+(defproc PP
+  "Poisson point process"
+  [rate] [dist (exponential rate)]
+  (produce [this] dist)
+  (absorb [this value] this))
+\end{lstlisting}
+While distributions are random and non-functional,
 random processes are deterministic and functional, hence
 \texttt{produce} and \texttt{absorb} are called directly and do
-not have corresponding special forms in Anglican.
+not have corresponding special forms in Anglican: 
+\begin{itemize}
+    \item Values produced by the process are sampled from the distribution
+        returned by \texttt{produce}.
+    \item When a value is observed from the distribution, a new
+        process, with the observed value absorbed, is returned
+        by \texttt{absorb}.
+\end{itemize}
+A common pattern of programming with processes is Anglican is
+recursively computing a new process based on the observed value:
+\begin{lstlisting}[style=default]
+(loop [process (HawkesPointProcess)
+       [value & values] values]
+  (let [dist (produce process)]
+    (observe dist value)
+    (recur (absorb process value) values)))
+\end{lstlisting}
+Unlike conventional implementations of random processes,
+Anglican's random processes do not have mutable state.
 

### 2016-07-31 paper/paper.tex
local changes, probably conflicting

--- 2016-07-31 paper/paper.tex
@@
 
+\usepackage{algorithm}
+\usepackage{algorithmic}
 \usepackage{listings}
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}
+\lstdefinestyle{default}{frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}
 
@@
 \title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan Willem van de Meent \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and fwood@robots.ox.ac.uk}
+\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
 
@@
 involved, and normal, linear execution of the program is
-disrupted. In Anglican and similar languages there are two types
-of such checkpoints:
-
-\begin{itemize}
-	\item drawing a value from a random source (\texttt{sample});
+disrupted. In Anglican and similar languages there are two types of such checkpoints:
+\begin{itemize} \item drawing a value from a random source (\texttt{sample});
     \item conditioning a computed value on a random source
@@
 \end{itemize}
-The CPS transformation of constant \texttt{1} with continutation
-\texttt{cont} thus takes the following form:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression 1 'cont)
-(cont 1 $state)
-\end{lstlisting}
-
 The state (\texttt{\$state}) is threaded through the computation
 and contains data used by inference algorithm. \texttt{\$state}
-is a Clojure hash map:
-\begin{lstlisting}[style=default]
-(def initial-state
-  "initial program state"
-  {:log-weight 0.0
-   :predicts []
-   :result nil
-   ::mem {}
-   ::store nil
-   ... })
-\end{lstlisting}
-The full list of map entries depends on the inference
-algorithm. Except for transformation of
+is a Clojure hash map, and the map entries are
+algorithm-dependent. Except for transformation of
 \texttt{mem}, the memoization form, CPS transformation routines
@@
 (\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
-corresponding calls:
-\begin{lstlisting}[style=default]
-=> (cps-of-vector [0 1 2] 'cont)
-(cont (vector 0 1 2) $state)
-=> (cps-of-hash-map {:a 1 :b 2} 'cont)
-(cont (hash-map :a 1, :b 2) $state)
-=> (cps-of-set #{0 1} 'cont)
-(cont (set (list 0 1)) $state)
-\end{lstlisting}
+corresponding calls.
 
@@
 \texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps},
-\texttt{fn-cps}, and \texttt{mem-cps}, correspondingly: a
-slightly simplified CPS form of expression
-\begin{lstlisting}[style=default]
-(fn [x y]
-  (+ x y))
-\end{lstlisting}
-would be
-\begin{lstlisting}[style=default]
-(fn [cont $state x y] 
-  (cont (+ x y) $state))
-\end{lstlisting}
-In the actual code an automatically generated fresh symbol is
-used instead of \texttt{cont}.
+\texttt{fn-cps}, and \texttt{mem-cps}, correspondingly.
 
@@
 
-\paragraph{Managing stack size}
-
 In implementations of functional programming languages stack
@@
 continuation call. Function \texttt{continue} implements the
-wrapping and is invoked on every continuation call:
-\begin{lstlisting}[style=default]
-=> (continue 'cont 'value 'state)
-(fn [] (cont value state))
-\end{lstlisting}
-Correspondingly, the full, wrapped CPS form of
-\begin{lstlisting}[style=default]
-(fn [x y] (+ x y))
-\end{lstlisting}
-becomes
-\begin{lstlisting}[style=default]
-(fn [cont $state x y] 
-  (fn [] 
-    (cont (+ x y) $state)))
-\end{lstlisting}
-When the CPS-transformed function is called, it returns a
-\textit{thunk} (a parameterless function) which is then
-re-invoked through the trampoline, with the stack collapsed.
-
-\paragraph{Avoiding exponential code growth}
+wrapping.
 
@@
 \end{lstlisting}
-The code of the continuation, represented by the \texttt{fn}
+The code of the continuation, represented by an \texttt{fn}
 form, will be repeated three times. In general, CPS code can
@@
 the continuation to a fresh symbol if it is not yet a symbol.
-Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
-establishes the binding automatically:
-\begin{lstlisting}[style=default]
-=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
-(let [cont (fn [x] (* 2 x))]
-  (if c
-      (fn [] (cont t $state))
-      (fn [] (cont f $state))))
-\end{lstlisting}
+Macro \texttt{defn-with-named-cont} establishes the binding
+automatically.
 
@@
 bound variables \texttt{*primitive-procedures*} and
-\texttt{*primitive-namespace*} contain the initial lists of
+\texttt{*primitive-namespaces*} contain the initial lists of
 names and namespaces, correspondingly. Of course, local bindings

### 2016-07-31 paper/paper.tex

--- 2016-07-31 paper/paper.tex
@@
 \end{itemize}
-A common pattern of programming with processes is Anglican is
+A common pattern of programming with processes in Anglican is
 recursively computing a new process based on the observed value:

### 2016-07-31 paper/paper.tex
another typo fixed
language fix

@@
 \end{itemize}
-The CPS transformation of constant \texttt{1} with continutation
+The CPS transformation of constant \texttt{1} with continuation
 \texttt{cont} thus takes the following form:
@@
 functional approach as much as to rich inference opportunities
-by the Anglican environment.
+of the Anglican environment.
 

### 2016-08-03 paper/paper.tex
minor

@@
 \setlength{\pdfpagewidth}{\paperwidth}
+
+\newcommand*{\dt}[1]{{ \bf [}{\bf DT:} {\it #1}{\bf ]}}
+\newcommand*{\fw}[1]{{ \bf [}{\bf FW:} {\it #1}{\bf ]}}
+\newcommand*{\hy}[1]{{ \bf [}{\bf HY:} {\it #1}{\bf ]}}
+\newcommand*{\jw}[1]{{ \bf [}{\bf JW:} {\it #1}{\bf ]}}
 
@@
 For data science practitioners, statistical inference is typically
-but one step in a more elaborate analysis workflow. The first stage
+just one step in a more elaborate analysis workflow. The first stage
 of this work involves data acquisition, pre-processing and cleaning.
@@
 design and testing of inference algorithms. Once a sufficiently
-robust statistical model and corresponding inference algorithm have
+robust statistical model and a corresponding inference algorithm have
 been identified, analysis results must be post-processed, visualized,
@@
 terms of the dimensionality of its latent variables.
+
+\hy{Minor point. I got slightly confused by the last sentence. What does
+``efficient in terms of the dimensionality ..'' mean?}
 

### 2016-08-03 paper/paper.tex
update

@@
 In this demonstration, we describe the implementation of
-Anglican~(\cite{TMW15, Anglican}), a probabilistic
+Anglican~\cite{TMW15, Anglican}, a probabilistic
 programming language that tightly integrates with
@@
 sections.
+
+\hy{Great introduction. It very nicely explains why Anglican
+is born and how it is implemented. If I act as a really picky
+reviewer or a reader, I would also expect the instroduction
+to help me grasp in a high level what new insights or scientific problems 
+this development of Anglican have resulted in or solved so far.}
 

### 2016-08-04 paper/paper.tex
minor

@@
 \begin{lstlisting}[style=default]
-(defquery models data
-  "chooses a distribution
-   which describes the data"
+(defquery model data
+  "chooses a distribution which describes the data"
   (let [;; Guessing a distribution
         dist (sample (categorical
-                       [[uniform-continuous 0 1]
-                        [normal 0.5 1] 
-                        [gamma 1 1]]))
+                       [[normal 0.5] 
+                        [gamma 0.5]]))
         a (sample (gamma 1 1))
@@
 
-    ;; Predicting the distribution
-    (predict :d (type d))
-    (predict :a a) (predict :b b)))
+    ;; Predicting a, b and the distribution
+    (predict :a a) 
+    (predict :b b)
+    (predict :d d)))
 \end{lstlisting}
+This query defines a probability distribution on predicted
+variables, $\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
+for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
+positive parameters for the type. Samples from this distribution 
+can be obtained by running one of Anglican's inference algorithms
+on the query.
 

### 2016-08-04 paper/paper.tex
update

@@
 (defquery model data
-  "chooses a distribution which describes the data"
+  "chooses a distribution 
+   which describes the data"
   (let [;; Guessing a distribution
@@
 
-Clojure (and Anglican) run on the JVM and get access to a wide choice of Java
+Clojure (and Anglican) runs on the JVM and gets access to a wide choice of Java
 libraries for data processing, networking, presentation, and imaging.
@@
 operations and control changes. For example, \texttt{observe} in
-particle filtering inference algorithms \cite{WVM14} a non-deterministic
+particle filtering inference algorithms \cite{WVM14} is a non-deterministic
 control statement at which a particle can be either replicated
@@
 or rejection of a sample.
+
+\hy{What do you mean by input in ``both an input..''?}
 

### 2016-08-04 paper/paper.tex
minor

@@
 affecting control flow. This can be implemented through
-coroutines/cooperative multitasking, parallel execution/preemptive
-multitasking and shared memory, as well as through explicit
-maintenance of program continuations at checkpoints. Clojure is
+coroutines/cooperative multitasking, and parallel execution/preemptive
+multitasking, as well as through explicit
+maintenance of program continuations at checkpoints. Anglican
+follows the latter option. Clojure is
 a functional language, and continuation-passing style (CPS)

### 2016-08-04 paper/paper.tex
update

@@
 \end{lstlisting}
-This query defines a probability distribution on predicted
+This query defines a probability distribution on three predicted
 variables, $\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
@@
 positive parameters for the type. Samples from this distribution 
-can be obtained by running one of Anglican's inference algorithms
-on the query.
+can be obtained by running the query under one of Anglican's 
+inference algorithms.
 
@@
 \texttt{anglican.emit} namespace), which call the CPS
-transformations and define Clojure objects suitable for passing
+transformations and define Clojure values
+%Clojure objects
+suitable for passing
 as arguments to inference algorithms (\texttt{defquery},
@@
 Anglican promotes modularization of inference algorithms through
-definition of \textit{probabilistic functions} using
+the definitions of \textit{probabilistic functions} using
 \texttt{defm} and \texttt{fm} (Anglican counterparts of Clojure
@@
 inference algorithms, rather than `directly'.  The
-\texttt{anglican.inference} namespace supplies the
-\texttt{infer} multimethod, which accepts an Anglican query and
+\texttt{anglican.inference} namespace declares the polymorphic
+function \texttt{infer} using Clojure's multimethod mechanism.
+This function accepts an Anglican query and
 returns a lazy sequence of weighted samples from the
@@
 provide an implementation for \texttt{infer}, as well as
-override some of the methods of the \texttt{checkpoint}
-multimethod, called to handle \texttt{sample} and
-\texttt{observe} in an algorithmic-specific manner, as well
-as on termination of a probabilistic program.
+override the polymorphic function \texttt{checkpoint} (defined
+as a multimethod) so as to handle \texttt{sample} and
+\texttt{observe} in an algorithm-specific manner, as well
+as to construct an appropriate result on the 
+termination of a probabilistic program.
 
@@
 observations. Many primitive distributions are provided by the
-\texttt{anglican.runtime} namespace, and additional
-distributions can be defined by the user by implementing the
-\texttt{distribution} protocol. The \texttt{defdist} macro
+\texttt{anglican.runtime} namespace, and an additional
+distribution can be defined by the user by implementing a
+particular set of functions for the distribution 
+(via Clojure's protocol mechanism). The \texttt{defdist} macro
 provides a convenient syntax for defining primitive distributions.
+
+\hy{I think that the concepts like multimethod and protocol
+are specific to Clojure, and they are not well-known among PL
+people. I tried to rephase the text so that people can follow
+it even when they don't know these concepts. I didn't know
+``form'' before, but using it is perhaps ok because it is a
+standard terminology for Lisp programmers.}
 

### 2016-08-04 paper/paper.tex
update

@@
 the definitions of \textit{probabilistic functions} using
-\texttt{defm} and \texttt{fm} (Anglican counterparts of Clojure
-\texttt{defn} and \texttt{fn}). Probabilistic functions are
+\texttt{defm} 
+%and \texttt{fm} 
+(Anglican counterparts of Clojure \texttt{defn}).
+%and \texttt{fn}). 
+Probabilistic functions are
 written in Anglican, may include probabilistic forms
@@
 of macro-expansion of derived special forms is not well
-specified and implementation specific.}.  Within \texttt{defquery},
-\texttt{let}, \texttt{if}, \texttt{when}, \texttt{cond},
-\texttt{case}, \texttt{and}, \texttt{or}, \texttt{fn} forms are
-supported (others may be in the future but are not now). 
+specified and implementation specific.}.  Anglican queries
+are defined within \texttt{defquery} as shown in the previous section,
+using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case}, 
+\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
+other forms of Clojure may be in the future but are not now. 
 In \texttt{let} bindings and \texttt{fn} argument lists,
-vector destructuring (but not hash map destructuring) is
-supported. Compound literals for vectors, hash maps, and
+Clojure's pattern matching mechanism for vector data type
+(called vector destructuring) is supported. Also, compound 
+literals for vectors, hash maps, and
 sets are supported just like in Clojure.
 
-\subsection{\texttt{recur}}
+\subsection{Tail recursion}
 
-Anglican is stackless, therefore \texttt{recur} is
-unnecessary, no recursive call can lead to stack overflow;
-Recursive calls to functions should be used instead. However,
-\texttt{loop}/\texttt{recur} is provided for convenience as a way to express
-loops. \texttt{recur} outside of \texttt{loop} will lead to unpredictable
-behaviour and hard-to-catch errors.
+Clojure provides special constructs \texttt{loop} and \texttt{recur}
+for writing tail-recursive programs. Such constructs are not necessary
+in Anglican, because Anglican programs are CPS-converted and do not
+use stacks; no recursive call in Anglican can lead to stack overflow.
+In Anglican, recursive calls to functions should be used instead of
+\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
+for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop} 
+will lead to unpredictable behaviour and hard-to-catch errors.
 
@@
 available in Anglican. Higher-order functions cannot be reused
-from Clojure, they have to be re-implemented to accept
+from Clojure, as they have to be re-implemented to accept
 functional arguments in CPS form. The following higher-order

### 2016-08-05 paper/paper.tex
minor

@@
 as a multimethod) so as to handle \texttt{sample} and
-\texttt{observe} in an algorithm-specific manner, as well
-as to construct an appropriate result on the 
+\texttt{observe} in an algorithm-specific manner and
+to construct an appropriate result on the 
 termination of a probabilistic program.
@@
 \texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
-other forms of Clojure may be in the future but are not now. 
+other forms of Clojure may be supported in the future but are not now. 
 In \texttt{let} bindings and \texttt{fn} argument lists,
@@
 
-\subsection{Tail recursion}
-
-Clojure provides special constructs \texttt{loop} and \texttt{recur}
-for writing tail-recursive programs. Such constructs are not necessary
-in Anglican, because Anglican programs are CPS-converted and do not
-use stacks; no recursive call in Anglican can lead to stack overflow.
-In Anglican, recursive calls to functions should be used instead of
-\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
-for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop} 
-will lead to unpredictable behaviour and hard-to-catch errors.
-
-\subsection{Core library}
+\subsection{Core Library}
 
@@
 \texttt{comp}, \texttt{partial}.
+
+\subsection{Tail Call}
+
+Clojure provides special forms \texttt{loop} and \texttt{recur}
+for writing tail-recursive programs. Such forms are not necessary
+in Anglican, because Anglican programs are CPS-converted and do not
+use stacks. For instance, no recursive call in Anglican can lead to stack overflow.
+In fact, in Anglican, it is recommended to use recursive calls to functions instead of
+\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
+for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop} 
+will lead to unpredictable behaviour and hard-to-catch errors.
 

### 2016-08-05 paper/paper.tex
update

@@
 \section{Macro-based Compilation}
+
+\hy{I found this section confusing in the following sense. First,
+it does not highlight the most interesting aspects of the compilation
+of Anglican. It gives a good overview on how the compilation is
+implemented, but a reader might want to see something unique and
+interesting about Anglican. To me, the most interesting part of the
+compilation is the treatment of probabilistic constructs in Anglican's
+CPS transformation. How are sample,
+observe and stochastic memoization compiled? How should the standard
+CPS transformation be modified in order to handle these constructs?
+The use of state can be one answer for the second question.
+The second most interesting part is about the use of trampolining
+the mix with Clojure code.
+
+Second, the section assumes that a reader is already familiar 
+with Clojure or something specific to probabilistic programming. For 
+instance, it might be that a reader  has never heard about
+Clojure records and stochastic memoization. Then, she or he
+would have difficulties in following this section.
+
+What should we do? One option is not to do anything. Another is
+to take care of the second point. That is, we make the section
+easily followed by someone who does not know Clojure. The third
+is to restructure the section and to take care of my first criticism.}
 

### 2016-08-08 paper/paper.tex
update
minor

@@
 
-\newcommand*{\dt}[1]{{ \bf [}{\bf DT:} {\it #1}{\bf ]}}
-\newcommand*{\fw}[1]{{ \bf [}{\bf FW:} {\it #1}{\bf ]}}
-\newcommand*{\hy}[1]{{ \bf [}{\bf HY:} {\it #1}{\bf ]}}
-\newcommand*{\jw}[1]{{ \bf [}{\bf JW:} {\it #1}{\bf ]}}
+\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:} 
+        {\rm #1}\end{minipage}}}
+\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:} 
+        {\rm #1}\end{minipage}}}
+\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
+        {\rm #1}\end{minipage}}}
+\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
+        {\rm #1}\end{minipage}}}
 
@@
 the mix with Clojure code.
-
 Second, the section assumes that a reader is already familiar 
@@
 would have difficulties in following this section. 
-
 What should we do? One option is not to do anything. Another is
@@
 query, which accepts a value and a state. The state is supplied
-by the inference algorithm, however the value is provided as a
+by the inference algorithm, while the value is provided as a
 parameter of \texttt{infer}. A query does not have to have any
@@
 When a query is defined with a binding for the initial value,
-the value becomes available inside the query. The value is
-destructured in case of a structured binding, for example:
+the value becomes available inside the query. A query may
+accept multiple parameters as components of a Clojure vector,
+in which case the initial value is given as a vector and its
+components get bound to the corresponding parameters of the
+query (via destructuring).
+%The value is destructured via pattern match in case of a structured binding. 
+For example,
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-Finally, an indefinite number of auxiliary arguments can be
-passed to \texttt{infer}. By convention, the arguments should be
+Finally, any number of auxiliary arguments can be
+passed to \texttt{infer}. 
+%Finally, an indefinite number of auxiliary arguments can be
+%passed to \texttt{infer}. 
+By convention, the arguments should be
 keyword arguments, and are interpreted in the algorithm-specific

### 2016-08-08 paper/paper.tex
update

@@
 inference algorithms, rather than `directly'.  The
-\texttt{anglican.inference} namespace declares the polymorphic
+\texttt{anglican.inference} namespace declares the (ad-hoc) polymorphic
 function \texttt{infer} using Clojure's multimethod mechanism.
@@
 
-An inference algorithm is an implementation of \texttt{infer}
-multimethod from \texttt{anglican.inference} namespace. The
-multimethod accepts an algorithm identifier, a query --- the
+An inference algorithm is an implementation of the (ad-hoc) 
+polymorphic function \texttt{infer} declared as Clojure's multimethod
+in the \texttt{anglican.inference} namespace. The
+function accepts an algorithm identifier, a query --- the
 probabilistic program in which to perform the inference, an
@@
 
-\subsection{The \texttt{infer} multimethod}
+\subsection{The \texttt{infer} function}
 
-The sole purpose of the algorithm identifier is method dispatch
---- conventionally, the identifier is a keyword related to the
+The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
+an appropriate overloading or implementation of the function --- conventionally, 
+the identifier is a keyword related to the
 algorithm name (\texttt{:lmh} for Lightweight
@@
 etc.). The second parameter is a query as defined by
-\texttt{query} or \texttt{defquery} forms. If needed, the query
-can be defined anonymously, right in the argument list of a call
-to \texttt{infer}:
+the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
+the following Clojure code invokes \texttt{infer} on an Anglican query
+defined anonymously via the \texttt{query} form:
+%If needed, the query
+%can be defined anonymously, right in the argument list of a call
+%to \texttt{infer}:
 \begin{lstlisting}[style=default]
@@
 the value becomes available inside the query. A query may
-accept multiple parameters as components of a Clojure vector,
-in which case the initial value is given as a vector and its
-components get bound to the corresponding parameters of the
-query (via destructuring).
+accept multiple parameters using Clojure's structured binding.
+For instance, it may take multiple parameters as components 
+of an input vector. In this case, the initial value is given
+as a structured value, such as a vector, and the components
+of this value gets pattern matched to the corresponding
+parameters of the query via the destructuring mechanism of Clojure:
 %The value is destructured via pattern match in case of a structured binding. 

### 2016-08-08 paper/paper.tex
minor

@@
 of this value gets pattern matched to the corresponding
-parameters of the query via the destructuring mechanism of Clojure:
+parameters of the query via the destructuring mechanism of Clojure.
 %The value is destructured via pattern match in case of a structured binding. 
@@
 handler is overridden. In addition, SMC runs multiple particles
-(program instances) simultaneously, while LMH runs re-runs
+(program instances) simultaneously, while LMH re-runs
 programs from an intermediate continuation rather than from the
 beginning.
+
+\hy{This subsection is also not quite self-contained. Someone
+who is not familiar with Clojure or Anglican codebase is likely
+to get lost. On the other hand, it is a good overview or guide
+for those who want to study the implemenation of Anglican. One
+way to address this presentation issue while keeping the benefits
+of the current text is to include some extended examples. For instance,
+in the beginning of the paper, we might include several
+Anglican examples. Then, we show that how these examples get
+cps-transformed. Next, we present the implementation of inference
+algorithms in Anglican using semi pseudo code; this is not quite
+an implementation in Clojure but gives a good sketch on how
+various functions are implemented in each inference algorithm.
+Finally, we show how these algorithms work on cps-tranformed 
+examples.}
+
+\hy{I rewrote various parts so as to avoid the word multimethod.
+My strategy was to use the phrases, such as (ad-hoc) polymophic
+function and overloading. All of my rephrasing should be
+double-checked.}
 

### 2016-08-15 paper/paper.tex
update
added changebar and comments with brief instructions
simpler explanation of advantage of probabilistic programming

@@
 
+\usepackage[color]{changebar}
 \usepackage{enumitem}
@@
 
+% Authors: comment using these commands
+
 \newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:} 
@@
         {\rm #1}\end{minipage}}}
+
+% Authors: use changebars to mark changed regions
+% \cbstart
+% ...
+% \cbend
 
@@
 of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
-Although these generic techniques are not always as statistically
-efficient as techniques that take advantage of model-specific
-optimizations, probabilistic programming makes it easier to optimize
-models for a specific application in a manner that is efficient in
-terms of the dimensionality of its latent variables.
+% Although these generic techniques are not always as statistically
+% efficient as techniques that take advantage of model-specific
+% optimizations, probabilistic programming makes it easier to optimize
+% models for a specific application in a manner that is efficient in
+% terms of the dimensionality of its latent variables.
+\cbstart
+These generic techniques may be less statistically efficient
+than techniques tailored to a specific model, however
+probabilistic programming facilitates simpler implementations of
+models making the inference inherently faster.
+\cbend
 
-\hy{Minor point. I got slightly confused by the last sentence. What does
-``efficient in terms of the dimensionality ..'' mean?}
+%\hy{Minor point. I got slightly confused by the last sentence. What does
+%``efficient in terms of the dimensionality ..'' mean?}
 
@@
 In this paper, we presented design and implementation internals
-of the Probabilistic Programming System Anglican. Implementing
+of the \emph{probabilistic programming system} Anglican. Implementing
 a language is an interesting endeavour, in particular when the

### 2016-08-15 paper/paper.tex
rewriting in progress

@@
 
+\cbcolor{red}
 % Authors: use changebars to mark changed regions
@@
 \end{itemize}
+\cbstart
+Execution of probabilistic programs by inference algorithms is
+different from execution of deterministic programs.  A
+probabilistic program is executed multiple times, often hundreds
+of thousands or even millions of times for a single inference
+task. Random choices may affect which parts of the program code
+are executed and how often. Many inference algorithms require
+re-running the program multiple times partially, from a certain
+point on.  Different executions may employ different random
+choices, however for efficient inference a correspondence
+between random choices in different executions should be
+maintained. These are just some of the challenges which were
+faced and solved during development of Anglican.
+\cbend
 
@@
 
-\hy{Great introduction. It very nicely explains why Anglican
-is born and how it is implemented. If I act as a really picky
-reviewer or a reader, I would also expect the instroduction
-to help me grasp in a high level what new insights or scientific problems 
-this development of Anglican have resulted in or solved so far.}
+% \hy{Great introduction. It very nicely explains why Anglican
+% is born and how it is implemented. If I act as a really picky
+% reviewer or a reader, I would also expect the introduction
+% to help me grasp in a high level what new insights or scientific problems 
+% this development of Anglican have resulted in or solved so far.}
 
@@
 or terminated. Similarly, in Metropolis-Hastings \cite{WSG11},
-\texttt{sample} is both an input and a non-deterministic control
-statement (with delayed effect), eventually affecting acceptance
-or rejection of a sample.
+\texttt{sample} is both an input
+\cbstart
+statement which `reads' values from a random source,
+\cbend
+and a non-deterministic control statement
+(with delayed effect), eventually affecting acceptance or
+rejection of a sample.
 
-\hy{What do you mean by input in ``both an input..''?}
+%\hy{What do you mean by input in ``both an input..''?}
 

### 2016-08-18 paper/paper.tex
pre-submission draft

@@
         {\rm #1}\end{minipage}}}
-\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
-        {\rm #1}\end{minipage}}}
+\newcommand*{\hy}[1]{}
+% {\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
+%        {\rm #1}\end{minipage}}}
 \newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
@@
 
-% \hy{Great introduction. It very nicely explains why Anglican
-% is born and how it is implemented. If I act as a really picky
-% reviewer or a reader, I would also expect the introduction
-% to help me grasp in a high level what new insights or scientific problems 
-% this development of Anglican have resulted in or solved so far.}
+\hy{Great introduction. It very nicely explains why Anglican
+is born and how it is implemented. If I act as a really picky
+reviewer or a reader, I would also expect the introduction
+to help me grasp in a high level what new insights or scientific problems 
+this development of Anglican have resulted in or solved so far.}
 
@@
 
-Compilation of Anglican into Clojure relies on the Clojure
-\textit{macro} facility. However, the compilation algorithm is
-implemented as a library of \textit{functions} in namespace
-\texttt{anglican.trap}, which are invoked by macros.
-The CPS transformation is organized in top-down manner. The
-top-level function is  \texttt{cps-of-expression}, which
-receives an expression and a continuation, and returns the
-expression in the CPS form, with the computed result passed to
-the continuation. A continuation accepts \textit{two} arguments:
+\cbstart
+Compilation of Anglican into Clojure is built around a variant of CPS
+transformation. In a basic CPS-transformed program, a
+continuation receives a single argument --- the computed value.
+In Anglican, there are two flows of computation going in
+parallel: values are computed by functional code, and,
+at the same time, the state of the probabilistic program,
+used by inference algorithms, is updated by 
+probabilistic forms. Because of that, in Anglican a continuation
+accepts  \textit{two} arguments:
 \begin{itemize}
@@
 \end{itemize}
-The CPS transformation of constant \texttt{1} with continuation
-\texttt{cont} thus takes the following form:
+
+The compilation relies on the Clojure \textit{macro} facility,
+and implemented as a library of \textit{functions} in namespace
+\texttt{anglican.trap}, which are invoked by macros.  The CPS
+transformation is organized in top-down manner.  The top-level
+function is  \texttt{cps-of-expression}, which receives an
+expression and a continuation, and returns the expression in the
+CPS form.  For example, the CPS transformation of constant
+\texttt{1} with continuation \texttt{cont} thus takes the
+following form:
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
+
+\subsection {The State}
+\cbend
 
@@
 \subsection{Implementation Highlights}
+
+\cbstart
+So far we described the basics of Anglican compilation to
+Clojure. The described approaches and techniques are important
+for grasping the language implementation but relatively well-known.
+In the rest of the section we focus on challenges we met and
+resolved while implementing Anglican, as well as on
+implementation of unique features of Anglican as a probabilistic
+programming language.
+\cbend
 
@@
 Practically, this is achieved through returning
-\textbf{checkpoints} --- Clojure records of the corresponding
+\cbstart
+\textbf{checkpoints} --- Clojure values of the corresponding
 types (\texttt{anglican.trap.sample} or
-\texttt{anglican.trap.observe}). The records contain fields
+\texttt{anglican.trap.observe}). The types contain fields
+\cbend
 specific to each form, as well as the continuation; calling the
@@
 \texttt{anglican.state} namespace. The \texttt{mem} form, which
-implements stochastic memoization, deserves a more detailed
-explanation.
+implements memoization, deserves a more detailed explanation.
 
-\textbf{Memoization} is often implemented on top of a mutable
+\cbstart
+\paragraph{Memoization}
+The author of a probabilistic model might want to draw a
+feature of an entity randomly, but to retain the same drawn
+value for the same entity during a single run of the probabilistic
+program. For example, a person may have brown or green eyes
+with some probability, but the \textit{same} person will
+always have the same eye color. This can be achieved 
+memoized functions. In Anglican, one might write:
+\begin{lstlisting}[style=default]
+(let [eye-color (mem (fn [person] 
+                       (sample 
+                         (categorical
+                           ['brown 0.5]
+                           ['green 0.5]))))]
+  (if (not= (eye-color 'bill) (eye-color 'john))
+    (predict (eye-color 'bill))
+    (predict (eye-color 'john))))
+\end{lstlisting}
+\cbend
+Memoization is often implemented on top of a mutable
 dictionary, where the key is the argument list and the value is
@@
 
-\hy{I rewrote various parts so as to avoid the word multimethod.
-My strategy was to use the phrases, such as (ad-hoc) polymophic
-function and overloading. All of my rephrasing should be
-double-checked.}
-
 \section{Definitions and Runtime Library}
@@
 Macro \texttt{defdist} should be used to define distributions.
-\texttt{defdist} takes care of defining a separate record type
+\cbstart
+\texttt{defdist} takes care of defining a separate type
+\cbend
 for every distribution so that multimethods can be dispatched on

### 2016-08-18 paper/paper.tex
preparing for the next stage

@@
         {\rm #1}\end{minipage}}}
-\newcommand*{\hy}[1]{}
-% {\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
-%        {\rm #1}\end{minipage}}}
+\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
+        {\rm #1}\end{minipage}}}
 \newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
@@
 % terms of the dimensionality of its latent variables.
-\cbstart
 These generic techniques may be less statistically efficient
@@
 models making the inference inherently faster.
-\cbend
-
-%\hy{Minor point. I got slightly confused by the last sentence. What does
-%``efficient in terms of the dimensionality ..'' mean?}
 
@@
 \end{itemize}
-\cbstart
 Execution of probabilistic programs by inference algorithms is
@@
 faced and solved during development of Anglican.
-\cbend
 
@@
 sections.
-
-\hy{Great introduction. It very nicely explains why Anglican
-is born and how it is implemented. If I act as a really picky
-reviewer or a reader, I would also expect the introduction
-to help me grasp in a high level what new insights or scientific problems 
-this development of Anglican have resulted in or solved so far.}
 
@@
 \texttt{sample} is both an input
-\cbstart
 statement which `reads' values from a random source,
-\cbend
 and a non-deterministic control statement
@@
 rejection of a sample.
-
-%\hy{What do you mean by input in ``both an input..''?}
 
@@
 
-\hy{I think that the concepts like multimethod and protocol
-are specific to Clojure, and they are not well-known among PL
-people. I tried to rephase the text so that people can follow
-it even when they don't know these concepts. I didn't know
-``form'' before, but using it is perhaps ok because it is a
-standard terminology for Lisp programmers.}
-
 \section{Language}
@@
 
-\cbstart
+\dt{I believe I partially addressed the above comment, but need to
+think more about it.}
+
 Compilation of Anglican into Clojure is built around a variant of CPS
@@
 \subsection {The State}
-\cbend
 
@@
 
-\cbstart
-So far we described the basics of Anglican compilation to
+So far we introduced the basics of Anglican compilation to
 Clojure. The described approaches and techniques are important
@@
 programming language.
-\cbend
 
@@
 Practically, this is achieved through returning
-\cbstart
 \textbf{checkpoints} --- Clojure values of the corresponding
@@
 \texttt{anglican.trap.observe}). The types contain fields
-\cbend
 specific to each form, as well as the continuation; calling the
@@
 
-\cbstart
 \paragraph{Memoization}
@@
 \end{lstlisting}
-\cbend
+
 Memoization is often implemented on top of a mutable
@@
 Macro \texttt{defdist} should be used to define distributions.
-\cbstart
 \texttt{defdist} takes care of defining a separate type
-\cbend
 for every distribution so that multimethods can be dispatched on

### 2016-09-01 slides/ppdef.tex
in progress

\begin{frame}{Definition}

A \textbf{probabilistic program} is a stateful deterministic
computation $\mathcal{P}$:
\begin{itemize}
\item Initially, $\mathcal{P}$ expects no arguments.
\item On every call, $\mathcal{P}$ returns
    \begin{itemize}
        \item a distribution $F$,
        \item a distribution and a value $(G, y)$,
        \item a value $z$, 
        \item or $\bot$.
    \end{itemize}
\item Upon returning $F$, $\mathcal{P}$ expects $x\sim F$.
\item Upon returning $\bot$, $\mathcal{P}$ terminates.
\end{itemize}
A program is run by calling $\mathcal{P}$ repeatedly until termination.

The probability of each \textbf{trace} is $\propto\prod_{i=1}^{\left|\pmb{x}\right|} p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
\end{frame}

### 2016-09-01 slides/slides.tex
in progress

\documentclass{beamer}

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}
\usepackage{minted}

\newcommand {\I} {\ensuremath {\mathbf{1\hspace{-5.5pt}1}}}
\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}}

\title{Design and Implementation of\\Anglican Probabilistic Programming Language}
\author{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\begin{document}

\begin{frame}
\titlepage
\vfill
\center
Paper and slides: \url{https://bitbucket.org/probprog/anglican-white-paper}

\end{frame}

\section{Why probabilistic programming?}

\begin{frame}{Intuition}
    \textbf{Probabilistic program:}
\begin{itemize}
\item A program with random computations.
\item Distributions are conditioned by `observations'.
\item Values of certain expressions are `predicted' --- \textbf{the output}.
\end{itemize}

Can be written in any language (extended by \texttt{sample} and \texttt{observe}).
\end{frame}

\begin{frame}[fragile]{Example: Model Selection}
\begin{minted}[linenos, fontsize=\small]{clojure}
(let [;; Guessing a distribution
      dist (sample (categorical
                     [[normal 1] [gamma 1]
                      [uniform-continuous 1]
                      [uniform-discrete 1]]))
      a (sample (gamma 1 1))
      b (sample (gamma 1 1))
      d (dist a b)]

  ;; Observing samples from the distribution
  (loop [data data]
    (when (seq data)
      (let [[x & data] data]
        (observe d x))
      (recur data)))

  ;; Predicting a, b and the distribution
  (predict :a a) 
  (predict :b b)
  (predict :d d))
\end{minted}
\end{frame}

\begin{frame}{Definition}

A \textbf{probabilistic program} is a stateful deterministic
computation $\mathcal{P}$:
\pause
\begin{itemize}
\item Initially, $\mathcal{P}$ expects no arguments.
    \pause
\item On every call, $\mathcal{P}$ returns
    \begin{itemize}
        \item a distribution $F$,
        \item a distribution and a value $(G, y)$,
        \item a value $z$, 
        \item or $\bot$.
    \end{itemize}
    \pause
\item Upon returning $F$, $\mathcal{P}$ expects $x\sim F$.
    \pause
\item Upon returning $\bot$, $\mathcal{P}$ terminates.
\end{itemize}
\pause
A program is run by calling $\mathcal{P}$ repeatedly until termination.

The probability of each \textbf{trace} is $\propto\prod_{i=1}^{\left|\pmb{x}\right|} p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
\end{frame}

\begin{frame}{Evolution of Compiler Development}
    \begin{itemize} 
        \item 1950s: Ask a programmer to implement an algorithm efficiently: They'll write it on their own in assembly.
            \begin{itemize}
                \item   No good compilers; problem-dependent optimizations that only human expert could see.
            \end{itemize}
            \pause
        \item   1970s: Novice programmers use high-level languages and let compiler work out details, experts still write assembly.
            \begin{itemize}
                \item   Experts still write custom assembly when speed critical.
            \end{itemize}
            \pause
        \item   2000s: On most problems, even experts can't write faster assembly than optimizing compilers.
            \begin{itemize}
                \item   can automatically profile (JIT).
                \item   can take advantage of paralellization, complicated hardware, make appropriate choices w.r.t. caching.
                \item   Compilers embody decades of compiler research
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Evolution of Inference}

    \begin{itemize}
\item           2010: Ask a grad student to implement inference efficiently: They'll write it on their own.
    \begin{itemize}
\item   No good automatic inference engines; problem-dependent optimizations that only human expert can see.
    \end{itemize}
    \pause
\item   2015: Novice grad students use automatic inference engines and let compiler work out details, experts still write their own inference.
    \begin{itemize}
\item   Experts still write custom inference when speed critical.
    \end{itemize}
    \pause
\item   2020: On most problems, even experts can't write faster inference than mature automatic inference engines.
    \begin{itemize}
\item   Can use paralellization, sophisticated hardware
\item   Can automatically choose appropriate methods (meta-reasoning?).
\item   Inference engines will embody 1 decade (!) of PP research.
    \end{itemize}
    \end{itemize}
\end{frame}

\section{Why functional?}

\begin{frame}{Inference Objective}
\begin{itemize}
\item Suggest \textbf{most probable explanation} (MPE) - most likely assignment
  for all non-evidence variable given evidence.
  \pause
\item Approximately \textbf{compute integral} of the
  form $$\Phi=\int_{-\infty}^{\infty} \varphi(x)p(x) dx$$
  \pause
\item Continuously and \textbf{infinitely generate a sequence of samples} drawn
  from the distribution of the output expression --- so that someone
  else puts it in good use (vague but common). $\checkmark$
\end{itemize}
\end{frame}

\begin{frame}{Example: Inference Results}
  \begin{figure}[H]
      \includegraphics[scale=0.6]{models-results.pdf}
  \end{figure}
\end{frame}

\begin{frame}{Importance Sampling}
    \begin{algorithmic}
        \Loop
            \State Run $\mathcal{P}$, computing weight $w=\prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
            \State output $\pmb{z}, w$.
        \EndLoop
    \end{algorithmic}
    \begin{itemize}
        \item[] 
         \item Simple --- good.
         \item Slow convergence (unless one knows the answer) --- bad.
     \end{itemize}
     \vfill
     Can we do better?
\end{frame}

\begin{frame}{Lightweight Metropolis-Hastings (LMH)}
\begin{algorithmic}
    \State {Run $\mathcal{P}$ once, remember $\pmb{x}, \pmb{z}$.}
    \Loop
         \State {Uniformly select $x_i$}.
         \State {Propose a value for $x_i$.}
         \State {Run $\mathcal{P}$, remember $\pmb{x'}, \pmb{z'}$}.
         \State {Accept ($\pmb{x,z}=\pmb{x',z'}$) or reject with MH probability}.
         \State {Output $\pmb{z}$.}
    \EndLoop
\end{algorithmic}
\vfill
Can we do better?
\begin{itemize}
    \item Particle Markov Chain Monte Carlo
    \item Variational Inference
    \item ...

\end{itemize}
\end{frame}

\begin{frame}{Challenges}
    \begin{itemize}
        \item `Transformational compilation' --- limited languages, some less ugly than others.
        \item Slow inference --- Markov Chain Monte Carlo is \textbf{always slow} (but there are Hybrid Monte Carlo, Variational Inference, ...)
        \item How to learn persistently --- what is the learned model? Transfer learning?
        \item Handling indeterminism (learning policies).
    \end{itemize}
\end{frame}

\begin{frame}
    \LARGE
    \center
    Thank you!\\Questions?
\end{frame}

\end{document}

### 2016-09-01 slides/slides.tex
in progress

@@
 \center
-Paper and slides: \url{https://bitbucket.org/probprog/anglican-white-paper}
+\url{https://bitbucket.org/probprog/anglican-white-paper}\\
+\url{https://bitbucket.org/probprog/anglican} \\
+\url{http://www.robots.ox.ac.uk/~fwood/anglican/index.html}
 
@@
 
-\section{Why probabilistic programming?}
+\section{Motivation}
 
@@
 
-\begin{frame}{Definition}
-
-A \textbf{probabilistic program} is a stateful deterministic
-computation $\mathcal{P}$:
+\begin{frame}{More examples}
+    \begin{itemize}
+        \item \textit{Intruder detection} --- 
+            given a log of \textbf{times} and \textbf{amounts} of
+            payments in a bank account, how likely that the baccount was
+            compromised?
         \pause
-\begin{itemize}
-\item Initially, $\mathcal{P}$ expects no arguments.
+        \item \textit{Counterfactual reasoning} --- There are
+            \textbf{two routes} from Jerusalem to Tel Aviv:
+            1 and 443.  Based on traffic reports, I chose route
+            1 and was late. Would I arrive on time If I chose 443
+            instead?
         \pause 
-\item On every call, $\mathcal{P}$ returns
-    \begin{itemize}
-        \item a distribution $F$,
-        \item a distribution and a value $(G, y)$,
-        \item a value $z$, 
-        \item or $\bot$.
-    \end{itemize}
-    \pause
-\item Upon returning $F$, $\mathcal{P}$ expects $x\sim F$.
-    \pause
-\item Upon returning $\bot$, $\mathcal{P}$ terminates.
-\end{itemize}
-\pause
-A program is run by calling $\mathcal{P}$ repeatedly until termination.
-
-The probability of each \textbf{trace} is $\propto\prod_{i=1}^{\left|\pmb{x}\right|} p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
-\end{frame}
-
-\begin{frame}{Evolution of Compiler Development}
-    \begin{itemize} 
-        \item 1950s: Ask a programmer to implement an algorithm efficiently: They'll write it on their own in assembly.
-            \begin{itemize}
-                \item   No good compilers; problem-dependent optimizations that only human expert could see.
-            \end{itemize}
-            \pause
-        \item   1970s: Novice programmers use high-level languages and let compiler work out details, experts still write assembly.
-            \begin{itemize}
-                \item   Experts still write custom assembly when speed critical.
-            \end{itemize}
-            \pause
-        \item   2000s: On most problems, even experts can't write faster assembly than optimizing compilers.
-            \begin{itemize}
-                \item   can automatically profile (JIT).
-                \item   can take advantage of paralellization, complicated hardware, make appropriate choices w.r.t. caching.
-                \item   Compilers embody decades of compiler research
-            \end{itemize}
+        \item \textit{(Due to Stuart Russell)} If you observe that a
+            student GPA is exactly 4.0 in a model of
+            transcripts of students from the USA (GPA's from
+            0.0 to 4.0) and India (GPA's from 0.0 to
+            10.0) what is the probability that the student
+            is from India? 
     \end{itemize}
 \end{frame}
-
-\begin{frame}{Evolution of Inference}
-
-    \begin{itemize}
-\item           2010: Ask a grad student to implement inference efficiently: They'll write it on their own.
-    \begin{itemize}
-\item   No good automatic inference engines; problem-dependent optimizations that only human expert can see.
-    \end{itemize}
-    \pause
-\item   2015: Novice grad students use automatic inference engines and let compiler work out details, experts still write their own inference.
-    \begin{itemize}
-\item   Experts still write custom inference when speed critical.
-    \end{itemize}
-    \pause
-\item   2020: On most problems, even experts can't write faster inference than mature automatic inference engines.
-    \begin{itemize}
-\item   Can use paralellization, sophisticated hardware
-\item   Can automatically choose appropriate methods (meta-reasoning?).
-\item   Inference engines will embody 1 decade (!) of PP research.
-    \end{itemize}
-    \end{itemize}
-\end{frame}
-
-\section{Why functional?}
 
@@
         \Loop
-            \State Run $\mathcal{P}$, computing weight $w=\prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})$.
-            \State output $\pmb{z}, w$.
+            \State Run program, computing weight based on observations.
+            \State Output result and weight.
         \EndLoop
@@
 \begin{algorithmic}
-    \State {Run $\mathcal{P}$ once, remember $\pmb{x}, \pmb{z}$.}
+    \State Run program once, remembering random choices.
     \Loop
-         \State {Uniformly select $x_i$}.
-         \State {Propose a value for $x_i$.}
-         \State {Run $\mathcal{P}$, remember $\pmb{x'}, \pmb{z'}$}.
-         \State {Accept ($\pmb{x,z}=\pmb{x',z'}$) or reject with MH probability}.
-         \State {Output $\pmb{z}$.}
+         \State Uniformly select one random choice.
+         \State Propose a new value for the choice.
+         \State Re-run the program.
+         \State Accept or reject with MH probability.
+         \State Output result.
     \EndLoop
@@
 \begin{itemize}
-    \item Particle Markov Chain Monte Carlo
-    \item Variational Inference
+    \item Particle methods
+    \item Variational inference
     \item ...
-
 \end{itemize}
@@
 
-\begin{frame}{Challenges}
+\begin{frame}{Why functional?}
+    We want a functional language because an inference algorithm
+    controls the execution:
     \begin{itemize}
-        \item `Transformational compilation' --- limited languages, some less ugly than others.
-        \item Slow inference --- Markov Chain Monte Carlo is \textbf{always slow} (but there are Hybrid Monte Carlo, Variational Inference, ...)
-        \item How to learn persistently --- what is the learned model? Transfer learning?
-        \item Handling indeterminism (learning policies).
+        \item A program is run many (often many hundreds of
+            thousands) of times (with almost any algorithm).
+        \item A program must be partially re-executed multiple
+            times from different positions (particle methods).
+        \item We want to reason about the distribution defined
+            by the program.
     \end{itemize}
 \end{frame}
+
+\begin{frame}{Why Closure?}
+\begin{itemize}
+    \item Runs on JVM --- easy deployment and access to libraries.
+        \pause
+    \item A Lisp --- we (ab)use the macro facility.
+        \pause
+    \item Church (\url{https://en.wikipedia.org/wiki/Church_(programming_language)}) is derived from Scheme.
+\end{itemize}
+\pause
+Others use:
+    \begin{itemize}
+        \item Scheme --- (Church, Venture).
+        \item Scala --- Figaro.
+        \item Haskell --- Hakaru, Model-Bayes.
+        \item ...
+    \end{itemize}
+    \pause
+    As well as Python, C\#,  and other languages.
+\end{frame}
+
+\section{Design Outline}
+
+\subsection{Language}
+
+\subsection{Macro-based compilation}
+
+\subsection{Implementation Highlights}
+
+\subsection{Managing stack size}
+
+\subsection{Probabilistic forms}
+
+\subsection{Memoization}
+
+\section{Inference Algorithms}
+
+\section{Definitions and Runtime Library}
 

### 2016-09-01 slides/slides.tex
proper structure

@@
       d (dist a b)]
-
   ;; Observing samples from the distribution
@@
       (recur data)))
-
   ;; Predicting a, b and the distribution
@@
 
-\subsection{Language}
+\begin{frame}{Language}
+\end{frame}
 
-\subsection{Macro-based compilation}
+\begin{frame}{Macro-based compilation}
+\end{frame}
 
-\subsection{Implementation Highlights}
+\section{Implementation Highlights}
 
-\subsection{Managing stack size}
+\begin{frame}{Managing stack size}
+\end{frame}
 
-\subsection{Probabilistic forms}
+\begin{frame}{Probabilistic forms}
+\end{frame}
 
-\subsection{Memoization}
+\begin{frame}{Memoization}
+\end{frame}
 

### 2016-09-01 slides/slides.tex
added worksheet

@@
 
-\begin{frame}{Language}
+\begin{frame}{Anglican and Clojure}
+    On language on top of (or besides) another:
+    \begin{itemize}
+        \item Interpeter.
+        \item Source-to-source compiler.
+        \item Integrated language. $\checkmark$
+    \end{itemize}
+    \pause
+    \bigskip
+    Integrated language
+    \begin{itemize}
+        \item either extends the syntax of the host language;
+        \item or changes the semantics of the host language
+            constructs $\checkmark$.
+    \end{itemize}
+    \pause
+    \bigskip
+    Anglican is 
+    \begin{itemize}
+        \item Integrated with Clojure.
+        \item Shares syntax.
+        \item Alters operational semantics.
+    \end{itemize}
 \end{frame}
 
-\begin{frame}{Macro-based compilation}
+\begin{frame}{Anglican}
+    A subset of Clojure, wrapped inside \texttt{defquery}:
+    \begin{itemize}
+        \item \texttt{if}, \texttt{when}, \texttt{cond},
+            \texttt{case}, \texttt{let}, \texttt{and}, \texttt{or},
+            \texttt{fn}. 
+        \item Vector destructuring in bindings of \texttt{let} and \texttt{fn}.
+        \item Compound literals for vectors, hash maps, and sets.
+        \item \texttt{loop}/\texttt{recur} --- a convenience.
+    \end{itemize}
+    \bigskip
+    Core library:
+    \begin{itemize}
+        \item All of Clojure core library, except for
+            higher-order functions.
+        \item  \texttt{map}, \texttt{reduce},
+\texttt{filter}, \texttt{some}, \texttt{repeatedly},
+\texttt{comp}, \texttt{partial}.
+    \end{itemize}
+    \bigskip
+    Any Clojure function can be called from Anglican.
+\end{frame}
+
+\begin{frame}[fragile]{Macro-based compilation}
+    Anglican code macro-compiled into Clojure:
+    \begin{minipage}{0.48\textwidth}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+(loop [data data]
+  (if (seq data)
+    (let [[x & data] data]
+      (observe (gamma a b) x)
+      (recur data))
+    (predict :a a)))
+\end{minted}
+    \end{minipage}
+    \begin{minipage}{0.44\textwidth}
+    \vspace{2em}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+(fn loop [C23151 state data]
+   (if (seq data)
+     (let [[x & data] data]
+       (->observe 'O23153 (gamma a b) x
+         (fn do23152 [_ state]
+           (loop C23151 state data))
+         state))
+     (fn []
+       (C23151
+         nil
+         (add-predict state :a a)))))
+\end{minted}
+    \end{minipage}
 \end{frame}

### 2016-09-01 paper/paper.tex
typo fixed
added worksheet
in good progres

@@
 of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
-% Although these generic techniques are not always as statistically
-% efficient as techniques that take advantage of model-specific
-% optimizations, probabilistic programming makes it easier to optimize
-% models for a specific application in a manner that is efficient in
-% terms of the dimensionality of its latent variables.
 These generic techniques may be less statistically efficient
@@
 always have the same eye color. This can be achieved
-memoized functions. In Anglican, one might write:
+through the use of memoized functions. In Anglican, one
+might write:
 \begin{lstlisting}[style=default]
@@
         foo
-        ;; memoize result in continuation
+        ;; memoize result in state
         (fn [V $state]

### 2016-09-01 slides/slides.tex
in good progres

@@
 \begin{frame}{Anglican and Clojure}
-    On language on top of (or besides) another:
+    One language on top of (or besides) another:
     \begin{itemize}
@@
 
-\begin{frame}{Managing stack size}
+\begin{frame}[fragile]{Probabilistic forms}
+    Two \textbf{probabilistic} forms:
+    \begin{itemize}
+        \item \texttt{sample} --- draws a value from a
+            distribution;
+        \item \texttt{observe} --- conditions the \textit{a
+            posteriori} distribution by observing a value from a
+            distribution.
+    \end{itemize}
+    \bigskip
+    This is where the inference algorithm steps in:
+\begin{minted}[fontsize=\small]{clojure}
+=> (cps-of-expression '(sample dist) 'cont)
+(->sample dist cont $state)
+
+=> (cps-of-expression '(observe dist val) 'cont)
+(->observe dist val cont $state)
+\end{minted}
 \end{frame}
 
-\begin{frame}{Probabilistic forms}
+\begin{frame}[fragile]{Memoization}
+Memoization of random choice:
+\begin{itemize}
+    \item A person has a random eye color.
+    \item But the \textit{same} person has a fixed eye color.
+\end{itemize}
+\begin{minted}[linenos, fontsize=\small]{clojure}
+(let [eye-color (mem (fn [person] 
+                       (sample 
+                         (categorical
+                           ['brown 0.5]
+                           ['green 0.5]))))]
+  (if (not= (eye-color 'bill) (eye-color 'john))
+    (predict (eye-color 'bill))
+    (predict (eye-color 'john))))
+\end{minted}
 \end{frame}
 
-\begin{frame}{Memoization}
+\begin{frame}[fragile]{Compiling a memoized function}
+    \vspace{-2em}
+    \begin{minipage}{0.38\textwidth}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+(mem (fn [person] ...))
+\end{minted}
+    \end{minipage}
+    \begin{minipage}{0.52\textwidth}
+    \vspace{2em}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+(fn []
+ (cont
+  ;; every memoization gets a unique key  
+  (let [M23145 (gensym "M")]
+   (fn [C23144 $state & P23147]
+    (if (in-mem? $state M23145 P23147)
+     ;; previously memoized result
+     (fn []
+       (C23144
+         (get-mem $state M23145 P23147)
+         $state))
+     ;; new computation
+     (clojure.core/apply
+      (fn [C23150 $state person]
+        (fn [] (C23150 ... $state)))
+      ;; memoize result in state
+      (fn [V23146 $state]
+       (fn []
+        (C23144
+          V23146
+          (set-mem $state M23145 P23147 V23146))))
+      $state
+      P23147))))
+  $state))
+\end{minted}
+    \end{minipage}
+\end{frame}
+
+\begin{frame}[fragile]{Managing stack size}
+    \begin{itemize}
+        \item Clojure does not support tail-call optimization (TCO).
+        \item Under CPS transformation the stack will explode.
+            \pause
+        \item Anglican uses \textit{trampolining}: every continuation
+            call is wrapped into a \textit{thunk}.
+    \end{itemize}
+
+    The thunk is returned ...
+
+    \begin{minipage}{0.48\textwidth}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+    (fn [x y] (+ x y))
+\end{minted}
+    \end{minipage}
+    \begin{minipage}{0.48\textwidth}
+        \vspace{0.5em}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+    (fn [cont $state x y] 
+      (fn []
+        (cont (+ x y) $state)))
+\end{minted}
+      \vspace{0.5em}
+    \end{minipage}
+
+    ... and  called by \texttt{trampoline}:
+
+\begin{minted}[linenos, fontsize=\small]{clojure}
+        (defn exec
+          [algorithm prog value state]
+          (loop [step (trampoline prog value state)]
+            (let [next (checkpoint algorithm step)]
+              (if (fn? next)
+                (recur (trampoline next))
+                next))))
+\end{minted}
+
 \end{frame}

### 2016-09-01 slides/slides.tex
drafted, pause

@@
 
-\section{Definitions and Runtime Library}
+\begin{frame}[fragile]{Importance sampling}
+
+Importance sampling is the simplest:
+\begin{minted}[linenos, fontsize=\small]{clojure}
+(defmethod infer :importance [_ prog value & {}]
+  (letfn
+    [(sample-seq []
+       (lazy-seq
+         (cons
+           (:state (exec ::algorithm prog value
+                         initial-state))
+           (sample-seq))))]
+    (sample-seq)))
+\end{minted}
+Default checkpoint handlers are called:
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+(defmethod checkpoint [::algorithm anglican.trap.observe] [_ obs]
+  #((:cont obs) nil (add-log-weight (:state obs)
+                                    (observe* (:dist obs) (:value obs)))))
+
+(defmethod checkpoint [::algorithm anglican.trap.sample] [_ smp]
+  #((:cont smp) (sample* (:dist smp)) (:state smp)))
+\end{minted}
+\end{frame}
+
+\begin{frame}{Other inference algorithms}
+    \begin{itemize}
+        \item Implement \texttt{infer}.
+        \item Redefine one or both checkpoints.
+        \item An average implementation is $\approx 150$ lines of source
+            code.
+        \item A dozen different inference algorithms are in the
+            code base.
+        \item Half of them are really useful.
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Recap}
+    \begin{itemize}
+        \item Anglican is integrated with Clojure.
+        \item Shares syntax but alters semantics.
+        \item Macro-compiled.
+        \item Efficient.
+        \item Makes implementing AND using inference easy.
+    \end{itemize}
+\end{frame}
 

### 2016-09-02 slides/slides.tex
more slides

@@
 
-\begin{frame}{Anglican}
+\begin{frame}{Design challenges and choices}
+\begin{itemize}
+    \item Anglican syntax: Clojure + probabilistic constructs.
+		\begin{itemize}
+			\item Subset of Clojure.
+			\item Special forms \texttt{sample} and \texttt{observe}.
+			\item Macros to delimit Anglican code within a Clojure program.
+		\end{itemize}
+		\pause
+		\bigskip
+    \item Source-to-source compilation of Anglican into Clojure.
+		\begin{itemize}
+			\item CPS Transformation, with some tricks.
+			\item Transparent use of Clojure functions from Anglican.
+		\end{itemize}
+		\pause
+		\bigskip
+    \item Inference algorithms.
+		\begin{itemize}
+            \item Required to run an Anglican program.
+			\item Step in at checkpoints (\texttt{sample} and \texttt{observe}).
+			\item Execute programs by calling continuations repeatedly.
+		\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{The language}
     A subset of Clojure, wrapped inside \texttt{defquery}:

### 2016-09-02 slides/slides.tex
seems to becoming ready

@@
             0.0 to 4.0) and India (GPA's from 0.0 to
-            10.0) what is the probability that the student
+            10.0), what is the probability that the student
             is from India? 
@@
 		\begin{itemize}
-			\item CPS Transformation, with some tricks.
+			\item CPS transformation, with some tricks.
 			\item Transparent use of Clojure functions from Anglican.
@@
 \begin{minted}[linenos, fontsize=\scriptsize]{clojure}
-(fn loop [C23151 state data]
+(fn loop [C23151 $state data]
    (if (seq data)
@@
        (->observe 'O23153 (gamma a b) x
-         (fn do23152 [_ state]
-           (loop C23151 state data))
-         state))
+         (fn do23152 [_ $state]
+           (loop C23151 $state data))
+         $state))
      (fn []
@@
          nil
-         (add-predict state :a a)))))
+         (add-predict $state :a a)))))
 \end{minted}
@@
 \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Managing stack size}
+    \begin{itemize}
+        \item Clojure does not support tail-call optimization (TCO).
+        \item Under CPS transformation the stack will explode.
+            \pause
+        \item Anglican uses \textit{trampolining}: every continuation
+            call is wrapped into a \textit{thunk}.
+    \end{itemize}
+
+    The thunk is returned ...
+
+    \begin{minipage}{0.48\textwidth}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+    (fn [x y] (+ x y))
+\end{minted}
+    \end{minipage}
+    \begin{minipage}{0.48\textwidth}
+        \vspace{0.5em}
+\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
+    (fn [cont $state x y] 
+      (fn []
+        (cont (+ x y) $state)))
+\end{minted}
+      \vspace{0.5em}
+    \end{minipage}
+    \pause
+    ... and  called by \texttt{trampoline}:
+
+\begin{minted}[linenos, fontsize=\small]{clojure}
+        (defn exec
+          [algorithm prog value state]
+          (loop [step (trampoline prog value state)]
+            (let [next (checkpoint algorithm step)]
+              (if (fn? next)
+                (recur (trampoline next))
+                next))))
+\end{minted}
+
 \end{frame}
@@
 
-\begin{frame}[fragile]{Managing stack size}
-    \begin{itemize}
-        \item Clojure does not support tail-call optimization (TCO).
-        \item Under CPS transformation the stack will explode.
-            \pause
-        \item Anglican uses \textit{trampolining}: every continuation
-            call is wrapped into a \textit{thunk}.
-    \end{itemize}
-
-    The thunk is returned ...
-
-    \begin{minipage}{0.48\textwidth}
-\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
-    (fn [x y] (+ x y))
-\end{minted}
-    \end{minipage}
-    \begin{minipage}{0.48\textwidth}
-        \vspace{0.5em}
-\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
-    (fn [cont $state x y] 
-      (fn []
-        (cont (+ x y) $state)))
-\end{minted}
-      \vspace{0.5em}
-    \end{minipage}
-
-    ... and  called by \texttt{trampoline}:
-
-\begin{minted}[linenos, fontsize=\small]{clojure}
-        (defn exec
-          [algorithm prog value state]
-          (loop [step (trampoline prog value state)]
-            (let [next (checkpoint algorithm step)]
-              (if (fn? next)
-                (recur (trampoline next))
-                next))))
-\end{minted}
-
-\end{frame}
-
 \section{Inference Algorithms}

### 2016-09-02 slides/slides.tex
for presentation

@@
 \begin{frame}[fragile]{Macro-based compilation}
-    Anglican code macro-compiled into Clojure:
+    Anglican code macro-compiled into Clojure, in CPS:
     \begin{minipage}{0.48\textwidth}
@@
     \end{minipage}
+	\bigskip
+
+	A continuation accepts \textit{two} arguments:
+	\begin{itemize}
+		\item The computed value.
+		\item The internal state (\texttt{\$state}).
+	\end{itemize}
+	Functions accept continuation and state as two first arguments.
 \end{frame}
@@
         \item Redefine one or both checkpoints.
-        \item An average implementation is $\approx 150$ lines of source
+        \item An average implementation is $\approx 150$ lines of Clojure
             code.

### 2016-09-02 slides/slides.tex
as presented

@@
     \end{itemize}
+	\bigskip
+	We need
+	\begin{itemize}
+		\item First-class functions.
+		\item No side effects.
+	\end{itemize}
 \end{frame}
@@
         \item or changes the semantics of the host language
-            constructs $\checkmark$.
+            constructs. $\checkmark$
     \end{itemize}
@@
 \begin{minted}[linenos, fontsize=\scriptsize]{clojure}
-(fn loop [C23151 $state data]
+((fn loop [C23151 $state data]
    (if (seq data)
@@
          (add-predict $state :a a)))))
+  cont $state data)
 \end{minted}

### 2016-11-20 paper/paper.tex
typo fixed and simpler language

@@
 as a structured value, such as a vector, and the components
-of this value gets pattern matched to the corresponding
+of this value are matched to the corresponding
 parameters of the query via the destructuring mechanism of Clojure.

### 2016-11-20 paper/paper.tex
added self-citation to Output-Sensitive
describing the addressing scheme

@@
 to get lost. On the other hand, it is a good overview or guide
-for those who want to study the implemenation of Anglican. One
+for those who want to study the implementation of Anglican. One
 way to address this presentation issue while keeping the benefits
@@
 various functions are implemented in each inference algorithm.
-Finally, we show how these algorithms work on cps-tranformed 
+Finally, we show how these algorithms work on cps-transformed 
 examples.}
+
+\subsection{Addressing of checkpoints}
+
+Many inference algorithms memoize and reuse earlier computations
+at checkpoints: variants of Metropolis-Hastings reuse previously
+drawn values~\cite{WSG11}, as well as additional information
+used for adaptive sampling~\cite{TMP+15} at \texttt{sample}
+checkpoints, Asynchronous SMC (Particle Cascade) computes
+average particle weights at \texttt{observe}
+checkpoints~\cite{PWD+14}. Implementations of black-box
+variational inference~\cite{WW13,MPT+16} associate with random
+variables the learned parameters of variational posterior
+distribution.
+
+Checkpoints can be uniquely named at the compilation time;
+however, at runtime a checkpoint corresponding to a single code
+location may occur multiple time due to recurrent invocation of
+the function containing the checkpoint. Every unique occurrence
+of a checkpoint must receive a different address. An addressing
+scheme based on computing of stack addresses of checkpoints was
+described in the context of Lightweight
+Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
+over the naive scheme where dynamic occurrences of checkpoints
+are numbered sequentially, however it impacts the 
+performance of probabilistic programs because of the computation
+cost of computing the stack addresses:
+\begin{enumerate}
+    \item The address size is linear in stack depth.
+    \item Every function call must be augmented by symbolic
+        information required to compute stack addresses.
+\end{enumerate}
+In addition, the above scheme can still lead to inferior
+correspondence between checkpoints in different traces: in
+Anglican and other probabilistic languages where distributions
+are first-class objects checkpoints with incompatible arguments 
+can correspond to the same stack address. Consider the following
+program fragment:
+\begin{lstlisting}[style=default]
+(let [dist (if use-gamma
+               (gamma 2. 2.)
+               (normal 0. 1.))
+      x (sample dist)]
+  (predict x))
+\end{lstlisting}
+The \texttt{sample} checkpoint has the same stack address in
+different traces, however the random values should not be
+re-used between different distributions. Further on, in some
+algorithms, such as black-box variational inference, the role of 
+checkpoint addresses is semantic rather than heuristic ---
+appropriate correspondence must be established
+between checkpoints in different traces for the algorithm to
+work.
+
+To overcome the above problems, Anglican uses a different scheme
+which is almost as efficient as the scheme based on stack addresses for
+re-use of previously drawn values, produces addresses of constant size,
+and allows manual computation of checkpoint addresses at runtime
+when the default automatic scheme is insufficient.
+
+\begin{verbatim}
+A simple heuristic gives the same acceptance rate as with interpreted Anglican on all examples,
+including Neil’s Locomotion, as follows:
+
+  a) each sample form has a unique id, generated at compile time. It does anyway, and this 
+      has almost zero cost.
+  b) address is a pair 
+        [id-of-sample-form number-of-previous-occurences]
+      the occurences are of the same form.
+  c) if a sequence of random choices with the same id is interrupted, the count is padded up 
+  by a small power of 2, for example 16.
+
+For example, if random choices
+
+(sample S1)
+(sample ‘S2 …
+(sample ‘S2 …
+(sample ‘S1 …
+(sample ‘S1 …
+(sample ‘S1 …
+(sample ‘S2 …
+
+are encountered in the trace, then the addresses are:
+
+[S1 1]
+[S2 1]
+[S2 2]
+[S1 16]
+[S1 17]
+[S1 18]
+[S2 16]
+
+This way, every occurence of random choice has a unique address, but small disturbances (like
+removal or addition of a single choice of a particular kind) do not derail the entire sequence.
+This aligns well with recursive patterns used in DPmem, for example.
+
+I have some thoughts about stopping early in addition to starting late, but will do it when there
+are less deadlines. Right now, the code runs MUCH MUCH faster, it should take roughly 10
+hours to produce the same amount of data for Neil’s Locomotion example which was produced
+during 14 DAYS of the holidays. 
+\end{verbatim}
 

### 2016-11-20 paper/paper.tex
in progress

@@
 \begin{enumerate}
-    \item The address size is linear in stack depth.
+    \item On every function call, a component is added to the
+        address, hence the address size is linear in stack depth.
     \item Every function call must be augmented by symbolic
@@
 which is almost as efficient as the scheme based on stack addresses for
-re-use of previously drawn values, produces addresses of constant size,
+re-use of previously drawn values, as produces addresses of constant size,
 and allows manual computation of checkpoint addresses at runtime
-when the default automatic scheme is insufficient.
-
+when the default automatic scheme is insufficient. According to
+the scheme:
+\begin{itemize}
+    \item A checkpoint may accept an auxiliary argument --- the
+        checkpoint identifier. If specified, the identifier is the first
+        argument of a checkpoint. For example \texttt{(sample
+        'x1  (normal 0 1)} defines a \texttt{sample} checkpoint
+        with identifier \texttt{x1}.
+    \item If the identifier is omitted, a unique identifier is
+        generated automatically as a fresh symbol.
+    \item At runtime, the address of a checkpoint invocation has
+        the form
+        \texttt{[}\textit{checkpoint-identifier
+        number-of-previous-occurrences}\texttt{]}, where the
+        occurrences are of a checkpoint with the same checkpoint
+        identifier.
+    \item If a sequence of invocations of the same checkpoint is
+        interrupted by a different checkpoint, the number of
+        previous occurrences is rounded up to a multiple of
+        an integer. For efficiency, a small power of 2 is used
+        such as $2^4 = 16$.
+\end{itemize}
+
+For example, if checkpoint invocations
+\begin{lstlisting}[style=default]
+    (sample 'C1 ...)
+    (sample 'C2 ...) 
+    (sample 'C2 ...) 
+    (sample 'C1 ...)
+    (sample 'C1 ...)
+    (sample 'C1 ...)
+    (sample 'C2 ...) 
+    (sample 'C3 ...) 
+\end{lstlisting}
+are encountered in the trace, then the addresses are:
 \begin{verbatim}
-A simple heuristic gives the same acceptance rate as with interpreted Anglican on all examples,
-including Neil’s Locomotion, as follows:
-
-  a) each sample form has a unique id, generated at compile time. It does anyway, and this 
-      has almost zero cost.
-  b) address is a pair 
-        [id-of-sample-form number-of-previous-occurences]
-      the occurences are of the same form.
-  c) if a sequence of random choices with the same id is interrupted, the count is padded up 
-  by a small power of 2, for example 16.
-
-For example, if random choices
-
-(sample S1)
-(sample ‘S2 …
-(sample ‘S2 …
-(sample ‘S1 …
-(sample ‘S1 …
-(sample ‘S1 …
-(sample ‘S2 …
-
-are encountered in the trace, then the addresses are:
-
-[S1 1]
-[S2 1]
-[S2 2]
-[S1 16]
-[S1 17]
-[S1 18]
-[S2 16]
-
-This way, every occurence of random choice has a unique address, but small disturbances (like
-removal or addition of a single choice of a particular kind) do not derail the entire sequence.
-This aligns well with recursive patterns used in DPmem, for example.
-
-I have some thoughts about stopping early in addition to starting late, but will do it when there
-are less deadlines. Right now, the code runs MUCH MUCH faster, it should take roughly 10
-hours to produce the same amount of data for Neil’s Locomotion example which was produced
-during 14 DAYS of the holidays. 
+    [C1 0]
+    [C2 0]
+    [C2 1]
+    [C1 16]
+    [C1 17]
+    [C1 18]
+    [C2 16]
+    [C3 0]
 \end{verbatim}
+This way, each occurrence of a checkpoint has unique address,
+but small disturbances (like removal or addition of a single
+checkpoint with a particular identifier) are
+unlikely to derail the entire sequence.
 

### 2016-11-20 paper/paper.tex
removed typo

@@
 
-Checkpoints can be uniquely named at the compilation time;
+Checkpoints can be uniquely named at compilation time;
 however, at runtime a checkpoint corresponding to a single code

### 2016-11-21 paper/paper.tex
details on addressing scheme

@@
         previous occurrences is rounded up to a multiple of
-        an integer. For efficiency, a small power of 2 is used
+        an integer. For efficiency, a small power of 2 is used,
         such as $2^4 = 16$.
@@
 
-For example, if checkpoint invocations
+Consider the following simplified Anglican query:
+
+\begin{lstlisting}[style=default]
+(defm foo []
+  (if (sample 'C1 (flip 0.5))
+    (foo)
+    (bar)))
+
+(defm bar []
+   (case (sample 'C2 (uniform-discrete 0 3))
+     0 (bar)
+     1 (foo)
+     2 (sample 'C3 (normal 0. 1.))))
+
+(defquery baz
+   (predict (foo)))
+\end{lstlisting}
+
+An execution of the query may result in the following
+sequence of checkpoint invocations:
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-are encountered in the trace, then the addresses are:
+According to the addressing scheme, the addresses generated
+for this invocations are
 \begin{verbatim}
@@
 \end{verbatim}
+If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
+\begin{lstlisting}[style=default]
+    (sample 'C1 ...)
+    (sample 'C2 ...) 
+    (sample 'C1 ...)
+    (sample 'C1 ...)
+    (sample 'C2 ...) 
+    (sample 'C2 ...) 
+    (sample 'C3 ...) 
+\end{lstlisting}
+The addresses for the new sequence are
+\begin{verbatim}
+    [C1 0]
+    [C2 0]
+    [C1 16]
+    [C1 17]
+    [C2 16]
+    [C2 17]
+    [C3 0]
+\end{verbatim}
+Despite the change, the correspondence between checkpoints of
+similar types occurring in similar positions (relative positions
+in contiguous subsequences of a certain type) is preserved, and
+drawn values can be reused efficiently:
+
+\begin{center}
+\begin{tabular}{l l}
+    \textbf{old} & \textbf{new} \\ \hline
+    {}\texttt{[C1 0]} & \texttt{[C1 0]} \\
+    {}\texttt{[C2 0]} & \texttt{[C2 0]} \\
+    {}\texttt{[C2 1]} & \textit{unused} \\
+    {}\texttt{[C1 16]} & \texttt{[C1 16]} \\
+    {}\texttt{[C1 17]} & \texttt{[C1 17]} \\
+    {}\texttt{[C1 18]} & \textit{unused} \\
+    {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
+    \textit{redrawn} & \texttt{[C2 17]} \\
+    {}\texttt{[C3 0]} & \texttt{[C3 0]}
+\end{tabular}
+\end{center}
+
+Note that correspondence between checkpoints in different
+traces plays the role of an heuristic in Metropolis-Hastings
+family of algorithms. Any correspondence (or no correspondence,
+meaning all values must be re-drawn from their proposal
+distributions) is valid, and reused values from the previous
+invocation which are not in support or have zero probability in
+the new invocation are simply ignored and new values are
+re-drawn.
+
 This way, each occurrence of a checkpoint has unique address,

### 2016-11-21 paper/paper.tex
addressing of checkpoints

@@
 This way, each occurrence of a checkpoint has unique address,
-but small disturbances (like removal or addition of a single
-checkpoint with a particular identifier) are
-unlikely to derail the entire sequence.
+but small disturbances --- removal or addition of a single or just
+a few checkpoints --- are unlikely to derail the entire sequence.
+The probability of derailment depends on the padding. The
+padding can be safely, and without any impact on performance,
+set to rather large numbers. However, rounding up to a multiple
+of 16 proved to be appropriate for all practical purposes.
+
+Function \texttt{checkpoint-id} in the
+\texttt{anglican.inference} namespace automates generation of
+checkpoint addresses and can be used from any inference
+algorithm.
 

### 2016-11-21 paper/paper.tex
missing and drawn
performance comparison citations, some small edits

@@
 
+Comparisons of Anglican with other implementations of
+probabilistic programming languages~\cite{SGG15}\cite[pp.
+32--33]{P16} demonstrate that Anglican achieves state-of-the-art
+computational efficiency without sacrificing expressiveness.
 Anglican language syntax, compilation, invocation, and runtime
@@
 sections.
-
 \section{Design Outline}
@@
 
-\hy{I found this section confusing in the following sense. First,
-it does not highlight the most interesting aspects of the compilation
-of Anglican. It gives a good overview on how the compilation is
-implemented, but a reader might want to see something unique and
-interesting about Anglican. To me, the most interesting part of the
-compilation is the treatment of probabilistic constructs in Anglican's
-CPS transformation. How are sample,
-observe and stochastic memoization compiled? How should the standard
-CPS transformation be modified in order to handle these constructs?
-The use of state can be one answer for the second question.
-The second most interesting part is about the use of trampolining
-the mix with Clojure code.
-Second, the section assumes that a reader is already familiar 
-with Clojure or something specific to probabilistic programming. For 
-instance, it might be that a reader  has never heard about
-Clojure records and stochastic memoization. Then, she or he
-would have difficulties in following this section. 
-What should we do? One option is not to do anything. Another is
-to take care of the second point. That is, we make the section
-easily followed by someone who does not know Clojure. The third
-is to restructure the section and to take care of my first criticism.}
-
-\dt{I believe I partially addressed the above comment, but need to
-think more about it.}
-
 Compilation of Anglican into Clojure is built around a variant of CPS
@@
 beginning.
-
-\hy{This subsection is also not quite self-contained. Someone
-who is not familiar with Clojure or Anglican codebase is likely
-to get lost. On the other hand, it is a good overview or guide
-for those who want to study the implementation of Anglican. One
-way to address this presentation issue while keeping the benefits
-of the current text is to include some extended examples. For instance,
-in the beginning of the paper, we might include several
-Anglican examples. Then, we show that how these examples get
-cps-transformed. Next, we present the implementation of inference
-algorithms in Anglican using semi pseudo code; this is not quite
-an implementation in Clojure but gives a good sketch on how
-various functions are implemented in each inference algorithm.
-Finally, we show how these algorithms work on cps-transformed 
-examples.}
 
@@
 
-\begin{center}
 \begin{tabular}{l l}
@@
     {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
-    \textit{redrawn} & \texttt{[C2 17]} \\
+    \textit{missing} & \texttt{[C2 17]} \textit{drawn}\\
     {}\texttt{[C3 0]} & \texttt{[C3 0]}
 \end{tabular}
-\end{center}
 
@@
 provided important insights on structure and semantics of
-probabilistic concepts incorporated in Anglican. Efficiency
-and expressive power of Anglican owe to adherence to the
-functional approach as much as to rich inference opportunities
-of the Anglican environment.
+probabilistic concepts incorporated in Anglican. Computational
+efficiency and expressive power of Anglican owe to adherence to
+the functional approach as much as to rich inference
+opportunities of the Anglican environment.
 

### 2016-11-21 paper/paper.tex
introduction to inference algorithms

@@
 
-An inference algorithm is an implementation of the (ad-hoc)
-polymorphic function \texttt{infer} declared as Clojure's multimethod
-in the \texttt{anglican.inference} namespace. The
-function accepts an algorithm identifier, a query --- the
-probabilistic program in which to perform the inference, an
-initial value for the query, and optional algorithm parameters.
+A probabilistic program in Anglican may look almost like a
+Clojure program, however the purpose of executing a
+probabilistic program is different from that of a `regular'
+program: instead of producing the result of a single execution,
+a probabilistic program computes, exactly or approximately, the
+distribution from which execution results are drawn.  Computing
+the distribution is facilitated by an inference algorithm.
+
+Probabilistic programming system Anglican provides a variety of
+approximate inference algorithms. Ideally, Anglican should
+automatically choose the most appropriate algorithm for each
+probabilistic program. In practice, selecting an inference
+algorithm, or a combination thereof, is still a challenging
+task, and program structure, intended use of the computation
+results, performance, approximation accuracy, and other factors
+must be taken into consideration. New algorithms are being
+developed and added to Anglican~\cite{TMP+15,MPT+16,RNL+2016},
+as a part of ongoing research.
+
+In the implementation of Anglican, inference algorithms are
+instantiations of the (ad-hoc) polymorphic function
+\texttt{infer} declared as Clojure's multimethod in the
+\texttt{anglican.{\linebreak[0]}inference} namespace. The function accepts an
+algorithm identifier, a query --- the probabilistic program in
+which to perform the inference, an initial value for the query,
+and optional algorithm parameters.
 

### 2016-11-21 paper/paper.tex
spaces around tabular
case study in progress

@@
 
+\vspace{\baselineskip}
 \begin{tabular}{l l}
@@
 \end{tabular}
+\vspace{\baselineskip}
 
@@
 
+\section{Case Study}
+
+The description of Anglican language and environment would be
+incomplete without a case study showing the process of building
+and executing a solution of an inference problem. This case
+study takes a problem for which a solution is not immediately
+obvious, \textit{the Deli dilemma}, and guides through
+writing an Anglican program for the problem, executing the
+program, and post-processing results.
+
+\subsection{The deli dilemma}
+
+Imagine that we are facing the following dilemma:
+
+\begin{quote}
+A customer wearing round sunglasses came to a deli at 1:13pm,
+and grabbed a sandwitch and a coffee. Later on the same day, a
+customer wearing round sunglasses came at 6:09pm and ordered a
+dinner.  Was it the same customer?
+\end{quote}
+
+In addition to the story above, we know that:
+
+\begin{itemize}
+    \item There is an adjacent office quarter, and it takes
+        between 5 and 15 minutes on average to walk from an
+        office to the deli, were different average times are  for
+        different buildings in the quarter.
+    \item Depending on traffic lights, the walking time varies by about 2
+        minutes.
+    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
+    \item The waiter's odds that this is the same customer are 2 to 1.
+\end{itemize}
+
+\subsection{Anglican query}
+
+We want to formalize the dilemma in an Anglican query, based on
+the knowledge we have. Let us formalize the knowledge in
+Clojure (the times are in minutes). First, we encode our prior
+information which holds true independently of the customer's
+visit:
+
+\begin{lstlisting}[style=default]
+(def mean-time-to-arrive
+     "average time to arrive"
+     10.)
+(def sd-time-to-arrive
+     "standard deviation of arrival time"
+     3.)
+(def time-sd
+     "walking time deviation"
+     1.)
+\end{lstlisting}
+
+Then, we encode our observations, based on which we want to
+solve the dilemma:
+
+\begin{lstlisting}[style=default]
+(def lunch-delay
+     "time between lunch break and lunch order"
+     13.)
+(def dinner-delay
+     "time between end of day and dinner order"
+     9.)
+(def p-same
+     "prior probability of the same customer"
+     (/ 2. 3.))
+\end{lstlisting}
+
+For inference, one often chooses known distribution to
+represent uncertainty. We choose the normal
+distribution for representing uncertainty about average arrival
+time.
+
+\begin{lstlisting}[style=default]
+(def time-to-arrive-prior 
+     "prior distribution of average arrival time"
+     (normal mean-time-to-arrive
+             sd-time-to-arrive))
+\end{lstlisting}
+
+There are two possibilities, either the same customer visited
+the deli twice, or two different customers came to the deli,
+one for lunch, one for dinner. We define an Anglican function
+for each case. Note that this is the first time we switch from
+Clojure to Anglican.
+
+\begin{lstlisting}[style=default]
+(defm same-customer 
+  "observe the same customer twice"
+  [time-to-arrive-prior
+   lunch-delay
+   dinner-delay]
+  (let [time-to-arrive
+          (sample time-to-arrive-prior)]
+    (observe (normal time-to-arrive time-sd) 
+             lunch-delay)
+    (observe (normal time-to-arrive time-sd)
+             dinner-delay)
+    (predict :time-to-arrive time-to-arrive)))
+
+(defm different-customers
+  "observe different customers"
+  [time-to-arrive-prior
+   lunch-delay
+   dinner-delay]
+  (let [time-to-arrive-1
+          (sample time-to-arrive-prior)
+        time-to-arrive-2
+          (sample time-to-arrive-prior)]
+    (observe (normal time-to-arrive-1 time-sd) 
+             lunch-delay)
+    (observe (normal time-to-arrive-2 time-sd)
+             dinner-delay)
+    (predict :times-to-arrive
+      [time-to-arrive-1 time-to-arrive-2])))
+\end{lstlisting}
+
+Both functions have the same structure: we first `guess' the
+average arrival time and then observe the actual time from a
+distribution parameterized by the guessed time. However, in
+\texttt{same-customer} the average arrival time is the same for
+both the lunch and the dinner, while for
+\texttt{different-customers} two average arrival times are
+guessed independently.
+
+We are finally ready to define the query.
+
+\begin{lstlisting}[style=default]
+(defquery deli [time-to-arrive-prior
+                lunch-delay
+                dinner-delay]
+ (let [is-same-customer (sample (flip p-same))
+       observe-customer (if is-same-customer
+                           same-customer
+                           different-customer)]
+    (predict :is-same-customer is-same-customer)
+    (observe-customer time-to-arrive-prior
+                      lunch-delay
+                      dinner-delay)))
+\end{lstlisting}
+
+Performing inference on query \texttt{deli} should predict the
+probability that the same customer visited the deli twice, as
+well as probability distributions of average arrival times for
+both cases.
+
+\subsection{Inference}
+
 \section{Summary}

### 2016-11-21 paper/paper.tex
case study in progress

@@
 
-Then, we encode our observations, based on which we want to
+Then, we record our observations, based on which we want to
 solve the dilemma:
@@
 
+\begin{lstlisting}[style=default]
+(def samples (doquery :lmh deli nil))
+
+(def N 1000)
+(def predicts (map get-predicts (take N (drop N samples))))
+
+(def p-same+ (/ (count (filter :same-customer predicts)) (double N)))
+
+;; single customer                      
+(def time-to-arrive+ (map :time-to-arrive (filter :same-customer predicts)))
+(def mean-to-arrive+ (mean time-to-arrive+))
+(def sd-to-arrive+ (std time-to-arrive+))
+
+;; two customers
+(def times-to-arrive+ (map :times-to-arrive 
+                           (filter (complement :same-customer) predicts)))
+(def mean-1-to-arrive+ (mean (map first times-to-arrive+)))
+(def sd-1-to-arrive+ (std (map first times-to-arrive+)))
+(def mean-2-to-arrive+ (mean (map second times-to-arrive+)))
+(def sd-2-to-arrive+ (std (map second times-to-arrive+)))
+
+(def samples (doquery :bbvb deli nil))
+
+(clojure.pprint/pprint (anglican.bbvb/get-variational (nth samples N)))
+\end{lstlisting}
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
+\end{figure}
+
+\begin{verbatim}
+{S28219
+ {(0 anglican.runtime.flip-distribution)
+  {:p 0.11671977016875924,
+   :dist
+   {:min 0.0,
+    :max 1.0}}},
+ S28209
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 10.99753360180663,
+   :sd 0.7290976433082352}},
+ S28217
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 12.668050292254202,
+   :sd 0.9446695174790353}},
+ S28215
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 9.104132559955836,
+   :sd 0.9479290526821788}}}
+\end{verbatim}
+
 \section{Summary}

### 2016-11-22 paper/paper.tex
case study in progress

--- 2016-11-21 paper/paper.tex
@@
 
+\begin{lstlisting}[style=default]
+(def samples (doquery :lmh deli nil))
+\end{lstlisting}
+
+\begin{lstlisting}[style=default]
+(def N 1000)
+(def predicts (map get-predicts
+                   (take N (drop N samples))))
+\end{lstlisting}
+
+\begin{lstlisting}[style=default]
+(def p-same+ 
+     (/ (count (filter :same-customer predicts))
+        (double N)))
+\end{lstlisting}
+
+\begin{lstlisting}[style=default]
+;; single customer                      
+(def time-to-arrive+ 
+     (map :time-to-arrive
+          (filter :same-customer predicts)))
+(def mean-to-arrive+ (mean time-to-arrive+))
+(def sd-to-arrive+ (std time-to-arrive+))
+
+;; two customers
+(def times-to-arrive+ 
+     (map :times-to-arrive 
+          (filter (complement :same-customer)
+                  predicts)))
+(def mean-1-to-arrive+
+     (mean (map first times-to-arrive+)))
+(def sd-1-to-arrive+
+     (std (map first times-to-arrive+)))
+(def mean-2-to-arrive+
+     (mean (map second times-to-arrive+)))
+(def sd-2-to-arrive+
+     (std (map second times-to-arrive+)))
+\end{lstlisting}
+
+\begin{lstlisting}[style=default]
+(def samples (doquery :bbvb deli nil))
+\end{lstlisting}
+
+\begin{lstlisting}[style=default]
+(clojure.pprint/pprint
+  (anglican.bbvb/get-variational (nth samples N)))
+\end{lstlisting}
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
+\end{figure}
+
+\begin{verbatim}
+{S28209
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 10.99753360180663,
+   :sd 0.7290976433082352}},
+ S28217
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 12.668050292254202,
+   :sd 0.9446695174790353}},
+ S28215
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 9.104132559955836,
+   :sd 0.9479290526821788}},
+ S28219
+ {(0 anglican.runtime.flip-distribution)
+  {:p 0.11671977016875924,
+   :dist
+   {:min 0.0,
+    :max 1.0}}}}
+\end{verbatim}
+
 \section{Summary}

### 2016-11-22 paper/paper.tex
case study completed

@@
 \section{Introduction}
+\label{sec:intro}
+
 For data science practitioners, statistical inference is typically
@@
 sections.
+
 \section{Design Outline}
+\label{sec:design}
 
@@
 \section{Language}
+\label{sec:language}
 
@@
 \subsection{Core Library}
+\label{sec:core}
 
@@
 \subsection{Tail Call}
+\label{sec:tail}
 
@@
 \section{Macro-based Compilation}
+\label{sec:compilation}
 
@@
 \subsection {The State}
+\label{sec:state}
 
@@
 \subsection{Expression Kinds}
+\label{sec:expr}
 
@@
 \subsubsection{Literals}
+\label{sec:literals}
 
@@
 \subsubsection{Opaque Expressions}
+\label{sec:opaque}
 
@@
 \subsubsection{General Expressions}
+\label{seq:general}
 
@@
 \subsection{Implementation Highlights}
+\label{seq:highlights}
 
@@
 \subsubsection{Continuations}
+\label{seq:continuations}
 
@@
 \subsubsection{Primitive Procedures}
+\label{seq:primitive}
 
@@
 
-\subsubsection{Probabilistic forms}
+\subsubsection{Probabilistic Forms}
+\label{seq:forms}
 
@@
 \section{Inference Algorithms}
+\label{sec:inference}
 
@@
 
-\subsection{The \texttt{infer} function}
+\subsection{The \texttt{infer} Function}
+\label{sec:infer}
 
@@
 
-\subsection{Internals of an inference algorithm}
+\subsection{Internals of an Inference Algorithm}
+\label{seq:internals}
 
@@
 
-\subsection{Addressing of checkpoints}
+\subsection{Addressing of Checkpoints}
+\label{sec:addressing}
 
@@
 \section{Definitions and Runtime Library}
+\label{seq:runtime}
 
@@
 \section{Case Study}
+\label{seq:study}
 
@@
 
-\subsection{The deli dilemma}
+The program presented in this section is intentionally short and
+simple. Aglican is capable of compiling and running elaborated
+programs handling large amounts of data. Advanced examples of
+Anglican programs and inference can be found in literature on
+applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.
+
+\subsection{The Deli Dilemma}
+\label{seq:deli}
 
@@
 
-\subsection{Anglican query}
+\subsection{Anglican Query}
+\label{seq:deli-query}
 
@@
 
-There are two possibilities, either the same customer visited
+There are two possibilities: either the same customer visited
 the deli twice, or two different customers came to the deli,
-one for lunch, one for dinner. We define an Anglican function
+one for lunch, one for dinner. We define an \textbf{Anglican} function
 for each case. Note that this is the first time we switch from
-Clojure to Anglican.
+Clojure to Anglican. The functions must be written in Anglican
+(and hence defined using \texttt{defm} instead of \texttt{defn})
+because they contain probabilistic forms \texttt{sample} and
+\texttt{observe}.
 
@@
 
-We are finally ready to define the query.
+We are finally ready to define the query in \textbf{Anglican}.
 
@@
 
-Performing inference on query \texttt{deli} should predict the
+Performing inference on query \texttt{deli} predicts the
 probability that the same customer visited the deli twice, as
@@
 \subsection{Inference}
+\label{seq:deli-inference}
+
+Having defined the query, we are now ready to run the query
+using an inference algorithm. Function \texttt{doquery} provided
+by the \texttt{anglican.core} namespace accepts the inference
+algorithm, the query, and optional parameters, and returns a
+lazy sequence of samples. We use here the inference algorithm
+called Lightweight Metropolis-Hastings (LMH). The algorithm is
+somewhat slow to converge but can be used with any Anglican
+query, and should be robust enough for our simple problem.
+
+We bind the results of \texttt{doquery} to variable
+\texttt{samples}, to analyse the results later. However, since
+the sequence is lazy, no inference is performed and no samples
+are generated until they are retrieved and processed.
 
@@
 
-\begin{lstlisting}[style=default]
-(def N 1000)
+To approximate the inferred distribution, we extract a finite
+subsequence of samples. Many algorithms use an initial
+subsequence of samples to converge to the target distribution,
+hence we drop initial $N$ samples ($N$ is 5000 in the code
+snippet below), and collect the next $N$ samples.
+
+\begin{lstlisting}[style=default]
+(def N 5000)
 (def predicts (map get-predicts
@@
 \end{lstlisting}
+
+Based on the collected samples we compute an approximation of
+the posterior probability \texttt{p-same+} that the same
+customer visited the deli twice.
 
@@
 \end{lstlisting}
+
+With the specified observations, \texttt{-same+} is $\approx
+0.12$. The results may vary from run to run, and, for given
+algorithm, the accuracy depends on the number of samples we
+chose to retrieve. The probability is much lower than the
+waiter's guess \texttt{p-same} ($\frac 2 3$).
+
+In addition to computing the posterior probability that the same
+customer visited the deli twice, we may want to know the average
+time (or times) to arrive. In Bayesian inference, it is common
+to report distributions instead of `most likely' values. We use
+predicted values in retrieved samples to approximate the
+distributions, and plot distribution histograms for the same
+customer visiting twice (Figure~\ref{fig:dist-same}) and
+two different customers (Figure~\ref{fig:dist-diff})
 
@@
 
+In addition to plotting the distribution histograms, we use
+functions \texttt{mean}, \texttt{std} provided, along with other
+useful statistical functions, in the \texttt{anglican.{\linebreak[0]}stat}
+namespace. This is another illustration of advantage of tight
+integration between Clojure and Anglican --- probabilistic
+models are expressed in Anglican, however processing of data and
+results can rely on the full power of Clojure.
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
+    \caption{Arrival time distribution for a single customer.}
+    \label{fig:dist-same}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
+    \caption{Arrival time distributions for two customers.}
+    \label{fig:dist-diff}
+\end{figure}
+
+For inference, we chose to use  Lightweight
+Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
+probabilistic programming is that models are separated from
+inference. To switch to a different inference algorithm we just
+need to pass a different value to \texttt{doquery}. For example,
+we may decide to use Black-Box Variational Bayes (BBVB), which
+may not work equally well for all probabilistic programs, but is
+much faster to converge.
+
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
+
+We can still use samples and summary statistics with BBVB to
+approximate the posterior distribution, however variational
+inferences approximates the posterior by known distributions,
+and we can directly retrieve the distribution parameters.
 
@@
 \end{lstlisting}
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
-\end{figure}
 
@@
 
+We can guess that the variational distributions correspond to the
+prior distributions in the \texttt{sample} forms, in the order
+of appearance. However, it would help if we could use more
+informative labels instead of automatically generated symbols
+\texttt{S28209}, \texttt{S28217}, etc. Here the option to
+specify identifiers for probabilistic forms explicitly comes
+handy. If we modify the \texttt{sample} forms to use explicit identifiers
+(only the forms are shown for brevity), the output becomes much
+easier to analyse.
+
+\begin{lstlisting}[style=default]
+          (sample :arrival-time-same
+                  time-to-arrive-prior)]
+
+          (sample :arrival-time-first
+                  time-to-arrive-prior)
+
+          (sample :arrival-time-second
+                  time-to-arrive-prior)]
+
+  (let [is-same-customer (sample :same-or-different
+                                 (flip p-same))
+\end{lstlisting}
+
+The output becomes much easier to interpret and analyse
+programmatically.
+
+\begin{verbatim}
+{:arrivate-time-same
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 10.99753360180663,
+   :sd 0.7290976433082352}},
+ :arrival-time-first
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 12.668050292254202,
+   :sd 0.9446695174790353}},
+ :arrival-time-second
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 9.104132559955836,
+   :sd 0.9479290526821788}},
+ :same-or-different
+ {(0 anglican.runtime.flip-distribution)
+  {:p 0.11671977016875924,
+   :dist
+   {:min 0.0,
+    :max 1.0}}}}
+\end{verbatim}
+
+This completes the case study, where we showed a probabilistic
+programming solution of a problem, implemented and analysed in
+Anglican and Clojure, using two different inference algorithms.
+
 \section{Summary}
+\label{seq:summary}
 

### 2016-11-22 paper/paper.tex
case study completed

@@
 approximate the posterior distribution, however variational
-inferences approximates the posterior by known distributions,
+inference approximates the posterior by known distributions,
 and we can directly retrieve the distribution parameters.

### 2016-11-22 paper/paper.tex
no predict in the paper

@@
 
-An Anglican program, or \textit{query}, is compiled into a Clojure
-function. When inference is performed with a provided algorithm, this
-produces a sequence of output values, or \textit{predicts}. Anglican
-shares a common syntax with Clojure; Clojure functions can be called
-from Anglican code and vice versa. A simple program in Anglican can
-look like the following code:
+An Anglican program, or \textit{query}, is compiled into a
+Clojure function. When inference is performed with a provided
+algorithm, this produces a sequence of samples. Anglican shares
+a common syntax with Clojure; Clojure functions can be called
+from Anglican code and vice versa. A simple program in Anglican
+can look like the following code:
 \begin{lstlisting}[style=default]
@@
 
-    ;; Predicting a, b and the distribution
-    (predict :a a)
-    (predict :b b)
-    (predict :d d)))
-\end{lstlisting}
-This query defines a probability distribution on three predicted
+    ;; Returning the distribution and parameters
+    [d a b]))
+\end{lstlisting}
+This query defines a probability distribution on three
 variables, $\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
@@
 written in Anglican, may include probabilistic forms
-\texttt{sample} and \texttt{observe} (as well \texttt{predict}
-for the output), and can be seamlessly called from inside
-Anglican queries, just like functions locally defined within the
-same query.
+\texttt{sample} and \texttt{observe}, and can be seamlessly
+called from inside Anglican queries, just like functions locally
+defined within the same query.
 
@@
   {:log-weight 0.0
-   :predicts []
    :result nil
@@
     \item function applications and \texttt{apply};
-    \item probabilistic forms --- \texttt{predict},
-        \texttt{observe}, \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
+    \item probabilistic forms --- \texttt{observe},
+        \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
 \end{itemize}
@@
 In addition to checkpoints, there are a few other special forms
---- \texttt{predict}, \texttt{store}, \texttt{retrieve},
-\texttt{mem} --- which modify program state. These forms are
-translated into expressions involving calls of functions from the
+---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
+modify program state. These forms are translated into
+expressions involving calls of functions from the
 \texttt{anglican.state} namespace. The \texttt{mem} form, which
@@
   (if (not= (eye-color 'bill) (eye-color 'john))
-    (predict (eye-color 'bill))
-    (predict (eye-color 'john))))
+    (eye-color 'bill)
+    (eye-color 'john)))
 \end{lstlisting}
@@
 (let [x 1]
-  (infer :pgibbs (query (predict x)) nil))
+  (infer :pgibbs (query x) nil))
 \end{lstlisting}
@@
 (defquery my-query [mean sd]
-  (predict (sample (normal mean sd))))
+  (sample (normal mean sd)))
 
@@
                (gamma 2. 2.)
-               (normal 0. 1.))
-      x (sample dist)]
-  (predict x))
+               (normal 0. 1.))]
+  (sample dist))
 \end{lstlisting}
@@
 (defquery baz
-   (predict (foo)))
+  (foo))
 \end{lstlisting}
@@
              dinner-delay)
-    (predict :time-to-arrive time-to-arrive)))
+    [time-to-arrive]))
 
@@
              dinner-delay)
-    (predict :times-to-arrive
-      [time-to-arrive-1 time-to-arrive-2])))
+    [time-to-arrive-1 time-to-arrive-2]))
 \end{lstlisting}
@@
                             different-customer)]
-    (predict :is-same-customer is-same-customer)
-    (observe-customer time-to-arrive-prior
+    {:is-same-customer is-same-customer
+     :times-to-arrive (observe-customer 
+                        time-to-arrive-prior
                         lunch-delay
-                      dinner-delay)))
-\end{lstlisting}
-
-Performing inference on query \texttt{deli} predicts the
+                        dinner-delay)}))
+\end{lstlisting}
+
+Performing inference on query \texttt{deli} computes the
 probability that the same customer visited the deli twice, as
@@
 (def N 5000)
-(def predicts (map get-predicts
+(def results (map get-result
                    (take N (drop N samples))))
@@
 (def p-same+ 
-     (/ (count (filter :same-customer predicts))
+     (/ (count (filter :same-customer results))
         (double N)))
@@
 to report distributions instead of `most likely' values. We use
-predicted values in retrieved samples to approximate the
+query results in retrieved samples to approximate the
 distributions, and plot distribution histograms for the same
@@
 (def time-to-arrive+ 
-     (map :time-to-arrive
-          (filter :same-customer predicts)))
+     (map (comp first :times-to-arrive)
+          (filter :same-customer results)))
 (def mean-to-arrive+ (mean time-to-arrive+))
@@
           (filter (complement :same-customer)
-                  predicts)))
+                  results)))
 (def mean-1-to-arrive+

### 2016-11-26 paper/paper.tex
minor

@@
 These generic techniques may be less statistically efficient
-than techniques tailored to a specific model, however
+than techniques tailored to a specific model. However,
 probabilistic programming facilitates simpler implementations of
@@
 point on.  Different executions may employ different random
-choices, however for efficient inference a correspondence
+choices. However, for efficient inference a correspondence
 between random choices in different executions should be
@@
 possible to support almost full Clojure by expanding all macros
-in the Anglican source code, however in Clojure, unlike in
+in the Anglican source code. However, in Clojure, unlike in
 Scheme~\cite{SDF+10}, or Common Lisp~\cite{PC94}, the result
@@
 current continuation and the state are passed to the called
-function. Anglican can also call Clojure functions; however
+function. Anglican can also call Clojure functions; however,
 Clojure functions do not expect these auxiliary parameters.
@@
 option would be to use meta-data to identify Anglican function
-calls at runtime, however this would impact performance, and
+calls at runtime. However, this would impact performance, and
 a good runtime performance is critical for probabilistic
@@
 A probabilistic program in Anglican may look almost like a
-Clojure program, however the purpose of executing a
+Clojure program. However, the purpose of executing a
 probabilistic program is different from that of a `regular'
@@
 over the naive scheme where dynamic occurrences of checkpoints
-are numbered sequentially, however it impacts the
+are numbered sequentially. However, it impacts the
 performance of probabilistic programs because of the computation
@@
 The \texttt{sample} checkpoint has the same stack address in
-different traces, however the random values should not be
+different traces. However, the random values should not be
 re-used between different distributions. Further on, in some
@@
 integration between Clojure and Anglican --- probabilistic
-models are expressed in Anglican, however processing of data and
+models are expressed in Anglican. However, processing of data and
 results can rely on the full power of Clojure.
@@
 We can still use samples and summary statistics with BBVB to
-approximate the posterior distribution, however variational
+approximate the posterior distribution. However, variational
 inference approximates the posterior by known distributions,

### 2016-11-26 paper/paper.tex
update

@@
 data I/O, pre-processing, and analysis and visualization of inference results.
-In this demonstration, we describe the implementation of
+In this paper, we describe the implementation of
 Anglican~\cite{TMW15, Anglican}, a probabilistic
@@
 make use of a rich set of libraries written in both Clojure and Java.
-Conversely Anglican allows intuitive and compact specification of models for
+Conversely, Anglican allows intuitive and compact specification of models for
 which inference may be performed as part of a larger Clojure project.
@@
 There are several ways to build a programming language
-on top or besides another language.  The easiest to grasp is an
+on top of or besides another language.  The easiest is an
 interpreter --- a program that reads a program, in its entirety

### 2016-11-26 paper/paper.tex
update

@@
     ;; Observing samples from the distribution
-    (loop [data data]
-      (when (seq data)
-        (let [[x & data] data]
+    (loop [data0 data]
+      (when (not-empty data0)
+        (let [[x & data1] data0]
           (observe d x))
-        (recur data)))
+        (recur data1)))
 
@@
 particle filtering inference algorithms \cite{WVM14} is a non-deterministic
-control statement at which a particle can be either replicated
-or terminated. Similarly, in Metropolis-Hastings \cite{WSG11},
+control statement at which a thread (also called particle) 
+can be either forked or terminated. Similarly, in 
+Metropolis-Hastings \cite{WSG11},
 \texttt{sample} is both an input

### 2016-11-26 paper/paper.tex
update

@@
 particle filtering inference algorithms \cite{WVM14} is a non-deterministic
-control statement at which a thread (also called particle) 
-can be either forked or terminated. Similarly, in 
+control statement at which a particle (corresponding to
+a thread executing a program)
+can be either replicated or terminated. Similarly, in 
 Metropolis-Hastings \cite{WSG11},

### 2016-11-26 paper/paper.tex
adding words for the first program

@@
 \end{lstlisting}
-This query defines a probability distribution on three
-variables, $\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
+The query builds a model for the input $\data$, a sequence of data points.
+It defines a probability distribution on three variables, 
+$\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
 for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
-positive parameters for the type. Samples from this distribution
+positive parameters for the type. Concretely,
+using the $\texttt{sample}$ statements, the query
+first defines so called a prior distribution on these three variables,
+and then it adjust this prior distribution based on data points in
+$\texttt{data}$ using the \texttt{observe} statement.
+Samples from this adapted distribution (also called posterior distribution)
 can be obtained by running the query under one of Anglican's

### 2016-11-26 paper/paper.tex
update
12 pages filled, program explained in the comments

@@
 \documentclass[preprint]{sigplanconf}
+\linespread{1.08}
 
@@
 can look like the following code:
+\begin{minipage}{\linewidth}
 \begin{lstlisting}[style=default]
@@
    which describes the data"
-  (let [;; Guessing a distribution
+  (let [;;; Guessing a distribution
         dist (sample (categorical
@@
 
-    ;; Observing samples from the distribution
-    (loop [data0 data]
-      (when (not-empty data0)
-        (let [[x & data1] data0]
-          (observe d x))
-        (recur data1)))
-
-    ;; Returning the distribution and parameters
+    ;;; Observing samples from the distribution
+    (loop [observations data]
+      (when (not-empty observations)
+        ;; Retrieve the first observation as `o',
+        ;; and store the rest of observations in 
+        ;; 'observations*'.
+        (let [[o & observations*] observations]
+          ;; Observe 'o' from the guessed 
+          ;; distribution 'd'.
+          (observe d o))
+        ;; Proceed to the next iteration with
+        ;; the rest of observations.
+        (recur observations*)))
+
+    ;;; Finally, returning the distribution 
+    ;;; and parameters
     [d a b]))
 \end{lstlisting}
-The query builds a model for the input $\data$, a sequence of data points.
+\end{minipage}
+The query builds a model for the input $\texttt{data}$, 
+a sequence of data points.
 It defines a probability distribution on three variables, 
@@
 positive parameters for the type. Concretely,
-using the $\texttt{sample}$ statements, the query
-first defines so called a prior distribution on these three variables,
-and then it adjust this prior distribution based on data points in
-$\texttt{data}$ using the \texttt{observe} statement.
+using the $\texttt{sample}$ forms, the query
+first defines a so called prior distribution on these three variables,
+and then it adjusts this prior distribution based on observations in
+$\texttt{data}$ using the \texttt{observe} form.
 Samples from this adapted distribution (also called posterior distribution)

### 2016-11-26 paper/paper.tex
stacks -> stack

@@
    which describes the data"
-  (let [;;; Guessing a distribution
+  (let [;;; Guess a distribution.
         dist (sample (categorical
@@
 
-    ;;; Observing samples from the distribution
+    ;;; Observe samples from the distribution.
     (loop [observations data]
@@
 
-    ;;; Finally, returning the distribution 
-    ;;; and parameters
+    ;;; Return the distribution and parameters.
     [d a b]))
@@
 in Anglican, because Anglican programs are CPS-converted and do not
-use stacks. For instance, no recursive call in Anglican can lead to stack overflow.
+use the stack. For instance, no recursive call in Anglican can lead to stack overflow.
 In fact, in Anglican, it is recommended to use recursive calls to functions instead of

### 2016-11-27 paper/paper.tex
minor

--- 2016-11-26 paper/paper.tex
@@
 \end{lstlisting}
-The query builds a model for the input $\data$, a sequence of data points.
+The query builds a model for the input $\texttt{data}$, 
+a sequence of data points.
 It defines a probability distribution on three variables, 
@@
 using the $\texttt{sample}$ statements, the query
-first defines so called a prior distribution on these three variables,
-and then it adjust this prior distribution based on data points in
+first defines a so called prior distribution on these variables.
+Then, it adjusts this prior distribution based on data points in
 $\texttt{data}$ using the \texttt{observe} statement.

### 2016-11-27 paper/paper.tex
minor

@@
     ;; Observing samples from the distribution
-    (loop [data0 data]
+    (loop [data0 data] ;; Initializing data0
       (when (not-empty data0)
         (let [[x & data1] data0] 
-          (observe d x))
-        (recur data1)))
+          (observe d x)
+          (recur data1))))
 

### 2016-11-27 paper/paper.tex
minor

@@
 The state (\texttt{\$state}) is threaded through the computation
-and contains data used by inference algorithm. \texttt{\$state}
+and contains data used by inference algorithms. \texttt{\$state}
 is a Clojure hash map:
@@
   "initial program state"
-  {:log-weight 0.0
-   :result nil
-   ::mem {}
-   ::store nil
+  {:log-weight 0.0,
+   :result nil,
+   ::mem {},
+   ::store nil,
    ... })
 \end{lstlisting}
+which records inference-relevant information under various keys
+such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
 The full list of map entries depends on the inference
 algorithm. Except for transformation of
-\texttt{mem}, the memoization form, CPS transformation routines
+\texttt{mem} (which converts a function to one with memoization), 
+CPS transformation routines
 are not aware of contents of \texttt{\$state}, do not access or

### 2016-11-27 paper/paper.tex
update

@@
 algorithm. Except for transformation of
-\texttt{mem} (which converts a function to one with memoization), 
+the \texttt{mem} form (which converts a function to one with memoization), 
 CPS transformation routines
@@
 \begin{itemize}
-    \item literals, which are passed as an argument to the
-        continuation unmodified;
-    \item value expressions (e.g. the \texttt{fn} form)
+    \item Literals, which are constant expressions. They
+       are passed as an argument to the
+        continuation unmodified.
+    \item Value expressions such as the \texttt{fn} form
         (called \textit{opaque}
-        expressions in the code) which must be transformed to
+        expressions in the code). They must be transformed to
         CPS, but the transformed object is passed to the
-        continuation as a whole, opaquely;
-    \item general expressions (let's call them
-        \textit{transparent}), through which the continuation is
-        threaded in an expression-specific way, and can be
+        continuation as a whole, opaquely.
+    \item General expressions (which we call
+        \textit{transparent} expressions).
+        The continuation is threaded through such an expression
+        in an expression-specific way, and can be
         called in multiple locations of the CPS-transformed

### 2016-11-27 paper/paper.tex
I got confused about opaque expression initially, because I read 'the continuation' as a general referece to any conditinuation used in the transformed code. I replaced it by 'continuation parameter' so as to emphasise that we mean a particular continuation argument, not any continuation.

@@
 original and the CPS form. However, their CPS form follows the
-pattern \texttt{(continuation transformed-expression)}, and thus
-the transformation does not depend on the continuation, and
-can be accomplished without passing the continuation as a
+pattern \texttt{(continuation transformed-expression \$state)}, and thus
+the transformation does not depend on the continuation parameter, and
+can be accomplished without passing the parameter as a
 transformation argument. Primitive (non-CPS) procedures used in
@@
 The most general form of CPS transformation receives an
-expression and a continuation, and returns the expression
-in CPS form with the continuation potentially called in multiple
+expression and a continuation as parameters, and returns the expression
+in CPS form with the continuation parameter potentially called in multiple
 tail positions. General expressions can be somewhat voluntarily

### 2016-11-27 paper/paper.tex

@@
 \documentclass[preprint]{sigplanconf}
+\linespread{1.07}
 
@@
    which describes the data"
-  (let [;; Guessing a distribution
+  (let [;;; Guess a distribution.
         dist (sample (categorical
@@
 
-    ;; Observing samples from the distribution
-    (loop [data0 data] ;; Initializing data0
-      (when (not-empty data0)
-        (let [[x & data1] data0] 
-          (observe d x)
-          (recur data1))))
-
-    ;; Returning the distribution and parameters
+    ;;; Observe samples from the distribution.
+    (loop [observations data]
+      (when (not-empty observations)
+        ;; Retrieve the first observation as `o',
+        ;; and store the rest of observations in 
+        ;; 'observations*'.
+        (let [[o & observations*] observations]
+          ;; Observe 'o' from the guessed 
+          ;; distribution 'd'.
+          (observe d o))
+        ;; Proceed to the next iteration with
+        ;; the rest of observations.
+        (recur observations*)))
+
+    ;;; Return the distribution and parameters.
     [d a b]))
 \end{lstlisting}
+\end{minipage}
 The query builds a model for the input $\texttt{data}$, 
@@
 positive parameters for the type. Concretely,
-using the $\texttt{sample}$ statements, the query
-first defines a so called prior distribution on these variables.
-Then, it adjusts this prior distribution based on data points in
-$\texttt{data}$ using the \texttt{observe} statement.
+using the $\texttt{sample}$ forms, the query
+first defines a so called prior distribution on these three variables,
+and then it adjusts this prior distribution based on observations in
+$\texttt{data}$ using the \texttt{observe} form.
 Samples from this adapted distribution (also called posterior distribution)

### 2016-11-27 paper/paper.tex
update

--- 2016-11-27 paper/paper.tex
@@
 \end{lstlisting}
-The code of the continuation, represented by the \texttt{fn}
-form, will be repeated three times. In general, CPS code can
-grow extremely large if symbolic continuations are inserted
+During CPS transformation, if we substitute the code of this continuation 
+for all of its calls, the code will be repeated three times in 
+the CPS-transformed expression. In general, CPS code can
+grow extremely large if the code of continuations is substituted
 repeatedly.

### 2016-11-27 paper/paper.tex
update

@@
 can shade global primitive function names. For example,
-\texttt{first} is an Anglican function inside the let block in the
-following example:
+\texttt{first} is a Clojure primitive function that takes the first
+element from an ordered collection such as list and vector,
+but inside the let block in the
+following example, \texttt{first} is an Anglican function:
 \begin{lstlisting}[style=default]
@@
 (->sample dist cont $state)
-=> (cps-of-expression '(observe dist val) 'cont)
-(->observe dist val cont $state)
-\end{lstlisting}
+=> (cps-of-expression '(observe dist v) 'cont)
+(->observe dist v cont $state)
+\end{lstlisting}
+Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
+expressions constructors for Clojure records, and they take 
+values of their fields as arguments.
 

### 2016-11-27 paper/paper.tex
update

@@
 primitive functions by name --- if a function name is not in the
-list, the function is an Anglican function. Global dynamically
-bound variables \texttt{*primitive-procedures*} and
+list, the function is an Anglican function. Global dynamically-bound 
+variables \texttt{*primitive-procedures*} and
 \texttt{*primitive-namespaces*} contain the initial lists of
@@
 \paragraph{Memoization}
-The author of a probabilistic model might want to draw a
-feature of an entity randomly, but to retain the same drawn
+The author of a probabilistic model might want to 
+randomly draw a feature value of each entity in a collection,
+but to retain the same drawn
 value for the same entity during a single run of the probabilistic
@@
 \end{lstlisting}
+The \texttt{mem} form converts a function to a memoized variant,
+which remembers past inputs and the corresponding outputs, and 
+returns the remembered output when given such a past input, 
+instead of calling the original function with it. As a result,
+for every input, randon draws will be made only for the first time 
+that the memoized function is called with the input; all subsequent
+calls with the input will just reuse these draws return the same output.
 

### 2016-11-27 paper/paper.tex
update

@@
 memory is stored as a nested dictionary in the program state
-(function \texttt{mem-cps}).  Every memoized function gets a
+introduced during CPS transformation (function \texttt{mem-cps}).  
+Every memoized function gets a
 unique automatically generated identifier. Each time a memoized

### 2016-11-27 paper/paper.tex
update

@@
 an appropriate overloading or implementation of the function --- conventionally,
-the identifier is a keyword related to the
-algorithm name (\texttt{:lmh} for Lightweight
-Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
-etc.). The second parameter is a query as defined by
+the identifier is a Clojure keyword (a symbolic constant starting with colon)
+related to the algorithm name, such as \texttt{:lmh} for Lightweight
+Metropolis-Hasting and \texttt{:pcascade} for Particle Cascade.
+The second parameter is a query as defined by
 the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,

### 2016-11-27 paper/paper.tex

@@
 \documentclass[preprint]{sigplanconf}
+\linespread{1.06}
 

### 2016-11-27 paper/paper.tex
update

--- 2016-11-27 paper/paper.tex
@@
 
-The simplest inference algorithm is \textit{importance
-sampling}:
+Implementing an inference algorithm in Anglican amounts
+to defining a version of the \texttt{infer} function and 
+checkpoint handlers appropriate for the algorithm. The definitions may 
+just reuse default implementations or override them
+with new algorithm-specific treatment of \texttt{sample} and \texttt{observe} forms 
+and inference state.
+
+Let us illustrate this implementation process with
+\textit{importance sampling},
+the simplest inference algorithm.
+Here is an implementation of the \texttt{infer} function for the algorithm:
 \begin{lstlisting}[style=default]
@@
         :anglican.inference/algorithm)
-(defmethod infer :importance [_ prog value & {}]
+;; invoked when algo parameter is :importance
+(defmethod infer :importance 
+  [algo prog value & {}]
   (letfn
-    [(sample-seq []
+    ;; recursive function without any parameter
+    [sample-seq 
+      []
+      ;; lazy infinite sequence 
       (lazy-seq
@@
 \end{lstlisting}
-\texttt{anglican.importance/{\linebreak[0]}infer} just calls
-\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
-implementations of checkpoint handlers. A different inference
-algorithm would provide its own implementations of
+This implementation is dispatched 
+when \texttt{infer} receives \texttt{:importance} as its \texttt{algo} 
+parameter. (Anglican includes multiple implementations of \texttt{infer}
+for different inference algorithms.) Once dispatched, it lazily 
+constructs an infinite sequence of inference states by repeatedly 
+executing the program \texttt{prog} using \texttt{exec}
+and retrieving the final inference state of the execution 
+using \texttt{:state}. For checkpoint handlers, importance sampling simply
+relies on their default implementations.
+%\texttt{anglican.importance/{\linebreak[0]}infer} just calls
+%\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
+%implementations of checkpoint handlers.
+
+Typically, an inference
+algorithm has its own implementations of
 \texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
@@
 handler is overridden. In addition, SMC runs multiple particles
-(program instances) simultaneously, while LMH re-runs
+(threads executing a program) simultaneously, while LMH re-runs
 programs from an intermediate continuation rather than from the

### 2016-11-27 paper/paper.tex
update

@@
 control statement at which a particle (corresponding to
-a thread executing a program)
+a user-level thread executing a program)
 can be either replicated or terminated. Similarly, in 
@@
 handler is overridden. In addition, SMC runs multiple particles
-(threads executing a program) simultaneously, while LMH re-runs
+(corresponding to user-level threads) simultaneously, while LMH re-runs
 programs from an intermediate continuation rather than from the
@@
 Many inference algorithms memoize and reuse earlier computations
-at checkpoints: variants of Metropolis-Hastings reuse previously
+at checkpoints. Variants of Metropolis-Hastings reuse previously
 drawn values~\cite{WSG11}, as well as additional information
 used for adaptive sampling~\cite{TMP+15} at \texttt{sample}
-checkpoints, Asynchronous SMC (Particle Cascade) computes
+checkpoints. Asynchronous SMC (Particle Cascade) computes
 average particle weights at \texttt{observe}

### 2016-11-27 paper/paper.tex
Minor. I was impressed by the excellent explanation on checkpoint ids.

@@
 Implementing an inference algorithm in Anglican amounts
-to defining a version of the \texttt{infer} function and 
-checkpoint handlers appropriate for the algorithm. The definitions may 
-just reuse default implementations or override them
+to defining an appropriate version of the \texttt{infer} function and 
+checkpoint handlers for \texttt{sample} and \texttt{observe}.
+The definitions may just reuse default implementations or override them
 with new algorithm-specific treatment of \texttt{sample} and \texttt{observe} forms
@@
 
-Let us illustrate this implementation process with
+Let us illustrate this implementation step with
 \textit{importance sampling},
@@
 different traces. However, the random values should not be
-re-used between different distributions. Further on, in some
+reused between different distributions. Further on, in some
 algorithms, such as black-box variational inference, the role of
@@
 which is almost as efficient as the scheme based on stack addresses for
-re-use of previously drawn values, as produces addresses of constant size,
+reuse of previously drawn values, while producing addresses of constant size,
 and allows manual computation of checkpoint addresses at runtime
@@
 \texttt{sample} checkpoint, depending on whether the value is
-drawn or re-used.
+drawn or reused.
 

### 2016-11-27 paper/paper.tex
update

@@
 definitions of common distributions. Each distribution object
-implements the \texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution} protocol,
-with two methods: \texttt{sample*} and \texttt{observe*}.
+implements a distribution interface with the \texttt{sample*} and \texttt{observe*}
+methods; this interface is defined using Clojure's protocol mechanism
+(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
 The \texttt{sample*} \textit{method} returns a random sample and
@@
 \texttt{defdist} takes care of defining a separate type
-for every distribution so that multimethods can be dispatched on
-distribution types when needed, e.g. for custom proposals. The
+for every distribution so that Clojure multimethods (or overloaded
+methods) can be dispatched on
+distribution types when needed, e.g. for custom proposal distributions 
+used in an inference algorithm. The
 Bernoulli distribution could be defined as follows:
@@
 macro \texttt{defproc} and implement the
-\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
+\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol (that is,
+an interface for these random-sequence objects). The protocol
 has two methods --- \texttt{produce}, which returns a
@@
 (loop [process (HawkesPointProcess)
-       [value & values] values]
+       [value0 & values0] values] 
   (let [dist (produce process)]
-    (observe dist value)
-    (recur (absorb process value) values)))
+    (observe dist value0)
+    (recur (absorb process value0) values0)))
 \end{lstlisting}

### 2016-11-27 paper/paper.tex
small edit: notorious -> famous

@@
 level of abstraction. Different languages provide different
-means for this; Lisp is notorious for the macro facility
+means for this; Lisp is famous for the macro facility
 that allows to extend the language almost without

### 2016-11-27 paper/paper.tex
small edit: Jan Willem -> Jan-Willem

@@
 \title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
+\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
 

### 2016-11-27 paper/paper.tex
use color syntax highlighting
can be reverted by inputting anglican.tex instead of anglican-color.tex

@@
 
-\input{anglican}
+\input{anglican-color}
 
@@
 computation with access to continuations in probabilistic
-checkpoints. Anglican `compiler', represented by a set of
+checkpoints. The Anglican `compiler', represented by a set of
 functions in the \texttt{anglican.trap} namespace, accepts a

### 2016-11-27 paper/paper.tex
minor: value of each -> value for each

@@
 The author of a probabilistic model might want to 
-randomly draw a feature value of each entity in a collection,
+randomly draw a feature value for each entity in a collection,
 but to retain the same drawn

### 2016-11-27 paper/paper.tex
minor: indentation in code block off by 1 space
update

--- 2016-11-27 paper/paper.tex
@@
 \title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
+\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
 
@@
 level of abstraction. Different languages provide different
-means for this; Lisp is famous for the macro facility
+means for this; Lisp is notorious for the macro facility
 that allows to extend the language almost without
@@
 (cont (vector 0 1 2) $state)
-=> (cps-of-hash-map {:a 1 :b 2} 'cont)
+=> (cps-of-hash-map {:a 1, :b 2} 'cont)
 (cont (hash-map :a 1, :b 2) $state)
@@
 the returned value. However, there are no mutable data
-structures in a probabilistic program, hence \texttt{mem}'s
+structures in a probabilistic program. Hence, \texttt{mem}'s
 memory is stored as a nested dictionary in the program state
@@
 related to the algorithm name, such as \texttt{:lmh} for Lightweight
-Metropolis-Hasting and \texttt{:pcascade} for Particle Cascade.
+Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
 The second parameter is a query as defined by
@@
     \item On every function call, a component is added to the
-        address, hence the address size is linear in stack depth.
+        address. Hence, the address size is linear in stack depth.
     \item Every function call must be augmented by symbolic
@@
 While distributions are random and non-functional,
-random processes are deterministic and functional, hence
+random processes are deterministic and functional. Hence,
 \texttt{produce} and \texttt{absorb} are called directly and do
@@
         between 5 and 15 minutes on average to walk from an
-        office to the deli, were different average times are  for
+        office to the deli, where different average times are  for
         different buildings in the quarter.
@@
 
-For inference, one often chooses known distribution to
+For inference, one often chooses a known distribution to
 represent uncertainty. We choose the normal
@@
 the deli twice, or two different customers came to the deli,
-one for lunch, one for dinner. We define an \textbf{Anglican} function
+one for lunch and the other for dinner. We define an \textbf{Anglican} function
 for each case. Note that this is the first time we switch from
@@
 \texttt{same-customer} the average arrival time is the same for
-both the lunch and the dinner, while for
+both the lunch and the dinner, while in
 \texttt{different-customers} two average arrival times are
@@
                             same-customer
-                            different-customer)]
-    {:is-same-customer is-same-customer
+                            different-customers)]
+    {:same-customer is-same-customer,
      :times-to-arrive (observe-customer 
@@
 subsequence of samples. Many algorithms use an initial
-subsequence of samples to converge to the target distribution,
-hence we drop initial $N$ samples ($N$ is 5000 in the code
+subsequence of samples to converge to the target distribution.
+Hence, we drop initial $N$ samples ($N$ is 5000 in the code
 snippet below), and collect the next $N$ samples.
@@
 \end{lstlisting}
-
-With the specified observations, \texttt{-same+} is $\approx
-0.12$. The results may vary from run to run, and, for given
+The \texttt{:same-customer} here represents a function
+that looks up the entry with the same name in a given map.
+The map stores the result of a single sampled execution,
+and the \texttt{:same-customer} entry in the map
+records whether the same customer visits the deli in the execution.
+The \texttt{filter} function in \texttt{p-same+} 
+goes through all the maps in \texttt{results}
+and picks only the ones whose executions involve just one customer.
+
+With the specified observations, \texttt{p-same+} is $\approx
+0.12$. The probability is much lower than the
+waiter's guess \texttt{p-same} ($\frac 2 3$).
+Of course, the results may vary from run to run, and, for a given
 algorithm, the accuracy depends on the number of samples we
-chose to retrieve. The probability is much lower than the
-waiter's guess \texttt{p-same} ($\frac 2 3$).
+decide to retrieve.
 
@@
 (def time-to-arrive+ 
-     (map (comp first :times-to-arrive)
+    (map (fn [x] (first (:times-to-arrive x)))
          (filter :same-customer results)))
@@
      (map :times-to-arrive 
-          (filter (complement :same-customer)
+          (filter  
+            (fn [x] (not (:same-customer x)))
             results)))
@@
 In addition to plotting the distribution histograms, we use
-functions \texttt{mean}, \texttt{std} provided, along with other
-useful statistical functions, in the \texttt{anglican.{\linebreak[0]}stat}
-namespace. This is another illustration of advantage of tight
+functions \texttt{mean}, \texttt{std} provided
+in the \texttt{anglican.{\linebreak[0]}stat} namespace
+along with other
+useful statistical functions.
+This is another illustration of advantage of tight
 integration between Clojure and Anglican --- probabilistic
@@
 (clojure.pprint/pprint
-  (anglican.bbvb/get-variational (nth samples N)))
+  (anglican.bbvb/get-variational 
+    (nth samples N)))
 \end{lstlisting}

### 2016-11-27 paper/paper.tex
update

@@
 \begin{lstlisting}[style=default]
+(defm same-customer 
+      ...
       (sample :arrival-time-same
               time-to-arrive-prior)]
-
+      ...)
+
+(defm different-customers
+      ... 
       (sample :arrival-time-first
               time-to-arrive-prior)
-
+      ...
       (sample :arrival-time-second
               time-to-arrive-prior)]
-
-(let [is-same-customer (sample :same-or-different
+      ...)
+
+(defquery deli ...
+      ...
+      (sample :same-or-different
               (flip p-same))
+      ...)
 \end{lstlisting}

### 2016-11-27 paper/paper.tex
update

@@
 
-\section{Summary}
+\section{Conclusion}
 \label{seq:summary}
@@
 expressive specification of probabilistic generative models and
-efficient implementation of inference algorithm.
-
-Implementing a probabilistic language on top and in tight
+efficient implementation of inference algorithms.
+
+Implementing a probabilistic language on top of and in tight
 integration with a functional language, Clojure, both helped

### 2016-11-27 paper/anglican-color.tex
added a few more anglican primitives to color syntax highlighter

\usepackage{listings}
\usepackage{color}
\usepackage{xcolor}
\definecolor{blue}{rgb}{0,0.3,0.7}
\definecolor{red}{rgb}{0.60,0.0,0.0}
\definecolor{purple}{rgb}{0.5,0,0.7}
\definecolor{cyan}{rgb}{0.0,0.6,0.5}
\definecolor{orange}{rgb}{0.7,0.5,0.0}
\definecolor{gray}{rgb}{0.4,0.4,0.4}

\lstdefinelanguage{scheme}
{sensitive, %
 alsoletter={:,-,+,*,?,/,!,>,<}, %
 morecomment=[l];, %
}[comments]

% listings definition for Anglican programming language
\lstdefinelanguage{anglican}%
{% empty (strings and keywords)
 morekeywords=[1]{},
 % clojure declarations
 morekeywords=[2]{%
   def, def-, defn, defn-, defmacro, defmulti, defmethod, %
   defstruct, defonce, declare, definline, definterface, %
   defprotocol, defrecord, defstruct, deftype, defproject, ns, %
 }, %
 % clojure macros and special forms
 morekeywords=[3]{->, ->>, .., amap, and, areduce, as->, assert, binding, %
   bound-fn, case, comment, cond, cond->, cond->>, condp, declare, definline, %
   definterface, defmacro, defmethod, defmulti, defn, defn-, defonce, %
   defprotocol, defrecord, defstruct, deftype, delay, doseq, dosync, dotimes, %
   doto, extend-protocol, extend-type, fn, for, future, gen-class, %
   gen-interface, if, if-let, if-not, if-some, import, io!, lazy-cat, lazy-seq, let, %
   letfn, locking, loop, memfn, ns, or, proxy, proxy-super, pvalues, %
   recur, refer-clojure, reify, some->, some->>, sync, time, when, when-first, %
   when-let, when-not, when-some, while, with-bindings, with-in-str, %
   with-loading-context, with-local-vars, with-open, with-out-str, %
   with-precision, with-redefs}, %
  % clojure functions
  morekeywords=[4]{*, *', +, +', -, -', ->ArrayChunk, ->Vec, ->VecNode, %
    ->VecSeq, -cache-protocol-fn, -reset-methods, /, <, <=, =, ==, >, >=, %
    accessor, aclone, add-classpath, add-watch, agent, agent-error, %
    agent-errors, aget, alength, alias, all-ns, alter, alter-meta!, %
    alter-var-root, ancestors, apply, array-map, aset, aset-boolean, aset-byte, %
    aset-char, aset-double, aset-float, aset-int, aset-long, aset-short, assoc, %
    assoc!, assoc-in, associative?, atom, await, await-for, await1, bases, bean, %
    bigdec, bigint, biginteger, bit-and, bit-and-not, bit-clear, bit-flip, %
    bit-not, bit-or, bit-set, bit-shift-left, bit-shift-right, bit-test, %
    bit-xor, boolean, boolean-array, booleans, bound-fn*, bound?, butlast, byte, %
    byte-array, bytes, cast, char, char-array, char?, chars, chunk, %
    chunk-append, chunk-buffer, chunk-cons, chunk-first, chunk-next, chunk-rest, %
    chunked-seq?, class, class?, clear-agent-errors, clojure-version, coll?, %
    commute, comp, comparator, compare, compare-and-set!, compile, complement, %
    concat, conj, conj!, cons, constantly, construct-proxy, contains?, count, %
    counted?, create-ns, create-struct, cycle, dec, dec', decimal?, delay?, %
    deliver, denominator, deref, derive, descendants, destructure, disj, disj!, %
    dissoc, dissoc!, distinct, distinct?, doall, dorun, double, double-array, %
    doubles, drop, drop-last, drop-while, empty, empty?, ensure, %
    enumeration-seq, error-handler, error-mode, eval, even?, every-pred, every?, %
    ex-data, ex-info, extend, extenders, extends?, false?, ffirst, file-seq, %
    filter, filter-ns-publics, filterv, find, find-keyword, find-ns, %
    find-protocol-impl, find-protocol-method, find-var, first, flatten, float, %
    float-array, float?, floats, flush, fn?, fnext, fnil, force, format, %
    frequencies, future-call, future-cancel, future-cancelled?, future-done?, %
    future?, gensym, get, get-in, get-method, get-proxy-class, %
    get-thread-bindings, get-validator, group-by, hash, hash-combine, hash-map, %
    hash-ordered-coll, hash-set, hash-unordered-coll, identical?, identity, %
    ifn?, in-ns, inc, inc', init-proxy, instance?, int, int-array, integer?, %
    interleave, intern, interpose, into, into-array, ints, isa?, iterate, %
    iterator-seq, juxt, keep, keep-indexed, key, keys, keyword, keyword?, last, %
    line-seq, list, list*, list?, load, load-file, load-reader, load-string, %
    loaded-libs, long, long-array, longs, macroexpand, macroexpand-1, %
    make-array, make-hierarchy, map, map-indexed, map?, mapcat, mapv, max, %
    max-key, memoize, merge, merge-with, meta, method-sig, methods, min, %
    min-key, mix-collection-hash, mod, munge, name, namespace, namespace-munge, %
    neg?, newline, next, nfirst, nil?, nnext, not, not-any?, not-empty, %
    not-every?, not=, ns-aliases, ns-functions, ns-imports, ns-interns, %
    ns-macros, ns-map, ns-name, ns-publics, ns-refers, ns-resolve, ns-unalias, %
    ns-unmap, nth, nthnext, nthrest, num, number?, numerator, object-array, %
    odd?, parents, partial, partition, partition-all, partition-by, pcalls, %
    peek, persistent!, pmap, pop, pop!, pop-thread-bindings, pos?, pr, pr-str, %
    prefer-method, prefers, print, print-ctor, print-simple, print-str, printf, %
    println, println-str, prn, prn-str, promise, proxy-call-with-super, %
    proxy-mappings, proxy-name, push-thread-bindings, quot, rand, rand-int, %
    rand-nth, range, ratio?, rational?, rationalize, re-find, re-groups, %
    re-matcher, re-matches, re-pattern, re-seq, read, read-line, read-string, %
    realized?, record?, reduce, reduce-kv, reduced, reduced?, reductions, ref, %
    ref-history-count, ref-max-history, ref-min-history, ref-set, refer, %
    release-pending-sends, rem, remove, remove-all-methods, remove-method, %
    remove-ns, remove-watch, repeat, repeatedly, replace, replicate, require, %
    reset!, reset-meta!, resolve, rest, restart-agent, resultset-seq, reverse, %
    reversible?, rseq, rsubseq, satisfies?, second, select-keys, send, send-off, %
    send-via, seq, seq?, seque, sequence, sequential?, set, %
    set-agent-send-executor!, set-agent-send-off-executor!, set-error-handler!, %
    set-error-mode!, set-validator!, set?, short, short-array, shorts, shuffle, %
    shutdown-agents, slurp, some, some-fn, some?, sort, sort-by, sorted-map, %
    sorted-map-by, sorted-set, sorted-set-by, sorted?, special-symbol?, spit, %
    split-at, split-with, str, string?, struct, struct-map, subs, subseq, %
    subvec, supers, swap!, symbol, symbol?, take, take-last, take-nth, %
    take-while, test, the-ns, thread-bound?, to-array, to-array-2d, trampoline, %
    transient, tree-seq, true?, type, unchecked-add, unchecked-add-int, %
    unchecked-byte, unchecked-char, unchecked-dec, unchecked-dec-int, %
    unchecked-divide-int, unchecked-double, unchecked-float, unchecked-inc, %
    unchecked-inc-int, unchecked-int, unchecked-long, unchecked-multiply, %
    unchecked-multiply-int, unchecked-negate, unchecked-negate-int, %
    unchecked-remainder-int, unchecked-short, unchecked-subtract, %
    unchecked-subtract-int, underive, unsigned-bit-shift-right, update-in, %
    update-proxy, use, val, vals, var-get, var-set, var?, vary-meta, vec, %
    vector, vector-of, vector?, with-bindings*, with-meta, with-redefs-fn, %
    xml-seq, zero?, zipmap}, %
  % anglican declarations
  morekeywords=[5]{def-cps-fn, defanglican, defm, defquery, defun, defproc, defdist}, %
  % anglican macros
  morekeywords=[6]{cps-fn, fm, lambda, mem, query, with-primitive-procedures}, %
  % anglican functions
  morekeywords=[7]{%
    % anglican.core
    doquery, %
    % anglican.emit
    conditional, %
    % anglican.inference
    collect-by, equalize, exec, infer, log-marginal, print-predicts, %
    rand, rand-int, rand-nth, rand-roulette, stripdown, warmup, %
    % anglican.runtime
    ->CRP-process, ->DP-process, ->GP-process, %
    ->bernoulli-distribution, ->beta-distribution, ->binomial-distribution, %
    ->categorical-crp-distribution, ->categorical-distribution, %
    ->categorical-dp-distribution, ->chi-squared-distribution, %
    ->dirichlet-distribution, ->discrete-distribution, %
    ->exponential-distribution, ->flip-distribution, ->gamma-distribution, %
    ->mvn-distribution, ->normal-distribution, ->poisson-distribution, %
    ->sample, ->observe, sample*, observe*, %
    ->uniform-continuous-distribution, ->uniform-discrete-distribution, %
    ->wishart-distribution, CRP, DP, GP, abs, absorb, acos, asin, atan, %
    bernoulli, beta, binomial, categorical, categorical-crp, categorical-dp, %
    cbrt, ceil, chi-squared, cos, cosh, cov, dirichlet, discrete, exp, %
    exponential, flip, floor, gamma, gen-matrix, log, log-gamma-fn, %
    log-mv-gamma-fn, log-sum-exp, map->CRP-process, map->DP-process, %
    map->GP-process, map->bernoulli-distribution, map->beta-distribution, %
    map->binomial-distribution, map->categorical-crp-distribution, %
    map->categorical-distribution, map->categorical-dp-distribution, %
    map->chi-squared-distribution, map->dirichlet-distribution, %
    map->discrete-distribution, map->exponential-distribution, %
    map->flip-distribution, map->gamma-distribution, map->mvn-distribution, %
    map->normal-distribution, map->poisson-distribution, %
    map->uniform-continuous-distribution, map->uniform-discrete-distribution, %
    map->wishart-distribution, mvn, normal, poisson, pow, produce, %
    rint, round, signum, sin, sinh, sqrt, tag, tan, tanh, transform-sample, %
    uniform-continuous, uniform-discrete, wishart, %
    % anglican.state
    add-log-weight, add-predict, clear-predicts, get-log-weight, %
    get-mem, get-predicts, in-mem?, set-log-weight, set-mem, %     
  }, %
  % anglican special forms
  morekeywords=[8]{factor, observe, predict, retrieve, sample, store}, %
  sensitive, %
  alsoletter={:,-,+,*,?,/,!,>,<}, %
  morecomment=[l];, %
  morestring=[b]", %
  keywordsprefix=:, %
}[keywords,comments,strings]

\lstset{
  language=anglican,
  basicstyle=\small\ttfamily,
  commentstyle=\em\color{gray},
  % keywords and strings
  keywordstyle=[1]\color{orange},
  stringstyle=\color{orange},
  % clojure declarations
  keywordstyle=[2]\color{purple},
  % clojure macros and special forms
  keywordstyle=[3]\color{red},
  % clojure functions
  keywordstyle=[4]\color{cyan},
  % anglican declarations
  keywordstyle=[5]\color{purple},
  % anglican macros
  keywordstyle=[6]\color{red},
  % anglican functions
  keywordstyle=[7]\color{cyan},
  % anglican inference forms 
  keywordstyle=[8]\color{blue}
}

### 2016-11-27 paper/paper.tex
replaced poisson process with a beta-bernouilli process

--- 2016-11-27 paper/paper.tex
@@
 (cont (vector 0 1 2) $state)
-=> (cps-of-hash-map {:a 1, :b 2} 'cont)
+=> (cps-of-hash-map {:a 1 :b 2} 'cont)
 (cont (hash-map :a 1, :b 2) $state)
@@
 the returned value. However, there are no mutable data
-structures in a probabilistic program. Hence, \texttt{mem}'s
+structures in a probabilistic program, hence \texttt{mem}'s
 memory is stored as a nested dictionary in the program state
@@
 related to the algorithm name, such as \texttt{:lmh} for Lightweight
-Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
+Metropolis-Hasting and \texttt{:pcascade} for Particle Cascade.
 The second parameter is a query as defined by
@@
     \item On every function call, a component is added to the
-        address. Hence, the address size is linear in stack depth.
+        address, hence the address size is linear in stack depth.
     \item Every function call must be augmented by symbolic
@@
 random process object obtained by absorbing a value drawn from
-the process. Here is a definition of the Poisson process:
-\begin{lstlisting}[style=default]
-(defproc PP
-  "Poisson point process"
-  [rate] [dist (exponential rate)]
-  (produce [this] dist)
-  (absorb [this value] this))
+the process. Here is a definition of a beta Bernouilli process:
+\begin{lstlisting}[style=default]
+(defproc beta-bernoulli
+  "beta-Bernoulli process"
+  [a b]
+  (produce [this] (bernoulli (/ a (+ a b))))
+  (absorb [this x]
+    (case x
+      0 (beta-bernoulli a (inc b))
+      1 (beta-bernoulli (inc a) b))))
 \end{lstlisting}
 While distributions are random and non-functional,
-random processes are deterministic and functional. Hence,
+random processes are deterministic and functional, hence
 \texttt{produce} and \texttt{absorb} are called directly and do
@@
 \begin{lstlisting}[style=default]
-(loop [process (HawkesPointProcess)
+(loop [process (beta-bernoulli 1.0 1.0)
        [value0 & values0] values] 
@@
         between 5 and 15 minutes on average to walk from an
-        office to the deli, where different average times are  for
+        office to the deli, were different average times are  for
         different buildings in the quarter.
@@
 
-For inference, one often chooses a known distribution to
+For inference, one often chooses known distribution to
 represent uncertainty. We choose the normal
@@
 the deli twice, or two different customers came to the deli,
-one for lunch and the other for dinner. We define an \textbf{Anglican} function
+one for lunch, one for dinner. We define an \textbf{Anglican} function
 for each case. Note that this is the first time we switch from
@@
 \texttt{same-customer} the average arrival time is the same for
-both the lunch and the dinner, while in
+both the lunch and the dinner, while for
 \texttt{different-customers} two average arrival times are
@@
                             same-customer
-                            different-customers)]
-    {:same-customer is-same-customer,
+                            different-customer)]
+    {:is-same-customer is-same-customer
      :times-to-arrive (observe-customer 
@@
 subsequence of samples. Many algorithms use an initial
-subsequence of samples to converge to the target distribution.
-Hence, we drop initial $N$ samples ($N$ is 5000 in the code
+subsequence of samples to converge to the target distribution,
+hence we drop initial $N$ samples ($N$ is 5000 in the code
 snippet below), and collect the next $N$ samples.
@@
 \end{lstlisting}
-The \texttt{:same-customer} here represents a function
-that looks up the entry with the same name in a given map.
-The map stores the result of a single sampled execution,
-and the \texttt{:same-customer} entry in the map
-records whether the same customer visits the deli in the execution.
-The \texttt{filter} function in \texttt{p-same+} 
-goes through all the maps in \texttt{results}
-and picks only the ones whose executions involve just one customer.
-
-With the specified observations, \texttt{p-same+} is $\approx
-0.12$. The probability is much lower than the
+
+With the specified observations, \texttt{-same+} is $\approx
+0.12$. The results may vary from run to run, and, for given
+algorithm, the accuracy depends on the number of samples we
+chose to retrieve. The probability is much lower than the
 waiter's guess \texttt{p-same} ($\frac 2 3$).
-Of course, the results may vary from run to run, and, for a given
-algorithm, the accuracy depends on the number of samples we
-decide to retrieve.
 
@@
 (def time-to-arrive+ 
-    (map (fn [x] (first (:times-to-arrive x)))
+     (map (comp first :times-to-arrive)
           (filter :same-customer results)))
@@
      (map :times-to-arrive 
-          (filter  
-            (fn [x] (not (:same-customer x)))
+          (filter (complement :same-customer)
                   results)))
@@
 In addition to plotting the distribution histograms, we use
-functions \texttt{mean}, \texttt{std} provided
-in the \texttt{anglican.{\linebreak[0]}stat} namespace
-along with other
-useful statistical functions.
-This is another illustration of advantage of tight
+functions \texttt{mean}, \texttt{std} provided, along with other
+useful statistical functions, in the \texttt{anglican.{\linebreak[0]}stat}
+namespace. This is another illustration of advantage of tight
 integration between Clojure and Anglican --- probabilistic
@@
 (clojure.pprint/pprint
-  (anglican.bbvb/get-variational 
-    (nth samples N)))
+  (anglican.bbvb/get-variational (nth samples N)))
 \end{lstlisting}

### 2016-11-27 paper/paper.tex

--- 2016-11-27 paper/paper.tex
@@
 \documentclass[preprint]{sigplanconf}
+\linespread{1.03}
 
@@
 can look like the following code:
+%\begin{minipage}{\linewidth}
 \begin{lstlisting}[style=default]
@@
 
-    ;; Observing samples from the distribution
-    (loop [data0 data] ;; Initializing data0
-      (when (not-empty data0)
-        (let [[x & data1] data0] 
-          (observe d x)
-          (recur data1))))
-
-    ;; Returning the distribution and parameters
+    ;;; Observe samples from the distribution.
+    (loop [observations data]
+      (when (not-empty observations)
+        ;; Retrieve the first observation as `o',
+        ;; and store the rest of observations in 
+        ;; 'observations*'.
+        (let [[o & observations*] observations]
+          ;; Observe 'o' from the guessed 
+          ;; distribution 'd'.
+          (observe d o))
+        ;; Proceed to the next iteration with
+        ;; the rest of observations.
+        (recur observations*)))
+
+    ;;; Return the distribution and parameters.
     [d a b]))
 \end{lstlisting}
+%\end{minipage}
 The query builds a model for the input $\texttt{data}$, 

### 2016-11-27 paper/paper.tex

--- 2016-11-27 paper/paper.tex
@@
 random process object obtained by absorbing a value drawn from
-the process. Here is a definition of the Poisson process:
-\begin{lstlisting}[style=default]
-(defproc PP
-  "Poisson point process"
-  [rate] [dist (exponential rate)]
-  (produce [this] dist)
-  (absorb [this value] this))
+the process. Here is a definition of a beta Bernouilli process:
+\begin{lstlisting}[style=default]
+(defproc beta-bernoulli
+  "beta-Bernoulli process"
+  [a b]
+  (produce [this] (bernoulli (/ a (+ a b))))
+  (absorb [this x]
+    (case x
+      0 (beta-bernoulli a (inc b))
+      1 (beta-bernoulli (inc a) b))))
 \end{lstlisting}

### 2016-11-27 paper/paper.tex
edits to section on random processes

@@
 
-Similar to distributions, \textit{random processes}, related to
-so called `exchangeable random procedures'\footnote{However, random sequences
-generated by Anglican random processes are not required to be
-exchangeable.}, are defined using
-macro \texttt{defproc} and implement the
-\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol (that is,
-an interface for these random-sequence objects). The protocol
-has two methods --- \texttt{produce}, which returns a
-distribution object, and \texttt{absorb}, which returns a new
-random process object obtained by absorbing a value drawn from
-the process. Here is a definition of a beta Bernouilli process:
+In addition to distributions, Anglican provides \textit{random processes},
+which define sequences of random variables that are not independent and
+identically distributed. Random processes are similar to the so called
+`exchangeable random procedures' in Church \cite{GMR+08} and
+Venture \cite{MSP14}. However, random sequences generated by Anglican random 
+processes are not required to be exchangeable. Random processes are
+defined using the \texttt{defproc} macro and implement the
+\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
+protocol has two methods
+\begin{itemize}
+\item \texttt{produce}, which returns the distribution on the next
+random variable in the sequence, and
+\item \texttt{absorb}, which incorporates the value for the next random
+variable and returns an updated random process.
+\end{itemize}
+ As an example, here is a definition of a beta-Bernouilli process, in which
+each random variable is distributed according to a Bernoulli distribution with
+an unknown parameter that is drawn from a beta distribution:
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-While distributions are random and non-functional,
-random processes are deterministic and functional. Hence,
-\texttt{produce} and \texttt{absorb} are called directly and do
-not have corresponding special forms in Anglican:
-\begin{itemize}
-    \item Values produced by the process are sampled from the distribution
-        returned by \texttt{produce}.
-    \item When a value is observed from the distribution, a new
-        process, with the observed value absorbed, is returned
-        by \texttt{absorb}.
-\end{itemize}
-A common pattern of programming with processes in Anglican is
-recursively computing a new process based on the observed value:
+Unlike typical implementations of exchangeable random processes, Anglican's random processes do not have mutable state.
+The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican.
+Random values are generated using a recursion in which \texttt{produce} constructs the distribution on the next variable and \texttt{absorb} constructs the updated process for the next iteration.
+For example:
+\begin{lstlisting}[style=default]
+(defm sample-beta-binomial [n a b]
+  (loop [process (beta-bernoulli a bb)
+         values []]
+    (if (= (count values) n)
+      values
+      (let [dist (produce process)
+            value (sample dist)]
+      (recur (absorb process value)
+             (conj values value)))))))
+\end{lstlisting}
+Similarly, random processes can also be used to recursively observe a sequence of values:
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-Unlike conventional implementations of random processes,
-Anglican's random processes do not have mutable state.
+
+% \begin{itemize}
+%     \item Values produced by the process are sampled from the distribution
+%         returned by \texttt{produce}.
+%     \item When a value is observed from the distribution, a new
+%         process, with the observed value absorbed, is returned
+%         by \texttt{absorb}.
+% \end{itemize}
+
+% A common pattern when programming with random processes in 
+% Anglican is to generate a sequences of sampled values and 
+% constructing an updated process at each step
 

### 2016-11-28 paper/paper.tex
minor edits to section on random processes

@@
 \end{itemize}
- As an example, here is a definition of a beta-Bernouilli process, in which
+ As an example, here is a definition of a beta-Bernoulli process, in which
 each random variable is distributed according to a Bernoulli distribution with
@@
 The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican.
-Random values are generated using a recursion in which \texttt{produce} constructs the distribution on the next variable and \texttt{absorb} constructs the updated process for the next iteration.
+Random values are generated using a recursion in \texttt{absorb} constructs the updated process for the next iteration.
 For example:
@@
 (defm sample-beta-binomial [n a b]
-  (loop [process (beta-bernoulli a bb)
-         values []]
+  (loop [process (beta-bernoulli a b)]
     (if (= (count values) n)

### 2016-11-28 paper/paper.tex
fixed some typos in deli section

@@
 The program presented in this section is intentionally short and
-simple. Aglican is capable of compiling and running elaborated
+simple. Anglican is capable of compiling and running elaborated
 programs handling large amounts of data. Advanced examples of
@@
 Imagine that we are facing the following dilemma:
-
 \begin{quote}
 A customer wearing round sunglasses came to a deli at 1:13pm,
-and grabbed a sandwitch and a coffee. Later on the same day, a
+and grabbed a sandwich and a coffee. Later on the same day, a
 customer wearing round sunglasses came at 6:09pm and ordered a
@@
 \end{quote}
-
 In addition to the story above, we know that:
-
 \begin{itemize}

### 2016-11-28 paper/paper.tex
looking nicely

@@
 \usepackage{listings}
-\usepackage[lighttt]{lmodern}
+\usepackage{lmodern}
 \ttfamily
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}
 
@@
 for this invocations are
-\begin{verbatim}
+\begin{lstlisting}[style=default]
     [C1 0]
@@
     [C3 0]
-\end{verbatim}
+\end{lstlisting}
 If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
@@
 The addresses for the new sequence are
-\begin{verbatim}
+\begin{lstlisting}[style=default]
     [C1 0]
@@
     [C3 0]
-\end{verbatim}
+\end{lstlisting}
 Despite the change, the correspondence between checkpoints of
@@
 \vspace{\baselineskip}
+{\small
 \begin{tabular}{l l}
@@
 \end{tabular}
-\vspace{\baselineskip}
+\vspace{\baselineskip}}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}[style=default]
 {S28209
@@
     :max 1.0}}}}
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}[style=default]
 {:arrivate-time-same

### 2016-11-28 paper/paper.tex
normal font size
a few typos fixed

@@
 \documentclass[preprint]{sigplanconf}
-\linespread{1.03}
+\linespread{1.00}
 
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
 
@@
 in the Anglican source code. However, in Clojure, unlike in
-Scheme~\cite{SDF+10}, or Common Lisp~\cite{PC94}, the result
+Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
 of macro-expansion of derived special forms is not well
@@
 calls at runtime. However, this would impact performance, and
-a good runtime performance is critical for probabilistic
+good runtime performance is critical for probabilistic
 programs. The approach taken by Anglican is to maintain a list
@@
 instead of calling the original function with it. As a result,
-for every input, randon draws will be made only for the first time 
+for every input, random draws will be made only for the first time 
 that the memoized function is called with the input; all subsequent
@@
 \texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
-both, as well as invoke \texttt{exec} from an elaborated
+both, as well as invokes \texttt{exec} from an elaborated
 conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
@@
         argument of a checkpoint. For example \texttt{(sample
-        'x1  (normal 0 1)} defines a \texttt{sample} checkpoint
+        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
         with identifier \texttt{x1}.
@@
 \vspace{\baselineskip}
-{\small
+{
 \begin{tabular}{l l}

### 2016-11-28 paper/paper.tex
fixed a couple of typos

@@
 The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican.
-Random values are generated using a recursion in \texttt{absorb} constructs the updated process for the next iteration.
+A sequence of random values can be generated using a recursive loop in which \texttt{absorb} returns the updated process for the next iteration.
 For example:
@@
 (defm sample-beta-binomial [n a b]
-  (loop [process (beta-bernoulli a b)]
+  (loop [process (beta-bernoulli a b)
+         values []]
     (if (= (count values) n)
@@
       (recur (absorb process value)
-             (conj values value)))))))
+             (conj values value))))))
 \end{lstlisting}

### 2016-11-28 paper/paper.tex
typo

@@
 Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
-expressions constructors for Clojure records, and they take 
+expressions are constructors for Clojure records, and they take 
 values of their fields as arguments.
@@
 that the memoized function is called with the input; all subsequent
-calls with the input will just reuse these draws return the same output.
+calls with the input will just reuse these draws and return the same output.
 

### 2017-02-04 paper/paper.tex
new structure
full toggle, reorganization in progress

@@
 \usepackage{url}
+\usepackage{etoolbox}
 
@@
 \authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
+
+\newtoggle{full}
+\togglefalse{full}
 
@@
 
+\iftoggle{full}{%
 There are several ways to build a programming language
@@
 language is implemented `inside' another language of the same
-level of abstraction. Different languages provide different
+level of abstraction. 
+}{%
+One way to implement a programming language is to embed it within another language.
+}%
+Different languages provide different
 means for this; Lisp is famous for the macro facility
@@
 
+\section{Related Work}
+\label{sec:related}
+
+{\it TODO}
+
 \section{Language}
@@
 
+\subsection{Probabilistic Forms}
+
+{\it TODO}
+
 \subsection{Core Library}
@@
 for writing tail-recursive programs. Such forms are not necessary
-in Anglican, because Anglican programs are CPS-converted and do not
-use the stack. For instance, no recursive call in Anglican can lead to stack overflow.
-In fact, in Anglican, it is recommended to use recursive calls to functions instead of
-\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
-for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop}
-will lead to unpredictable behaviour and hard-to-catch errors.
+in Anglican, because Anglican programs are CPS-converted and do
+not use the stack. For instance, no recursive call in Anglican
+can lead to stack overflow.  In fact, in Anglican, it is
+recommended to use recursive calls to functions instead of
+\texttt{recur}. However, \texttt{loop}/\texttt{recur} is
+provided in Anglican for convenience as a way to express loops.
+\texttt{recur} outside of \texttt{loop} will lead to
+unpredictable behaviour and hard-to-catch errors.
+
+\section{Case Study}
+\label{seq:study}
+
+Before delving into implementation details, we provide a case
+study of using the Anglican language and environment. This case
+study takes a problem for which a solution is not immediately
+obvious, \textit{the Deli dilemma}, and guides through writing
+an Anglican program for the problem, executing the program, and
+post-processing results.
+
+The program presented in this section is intentionally short and
+simple. Anglican is capable of compiling and running elaborated
+programs handling large amounts of data. Advanced examples of
+Anglican programs and inference can be found in literature on
+applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.
+
+\subsection{The Deli Dilemma}
+\label{seq:deli}
+
+Imagine that we are facing the following dilemma:
+\begin{quote}
+A customer wearing round sunglasses came to a deli at 1:13pm,
+and grabbed a sandwich and a coffee. Later on the same day, a
+customer wearing round sunglasses came at 6:09pm and ordered a
+dinner.  Was it the same customer?
+\end{quote}
+In addition to the story above, we know that:
+\begin{itemize}
+    \item There is an adjacent office quarter, and it takes
+        between 5 and 15 minutes on average to walk from an
+        office to the deli, where different average times are  for
+        different buildings in the quarter.
+    \item Depending on traffic lights, the walking time varies by about 2
+        minutes.
+    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
+    \item The waiter's odds that this is the same customer are 2 to 1.
+\end{itemize}
+
+\subsection{Anglican Query}
+\label{seq:deli-query}
+
+We want to formalize the dilemma in an Anglican query, based on
+the knowledge we have. Let us formalize the knowledge in
+Clojure (the times are in minutes). First, we encode our prior
+information which holds true independently of the customer's
+visit:
+
+\begin{lstlisting}[style=default]
+(def mean-time-to-arrive
+     "average time to arrive"
+     10.)
+(def sd-time-to-arrive
+     "standard deviation of arrival time"
+     3.)
+(def time-sd
+     "walking time deviation"
+     1.)
+\end{lstlisting}
+
+Then, we record our observations, based on which we want to
+solve the dilemma:
+
+\begin{lstlisting}[style=default]
+(def lunch-delay
+     "time between lunch break and lunch order"
+     13.)
+(def dinner-delay
+     "time between end of day and dinner order"
+     9.)
+(def p-same
+     "prior probability of the same customer"
+     (/ 2. 3.))
+\end{lstlisting}
+
+For inference, one often chooses a known distribution to
+represent uncertainty. We choose the normal
+distribution for representing uncertainty about average arrival
+time.
+
+\begin{lstlisting}[style=default]
+(def time-to-arrive-prior 
+     "prior distribution of average arrival time"
+     (normal mean-time-to-arrive
+             sd-time-to-arrive))
+\end{lstlisting}
+
+There are two possibilities: either the same customer visited
+the deli twice, or two different customers came to the deli,
+one for lunch and the other for dinner. We define an \textbf{Anglican} function
+for each case. Note that this is the first time we switch from
+Clojure to Anglican. The functions must be written in Anglican
+(and hence defined using \texttt{defm} instead of \texttt{defn})
+because they contain probabilistic forms \texttt{sample} and
+\texttt{observe}.
+
+\begin{lstlisting}[style=default]
+(defm same-customer 
+  "observe the same customer twice"
+  [time-to-arrive-prior
+   lunch-delay
+   dinner-delay]
+  (let [time-to-arrive
+          (sample time-to-arrive-prior)]
+    (observe (normal time-to-arrive time-sd) 
+             lunch-delay)
+    (observe (normal time-to-arrive time-sd)
+             dinner-delay)
+    [time-to-arrive]))
+
+(defm different-customers
+  "observe different customers"
+  [time-to-arrive-prior
+   lunch-delay
+   dinner-delay]
+  (let [time-to-arrive-1
+          (sample time-to-arrive-prior)
+        time-to-arrive-2
+          (sample time-to-arrive-prior)]
+    (observe (normal time-to-arrive-1 time-sd) 
+             lunch-delay)
+    (observe (normal time-to-arrive-2 time-sd)
+             dinner-delay)
+    [time-to-arrive-1 time-to-arrive-2]))
+\end{lstlisting}
+
+Both functions have the same structure: we first `guess' the
+average arrival time and then observe the actual time from a
+distribution parameterized by the guessed time. However, in
+\texttt{same-customer} the average arrival time is the same for
+both the lunch and the dinner, while in
+\texttt{different-customers} two average arrival times are
+guessed independently.
+
+We are finally ready to define the query in \textbf{Anglican}.
+
+\begin{lstlisting}[style=default]
+(defquery deli [time-to-arrive-prior
+                lunch-delay
+                dinner-delay]
+  (let [is-same-customer (sample (flip p-same))
+        observe-customer (if is-same-customer
+                            same-customer
+                            different-customers)]
+    {:same-customer is-same-customer,
+     :times-to-arrive (observe-customer 
+                        time-to-arrive-prior
+                        lunch-delay
+                        dinner-delay)}))
+\end{lstlisting}
+
+Performing inference on query \texttt{deli} computes the
+probability that the same customer visited the deli twice, as
+well as probability distributions of average arrival times for
+both cases.
+
+\subsection{Inference}
+\label{seq:deli-inference}
+
+Having defined the query, we are now ready to run the query
+using an inference algorithm. Function \texttt{doquery} provided
+by the \texttt{anglican.core} namespace accepts the inference
+algorithm, the query, and optional parameters, and returns a
+lazy sequence of samples. We use here the inference algorithm
+called Lightweight Metropolis-Hastings (LMH). The algorithm is
+somewhat slow to converge but can be used with any Anglican
+query, and should be robust enough for our simple problem.
+
+We bind the results of \texttt{doquery} to variable
+\texttt{samples}, to analyse the results later. However, since
+the sequence is lazy, no inference is performed and no samples
+are generated until they are retrieved and processed.
+
+\begin{lstlisting}[style=default]
+(def samples (doquery :lmh deli nil))
+\end{lstlisting}
+
+To approximate the inferred distribution, we extract a finite
+subsequence of samples. Many algorithms use an initial
+subsequence of samples to converge to the target distribution.
+Hence, we drop initial $N$ samples ($N$ is 5000 in the code
+snippet below), and collect the next $N$ samples.
+
+\begin{lstlisting}[style=default]
+(def N 5000)
+(def results (map get-result
+                   (take N (drop N samples))))
+\end{lstlisting}
+
+Based on the collected samples we compute an approximation of
+the posterior probability \texttt{p-same+} that the same
+customer visited the deli twice.
+
+\begin{lstlisting}[style=default]
+(def p-same+ 
+     (/ (count (filter :same-customer results))
+        (double N)))
+\end{lstlisting}
+The \texttt{:same-customer} here represents a function
+that looks up the entry with the same name in a given map.
+The map stores the result of a single sampled execution,
+and the \texttt{:same-customer} entry in the map
+records whether the same customer visits the deli in the execution.
+The \texttt{filter} function in \texttt{p-same+} 
+goes through all the maps in \texttt{results}
+and picks only the ones whose executions involve just one customer.
+
+With the specified observations, \texttt{p-same+} is $\approx
+0.12$. The probability is much lower than the
+waiter's guess \texttt{p-same} ($\frac 2 3$).
+Of course, the results may vary from run to run, and, for a given
+algorithm, the accuracy depends on the number of samples we
+decide to retrieve.
+
+In addition to computing the posterior probability that the same
+customer visited the deli twice, we may want to know the average
+time (or times) to arrive. In Bayesian inference, it is common
+to report distributions instead of `most likely' values. We use
+query results in retrieved samples to approximate the
+distributions, and plot distribution histograms for the same
+customer visiting twice (Figure~\ref{fig:dist-same}) and
+two different customers (Figure~\ref{fig:dist-diff})
+
+\begin{lstlisting}[style=default]
+;; single customer                      
+(def time-to-arrive+ 
+    (map (fn [x] (first (:times-to-arrive x)))
+         (filter :same-customer results)))
+(def mean-to-arrive+ (mean time-to-arrive+))
+(def sd-to-arrive+ (std time-to-arrive+))
+
+;; two customers
+(def times-to-arrive+ 
+     (map :times-to-arrive 
+          (filter  
+            (fn [x] (not (:same-customer x)))
+            results)))
+(def mean-1-to-arrive+
+     (mean (map first times-to-arrive+)))
+(def sd-1-to-arrive+
+     (std (map first times-to-arrive+)))
+(def mean-2-to-arrive+
+     (mean (map second times-to-arrive+)))
+(def sd-2-to-arrive+
+     (std (map second times-to-arrive+)))
+\end{lstlisting}
+
+In addition to plotting the distribution histograms, we use
+functions \texttt{mean}, \texttt{std} provided
+in the \texttt{anglican.{\linebreak[0]}stat} namespace
+along with other
+useful statistical functions.
+This is another illustration of advantage of tight
+integration between Clojure and Anglican --- probabilistic
+models are expressed in Anglican. However, processing of data and
+results can rely on the full power of Clojure.
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
+    \caption{Arrival time distribution for a single customer.}
+    \label{fig:dist-same}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
+    \caption{Arrival time distributions for two customers.}
+    \label{fig:dist-diff}
+\end{figure}
+
+For inference, we chose to use  Lightweight
+Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
+probabilistic programming is that models are separated from
+inference. To switch to a different inference algorithm we just
+need to pass a different value to \texttt{doquery}. For example,
+we may decide to use Black-Box Variational Bayes (BBVB), which
+may not work equally well for all probabilistic programs, but is
+much faster to converge.
+
+\begin{lstlisting}[style=default]
+(def samples (doquery :bbvb deli nil))
+\end{lstlisting}
+
+We can still use samples and summary statistics with BBVB to
+approximate the posterior distribution. However, variational
+inference approximates the posterior by known distributions,
+and we can directly retrieve the distribution parameters.
+
+\begin{lstlisting}[style=default]
+(clojure.pprint/pprint
+  (anglican.bbvb/get-variational 
+    (nth samples N)))
+\end{lstlisting}
+
+\begin{lstlisting}[style=default]
+{S28209
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 10.99753360180663,
+   :sd 0.7290976433082352}},
+ S28217
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 12.668050292254202,
+   :sd 0.9446695174790353}},
+ S28215
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 9.104132559955836,
+   :sd 0.9479290526821788}},
+ S28219
+ {(0 anglican.runtime.flip-distribution)
+  {:p 0.11671977016875924,
+   :dist
+   {:min 0.0,
+    :max 1.0}}}}
+\end{lstlisting}
+
+We can guess that the variational distributions correspond to the
+prior distributions in the \texttt{sample} forms, in the order
+of appearance. However, it would help if we could use more
+informative labels instead of automatically generated symbols
+\texttt{S28209}, \texttt{S28217}, etc. Here the option to
+specify identifiers for probabilistic forms explicitly comes
+handy. If we modify the \texttt{sample} forms to use explicit identifiers
+(only the forms are shown for brevity), the output becomes much
+easier to analyse.
+
+\begin{lstlisting}[style=default]
+(defm same-customer 
+      ...
+      (sample :arrival-time-same
+              time-to-arrive-prior)]
+      ...)
+
+(defm different-customers
+      ... 
+      (sample :arrival-time-first
+              time-to-arrive-prior)
+      ...
+      (sample :arrival-time-second
+              time-to-arrive-prior)]
+      ...)
+
+(defquery deli ...
+      ...
+      (sample :same-or-different
+              (flip p-same))
+      ...)
+\end{lstlisting}
+
+The output becomes much easier to interpret and analyse
+programmatically.
+
+\begin{lstlisting}[style=default]
+{:arrivate-time-same
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 10.99753360180663,
+   :sd 0.7290976433082352}},
+ :arrival-time-first
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 12.668050292254202,
+   :sd 0.9446695174790353}},
+ :arrival-time-second
+ {(0 anglican.runtime.normal-distribution)
+  {:mean 9.104132559955836,
+   :sd 0.9479290526821788}},
+ :same-or-different
+ {(0 anglican.runtime.flip-distribution)
+  {:p 0.11671977016875924,
+   :dist
+   {:min 0.0,
+    :max 1.0}}}}
+\end{lstlisting}
+
+This completes the case study, where we showed a probabilistic
+programming solution of a problem, implemented and analysed in
+Anglican and Clojure, using two different inference algorithms.
 
@@
 
+\iftoggle{full}{%
 \subsection{Expression Kinds}
@@
 
+}{%
+}
+
 \subsection{Implementation Highlights}
@@
 
+\iftoggle{full}{%
 So far we introduced the basics of Anglican compilation to
@@
 for grasping the language implementation but relatively well-known.
+}{%
+Anglican compilation to Clojure is based on CPS transformation
+and compilation techniques, many of which are
+well known and widely used~\cite{AJ89,A07}.
+}%
 In the rest of the section we focus on challenges we met and
@@
 
+\iftoggle{full}{
 Continuations are functions that are called in tail positions
@@
 \paragraph{Managing stack size}
-
+}{%
+Continuations may lead to unbounded stack growth in
+recursive code.
+}%
 In implementations of functional programming languages stack
@@
 
+\iftoggle{full}{%
 \paragraph{Avoiding exponential code growth}
@@
 \end{lstlisting}
+}{
+}
 
@@
 
-When an Anglican function is transformed into a Clojure function
-by \texttt{fn-cps}, two auxiliary parameters are added to the
+When an Anglican function is transformed into a Clojure
+function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary parameters are added to the
 beginning of the parameter list --- continuation and state.
-Correspondingly, when a function \textit{call} is transformed
-(by \texttt{cps-of-application} or \texttt{cps-of-apply}), the
+Correspondingly, when a function \textit{call} is
+transformed\iftoggle{full}{ (by \texttt{cps-of-application} or
+\texttt{cps-of-apply})}{}, the
 current continuation and the state are passed to the called
@@
 The Anglican compiler takes care of the shading by rebinding
-\texttt{*primitive-procedures*} in every lexical scope
-(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
-\texttt{shading-primitive-procedures} automates the shading.
+\texttt{*primitive-procedures*} in every lexical
+scope\iftoggle{full}{ (\texttt{fn-cps}, \texttt{cps-of-let}). Macro
+\texttt{shading-primitive-procedures} automates the shading}{.}
 
@@
 % Anglican is to generate a sequences of sampled values and 
-% constructing an updated process at each step
-
-\section{Case Study}
-\label{seq:study}
-
-The description of Anglican language and environment would be
-incomplete without a case study showing the process of building
-and executing a solution of an inference problem. This case
-study takes a problem for which a solution is not immediately
-obvious, \textit{the Deli dilemma}, and guides through
-writing an Anglican program for the problem, executing the
-program, and post-processing results.
-
-The program presented in this section is intentionally short and
-simple. Anglican is capable of compiling and running elaborated
-programs handling large amounts of data. Advanced examples of
-Anglican programs and inference can be found in literature on
-applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.
-
-\subsection{The Deli Dilemma}
-\label{seq:deli}
-
-Imagine that we are facing the following dilemma:
-\begin{quote}
-A customer wearing round sunglasses came to a deli at 1:13pm,
-and grabbed a sandwich and a coffee. Later on the same day, a
-customer wearing round sunglasses came at 6:09pm and ordered a
-dinner.  Was it the same customer?
-\end{quote}
-In addition to the story above, we know that:
-\begin{itemize}
-    \item There is an adjacent office quarter, and it takes
-        between 5 and 15 minutes on average to walk from an
-        office to the deli, where different average times are  for
-        different buildings in the quarter.
-    \item Depending on traffic lights, the walking time varies by about 2
-        minutes.
-    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
-    \item The waiter's odds that this is the same customer are 2 to 1.
-\end{itemize}
-
-\subsection{Anglican Query}
-\label{seq:deli-query}
-
-We want to formalize the dilemma in an Anglican query, based on
-the knowledge we have. Let us formalize the knowledge in
-Clojure (the times are in minutes). First, we encode our prior
-information which holds true independently of the customer's
-visit:
-
-\begin{lstlisting}[style=default]
-(def mean-time-to-arrive
-     "average time to arrive"
-     10.)
-(def sd-time-to-arrive
-     "standard deviation of arrival time"
-     3.)
-(def time-sd
-     "walking time deviation"
-     1.)
-\end{lstlisting}
-
-Then, we record our observations, based on which we want to
-solve the dilemma:
-
-\begin{lstlisting}[style=default]
-(def lunch-delay
-     "time between lunch break and lunch order"
-     13.)
-(def dinner-delay
-     "time between end of day and dinner order"
-     9.)
-(def p-same
-     "prior probability of the same customer"
-     (/ 2. 3.))
-\end{lstlisting}
-
-For inference, one often chooses a known distribution to
-represent uncertainty. We choose the normal
-distribution for representing uncertainty about average arrival
-time.
-
-\begin{lstlisting}[style=default]
-(def time-to-arrive-prior 
-     "prior distribution of average arrival time"
-     (normal mean-time-to-arrive
-             sd-time-to-arrive))
-\end{lstlisting}
-
-There are two possibilities: either the same customer visited
-the deli twice, or two different customers came to the deli,
-one for lunch and the other for dinner. We define an \textbf{Anglican} function
-for each case. Note that this is the first time we switch from
-Clojure to Anglican. The functions must be written in Anglican
-(and hence defined using \texttt{defm} instead of \texttt{defn})
-because they contain probabilistic forms \texttt{sample} and
-\texttt{observe}.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-  "observe the same customer twice"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive time-sd)
-             dinner-delay)
-    [time-to-arrive]))
-
-(defm different-customers
-  "observe different customers"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive-1
-          (sample time-to-arrive-prior)
-        time-to-arrive-2
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive-1 time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive-2 time-sd)
-             dinner-delay)
-    [time-to-arrive-1 time-to-arrive-2]))
-\end{lstlisting}
-
-Both functions have the same structure: we first `guess' the
-average arrival time and then observe the actual time from a
-distribution parameterized by the guessed time. However, in
-\texttt{same-customer} the average arrival time is the same for
-both the lunch and the dinner, while in
-\texttt{different-customers} two average arrival times are
-guessed independently.
-
-We are finally ready to define the query in \textbf{Anglican}.
-
-\begin{lstlisting}[style=default]
-(defquery deli [time-to-arrive-prior
-                lunch-delay
-                dinner-delay]
-  (let [is-same-customer (sample (flip p-same))
-        observe-customer (if is-same-customer
-                            same-customer
-                            different-customers)]
-    {:same-customer is-same-customer,
-     :times-to-arrive (observe-customer 
-                        time-to-arrive-prior
-                        lunch-delay
-                        dinner-delay)}))
-\end{lstlisting}
-
-Performing inference on query \texttt{deli} computes the
-probability that the same customer visited the deli twice, as
-well as probability distributions of average arrival times for
-both cases.
-
-\subsection{Inference}
-\label{seq:deli-inference}
-
-Having defined the query, we are now ready to run the query
-using an inference algorithm. Function \texttt{doquery} provided
-by the \texttt{anglican.core} namespace accepts the inference
-algorithm, the query, and optional parameters, and returns a
-lazy sequence of samples. We use here the inference algorithm
-called Lightweight Metropolis-Hastings (LMH). The algorithm is
-somewhat slow to converge but can be used with any Anglican
-query, and should be robust enough for our simple problem.
-
-We bind the results of \texttt{doquery} to variable
-\texttt{samples}, to analyse the results later. However, since
-the sequence is lazy, no inference is performed and no samples
-are generated until they are retrieved and processed.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :lmh deli nil))
-\end{lstlisting}
-
-To approximate the inferred distribution, we extract a finite
-subsequence of samples. Many algorithms use an initial
-subsequence of samples to converge to the target distribution.
-Hence, we drop initial $N$ samples ($N$ is 5000 in the code
-snippet below), and collect the next $N$ samples.
-
-\begin{lstlisting}[style=default]
-(def N 5000)
-(def results (map get-result
-                   (take N (drop N samples))))
-\end{lstlisting}
-
-Based on the collected samples we compute an approximation of
-the posterior probability \texttt{p-same+} that the same
-customer visited the deli twice.
-
-\begin{lstlisting}[style=default]
-(def p-same+ 
-     (/ (count (filter :same-customer results))
-        (double N)))
-\end{lstlisting}
-The \texttt{:same-customer} here represents a function
-that looks up the entry with the same name in a given map.
-The map stores the result of a single sampled execution,
-and the \texttt{:same-customer} entry in the map
-records whether the same customer visits the deli in the execution.
-The \texttt{filter} function in \texttt{p-same+} 
-goes through all the maps in \texttt{results}
-and picks only the ones whose executions involve just one customer.
-
-With the specified observations, \texttt{p-same+} is $\approx
-0.12$. The probability is much lower than the
-waiter's guess \texttt{p-same} ($\frac 2 3$).
-Of course, the results may vary from run to run, and, for a given
-algorithm, the accuracy depends on the number of samples we
-decide to retrieve.
-
-In addition to computing the posterior probability that the same
-customer visited the deli twice, we may want to know the average
-time (or times) to arrive. In Bayesian inference, it is common
-to report distributions instead of `most likely' values. We use
-query results in retrieved samples to approximate the
-distributions, and plot distribution histograms for the same
-customer visiting twice (Figure~\ref{fig:dist-same}) and
-two different customers (Figure~\ref{fig:dist-diff})
-
-\begin{lstlisting}[style=default]
-;; single customer                      
-(def time-to-arrive+ 
-    (map (fn [x] (first (:times-to-arrive x)))
-         (filter :same-customer results)))
-(def mean-to-arrive+ (mean time-to-arrive+))
-(def sd-to-arrive+ (std time-to-arrive+))
-
-;; two customers
-(def times-to-arrive+ 
-     (map :times-to-arrive 
-          (filter  
-            (fn [x] (not (:same-customer x)))
-            results)))
-(def mean-1-to-arrive+
-     (mean (map first times-to-arrive+)))
-(def sd-1-to-arrive+
-     (std (map first times-to-arrive+)))
-(def mean-2-to-arrive+
-     (mean (map second times-to-arrive+)))
-(def sd-2-to-arrive+
-     (std (map second times-to-arrive+)))
-\end{lstlisting}
-
-In addition to plotting the distribution histograms, we use
-functions \texttt{mean}, \texttt{std} provided
-in the \texttt{anglican.{\linebreak[0]}stat} namespace
-along with other
-useful statistical functions.
-This is another illustration of advantage of tight
-integration between Clojure and Anglican --- probabilistic
-models are expressed in Anglican. However, processing of data and
-results can rely on the full power of Clojure.
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
-    \caption{Arrival time distribution for a single customer.}
-    \label{fig:dist-same}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
-    \caption{Arrival time distributions for two customers.}
-    \label{fig:dist-diff}
-\end{figure}
-
-For inference, we chose to use  Lightweight
-Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
-probabilistic programming is that models are separated from
-inference. To switch to a different inference algorithm we just
-need to pass a different value to \texttt{doquery}. For example,
-we may decide to use Black-Box Variational Bayes (BBVB), which
-may not work equally well for all probabilistic programs, but is
-much faster to converge.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :bbvb deli nil))
-\end{lstlisting}
-
-We can still use samples and summary statistics with BBVB to
-approximate the posterior distribution. However, variational
-inference approximates the posterior by known distributions,
-and we can directly retrieve the distribution parameters.
-
-\begin{lstlisting}[style=default]
-(clojure.pprint/pprint
-  (anglican.bbvb/get-variational 
-    (nth samples N)))
-\end{lstlisting}
-
-\begin{lstlisting}[style=default]
-{S28209
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- S28217
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- S28215
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- S28219
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
-\end{lstlisting}
-
-We can guess that the variational distributions correspond to the
-prior distributions in the \texttt{sample} forms, in the order
-of appearance. However, it would help if we could use more
-informative labels instead of automatically generated symbols
-\texttt{S28209}, \texttt{S28217}, etc. Here the option to
-specify identifiers for probabilistic forms explicitly comes
-handy. If we modify the \texttt{sample} forms to use explicit identifiers
-(only the forms are shown for brevity), the output becomes much
-easier to analyse.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-      ...
-      (sample :arrival-time-same
-              time-to-arrive-prior)]
-      ...)
-
-(defm different-customers
-      ... 
-      (sample :arrival-time-first
-              time-to-arrive-prior)
-      ...
-      (sample :arrival-time-second
-              time-to-arrive-prior)]
-      ...)
-
-(defquery deli ...
-      ...
-      (sample :same-or-different
-              (flip p-same))
-      ...)
-\end{lstlisting}
-
-The output becomes much easier to interpret and analyse
-programmatically.
-
-\begin{lstlisting}[style=default]
-{:arrivate-time-same
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- :arrival-time-first
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- :arrival-time-second
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- :same-or-different
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
-\end{lstlisting}
-
-This completes the case study, where we showed a probabilistic
-programming solution of a problem, implemented and analysed in
-Anglican and Clojure, using two different inference algorithms.
+% construct an updated process at each step.
+
+\section{Performance}
+
+{\it TODO}
 

### 2017-02-04 paper/code-explosion.tex
many changes in progress

--- 2016-11-28 paper/paper.tex
@@
-\documentclass[preprint]{sigplanconf}
-\linespread{1.00}
-
-\usepackage[color]{changebar}
-\usepackage{enumitem}
-\usepackage[lighttt]{lmodern}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-
-\usepackage{listings}
-\usepackage{lmodern}
-\ttfamily
-\DeclareFontShape{OT1}{lmtt}{m}{it}
-     {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
-
-\usepackage{hyperref}
-\usepackage{url}
-
-\input{anglican-color}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-% Authors: comment using these commands
-
-\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:}
-        {\rm #1}\end{minipage}}}
-
-\cbcolor{red}
-% Authors: use changebars to mark changed regions
-% \cbstart
-% ...
-% \cbend
-
-\title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-    Anglican is a probabilistic programming system designed to interoperate
-    with Clojure and other JVM languages. We introduce the
-    programming language Anglican, outline our design choices,
-    and discuss in depth the implementation of the Anglican language
-    and runtime, including macro-based compilation, extended
-    CPS-based evaluation model, and functional representations for
-    probabilistic paradigms, such as a distribution,
-    a random process, and an inference algorithm.
-
-    We show that a probabilistic functional language can be
-    implemented efficiently and integrated tightly with a
-    conventional functional language with only moderate
-    computational overhead. We also demonstrate how advanced
-    probabilistic modeling concepts are mapped naturally
-    to the functional foundation.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-
-For data science practitioners, statistical inference is typically
-just one step in a more elaborate analysis workflow. The first stage
-of this work involves data acquisition, pre-processing and cleaning.
-This is often followed by several iterations of exploratory model
-design and testing of inference algorithms. Once a sufficiently
-robust statistical model and a corresponding inference algorithm have
-been identified, analysis results must be post-processed, visualized,
-and in some cases integrated into a wider production system.
-
-Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
-represent generative models as programs written in a specialized
-language that provides syntax for the definition and conditioning of
-random variables. The code for such models is generally concise,
-modular, and easy to modify or extend. Typically inference can be
-performed for any probabilistic program using one or more generic
-inference techniques provided by the system backend, such as
-Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
-\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
-of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
-These generic techniques may be less statistically efficient
-than techniques tailored to a specific model. However,
-probabilistic programming facilitates simpler implementations of
-models making the inference inherently faster.
-
-While probabilistic programming systems shorten the iteration cycle in
-exploratory model design, they typically lack basic functionality needed for
-data I/O, pre-processing, and analysis and visualization of inference results.
-In this paper, we describe the implementation of
-Anglican~\cite{TMW15, Anglican}, a probabilistic
-programming language that tightly integrates with
-Clojure~\cite{H08,Clojure}, a general-purpose programming language
-that runs on the Java Virtual Machine (JVM). Both languages share a common
-syntax, and can be invoked from each other. This allows Anglican programs to
-make use of a rich set of libraries written in both Clojure and Java.
-Conversely, Anglican allows intuitive and compact specification of models for
-which inference may be performed as part of a larger Clojure project.
-
-There are several ways to build a programming language
-on top of or besides another language.  The easiest is an
-interpreter --- a program that reads a program, in its entirety
-or line-by-line, and executes it by applying operational
-semantics of a certain kind to the language. \textsc{Basic} is
-famous for line-by-line interpreted implementations.
-
-Another approach is to write a compiler, either to a virtual
-architecture, so called p-code or byte-code, or to real
-hardware. Here, the whole program is translated from the
-`higher-level' source language to a `lower-level' object
-language, which can be directly executed, either by hardware or
-by an interpreter --- but the latter interpreter can be made
-simpler and more efficient  than an interpreter for the source
-language.
-
-On top of these two approaches are methods in which a new
-language is implemented `inside' another language of the same
-level of abstraction. Different languages provide different
-means for this; Lisp is famous for the macro facility
-that allows to extend the language almost without
-restriction --- by writing \textit{macros}, one adds new
-constructs to the existing language. There are several uses of
-macros --- one is to extend the language \textit{syntax}, for
-example, by adding new control structures; another is to keep
-the existing syntax but alter the \textit{operational semantics}
---- the way programs are executed and compute their outputs.
-
-Anglican is implemented in just this way --- a macro facility
-provided by Clojure, a Lisp dialect, is used both to extend
-Clojure with constructs that delimit probabilistic code, and to
-alter the operational semantics of Clojure expressions inside
-probabilistic code fragments. Anglican claims its right to count
-as a separate language because of the ubiquitous probabilistic
-execution semantics rather than a different syntax,
-which is actually an advantage rather than a drawback ---
-Clojure programmers only need to know how to specify the
-boundaries of Anglican programs, but can use familiar Clojure
-syntax to write probabilistic code.
-
-An implementation of Anglican must therefore address three issues:
-\begin{itemize}
-    \item the Clojure syntax to introduce probabilistic Anglican
-        code inside Clojure modules;
-    \item source-to-source transformation of Anglican programs
-        into Clojure, so that probabilistic execution becomes
-        possible;
-    \item algorithms which run Clojure code, obtained by
-        transforming Anglican programs, according to the
-        probabilistic operational semantics.
-\end{itemize}
-Execution of probabilistic programs by inference algorithms is
-different from execution of deterministic programs.  A
-probabilistic program is executed multiple times, often hundreds
-of thousands or even millions of times for a single inference
-task. Random choices may affect which parts of the program code
-are executed and how often. Many inference algorithms require
-re-running the program multiple times partially, from a certain
-point on.  Different executions may employ different random
-choices. However, for efficient inference a correspondence
-between random choices in different executions should be
-maintained. These are just some of the challenges which were
-faced and solved during development of Anglican.
-
-Comparisons of Anglican with other implementations of
-probabilistic programming languages~\cite{SGG15}\cite[pp.
-32--33]{P16} demonstrate that Anglican achieves state-of-the-art
-computational efficiency without sacrificing expressiveness.
-Anglican language syntax, compilation, invocation, and runtime
-support of Anglican queries are discussed in detail in further
-sections.
-
-\section{Design Outline}
-\label{sec:design}
-
-An Anglican program, or \textit{query}, is compiled into a
-Clojure function. When inference is performed with a provided
-algorithm, this produces a sequence of samples. Anglican shares
-a common syntax with Clojure; Clojure functions can be called
-from Anglican code and vice versa. A simple program in Anglican
-can look like the following code:
-%\begin{minipage}{\linewidth}
-\begin{lstlisting}[style=default]
-(defquery model data
-  "chooses a distribution
-   which describes the data"
-  (let [;;; Guess a distribution.
-        dist (sample (categorical
-                       [[normal 0.5]
-                        [gamma 0.5]]))
-        a (sample (gamma 1 1))
-        b (sample (gamma 1 1))
-        d (dist a b)]
-
-    ;;; Observe samples from the distribution.
-    (loop [observations data]
-      (when (not-empty observations)
-        ;; Retrieve the first observation as `o',
-        ;; and store the rest of observations in 
-        ;; 'observations*'.
-        (let [[o & observations*] observations]
-          ;; Observe 'o' from the guessed 
-          ;; distribution 'd'.
-          (observe d o))
-        ;; Proceed to the next iteration with
-        ;; the rest of observations.
-        (recur observations*)))
-
-    ;;; Return the distribution and parameters.
-    [d a b]))
-\end{lstlisting}
-%\end{minipage}
-The query builds a model for the input $\texttt{data}$, 
-a sequence of data points.
-It defines a probability distribution on three variables, 
-$\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
-for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
-positive parameters for the type. Concretely,
-using the $\texttt{sample}$ forms, the query
-first defines a so called prior distribution on these three variables,
-and then it adjusts this prior distribution based on observations in
-$\texttt{data}$ using the \texttt{observe} form.
-Samples from this adapted distribution (also called posterior distribution)
-can be obtained by running the query under one of Anglican's
-inference algorithms.
-
-Internally, an Anglican query is represented by a computation in
-\textit{continuation passing style} (CPS)~\cite{AJ89}, and inference algorithms
-exploit the CPS structure of the code to intercept probabilistic
-operations in an algorithm-specific way\footnote{\cite{GS15}
-also describe a CPS-based implementation of a probabilistic
-programming language.}. Among the available
-inference algorithms there are Particle Cascade~\cite{PWD+14},
-Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
-Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
-Inference on Anglican queries generates a lazy sequence of
-samples, which can be processed asynchronously in Clojure
-code for analysis, integration, and decision making.
-
-Clojure (and Anglican) runs on the JVM and gets access to a wide choice of Java
-libraries for data processing, networking, presentation, and imaging.
-Conversely, Anglican queries can be called from Java and other JVM languages.
-Programs involving Anglican queries can be deployed as JVM
-\textit{jars}, and run without modification on any platform for
-which JVM is available.
-
-A probabilistic program, or query, mostly runs deterministic
-code, except for certain checkpoints, in which probabilities are
-involved, and normal, linear execution of the program is
-disrupted. In Anglican and similar languages there are two types
-of such checkpoints:
-
-\begin{itemize}
-    \item drawing a value from a random source (\texttt{sample});
-    \item conditioning a computed value on a random source
-        (\texttt{observe}).
-\end{itemize}
-
-Anglican can be mostly implemented as a regular programming
-language, except for the handling of these checkpoints.
-Depending on the \textit{inference algorithm}, \texttt{sample}
-and \texttt{observe} may result in implicit input/output
-operations and control changes. For example, \texttt{observe} in
-particle filtering inference algorithms \cite{WVM14} is a non-deterministic
-control statement at which a particle (corresponding to
-a user-level thread executing a program)
-can be either replicated or terminated. Similarly, in 
-Metropolis-Hastings \cite{WSG11},
-\texttt{sample} is both an input
-statement which `reads' values from a random source,
-and a non-deterministic control statement
-(with delayed effect), eventually affecting acceptance or
-rejection of a sample.
-
-Because of the checkpoints, Anglican programs must allow the
-inference algorithm to step in, recording information and
-affecting control flow. This can be implemented through
-coroutines/cooperative multitasking, and parallel execution/preemptive
-multitasking, as well as through explicit
-maintenance of program continuations at checkpoints. Anglican
-follows the latter option. Clojure is
-a functional language, and continuation-passing style (CPS)
-transformation is a well-developed technique in the area of
-functional languages. Implementing a variant of CPS
-transformation seemed to be the most flexible and lightweight
-option --- any other form of concurrency would put a higher burden
-on the underlying runtime (JVM) and the operating system.
-Consequently, Anglican has been implemented as a CPS-transformed
-computation with access to continuations in probabilistic
-checkpoints. The Anglican `compiler', represented by a set of
-functions in the \texttt{anglican.trap} namespace, accepts a
-Clojure subset and transforms it into a variant of CPS
-representation, which allows inference algorithms to intervene
-in the execution flow at probabilistic checkpoints.
-
-Anglican is intended to co-exist with Clojure and be a part of
-the source of a Clojure program. To facilitate this, Anglican
-programs, or queries, are wrapped by macros (defined in the
-\texttt{anglican.emit} namespace), which call the CPS
-transformations and define Clojure values
-%Clojure objects
-suitable for passing
-as arguments to inference algorithms (\texttt{defquery},
-\texttt{query}). In addition to defining entire queries,
-Anglican promotes modularization of inference algorithms through
-the definitions of \textit{probabilistic functions} using
-\texttt{defm}
-%and \texttt{fm}
-(Anglican counterparts of Clojure \texttt{defn}).
-%and \texttt{fn}).
-Probabilistic functions are
-written in Anglican, may include probabilistic forms
-\texttt{sample} and \texttt{observe}, and can be seamlessly
-called from inside Anglican queries, just like functions locally
-defined within the same query.
-
-Operational semantics of Anglican queries is different from that
-of Clojure code, therefore queries must be called through
-inference algorithms, rather than `directly'.  The
-\texttt{anglican.inference} namespace declares the (ad-hoc) polymorphic
-function \texttt{infer} using Clojure's multimethod mechanism.
-This function accepts an Anglican query and
-returns a lazy sequence of weighted samples from the
-distribution defined by the query.  When inference
-is performed on an Anglican query, the query is run by a
-particular inference algorithm. Inference algorithms must
-provide an implementation for \texttt{infer}, as well as
-override the polymorphic function \texttt{checkpoint} (defined
-as a multimethod) so as to handle \texttt{sample} and
-\texttt{observe} in an algorithm-specific manner and
-to construct an appropriate result on the
-termination of a probabilistic program.
-
-Finally, Anglican queries use `primitive', or commonly known
-and used, distributions, to draw random samples and condition
-observations. Many primitive distributions are provided by the
-\texttt{anglican.runtime} namespace, and an additional
-distribution can be defined by the user by implementing a
-particular set of functions for the distribution
-(via Clojure's protocol mechanism). The \texttt{defdist} macro
-provides a convenient syntax for defining primitive distributions.
-
-\section{Language}
-\label{sec:language}
-
-The Anglican language is a subset of Clojure\footnote{It would be
-possible to support almost full Clojure by expanding all macros
-in the Anglican source code. However, in Clojure, unlike in
-Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
-of macro-expansion of derived special forms is not well
-specified and implementation specific.}.  Anglican queries
-are defined within \texttt{defquery} as shown in the previous section,
-using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case},
-\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
-other forms of Clojure may be supported in the future but are not now.
-In \texttt{let} bindings and \texttt{fn} argument lists,
-Clojure's pattern matching mechanism for vector data type
-(called vector destructuring) is supported. Also, compound
-literals for vectors, hash maps, and
-sets are supported just like in Clojure.
-
-\subsection{Core Library}
-\label{sec:core}
-
-All of Clojure's core library except for higher-order functions
-(functions that accept other functions as arguments) is
-available in Anglican. Higher-order functions cannot be reused
-from Clojure, as they have to be re-implemented to accept
-functional arguments in CPS form. The following higher-order
-functions are implemented: \texttt{map}, \texttt{reduce},
-\texttt{filter}, \texttt{some}, \texttt{repeatedly},
-\texttt{comp}, \texttt{partial}.
-
-\subsection{Tail Call}
-\label{sec:tail}
-
-Clojure provides special forms \texttt{loop} and \texttt{recur}
-for writing tail-recursive programs. Such forms are not necessary
-in Anglican, because Anglican programs are CPS-converted and do not
-use the stack. For instance, no recursive call in Anglican can lead to stack overflow.
-In fact, in Anglican, it is recommended to use recursive calls to functions instead of
-\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
-for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop}
-will lead to unpredictable behaviour and hard-to-catch errors.
-
-\section{Macro-based Compilation}
-\label{sec:compilation}
-
-Compilation of Anglican into Clojure is built around a variant of CPS
-transformation. In a basic CPS-transformed program, a
-continuation receives a single argument --- the computed value.
-In Anglican, there are two flows of computation going in
-parallel: values are computed by functional code, and,
-at the same time, the state of the probabilistic program,
-used by inference algorithms, is updated by
-probabilistic forms. Because of that, in Anglican a continuation
-accepts  \textit{two} arguments:
-\begin{itemize}
-    \item the computed value;
-    \item the internal state, bound to the local variable
-        \texttt{\$state} in every lexical scope.
-\end{itemize}
-
-The compilation relies on the Clojure \textit{macro} facility,
-and implemented as a library of \textit{functions} in namespace
-\texttt{anglican.trap}, which are invoked by macros.  The CPS
-transformation is organized in top-down manner.  The top-level
-function is  \texttt{cps-of-expression}, which receives an
-expression and a continuation, and returns the expression in the
-CPS form.  For example, the CPS transformation of constant
-\texttt{1} with continuation \texttt{cont} thus takes the
-following form:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression 1 'cont)
-(cont 1 $state)
-\end{lstlisting}
-
-\subsection {The State}
-\label{sec:state}
-
-The state (\texttt{\$state}) is threaded through the computation
-and contains data used by inference algorithms. \texttt{\$state}
-is a Clojure hash map:
-\begin{lstlisting}[style=default]
-(def initial-state
-  "initial program state"
-  {:log-weight 0.0,
-   :result nil,
-   ::mem {},
-   ::store nil,
-   ... })
-\end{lstlisting}
-which records inference-relevant information under various keys
-such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
-The full list of map entries depends on the inference
-algorithm. Except for transformation of
-the \texttt{mem} form (which converts a function to one with memoization), 
-CPS transformation routines
-are not aware of contents of \texttt{\$state}, do not access or
-modify it directly, but rather just thread the state unmodified
-through the computation. Algorithm-specific handlers of
-checkpoints corresponding to the probabilistic forms
-(\texttt{sample}, \texttt{observe}) modify the
-state and reinject a new state into the computation.
-
-\subsection{Expression Kinds}
-\label{sec:expr}
-
-There are three different kinds of inputs to CPS transformation:
-\begin{itemize}
-    \item Literals, which are constant expressions. They
-       are passed as an argument to the
-        continuation unmodified.
-    \item Value expressions such as the \texttt{fn} form
-        (called \textit{opaque}
-        expressions in the code). They must be transformed to
-        CPS, but the transformed object is passed to the
-        continuation as a whole, opaquely.
-    \item General expressions (which we call
-        \textit{transparent} expressions).
-        The continuation is threaded through such an expression
-        in an expression-specific way, and can be
-        called in multiple locations of the CPS-transformed
-        code, such as in all branches of an \texttt{if}
-        statement.
-\end{itemize}
-
-\subsubsection{Literals}
-\label{sec:literals}
-
-Literals are the same in Anglican and Clojure. They are
-left unmodified; literals are a subset of opaque expressions.
-However, the Clojure syntax has a peculiarity
-of using the syntax of compound literals (vectors, hash maps,
-and sets) for data constructors. Hence, compound literals must
-be traversed recursively, and if there is a nested non-literal
-component, transformed into a call to the corresponding data
-constructor. Functions \texttt{cps-of-vector},
-\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
-\texttt{cps-of-expression}, transform Clojure constructor syntax
-(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
-corresponding calls:
-\begin{lstlisting}[style=default]
-=> (cps-of-vector [0 1 2] 'cont)
-(cont (vector 0 1 2) $state)
-=> (cps-of-hash-map {:a 1, :b 2} 'cont)
-(cont (hash-map :a 1, :b 2) $state)
-=> (cps-of-set #{0 1} 'cont)
-(cont (set (list 0 1)) $state)
-\end{lstlisting}
-
-\subsubsection{Opaque Expressions}
-\label{sec:opaque}
-
-Opaque, or value, expressions, have a different shape in the
-original and the CPS form. However, their CPS form follows the
-pattern \texttt{(continuation transformed-expression \$state)}, and thus
-the transformation does not depend on the continuation parameter, and
-can be accomplished without passing the parameter as a
-transformation argument. Primitive (non-CPS) procedures used in
-Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
-forms are opaque and transformed by
-\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps},
-\texttt{fn-cps}, and \texttt{mem-cps}, correspondingly: a
-slightly simplified CPS form of expression
-\begin{lstlisting}[style=default]
-(fn [x y]
-  (+ x y))
-\end{lstlisting}
-would be
-\begin{lstlisting}[style=default]
-(fn [cont $state x y]
-  (cont (+ x y) $state))
-\end{lstlisting}
-In the actual code an automatically generated fresh symbol is
-used instead of \texttt{cont}.
-
-\subsubsection{General Expressions}
-\label{seq:general}
-
-The most general form of CPS transformation receives an
-expression and a continuation as parameters, and returns the expression
-in CPS form with the continuation parameter potentially called in multiple
-tail positions. General expressions can be somewhat voluntarily
-divided into several groups:
-\begin{itemize}
-    \item binding forms --- \texttt{let} and
-        \texttt{loop/recur};
-    \item flow control --- \texttt{if}, \texttt{when},
-        \texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
-    \item function applications and \texttt{apply};
-    \item probabilistic forms --- \texttt{observe},
-        \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
-\end{itemize}
-Functions that transform general expressions accept the
-expression and the continuation as parameters, and are
-consistently named \texttt{cps-of-}\textit{form}, for example,
-\texttt{cps-of-do}, \texttt{cps-of-store}.
-
-\subsection{Implementation Highlights}
-\label{seq:highlights}
-
-So far we introduced the basics of Anglican compilation to
-Clojure. The described approaches and techniques are important
-for grasping the language implementation but relatively well-known.
-In the rest of the section we focus on challenges we met and
-resolved while implementing Anglican, as well as on
-implementation of unique features of Anglican as a probabilistic
-programming language.
-
-\subsubsection{Continuations}
-\label{seq:continuations}
-
-Continuations are functions that are called in tail positions
-with the computed value and state as their arguments --- in CPS
-there is always a function call in every tail position and never
-a value. Continuations are passed to CPS transformers, and when
-transformers are called recursively, the continuations are
-generated on the fly.
-
-There are two critical issues related to generation of
-continuations:
-\begin{itemize}
-    \item unbounded \textit{stack growth} in recursive code;
-    \item code size \textit{explosion} when a non-atomic
-        continuation is symbolically substituted in multiple
-        locations.
-\end{itemize}
-
-\paragraph{Managing stack size}
-
-In implementations of functional programming languages stack
-growth is avoided through \textit{tail call optimization} (TCO).
-However, Clojure does not support a general form of TCO, and
-CPS-transformed code that creates deeply nested calls will
-easily exhaust the stack. Anglican employs a workaround called
-\textit{trampolining} --- instead of inserting a continuation
-call directly, the transformer always wraps the call into a
-\textit{thunk}, or parameterless function. The thunk is returned
-and called by the trampoline (Clojure provides function
-\texttt{trampoline} for this purpose) --- this way the
-computation continues, but the stack is collapsed on every
-continuation call. Function \texttt{continue} implements the
-wrapping and is invoked on every continuation call:
-\begin{lstlisting}[style=default]
-=> (continue 'cont 'value 'state)
-(fn [] (cont value state))
-\end{lstlisting}
-Correspondingly, the full, wrapped CPS form of
-\begin{lstlisting}[style=default]
-(fn [x y] (+ x y))
-\end{lstlisting}
-becomes
-\begin{lstlisting}[style=default]
-(fn [cont $state x y]
-  (fn []
-    (cont (+ x y) $state)))
-\end{lstlisting}
-When the CPS-transformed function is called, it returns a
-\textit{thunk} (a parameterless function) which is then
-re-invoked through the trampoline, with the stack collapsed.
-
 \paragraph{Avoiding exponential code growth}
@@
 \end{lstlisting}
-
-\subsubsection{Primitive Procedures}
-\label{seq:primitive}
-
-When an Anglican function is transformed into a Clojure function
-by \texttt{fn-cps}, two auxiliary parameters are added to the
-beginning of the parameter list --- continuation and state.
-Correspondingly, when a function \textit{call} is transformed
-(by \texttt{cps-of-application} or \texttt{cps-of-apply}), the
-current continuation and the state are passed to the called
-function. Anglican can also call Clojure functions; however,
-Clojure functions do not expect these auxiliary parameters.
-To allow the mixing of Anglican (CPS-transformed) and Clojure
-function calls in Anglican code, the Anglican compiler must be
-able to recognize `primitive' (that is, implemented in Clojure
-rather than in Anglican) functions.
-
-Providing an explicit syntax for differentiating between
-Anglican and Clojure function calls would be cumbersome. Another
-option would be to use meta-data to identify Anglican function
-calls at runtime. However, this would impact performance, and
-good runtime performance is critical for probabilistic
-programs. The approach taken by Anglican is to maintain a list
-of unqualified names of primitive functions, as well of
-namespaces in which all functions are primitive, and recognize
-primitive functions by name --- if a function name is not in the
-list, the function is an Anglican function. Global dynamically-bound 
-variables \texttt{*primitive-procedures*} and
-\texttt{*primitive-namespaces*} contain the initial lists of
-names and namespaces, correspondingly. Of course, local bindings
-can shade global primitive function names. For example,
-\texttt{first} is a Clojure primitive function that takes the first
-element from an ordered collection such as list and vector,
-but inside the let block in the
-following example, \texttt{first} is an Anglican function:
-\begin{lstlisting}[style=default]
-(let [first (fn [[x & y]] x)]
-  (first '[1 2 3]))
-\end{lstlisting}
-The Anglican compiler takes care of the shading by rebinding
-\texttt{*primitive-procedures*} in every lexical scope
-(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
-\texttt{shading-primitive-procedures} automates the shading.
-
-\subsubsection{Probabilistic Forms}
-\label{seq:forms}
-
-There are two proper probabilistic forms turning Anglican into a
-probabilistic programming language --- \texttt{sample} and
-\texttt{observe}. Their purpose is to interrupt deterministic
-computation and transfer control to the inference algorithm.
-Practically, this is achieved through returning
-\textbf{checkpoints} --- Clojure values of the corresponding
-types (\texttt{anglican.trap.sample} or
-\texttt{anglican.trap.observe}). The types contain fields
-specific to each form, as well as the continuation; calling the
-continuation resumes the computation. Checkpoints expose the
-program state to the inference algorithm, and the updated state
-is re-injected into the computation when the continuation is
-called:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression '(sample dist) 'cont)
-(->sample dist cont $state)
-=> (cps-of-expression '(observe dist v) 'cont)
-(->observe dist v cont $state)
-\end{lstlisting}
-Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
-expressions constructors for Clojure records, and they take 
-values of their fields as arguments.
-
-In addition to checkpoints, there are a few other special forms
----  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
-modify program state. These forms are translated into
-expressions involving calls of functions from the
-\texttt{anglican.state} namespace. The \texttt{mem} form, which
-implements memoization, deserves a more detailed explanation.
-
-\paragraph{Memoization}
-The author of a probabilistic model might want to 
-randomly draw a feature value for each entity in a collection,
-but to retain the same drawn
-value for the same entity during a single run of the probabilistic
-program. For example, a person may have brown or green eyes
-with some probability, but the \textit{same} person will
-always have the same eye color. This can be achieved
-through the use of memoized functions. In Anglican, one
-might write:
-\begin{lstlisting}[style=default]
-(let [eye-color (mem (fn [person]
-                       (sample
-                         (categorical
-                           ['brown 0.5]
-                           ['green 0.5]))))]
-  (if (not= (eye-color 'bill) (eye-color 'john))
-    (eye-color 'bill)
-    (eye-color 'john)))
-\end{lstlisting}
-The \texttt{mem} form converts a function to a memoized variant,
-which remembers past inputs and the corresponding outputs, and 
-returns the remembered output when given such a past input, 
-instead of calling the original function with it. As a result,
-for every input, random draws will be made only for the first time 
-that the memoized function is called with the input; all subsequent
-calls with the input will just reuse these draws return the same output.
-
-Memoization is often implemented on top of a mutable
-dictionary, where the key is the argument list and the value is
-the returned value. However, there are no mutable data
-structures in a probabilistic program. Hence, \texttt{mem}'s
-memory is stored as a nested dictionary in the program state
-introduced during CPS transformation (function \texttt{mem-cps}).  
-Every memoized function gets a
-unique automatically generated identifier. Each time a memoized
-function is called, one of two continuations is chosen,
-depending on whether the same function (a function with the same
-identifier) was previously called in the same run of the
-probabilistic program with the same arguments. If the memoized
-result is available, the continuation of the memoized function
-call is immediately called with the stored result. Otherwise,
-the argument of \texttt{mem} is called with a continuation which
-first creates an updated state with the memoized result, and
-then calls the `outer' continuation with the result and the
-updated state:
-\begin{lstlisting}[style=default]
-=> (mem-cps '(foo))
-(let [M (gensym "M")]
-  (fn mem23623 [C $state & P]
-    (if (in-mem? $state M P)
-      ;; previously memoized result
-      (fn []
-        (C (get-mem $state M P) $state))
-      ;; new computation
-      (clojure.core/apply
-        foo
-        ;; memoize result in state
-        (fn [V $state]
-          (fn [] (C V (set-mem $state M P V))))
-        $state
-        P)))))
-\end{lstlisting}
-
-Memoized results are not shared among multiple runs of a
-probabilistic program, which is intended. Otherwise, it would be
-impossible to memoize functions with random results.
-
-\section{Inference Algorithms}
-\label{sec:inference}
-
-A probabilistic program in Anglican may look almost like a
-Clojure program. However, the purpose of executing a
-probabilistic program is different from that of a `regular'
-program: instead of producing the result of a single execution,
-a probabilistic program computes, exactly or approximately, the
-distribution from which execution results are drawn.  Computing
-the distribution is facilitated by an inference algorithm.
-
-Probabilistic programming system Anglican provides a variety of
-approximate inference algorithms. Ideally, Anglican should
-automatically choose the most appropriate algorithm for each
-probabilistic program. In practice, selecting an inference
-algorithm, or a combination thereof, is still a challenging
-task, and program structure, intended use of the computation
-results, performance, approximation accuracy, and other factors
-must be taken into consideration. New algorithms are being
-developed and added to Anglican~\cite{TMP+15,MPT+16,RNL+2016},
-as a part of ongoing research.
-
-In the implementation of Anglican, inference algorithms are
-instantiations of the (ad-hoc) polymorphic function
-\texttt{infer} declared as Clojure's multimethod in the
-\texttt{anglican.{\linebreak[0]}inference} namespace. The function accepts an
-algorithm identifier, a query --- the probabilistic program in
-which to perform the inference, an initial value for the query,
-and optional algorithm parameters.
-
-\subsection{The \texttt{infer} Function}
-\label{sec:infer}
-
-The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
-an appropriate overloading or implementation of the function --- conventionally,
-the identifier is a Clojure keyword (a symbolic constant starting with colon)
-related to the algorithm name, such as \texttt{:lmh} for Lightweight
-Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
-The second parameter is a query as defined by
-the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
-the following Clojure code invokes \texttt{infer} on an Anglican query
-defined anonymously via the \texttt{query} form:
-%If needed, the query
-%can be defined anonymously, right in the argument list of a call
-%to \texttt{infer}:
-\begin{lstlisting}[style=default]
-(let [x 1]
-  (infer :pgibbs (query x) nil))
-\end{lstlisting}
-A query is executed by calling the initial continuation of the
-query, which accepts a value and a state. The state is supplied
-by the inference algorithm, while the value is provided as a
-parameter of \texttt{infer}. A query does not have to have any
-parameters, in which case the value can be simply \texttt{nil}.
-When a query is defined with a binding for the initial value,
-the value becomes available inside the query. A query may
-accept multiple parameters using Clojure's structured binding.
-For instance, it may take multiple parameters as components
-of an input vector. In this case, the initial value is given
-as a structured value, such as a vector, and the components
-of this value are matched to the corresponding
-parameters of the query via the destructuring mechanism of Clojure.
-%The value is destructured via pattern match in case of a structured binding.
-For example,
-\begin{lstlisting}[style=default]
-(defquery my-query [mean sd]
-  (sample (normal mean sd)))
-
-(def samples (infer :lmh my-query [1.0 3.0]))
-\end{lstlisting}
-Finally, any number of auxiliary arguments can be
-passed to \texttt{infer}.
-%Finally, an indefinite number of auxiliary arguments can be
-%passed to \texttt{infer}.
-By convention, the arguments should be
-keyword arguments, and are interpreted in the algorithm-specific
-manner.
-
-\subsection{Internals of an Inference Algorithm}
-\label{seq:internals}
-
-Implementing an inference algorithm in Anglican amounts
-to defining an appropriate version of the \texttt{infer} function and 
-checkpoint handlers for \texttt{sample} and \texttt{observe}.
-The definitions may just reuse default implementations or override them
-with new algorithm-specific treatment of \texttt{sample} and \texttt{observe} forms
-and inference state.
-
-Let us illustrate this implementation step with
-\textit{importance sampling},
-the simplest inference algorithm.
-Here is an implementation of the \texttt{infer} function for the algorithm:
-\begin{lstlisting}[style=default]
-(derive ::algorithm
-        :anglican.inference/algorithm)
-;; invoked when algo parameter is :importance
-(defmethod infer :importance 
-  [algo prog value & {}]
-  (letfn
-    ;; recursive function without any parameter
-    [sample-seq 
-      []
-      ;; lazy infinite sequence 
-      (lazy-seq
-        (cons
-          (:state (exec ::algorithm prog value
-                        initial-state))
-          (sample-seq))))]
-    (sample-seq)))
-\end{lstlisting}
-This implementation is dispatched 
-when \texttt{infer} receives \texttt{:importance} as its \texttt{algo} 
-parameter. (Anglican includes multiple implementations of \texttt{infer}
-for different inference algorithms.) Once dispatched, it lazily 
-constructs an infinite sequence of inference states by repeatedly 
-executing the program \texttt{prog} using \texttt{exec}
-and retrieving the final inference state of the execution 
-using \texttt{:state}. For checkpoint handlers, importance sampling simply
-relies on their default implementations.
-%\texttt{anglican.importance/{\linebreak[0]}infer} just calls
-%\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
-%implementations of checkpoint handlers.
-
-Typically, an inference
-algorithm has its own implementations of
-\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
-both, as well as invokes \texttt{exec} from an elaborated
-conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
-(\texttt{anglican.smc}) are examples of inference algorithms
-where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
-handler is overridden. In addition, SMC runs multiple particles
-(corresponding to user-level threads) simultaneously, while LMH re-runs
-programs from an intermediate continuation rather than from the
-beginning.
-
-\subsection{Addressing of Checkpoints}
-\label{sec:addressing}
-
-Many inference algorithms memoize and reuse earlier computations
-at checkpoints. Variants of Metropolis-Hastings reuse previously
-drawn values~\cite{WSG11}, as well as additional information
-used for adaptive sampling~\cite{TMP+15} at \texttt{sample}
-checkpoints. Asynchronous SMC (Particle Cascade) computes
-average particle weights at \texttt{observe}
-checkpoints~\cite{PWD+14}. Implementations of black-box
-variational inference~\cite{WW13,MPT+16} associate with random
-variables the learned parameters of variational posterior
-distribution.
-
-Checkpoints can be uniquely named at compilation time;
-however, at runtime a checkpoint corresponding to a single code
-location may occur multiple time due to recurrent invocation of
-the function containing the checkpoint. Every unique occurrence
-of a checkpoint must receive a different address. An addressing
-scheme based on computing of stack addresses of checkpoints was
-described in the context of Lightweight
-Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
-over the naive scheme where dynamic occurrences of checkpoints
-are numbered sequentially. However, it impacts the
-performance of probabilistic programs because of the computation
-cost of computing the stack addresses:
-\begin{enumerate}
-    \item On every function call, a component is added to the
-        address. Hence, the address size is linear in stack depth.
-    \item Every function call must be augmented by symbolic
-        information required to compute stack addresses.
-\end{enumerate}
-In addition, the above scheme can still lead to inferior
-correspondence between checkpoints in different traces: in
-Anglican and other probabilistic languages where distributions
-are first-class objects checkpoints with incompatible arguments
-can correspond to the same stack address. Consider the following
-program fragment:
-\begin{lstlisting}[style=default]
-(let [dist (if use-gamma
-               (gamma 2. 2.)
-               (normal 0. 1.))]
-  (sample dist))
-\end{lstlisting}
-The \texttt{sample} checkpoint has the same stack address in
-different traces. However, the random values should not be
-reused between different distributions. Further on, in some
-algorithms, such as black-box variational inference, the role of
-checkpoint addresses is semantic rather than heuristic ---
-appropriate correspondence must be established
-between checkpoints in different traces for the algorithm to
-work.
-
-To overcome the above problems, Anglican uses a different scheme
-which is almost as efficient as the scheme based on stack addresses for
-reuse of previously drawn values, while producing addresses of constant size,
-and allows manual computation of checkpoint addresses at runtime
-when the default automatic scheme is insufficient. According to
-the scheme:
-\begin{itemize}
-    \item A checkpoint may accept an auxiliary argument --- the
-        checkpoint identifier. If specified, the identifier is the first
-        argument of a checkpoint. For example \texttt{(sample
-        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
-        with identifier \texttt{x1}.
-    \item If the identifier is omitted, a unique identifier is
-        generated automatically as a fresh symbol.
-    \item At runtime, the address of a checkpoint invocation has
-        the form
-        \texttt{[}\textit{checkpoint-identifier
-        number-of-previous-occurrences}\texttt{]}, where the
-        occurrences are of a checkpoint with the same checkpoint
-        identifier.
-    \item If a sequence of invocations of the same checkpoint is
-        interrupted by a different checkpoint, the number of
-        previous occurrences is rounded up to a multiple of
-        an integer. For efficiency, a small power of 2 is used,
-        such as $2^4 = 16$.
-\end{itemize}
-
-Consider the following simplified Anglican query:
-
-\begin{lstlisting}[style=default]
-(defm foo []
-  (if (sample 'C1 (flip 0.5))
-    (foo)
-    (bar)))
-
-(defm bar []
-  (case (sample 'C2 (uniform-discrete 0 3))
-    0 (bar)
-    1 (foo)
-    2 (sample 'C3 (normal 0. 1.))))
-
-(defquery baz
-  (foo))
-\end{lstlisting}
-
-An execution of the query may result in the following
-sequence of checkpoint invocations:
-\begin{lstlisting}[style=default]
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C2 ...)
-    (sample 'C1 ...)
-    (sample 'C1 ...)
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C3 ...)
-\end{lstlisting}
-According to the addressing scheme, the addresses generated
-for this invocations are
-\begin{lstlisting}[style=default]
-    [C1 0]
-    [C2 0]
-    [C2 1]
-    [C1 16]
-    [C1 17]
-    [C1 18]
-    [C2 16]
-    [C3 0]
-\end{lstlisting}
-If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
-\begin{lstlisting}[style=default]
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C1 ...)
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C2 ...)
-    (sample 'C3 ...)
-\end{lstlisting}
-The addresses for the new sequence are
-\begin{lstlisting}[style=default]
-    [C1 0]
-    [C2 0]
-    [C1 16]
-    [C1 17]
-    [C2 16]
-    [C2 17]
-    [C3 0]
-\end{lstlisting}
-Despite the change, the correspondence between checkpoints of
-similar types occurring in similar positions (relative positions
-in contiguous subsequences of a certain type) is preserved, and
-drawn values can be reused efficiently:
-
-\vspace{\baselineskip}
-{
-\begin{tabular}{l l}
-    \textbf{old} & \textbf{new} \\ \hline
-    {}\texttt{[C1 0]} & \texttt{[C1 0]} \\
-    {}\texttt{[C2 0]} & \texttt{[C2 0]} \\
-    {}\texttt{[C2 1]} & \textit{unused} \\
-    {}\texttt{[C1 16]} & \texttt{[C1 16]} \\
-    {}\texttt{[C1 17]} & \texttt{[C1 17]} \\
-    {}\texttt{[C1 18]} & \textit{unused} \\
-    {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
-    \textit{missing} & \texttt{[C2 17]} \textit{drawn}\\
-    {}\texttt{[C3 0]} & \texttt{[C3 0]}
-\end{tabular}
-\vspace{\baselineskip}}
-
-Note that correspondence between checkpoints in different
-traces plays the role of an heuristic in Metropolis-Hastings
-family of algorithms. Any correspondence (or no correspondence,
-meaning all values must be re-drawn from their proposal
-distributions) is valid, and reused values from the previous
-invocation which are not in support or have zero probability in
-the new invocation are simply ignored and new values are
-re-drawn.
-
-This way, each occurrence of a checkpoint has unique address,
-but small disturbances --- removal or addition of a single or just
-a few checkpoints --- are unlikely to derail the entire sequence.
-The probability of derailment depends on the padding. The
-padding can be safely, and without any impact on performance,
-set to rather large numbers. However, rounding up to a multiple
-of 16 proved to be appropriate for all practical purposes.
-
-Function \texttt{checkpoint-id} in the
-\texttt{anglican.inference} namespace automates generation of
-checkpoint addresses and can be used from any inference
-algorithm.
-
-\section{Definitions and Runtime Library}
-\label{seq:runtime}
-
-A Clojure namespace that includes a definition of an Anglican
-program imports (`requires') two essential namespaces:
-\texttt{anglican.emit} and \texttt{anglican.runtime}. The former
-provides macros for defining Anglican programs
-(\texttt{defquery}, \texttt{query}) and functions
-(\texttt{defm}, \texttt{fm}, \texttt{mem}), as well as Anglican
-bootstrap definitions that must be included with every program
---- first of all, CPS implementations of higher-order functions.
-\texttt{anglican.emit} can be viewed as the Anglican
-\textit{compiler tool}, which helps transform Anglican code into
-Clojure before any inference is performed.
-
-\texttt{anglican.runtime} is the Anglican runtime library. For
-convenience, it exposes common mathematical functions
-(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
-\texttt{exp}, etc.), but most importantly, it provides
-definitions of common distributions. Each distribution object
-implements a distribution interface with the \texttt{sample*} and \texttt{observe*}
-methods; this interface is defined using Clojure's protocol mechanism
-(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
-The \texttt{sample*} \textit{method} returns a random sample and
-roughly corresponds to the \texttt{sample} checkpoint,
-the \texttt{observe*} \textit{method} returns the log probability
-of the value and roughly corresponds to the \texttt{observe}
-checkpoint. The methods can be, and sometimes are called from
-handlers of the corresponding checkpoints, but do not have to
-be. For example, in LMH either the \texttt{sample*} or
-the \texttt{observe*} \textit{method} is called for a
-\texttt{sample} checkpoint, depending on whether the value is
-drawn or reused.
-
-Macro \texttt{defdist} should be used to define distributions.
-\texttt{defdist} takes care of defining a separate type
-for every distribution so that Clojure multimethods (or overloaded
-methods) can be dispatched on
-distribution types when needed, e.g. for custom proposal distributions 
-used in an inference algorithm. The
-Bernoulli distribution could be defined as follows:
-\begin{lstlisting}[style=default]
-(defdist bernoulli
-  "Bernoulli distribution"
-  [p]
-  (sample* [this] (if (< (rand) p) 1 0))
-  (observe* [this value]
-    (Math/log (case value
-                1 p
-                0 (- 1. p)
-                0.))))
-\end{lstlisting}
-
-In addition to distributions, Anglican provides \textit{random processes},
-which define sequences of random variables that are not independent and
-identically distributed. Random processes are similar to the so called
-`exchangeable random procedures' in Church \cite{GMR+08} and
-Venture \cite{MSP14}. However, random sequences generated by Anglican random 
-processes are not required to be exchangeable. Random processes are
-defined using the \texttt{defproc} macro and implement the
-\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
-protocol has two methods
-\begin{itemize}
-\item \texttt{produce}, which returns the distribution on the next
-random variable in the sequence, and
-\item \texttt{absorb}, which incorporates the value for the next random
-variable and returns an updated random process.
-\end{itemize}
- As an example, here is a definition of a beta-Bernoulli process, in which
-each random variable is distributed according to a Bernoulli distribution with
-an unknown parameter that is drawn from a beta distribution:
-\begin{lstlisting}[style=default]
-(defproc beta-bernoulli
-  "beta-Bernoulli process"
-  [a b]
-  (produce [this] (bernoulli (/ a (+ a b))))
-  (absorb [this x]
-    (case x
-      0 (beta-bernoulli a (inc b))
-      1 (beta-bernoulli (inc a) b))))
-\end{lstlisting}
-Unlike typical implementations of exchangeable random processes, Anglican's random processes do not have mutable state.
-The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican.
-A sequence of random values can be generated using a recursive loop in which \texttt{absorb} returns the updated process for the next iteration.
-For example:
-\begin{lstlisting}[style=default]
-(defm sample-beta-binomial [n a b]
-  (loop [process (beta-bernoulli a b)
-         values []]
-    (if (= (count values) n)
-      values
-      (let [dist (produce process)
-            value (sample dist)]
-      (recur (absorb process value)
-             (conj values value))))))
-\end{lstlisting}
-Similarly, random processes can also be used to recursively observe a sequence of values:
-\begin{lstlisting}[style=default]
-(loop [process (beta-bernoulli 1.0 1.0)
-       [value0 & values0] values] 
-  (let [dist (produce process)]
-    (observe dist value0)
-    (recur (absorb process value0) values0)))
-\end{lstlisting}
-
-% \begin{itemize}
-%     \item Values produced by the process are sampled from the distribution
-%         returned by \texttt{produce}.
-%     \item When a value is observed from the distribution, a new
-%         process, with the observed value absorbed, is returned
-%         by \texttt{absorb}.
-% \end{itemize}
-
-% A common pattern when programming with random processes in 
-% Anglican is to generate a sequences of sampled values and 
-% constructing an updated process at each step
-
-\section{Case Study}
-\label{seq:study}
-
-The description of Anglican language and environment would be
-incomplete without a case study showing the process of building
-and executing a solution of an inference problem. This case
-study takes a problem for which a solution is not immediately
-obvious, \textit{the Deli dilemma}, and guides through
-writing an Anglican program for the problem, executing the
-program, and post-processing results.
-
-The program presented in this section is intentionally short and
-simple. Anglican is capable of compiling and running elaborated
-programs handling large amounts of data. Advanced examples of
-Anglican programs and inference can be found in literature on
-applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.
-
-\subsection{The Deli Dilemma}
-\label{seq:deli}
-
-Imagine that we are facing the following dilemma:
-\begin{quote}
-A customer wearing round sunglasses came to a deli at 1:13pm,
-and grabbed a sandwich and a coffee. Later on the same day, a
-customer wearing round sunglasses came at 6:09pm and ordered a
-dinner.  Was it the same customer?
-\end{quote}
-In addition to the story above, we know that:
-\begin{itemize}
-    \item There is an adjacent office quarter, and it takes
-        between 5 and 15 minutes on average to walk from an
-        office to the deli, where different average times are  for
-        different buildings in the quarter.
-    \item Depending on traffic lights, the walking time varies by about 2
-        minutes.
-    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
-    \item The waiter's odds that this is the same customer are 2 to 1.
-\end{itemize}
-
-\subsection{Anglican Query}
-\label{seq:deli-query}
-
-We want to formalize the dilemma in an Anglican query, based on
-the knowledge we have. Let us formalize the knowledge in
-Clojure (the times are in minutes). First, we encode our prior
-information which holds true independently of the customer's
-visit:
-
-\begin{lstlisting}[style=default]
-(def mean-time-to-arrive
-     "average time to arrive"
-     10.)
-(def sd-time-to-arrive
-     "standard deviation of arrival time"
-     3.)
-(def time-sd
-     "walking time deviation"
-     1.)
-\end{lstlisting}
-
-Then, we record our observations, based on which we want to
-solve the dilemma:
-
-\begin{lstlisting}[style=default]
-(def lunch-delay
-     "time between lunch break and lunch order"
-     13.)
-(def dinner-delay
-     "time between end of day and dinner order"
-     9.)
-(def p-same
-     "prior probability of the same customer"
-     (/ 2. 3.))
-\end{lstlisting}
-
-For inference, one often chooses a known distribution to
-represent uncertainty. We choose the normal
-distribution for representing uncertainty about average arrival
-time.
-
-\begin{lstlisting}[style=default]
-(def time-to-arrive-prior 
-     "prior distribution of average arrival time"
-     (normal mean-time-to-arrive
-             sd-time-to-arrive))
-\end{lstlisting}
-
-There are two possibilities: either the same customer visited
-the deli twice, or two different customers came to the deli,
-one for lunch and the other for dinner. We define an \textbf{Anglican} function
-for each case. Note that this is the first time we switch from
-Clojure to Anglican. The functions must be written in Anglican
-(and hence defined using \texttt{defm} instead of \texttt{defn})
-because they contain probabilistic forms \texttt{sample} and
-\texttt{observe}.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-  "observe the same customer twice"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive time-sd)
-             dinner-delay)
-    [time-to-arrive]))
-
-(defm different-customers
-  "observe different customers"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive-1
-          (sample time-to-arrive-prior)
-        time-to-arrive-2
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive-1 time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive-2 time-sd)
-             dinner-delay)
-    [time-to-arrive-1 time-to-arrive-2]))
-\end{lstlisting}
-
-Both functions have the same structure: we first `guess' the
-average arrival time and then observe the actual time from a
-distribution parameterized by the guessed time. However, in
-\texttt{same-customer} the average arrival time is the same for
-both the lunch and the dinner, while in
-\texttt{different-customers} two average arrival times are
-guessed independently.
-
-We are finally ready to define the query in \textbf{Anglican}.
-
-\begin{lstlisting}[style=default]
-(defquery deli [time-to-arrive-prior
-                lunch-delay
-                dinner-delay]
-  (let [is-same-customer (sample (flip p-same))
-        observe-customer (if is-same-customer
-                            same-customer
-                            different-customers)]
-    {:same-customer is-same-customer,
-     :times-to-arrive (observe-customer 
-                        time-to-arrive-prior
-                        lunch-delay
-                        dinner-delay)}))
-\end{lstlisting}
-
-Performing inference on query \texttt{deli} computes the
-probability that the same customer visited the deli twice, as
-well as probability distributions of average arrival times for
-both cases.
-
-\subsection{Inference}
-\label{seq:deli-inference}
-
-Having defined the query, we are now ready to run the query
-using an inference algorithm. Function \texttt{doquery} provided
-by the \texttt{anglican.core} namespace accepts the inference
-algorithm, the query, and optional parameters, and returns a
-lazy sequence of samples. We use here the inference algorithm
-called Lightweight Metropolis-Hastings (LMH). The algorithm is
-somewhat slow to converge but can be used with any Anglican
-query, and should be robust enough for our simple problem.
-
-We bind the results of \texttt{doquery} to variable
-\texttt{samples}, to analyse the results later. However, since
-the sequence is lazy, no inference is performed and no samples
-are generated until they are retrieved and processed.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :lmh deli nil))
-\end{lstlisting}
-
-To approximate the inferred distribution, we extract a finite
-subsequence of samples. Many algorithms use an initial
-subsequence of samples to converge to the target distribution.
-Hence, we drop initial $N$ samples ($N$ is 5000 in the code
-snippet below), and collect the next $N$ samples.
-
-\begin{lstlisting}[style=default]
-(def N 5000)
-(def results (map get-result
-                   (take N (drop N samples))))
-\end{lstlisting}
-
-Based on the collected samples we compute an approximation of
-the posterior probability \texttt{p-same+} that the same
-customer visited the deli twice.
-
-\begin{lstlisting}[style=default]
-(def p-same+ 
-     (/ (count (filter :same-customer results))
-        (double N)))
-\end{lstlisting}
-The \texttt{:same-customer} here represents a function
-that looks up the entry with the same name in a given map.
-The map stores the result of a single sampled execution,
-and the \texttt{:same-customer} entry in the map
-records whether the same customer visits the deli in the execution.
-The \texttt{filter} function in \texttt{p-same+} 
-goes through all the maps in \texttt{results}
-and picks only the ones whose executions involve just one customer.
-
-With the specified observations, \texttt{p-same+} is $\approx
-0.12$. The probability is much lower than the
-waiter's guess \texttt{p-same} ($\frac 2 3$).
-Of course, the results may vary from run to run, and, for a given
-algorithm, the accuracy depends on the number of samples we
-decide to retrieve.
-
-In addition to computing the posterior probability that the same
-customer visited the deli twice, we may want to know the average
-time (or times) to arrive. In Bayesian inference, it is common
-to report distributions instead of `most likely' values. We use
-query results in retrieved samples to approximate the
-distributions, and plot distribution histograms for the same
-customer visiting twice (Figure~\ref{fig:dist-same}) and
-two different customers (Figure~\ref{fig:dist-diff})
-
-\begin{lstlisting}[style=default]
-;; single customer                      
-(def time-to-arrive+ 
-    (map (fn [x] (first (:times-to-arrive x)))
-         (filter :same-customer results)))
-(def mean-to-arrive+ (mean time-to-arrive+))
-(def sd-to-arrive+ (std time-to-arrive+))
-
-;; two customers
-(def times-to-arrive+ 
-     (map :times-to-arrive 
-          (filter  
-            (fn [x] (not (:same-customer x)))
-            results)))
-(def mean-1-to-arrive+
-     (mean (map first times-to-arrive+)))
-(def sd-1-to-arrive+
-     (std (map first times-to-arrive+)))
-(def mean-2-to-arrive+
-     (mean (map second times-to-arrive+)))
-(def sd-2-to-arrive+
-     (std (map second times-to-arrive+)))
-\end{lstlisting}
-
-In addition to plotting the distribution histograms, we use
-functions \texttt{mean}, \texttt{std} provided
-in the \texttt{anglican.{\linebreak[0]}stat} namespace
-along with other
-useful statistical functions.
-This is another illustration of advantage of tight
-integration between Clojure and Anglican --- probabilistic
-models are expressed in Anglican. However, processing of data and
-results can rely on the full power of Clojure.
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
-    \caption{Arrival time distribution for a single customer.}
-    \label{fig:dist-same}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
-    \caption{Arrival time distributions for two customers.}
-    \label{fig:dist-diff}
-\end{figure}
-
-For inference, we chose to use  Lightweight
-Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
-probabilistic programming is that models are separated from
-inference. To switch to a different inference algorithm we just
-need to pass a different value to \texttt{doquery}. For example,
-we may decide to use Black-Box Variational Bayes (BBVB), which
-may not work equally well for all probabilistic programs, but is
-much faster to converge.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :bbvb deli nil))
-\end{lstlisting}
-
-We can still use samples and summary statistics with BBVB to
-approximate the posterior distribution. However, variational
-inference approximates the posterior by known distributions,
-and we can directly retrieve the distribution parameters.
-
-\begin{lstlisting}[style=default]
-(clojure.pprint/pprint
-  (anglican.bbvb/get-variational 
-    (nth samples N)))
-\end{lstlisting}
-
-\begin{lstlisting}[style=default]
-{S28209
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- S28217
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- S28215
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- S28219
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
-\end{lstlisting}
-
-We can guess that the variational distributions correspond to the
-prior distributions in the \texttt{sample} forms, in the order
-of appearance. However, it would help if we could use more
-informative labels instead of automatically generated symbols
-\texttt{S28209}, \texttt{S28217}, etc. Here the option to
-specify identifiers for probabilistic forms explicitly comes
-handy. If we modify the \texttt{sample} forms to use explicit identifiers
-(only the forms are shown for brevity), the output becomes much
-easier to analyse.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-      ...
-      (sample :arrival-time-same
-              time-to-arrive-prior)]
-      ...)
-
-(defm different-customers
-      ... 
-      (sample :arrival-time-first
-              time-to-arrive-prior)
-      ...
-      (sample :arrival-time-second
-              time-to-arrive-prior)]
-      ...)
-
-(defquery deli ...
-      ...
-      (sample :same-or-different
-              (flip p-same))
-      ...)
-\end{lstlisting}
-
-The output becomes much easier to interpret and analyse
-programmatically.
-
-\begin{lstlisting}[style=default]
-{:arrivate-time-same
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- :arrival-time-first
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- :arrival-time-second
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- :same-or-different
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
-\end{lstlisting}
-
-This completes the case study, where we showed a probabilistic
-programming solution of a problem, implemented and analysed in
-Anglican and Clojure, using two different inference algorithms.
-
-\section{Conclusion}
-\label{seq:summary}
-
-In this paper, we presented design and implementation internals
-of the \emph{probabilistic programming system} Anglican. Implementing
-a language is an interesting endeavour, in particular when the
-language implements a new paradigm, in this case probabilistic
-programming. Functional programming is a natural complement of
-probabilistic programming --- the latter allows both concise and
-expressive specification of probabilistic generative models and
-efficient implementation of inference algorithms.
-
-Implementing a probabilistic language on top of and in tight
-integration with a functional language, Clojure, both helped
-us to accomplish an ambitious goal in a short time span, and
-provided important insights on structure and semantics of
-probabilistic concepts incorporated in Anglican. Computational
-efficiency and expressive power of Anglican owe to adherence to
-the functional approach as much as to rich inference
-opportunities of the Anglican environment.
-
-\acks
-
-This work was partially supported under DARPA PPAML through the
-U.S. AFRL under Cooperative Agreement number FA8750- 14-2-0006,
-Sub Award number 61160290-111668.
-
-\bibliographystyle{abbrvnat}
-\bibliography{refs}
-
-\end{document}

### 2017-02-04 paper/expression-kinds.tex
many changes in progress

--- 2016-11-28 paper/paper.tex
@@
-\documentclass[preprint]{sigplanconf}
-\linespread{1.00}
-
-\usepackage[color]{changebar}
-\usepackage{enumitem}
-\usepackage[lighttt]{lmodern}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-
-\usepackage{listings}
-\usepackage{lmodern}
-\ttfamily
-\DeclareFontShape{OT1}{lmtt}{m}{it}
-     {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
-
-\usepackage{hyperref}
-\usepackage{url}
-
-\input{anglican-color}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-% Authors: comment using these commands
-
-\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:}
-        {\rm #1}\end{minipage}}}
-
-\cbcolor{red}
-% Authors: use changebars to mark changed regions
-% \cbstart
-% ...
-% \cbend
-
-\title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-    Anglican is a probabilistic programming system designed to interoperate
-    with Clojure and other JVM languages. We introduce the
-    programming language Anglican, outline our design choices,
-    and discuss in depth the implementation of the Anglican language
-    and runtime, including macro-based compilation, extended
-    CPS-based evaluation model, and functional representations for
-    probabilistic paradigms, such as a distribution,
-    a random process, and an inference algorithm.
-
-    We show that a probabilistic functional language can be
-    implemented efficiently and integrated tightly with a
-    conventional functional language with only moderate
-    computational overhead. We also demonstrate how advanced
-    probabilistic modeling concepts are mapped naturally
-    to the functional foundation.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-
-For data science practitioners, statistical inference is typically
-just one step in a more elaborate analysis workflow. The first stage
-of this work involves data acquisition, pre-processing and cleaning.
-This is often followed by several iterations of exploratory model
-design and testing of inference algorithms. Once a sufficiently
-robust statistical model and a corresponding inference algorithm have
-been identified, analysis results must be post-processed, visualized,
-and in some cases integrated into a wider production system.
-
-Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
-represent generative models as programs written in a specialized
-language that provides syntax for the definition and conditioning of
-random variables. The code for such models is generally concise,
-modular, and easy to modify or extend. Typically inference can be
-performed for any probabilistic program using one or more generic
-inference techniques provided by the system backend, such as
-Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
-\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
-of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
-These generic techniques may be less statistically efficient
-than techniques tailored to a specific model. However,
-probabilistic programming facilitates simpler implementations of
-models making the inference inherently faster.
-
-While probabilistic programming systems shorten the iteration cycle in
-exploratory model design, they typically lack basic functionality needed for
-data I/O, pre-processing, and analysis and visualization of inference results.
-In this paper, we describe the implementation of
-Anglican~\cite{TMW15, Anglican}, a probabilistic
-programming language that tightly integrates with
-Clojure~\cite{H08,Clojure}, a general-purpose programming language
-that runs on the Java Virtual Machine (JVM). Both languages share a common
-syntax, and can be invoked from each other. This allows Anglican programs to
-make use of a rich set of libraries written in both Clojure and Java.
-Conversely, Anglican allows intuitive and compact specification of models for
-which inference may be performed as part of a larger Clojure project.
-
-There are several ways to build a programming language
-on top of or besides another language.  The easiest is an
-interpreter --- a program that reads a program, in its entirety
-or line-by-line, and executes it by applying operational
-semantics of a certain kind to the language. \textsc{Basic} is
-famous for line-by-line interpreted implementations.
-
-Another approach is to write a compiler, either to a virtual
-architecture, so called p-code or byte-code, or to real
-hardware. Here, the whole program is translated from the
-`higher-level' source language to a `lower-level' object
-language, which can be directly executed, either by hardware or
-by an interpreter --- but the latter interpreter can be made
-simpler and more efficient  than an interpreter for the source
-language.
-
-On top of these two approaches are methods in which a new
-language is implemented `inside' another language of the same
-level of abstraction. Different languages provide different
-means for this; Lisp is famous for the macro facility
-that allows to extend the language almost without
-restriction --- by writing \textit{macros}, one adds new
-constructs to the existing language. There are several uses of
-macros --- one is to extend the language \textit{syntax}, for
-example, by adding new control structures; another is to keep
-the existing syntax but alter the \textit{operational semantics}
---- the way programs are executed and compute their outputs.
-
-Anglican is implemented in just this way --- a macro facility
-provided by Clojure, a Lisp dialect, is used both to extend
-Clojure with constructs that delimit probabilistic code, and to
-alter the operational semantics of Clojure expressions inside
-probabilistic code fragments. Anglican claims its right to count
-as a separate language because of the ubiquitous probabilistic
-execution semantics rather than a different syntax,
-which is actually an advantage rather than a drawback ---
-Clojure programmers only need to know how to specify the
-boundaries of Anglican programs, but can use familiar Clojure
-syntax to write probabilistic code.
-
-An implementation of Anglican must therefore address three issues:
-\begin{itemize}
-    \item the Clojure syntax to introduce probabilistic Anglican
-        code inside Clojure modules;
-    \item source-to-source transformation of Anglican programs
-        into Clojure, so that probabilistic execution becomes
-        possible;
-    \item algorithms which run Clojure code, obtained by
-        transforming Anglican programs, according to the
-        probabilistic operational semantics.
-\end{itemize}
-Execution of probabilistic programs by inference algorithms is
-different from execution of deterministic programs.  A
-probabilistic program is executed multiple times, often hundreds
-of thousands or even millions of times for a single inference
-task. Random choices may affect which parts of the program code
-are executed and how often. Many inference algorithms require
-re-running the program multiple times partially, from a certain
-point on.  Different executions may employ different random
-choices. However, for efficient inference a correspondence
-between random choices in different executions should be
-maintained. These are just some of the challenges which were
-faced and solved during development of Anglican.
-
-Comparisons of Anglican with other implementations of
-probabilistic programming languages~\cite{SGG15}\cite[pp.
-32--33]{P16} demonstrate that Anglican achieves state-of-the-art
-computational efficiency without sacrificing expressiveness.
-Anglican language syntax, compilation, invocation, and runtime
-support of Anglican queries are discussed in detail in further
-sections.
-
-\section{Design Outline}
-\label{sec:design}
-
-An Anglican program, or \textit{query}, is compiled into a
-Clojure function. When inference is performed with a provided
-algorithm, this produces a sequence of samples. Anglican shares
-a common syntax with Clojure; Clojure functions can be called
-from Anglican code and vice versa. A simple program in Anglican
-can look like the following code:
-%\begin{minipage}{\linewidth}
-\begin{lstlisting}[style=default]
-(defquery model data
-  "chooses a distribution
-   which describes the data"
-  (let [;;; Guess a distribution.
-        dist (sample (categorical
-                       [[normal 0.5]
-                        [gamma 0.5]]))
-        a (sample (gamma 1 1))
-        b (sample (gamma 1 1))
-        d (dist a b)]
-
-    ;;; Observe samples from the distribution.
-    (loop [observations data]
-      (when (not-empty observations)
-        ;; Retrieve the first observation as `o',
-        ;; and store the rest of observations in 
-        ;; 'observations*'.
-        (let [[o & observations*] observations]
-          ;; Observe 'o' from the guessed 
-          ;; distribution 'd'.
-          (observe d o))
-        ;; Proceed to the next iteration with
-        ;; the rest of observations.
-        (recur observations*)))
-
-    ;;; Return the distribution and parameters.
-    [d a b]))
-\end{lstlisting}
-%\end{minipage}
-The query builds a model for the input $\texttt{data}$, 
-a sequence of data points.
-It defines a probability distribution on three variables, 
-$\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
-for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
-positive parameters for the type. Concretely,
-using the $\texttt{sample}$ forms, the query
-first defines a so called prior distribution on these three variables,
-and then it adjusts this prior distribution based on observations in
-$\texttt{data}$ using the \texttt{observe} form.
-Samples from this adapted distribution (also called posterior distribution)
-can be obtained by running the query under one of Anglican's
-inference algorithms.
-
-Internally, an Anglican query is represented by a computation in
-\textit{continuation passing style} (CPS)~\cite{AJ89}, and inference algorithms
-exploit the CPS structure of the code to intercept probabilistic
-operations in an algorithm-specific way\footnote{\cite{GS15}
-also describe a CPS-based implementation of a probabilistic
-programming language.}. Among the available
-inference algorithms there are Particle Cascade~\cite{PWD+14},
-Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
-Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
-Inference on Anglican queries generates a lazy sequence of
-samples, which can be processed asynchronously in Clojure
-code for analysis, integration, and decision making.
-
-Clojure (and Anglican) runs on the JVM and gets access to a wide choice of Java
-libraries for data processing, networking, presentation, and imaging.
-Conversely, Anglican queries can be called from Java and other JVM languages.
-Programs involving Anglican queries can be deployed as JVM
-\textit{jars}, and run without modification on any platform for
-which JVM is available.
-
-A probabilistic program, or query, mostly runs deterministic
-code, except for certain checkpoints, in which probabilities are
-involved, and normal, linear execution of the program is
-disrupted. In Anglican and similar languages there are two types
-of such checkpoints:
-
-\begin{itemize}
-    \item drawing a value from a random source (\texttt{sample});
-    \item conditioning a computed value on a random source
-        (\texttt{observe}).
-\end{itemize}
-
-Anglican can be mostly implemented as a regular programming
-language, except for the handling of these checkpoints.
-Depending on the \textit{inference algorithm}, \texttt{sample}
-and \texttt{observe} may result in implicit input/output
-operations and control changes. For example, \texttt{observe} in
-particle filtering inference algorithms \cite{WVM14} is a non-deterministic
-control statement at which a particle (corresponding to
-a user-level thread executing a program)
-can be either replicated or terminated. Similarly, in 
-Metropolis-Hastings \cite{WSG11},
-\texttt{sample} is both an input
-statement which `reads' values from a random source,
-and a non-deterministic control statement
-(with delayed effect), eventually affecting acceptance or
-rejection of a sample.
-
-Because of the checkpoints, Anglican programs must allow the
-inference algorithm to step in, recording information and
-affecting control flow. This can be implemented through
-coroutines/cooperative multitasking, and parallel execution/preemptive
-multitasking, as well as through explicit
-maintenance of program continuations at checkpoints. Anglican
-follows the latter option. Clojure is
-a functional language, and continuation-passing style (CPS)
-transformation is a well-developed technique in the area of
-functional languages. Implementing a variant of CPS
-transformation seemed to be the most flexible and lightweight
-option --- any other form of concurrency would put a higher burden
-on the underlying runtime (JVM) and the operating system.
-Consequently, Anglican has been implemented as a CPS-transformed
-computation with access to continuations in probabilistic
-checkpoints. The Anglican `compiler', represented by a set of
-functions in the \texttt{anglican.trap} namespace, accepts a
-Clojure subset and transforms it into a variant of CPS
-representation, which allows inference algorithms to intervene
-in the execution flow at probabilistic checkpoints.
-
-Anglican is intended to co-exist with Clojure and be a part of
-the source of a Clojure program. To facilitate this, Anglican
-programs, or queries, are wrapped by macros (defined in the
-\texttt{anglican.emit} namespace), which call the CPS
-transformations and define Clojure values
-%Clojure objects
-suitable for passing
-as arguments to inference algorithms (\texttt{defquery},
-\texttt{query}). In addition to defining entire queries,
-Anglican promotes modularization of inference algorithms through
-the definitions of \textit{probabilistic functions} using
-\texttt{defm}
-%and \texttt{fm}
-(Anglican counterparts of Clojure \texttt{defn}).
-%and \texttt{fn}).
-Probabilistic functions are
-written in Anglican, may include probabilistic forms
-\texttt{sample} and \texttt{observe}, and can be seamlessly
-called from inside Anglican queries, just like functions locally
-defined within the same query.
-
-Operational semantics of Anglican queries is different from that
-of Clojure code, therefore queries must be called through
-inference algorithms, rather than `directly'.  The
-\texttt{anglican.inference} namespace declares the (ad-hoc) polymorphic
-function \texttt{infer} using Clojure's multimethod mechanism.
-This function accepts an Anglican query and
-returns a lazy sequence of weighted samples from the
-distribution defined by the query.  When inference
-is performed on an Anglican query, the query is run by a
-particular inference algorithm. Inference algorithms must
-provide an implementation for \texttt{infer}, as well as
-override the polymorphic function \texttt{checkpoint} (defined
-as a multimethod) so as to handle \texttt{sample} and
-\texttt{observe} in an algorithm-specific manner and
-to construct an appropriate result on the
-termination of a probabilistic program.
-
-Finally, Anglican queries use `primitive', or commonly known
-and used, distributions, to draw random samples and condition
-observations. Many primitive distributions are provided by the
-\texttt{anglican.runtime} namespace, and an additional
-distribution can be defined by the user by implementing a
-particular set of functions for the distribution
-(via Clojure's protocol mechanism). The \texttt{defdist} macro
-provides a convenient syntax for defining primitive distributions.
-
-\section{Language}
-\label{sec:language}
-
-The Anglican language is a subset of Clojure\footnote{It would be
-possible to support almost full Clojure by expanding all macros
-in the Anglican source code. However, in Clojure, unlike in
-Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
-of macro-expansion of derived special forms is not well
-specified and implementation specific.}.  Anglican queries
-are defined within \texttt{defquery} as shown in the previous section,
-using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case},
-\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
-other forms of Clojure may be supported in the future but are not now.
-In \texttt{let} bindings and \texttt{fn} argument lists,
-Clojure's pattern matching mechanism for vector data type
-(called vector destructuring) is supported. Also, compound
-literals for vectors, hash maps, and
-sets are supported just like in Clojure.
-
-\subsection{Core Library}
-\label{sec:core}
-
-All of Clojure's core library except for higher-order functions
-(functions that accept other functions as arguments) is
-available in Anglican. Higher-order functions cannot be reused
-from Clojure, as they have to be re-implemented to accept
-functional arguments in CPS form. The following higher-order
-functions are implemented: \texttt{map}, \texttt{reduce},
-\texttt{filter}, \texttt{some}, \texttt{repeatedly},
-\texttt{comp}, \texttt{partial}.
-
-\subsection{Tail Call}
-\label{sec:tail}
-
-Clojure provides special forms \texttt{loop} and \texttt{recur}
-for writing tail-recursive programs. Such forms are not necessary
-in Anglican, because Anglican programs are CPS-converted and do not
-use the stack. For instance, no recursive call in Anglican can lead to stack overflow.
-In fact, in Anglican, it is recommended to use recursive calls to functions instead of
-\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
-for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop}
-will lead to unpredictable behaviour and hard-to-catch errors.
-
-\section{Macro-based Compilation}
-\label{sec:compilation}
-
-Compilation of Anglican into Clojure is built around a variant of CPS
-transformation. In a basic CPS-transformed program, a
-continuation receives a single argument --- the computed value.
-In Anglican, there are two flows of computation going in
-parallel: values are computed by functional code, and,
-at the same time, the state of the probabilistic program,
-used by inference algorithms, is updated by
-probabilistic forms. Because of that, in Anglican a continuation
-accepts  \textit{two} arguments:
-\begin{itemize}
-    \item the computed value;
-    \item the internal state, bound to the local variable
-        \texttt{\$state} in every lexical scope.
-\end{itemize}
-
-The compilation relies on the Clojure \textit{macro} facility,
-and implemented as a library of \textit{functions} in namespace
-\texttt{anglican.trap}, which are invoked by macros.  The CPS
-transformation is organized in top-down manner.  The top-level
-function is  \texttt{cps-of-expression}, which receives an
-expression and a continuation, and returns the expression in the
-CPS form.  For example, the CPS transformation of constant
-\texttt{1} with continuation \texttt{cont} thus takes the
-following form:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression 1 'cont)
-(cont 1 $state)
-\end{lstlisting}
-
-\subsection {The State}
-\label{sec:state}
-
-The state (\texttt{\$state}) is threaded through the computation
-and contains data used by inference algorithms. \texttt{\$state}
-is a Clojure hash map:
-\begin{lstlisting}[style=default]
-(def initial-state
-  "initial program state"
-  {:log-weight 0.0,
-   :result nil,
-   ::mem {},
-   ::store nil,
-   ... })
-\end{lstlisting}
-which records inference-relevant information under various keys
-such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
-The full list of map entries depends on the inference
-algorithm. Except for transformation of
-the \texttt{mem} form (which converts a function to one with memoization), 
-CPS transformation routines
-are not aware of contents of \texttt{\$state}, do not access or
-modify it directly, but rather just thread the state unmodified
-through the computation. Algorithm-specific handlers of
-checkpoints corresponding to the probabilistic forms
-(\texttt{sample}, \texttt{observe}) modify the
-state and reinject a new state into the computation.
-
 \subsection{Expression Kinds}
@@
 \texttt{cps-of-do}, \texttt{cps-of-store}.
-
-\subsection{Implementation Highlights}
-\label{seq:highlights}
-
-So far we introduced the basics of Anglican compilation to
-Clojure. The described approaches and techniques are important
-for grasping the language implementation but relatively well-known.
-In the rest of the section we focus on challenges we met and
-resolved while implementing Anglican, as well as on
-implementation of unique features of Anglican as a probabilistic
-programming language.
-
-\subsubsection{Continuations}
-\label{seq:continuations}
-
-Continuations are functions that are called in tail positions
-with the computed value and state as their arguments --- in CPS
-there is always a function call in every tail position and never
-a value. Continuations are passed to CPS transformers, and when
-transformers are called recursively, the continuations are
-generated on the fly.
-
-There are two critical issues related to generation of
-continuations:
-\begin{itemize}
-    \item unbounded \textit{stack growth} in recursive code;
-    \item code size \textit{explosion} when a non-atomic
-        continuation is symbolically substituted in multiple
-        locations.
-\end{itemize}
-
-\paragraph{Managing stack size}
-
-In implementations of functional programming languages stack
-growth is avoided through \textit{tail call optimization} (TCO).
-However, Clojure does not support a general form of TCO, and
-CPS-transformed code that creates deeply nested calls will
-easily exhaust the stack. Anglican employs a workaround called
-\textit{trampolining} --- instead of inserting a continuation
-call directly, the transformer always wraps the call into a
-\textit{thunk}, or parameterless function. The thunk is returned
-and called by the trampoline (Clojure provides function
-\texttt{trampoline} for this purpose) --- this way the
-computation continues, but the stack is collapsed on every
-continuation call. Function \texttt{continue} implements the
-wrapping and is invoked on every continuation call:
-\begin{lstlisting}[style=default]
-=> (continue 'cont 'value 'state)
-(fn [] (cont value state))
-\end{lstlisting}
-Correspondingly, the full, wrapped CPS form of
-\begin{lstlisting}[style=default]
-(fn [x y] (+ x y))
-\end{lstlisting}
-becomes
-\begin{lstlisting}[style=default]
-(fn [cont $state x y]
-  (fn []
-    (cont (+ x y) $state)))
-\end{lstlisting}
-When the CPS-transformed function is called, it returns a
-\textit{thunk} (a parameterless function) which is then
-re-invoked through the trampoline, with the stack collapsed.
-
-\paragraph{Avoiding exponential code growth}
-
-To realize potential danger of code size explosion, consider
-CPS transformation of code
-\begin{lstlisting}[style=default]
-(if (adult? person)
-  (if (male? person)
-     (choose-beer)
-     (choose-wine))
-  (choose-juice))
-\end{lstlisting}
-with continuation
-\begin{lstlisting}[style=default]
-(fn [choice _]
-  (case (kind choice)
-    :beer (beer-jar choice)
-    :wine (wine-glass choice)
-    :juice (juice-bottle choice)))
-\end{lstlisting}
-During CPS transformation, if we substitute the code of this continuation 
-for all of its calls, the code will be repeated three times in 
-the CPS-transformed expression. In general, CPS code can
-grow extremely large if the code of continuations is substituted
-repeatedly.
-
-To circumvent this inefficiency, CPS transformers for
-expressions with multiple continuation points (\texttt{if} and
-derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
-the continuation to a fresh symbol if it is not yet a symbol.
-Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
-establishes the binding automatically:
-\begin{lstlisting}[style=default]
-=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
-(let [cont (fn [x] (* 2 x))]
-  (if c
-      (fn [] (cont t $state))
-      (fn [] (cont f $state))))
-\end{lstlisting}
-
-\subsubsection{Primitive Procedures}
-\label{seq:primitive}
-
-When an Anglican function is transformed into a Clojure function
-by \texttt{fn-cps}, two auxiliary parameters are added to the
-beginning of the parameter list --- continuation and state.
-Correspondingly, when a function \textit{call} is transformed
-(by \texttt{cps-of-application} or \texttt{cps-of-apply}), the
-current continuation and the state are passed to the called
-function. Anglican can also call Clojure functions; however,
-Clojure functions do not expect these auxiliary parameters.
-To allow the mixing of Anglican (CPS-transformed) and Clojure
-function calls in Anglican code, the Anglican compiler must be
-able to recognize `primitive' (that is, implemented in Clojure
-rather than in Anglican) functions.
-
-Providing an explicit syntax for differentiating between
-Anglican and Clojure function calls would be cumbersome. Another
-option would be to use meta-data to identify Anglican function
-calls at runtime. However, this would impact performance, and
-good runtime performance is critical for probabilistic
-programs. The approach taken by Anglican is to maintain a list
-of unqualified names of primitive functions, as well of
-namespaces in which all functions are primitive, and recognize
-primitive functions by name --- if a function name is not in the
-list, the function is an Anglican function. Global dynamically-bound 
-variables \texttt{*primitive-procedures*} and
-\texttt{*primitive-namespaces*} contain the initial lists of
-names and namespaces, correspondingly. Of course, local bindings
-can shade global primitive function names. For example,
-\texttt{first} is a Clojure primitive function that takes the first
-element from an ordered collection such as list and vector,
-but inside the let block in the
-following example, \texttt{first} is an Anglican function:
-\begin{lstlisting}[style=default]
-(let [first (fn [[x & y]] x)]
-  (first '[1 2 3]))
-\end{lstlisting}
-The Anglican compiler takes care of the shading by rebinding
-\texttt{*primitive-procedures*} in every lexical scope
-(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
-\texttt{shading-primitive-procedures} automates the shading.
-
-\subsubsection{Probabilistic Forms}
-\label{seq:forms}
-
-There are two proper probabilistic forms turning Anglican into a
-probabilistic programming language --- \texttt{sample} and
-\texttt{observe}. Their purpose is to interrupt deterministic
-computation and transfer control to the inference algorithm.
-Practically, this is achieved through returning
-\textbf{checkpoints} --- Clojure values of the corresponding
-types (\texttt{anglican.trap.sample} or
-\texttt{anglican.trap.observe}). The types contain fields
-specific to each form, as well as the continuation; calling the
-continuation resumes the computation. Checkpoints expose the
-program state to the inference algorithm, and the updated state
-is re-injected into the computation when the continuation is
-called:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression '(sample dist) 'cont)
-(->sample dist cont $state)
-=> (cps-of-expression '(observe dist v) 'cont)
-(->observe dist v cont $state)
-\end{lstlisting}
-Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
-expressions constructors for Clojure records, and they take 
-values of their fields as arguments.
-
-In addition to checkpoints, there are a few other special forms
----  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
-modify program state. These forms are translated into
-expressions involving calls of functions from the
-\texttt{anglican.state} namespace. The \texttt{mem} form, which
-implements memoization, deserves a more detailed explanation.
-
-\paragraph{Memoization}
-The author of a probabilistic model might want to 
-randomly draw a feature value for each entity in a collection,
-but to retain the same drawn
-value for the same entity during a single run of the probabilistic
-program. For example, a person may have brown or green eyes
-with some probability, but the \textit{same} person will
-always have the same eye color. This can be achieved
-through the use of memoized functions. In Anglican, one
-might write:
-\begin{lstlisting}[style=default]
-(let [eye-color (mem (fn [person]
-                       (sample
-                         (categorical
-                           ['brown 0.5]
-                           ['green 0.5]))))]
-  (if (not= (eye-color 'bill) (eye-color 'john))
-    (eye-color 'bill)
-    (eye-color 'john)))
-\end{lstlisting}
-The \texttt{mem} form converts a function to a memoized variant,
-which remembers past inputs and the corresponding outputs, and 
-returns the remembered output when given such a past input, 
-instead of calling the original function with it. As a result,
-for every input, random draws will be made only for the first time 
-that the memoized function is called with the input; all subsequent
-calls with the input will just reuse these draws return the same output.
-
-Memoization is often implemented on top of a mutable
-dictionary, where the key is the argument list and the value is
-the returned value. However, there are no mutable data
-structures in a probabilistic program. Hence, \texttt{mem}'s
-memory is stored as a nested dictionary in the program state
-introduced during CPS transformation (function \texttt{mem-cps}).  
-Every memoized function gets a
-unique automatically generated identifier. Each time a memoized
-function is called, one of two continuations is chosen,
-depending on whether the same function (a function with the same
-identifier) was previously called in the same run of the
-probabilistic program with the same arguments. If the memoized
-result is available, the continuation of the memoized function
-call is immediately called with the stored result. Otherwise,
-the argument of \texttt{mem} is called with a continuation which
-first creates an updated state with the memoized result, and
-then calls the `outer' continuation with the result and the
-updated state:
-\begin{lstlisting}[style=default]
-=> (mem-cps '(foo))
-(let [M (gensym "M")]
-  (fn mem23623 [C $state & P]
-    (if (in-mem? $state M P)
-      ;; previously memoized result
-      (fn []
-        (C (get-mem $state M P) $state))
-      ;; new computation
-      (clojure.core/apply
-        foo
-        ;; memoize result in state
-        (fn [V $state]
-          (fn [] (C V (set-mem $state M P V))))
-        $state
-        P)))))
-\end{lstlisting}
-
-Memoized results are not shared among multiple runs of a
-probabilistic program, which is intended. Otherwise, it would be
-impossible to memoize functions with random results.
-
-\section{Inference Algorithms}
-\label{sec:inference}
-
-A probabilistic program in Anglican may look almost like a
-Clojure program. However, the purpose of executing a
-probabilistic program is different from that of a `regular'
-program: instead of producing the result of a single execution,
-a probabilistic program computes, exactly or approximately, the
-distribution from which execution results are drawn.  Computing
-the distribution is facilitated by an inference algorithm.
-
-Probabilistic programming system Anglican provides a variety of
-approximate inference algorithms. Ideally, Anglican should
-automatically choose the most appropriate algorithm for each
-probabilistic program. In practice, selecting an inference
-algorithm, or a combination thereof, is still a challenging
-task, and program structure, intended use of the computation
-results, performance, approximation accuracy, and other factors
-must be taken into consideration. New algorithms are being
-developed and added to Anglican~\cite{TMP+15,MPT+16,RNL+2016},
-as a part of ongoing research.
-
-In the implementation of Anglican, inference algorithms are
-instantiations of the (ad-hoc) polymorphic function
-\texttt{infer} declared as Clojure's multimethod in the
-\texttt{anglican.{\linebreak[0]}inference} namespace. The function accepts an
-algorithm identifier, a query --- the probabilistic program in
-which to perform the inference, an initial value for the query,
-and optional algorithm parameters.
-
-\subsection{The \texttt{infer} Function}
-\label{sec:infer}
-
-The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
-an appropriate overloading or implementation of the function --- conventionally,
-the identifier is a Clojure keyword (a symbolic constant starting with colon)
-related to the algorithm name, such as \texttt{:lmh} for Lightweight
-Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
-The second parameter is a query as defined by
-the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
-the following Clojure code invokes \texttt{infer} on an Anglican query
-defined anonymously via the \texttt{query} form:
-%If needed, the query
-%can be defined anonymously, right in the argument list of a call
-%to \texttt{infer}:
-\begin{lstlisting}[style=default]
-(let [x 1]
-  (infer :pgibbs (query x) nil))
-\end{lstlisting}
-A query is executed by calling the initial continuation of the
-query, which accepts a value and a state. The state is supplied
-by the inference algorithm, while the value is provided as a
-parameter of \texttt{infer}. A query does not have to have any
-parameters, in which case the value can be simply \texttt{nil}.
-When a query is defined with a binding for the initial value,
-the value becomes available inside the query. A query may
-accept multiple parameters using Clojure's structured binding.
-For instance, it may take multiple parameters as components
-of an input vector. In this case, the initial value is given
-as a structured value, such as a vector, and the components
-of this value are matched to the corresponding
-parameters of the query via the destructuring mechanism of Clojure.
-%The value is destructured via pattern match in case of a structured binding.
-For example,
-\begin{lstlisting}[style=default]
-(defquery my-query [mean sd]
-  (sample (normal mean sd)))
-
-(def samples (infer :lmh my-query [1.0 3.0]))
-\end{lstlisting}
-Finally, any number of auxiliary arguments can be
-passed to \texttt{infer}.
-%Finally, an indefinite number of auxiliary arguments can be
-%passed to \texttt{infer}.
-By convention, the arguments should be
-keyword arguments, and are interpreted in the algorithm-specific
-manner.
-
-\subsection{Internals of an Inference Algorithm}
-\label{seq:internals}
-
-Implementing an inference algorithm in Anglican amounts
-to defining an appropriate version of the \texttt{infer} function and 
-checkpoint handlers for \texttt{sample} and \texttt{observe}.
-The definitions may just reuse default implementations or override them
-with new algorithm-specific treatment of \texttt{sample} and \texttt{observe} forms
-and inference state.
-
-Let us illustrate this implementation step with
-\textit{importance sampling},
-the simplest inference algorithm.
-Here is an implementation of the \texttt{infer} function for the algorithm:
-\begin{lstlisting}[style=default]
-(derive ::algorithm
-        :anglican.inference/algorithm)
-;; invoked when algo parameter is :importance
-(defmethod infer :importance 
-  [algo prog value & {}]
-  (letfn
-    ;; recursive function without any parameter
-    [sample-seq 
-      []
-      ;; lazy infinite sequence 
-      (lazy-seq
-        (cons
-          (:state (exec ::algorithm prog value
-                        initial-state))
-          (sample-seq))))]
-    (sample-seq)))
-\end{lstlisting}
-This implementation is dispatched 
-when \texttt{infer} receives \texttt{:importance} as its \texttt{algo} 
-parameter. (Anglican includes multiple implementations of \texttt{infer}
-for different inference algorithms.) Once dispatched, it lazily 
-constructs an infinite sequence of inference states by repeatedly 
-executing the program \texttt{prog} using \texttt{exec}
-and retrieving the final inference state of the execution 
-using \texttt{:state}. For checkpoint handlers, importance sampling simply
-relies on their default implementations.
-%\texttt{anglican.importance/{\linebreak[0]}infer} just calls
-%\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
-%implementations of checkpoint handlers.
-
-Typically, an inference
-algorithm has its own implementations of
-\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
-both, as well as invokes \texttt{exec} from an elaborated
-conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
-(\texttt{anglican.smc}) are examples of inference algorithms
-where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
-handler is overridden. In addition, SMC runs multiple particles
-(corresponding to user-level threads) simultaneously, while LMH re-runs
-programs from an intermediate continuation rather than from the
-beginning.
-
-\subsection{Addressing of Checkpoints}
-\label{sec:addressing}
-
-Many inference algorithms memoize and reuse earlier computations
-at checkpoints. Variants of Metropolis-Hastings reuse previously
-drawn values~\cite{WSG11}, as well as additional information
-used for adaptive sampling~\cite{TMP+15} at \texttt{sample}
-checkpoints. Asynchronous SMC (Particle Cascade) computes
-average particle weights at \texttt{observe}
-checkpoints~\cite{PWD+14}. Implementations of black-box
-variational inference~\cite{WW13,MPT+16} associate with random
-variables the learned parameters of variational posterior
-distribution.
-
-Checkpoints can be uniquely named at compilation time;
-however, at runtime a checkpoint corresponding to a single code
-location may occur multiple time due to recurrent invocation of
-the function containing the checkpoint. Every unique occurrence
-of a checkpoint must receive a different address. An addressing
-scheme based on computing of stack addresses of checkpoints was
-described in the context of Lightweight
-Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
-over the naive scheme where dynamic occurrences of checkpoints
-are numbered sequentially. However, it impacts the
-performance of probabilistic programs because of the computation
-cost of computing the stack addresses:
-\begin{enumerate}
-    \item On every function call, a component is added to the
-        address. Hence, the address size is linear in stack depth.
-    \item Every function call must be augmented by symbolic
-        information required to compute stack addresses.
-\end{enumerate}
-In addition, the above scheme can still lead to inferior
-correspondence between checkpoints in different traces: in
-Anglican and other probabilistic languages where distributions
-are first-class objects checkpoints with incompatible arguments
-can correspond to the same stack address. Consider the following
-program fragment:
-\begin{lstlisting}[style=default]
-(let [dist (if use-gamma
-               (gamma 2. 2.)
-               (normal 0. 1.))]
-  (sample dist))
-\end{lstlisting}
-The \texttt{sample} checkpoint has the same stack address in
-different traces. However, the random values should not be
-reused between different distributions. Further on, in some
-algorithms, such as black-box variational inference, the role of
-checkpoint addresses is semantic rather than heuristic ---
-appropriate correspondence must be established
-between checkpoints in different traces for the algorithm to
-work.
-
-To overcome the above problems, Anglican uses a different scheme
-which is almost as efficient as the scheme based on stack addresses for
-reuse of previously drawn values, while producing addresses of constant size,
-and allows manual computation of checkpoint addresses at runtime
-when the default automatic scheme is insufficient. According to
-the scheme:
-\begin{itemize}
-    \item A checkpoint may accept an auxiliary argument --- the
-        checkpoint identifier. If specified, the identifier is the first
-        argument of a checkpoint. For example \texttt{(sample
-        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
-        with identifier \texttt{x1}.
-    \item If the identifier is omitted, a unique identifier is
-        generated automatically as a fresh symbol.
-    \item At runtime, the address of a checkpoint invocation has
-        the form
-        \texttt{[}\textit{checkpoint-identifier
-        number-of-previous-occurrences}\texttt{]}, where the
-        occurrences are of a checkpoint with the same checkpoint
-        identifier.
-    \item If a sequence of invocations of the same checkpoint is
-        interrupted by a different checkpoint, the number of
-        previous occurrences is rounded up to a multiple of
-        an integer. For efficiency, a small power of 2 is used,
-        such as $2^4 = 16$.
-\end{itemize}
-
-Consider the following simplified Anglican query:
-
-\begin{lstlisting}[style=default]
-(defm foo []
-  (if (sample 'C1 (flip 0.5))
-    (foo)
-    (bar)))
-
-(defm bar []
-  (case (sample 'C2 (uniform-discrete 0 3))
-    0 (bar)
-    1 (foo)
-    2 (sample 'C3 (normal 0. 1.))))
-
-(defquery baz
-  (foo))
-\end{lstlisting}
-
-An execution of the query may result in the following
-sequence of checkpoint invocations:
-\begin{lstlisting}[style=default]
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C2 ...)
-    (sample 'C1 ...)
-    (sample 'C1 ...)
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C3 ...)
-\end{lstlisting}
-According to the addressing scheme, the addresses generated
-for this invocations are
-\begin{lstlisting}[style=default]
-    [C1 0]
-    [C2 0]
-    [C2 1]
-    [C1 16]
-    [C1 17]
-    [C1 18]
-    [C2 16]
-    [C3 0]
-\end{lstlisting}
-If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
-\begin{lstlisting}[style=default]
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C1 ...)
-    (sample 'C1 ...)
-    (sample 'C2 ...)
-    (sample 'C2 ...)
-    (sample 'C3 ...)
-\end{lstlisting}
-The addresses for the new sequence are
-\begin{lstlisting}[style=default]
-    [C1 0]
-    [C2 0]
-    [C1 16]
-    [C1 17]
-    [C2 16]
-    [C2 17]
-    [C3 0]
-\end{lstlisting}
-Despite the change, the correspondence between checkpoints of
-similar types occurring in similar positions (relative positions
-in contiguous subsequences of a certain type) is preserved, and
-drawn values can be reused efficiently:
-
-\vspace{\baselineskip}
-{
-\begin{tabular}{l l}
-    \textbf{old} & \textbf{new} \\ \hline
-    {}\texttt{[C1 0]} & \texttt{[C1 0]} \\
-    {}\texttt{[C2 0]} & \texttt{[C2 0]} \\
-    {}\texttt{[C2 1]} & \textit{unused} \\
-    {}\texttt{[C1 16]} & \texttt{[C1 16]} \\
-    {}\texttt{[C1 17]} & \texttt{[C1 17]} \\
-    {}\texttt{[C1 18]} & \textit{unused} \\
-    {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
-    \textit{missing} & \texttt{[C2 17]} \textit{drawn}\\
-    {}\texttt{[C3 0]} & \texttt{[C3 0]}
-\end{tabular}
-\vspace{\baselineskip}}
-
-Note that correspondence between checkpoints in different
-traces plays the role of an heuristic in Metropolis-Hastings
-family of algorithms. Any correspondence (or no correspondence,
-meaning all values must be re-drawn from their proposal
-distributions) is valid, and reused values from the previous
-invocation which are not in support or have zero probability in
-the new invocation are simply ignored and new values are
-re-drawn.
-
-This way, each occurrence of a checkpoint has unique address,
-but small disturbances --- removal or addition of a single or just
-a few checkpoints --- are unlikely to derail the entire sequence.
-The probability of derailment depends on the padding. The
-padding can be safely, and without any impact on performance,
-set to rather large numbers. However, rounding up to a multiple
-of 16 proved to be appropriate for all practical purposes.
-
-Function \texttt{checkpoint-id} in the
-\texttt{anglican.inference} namespace automates generation of
-checkpoint addresses and can be used from any inference
-algorithm.
-
-\section{Definitions and Runtime Library}
-\label{seq:runtime}
-
-A Clojure namespace that includes a definition of an Anglican
-program imports (`requires') two essential namespaces:
-\texttt{anglican.emit} and \texttt{anglican.runtime}. The former
-provides macros for defining Anglican programs
-(\texttt{defquery}, \texttt{query}) and functions
-(\texttt{defm}, \texttt{fm}, \texttt{mem}), as well as Anglican
-bootstrap definitions that must be included with every program
---- first of all, CPS implementations of higher-order functions.
-\texttt{anglican.emit} can be viewed as the Anglican
-\textit{compiler tool}, which helps transform Anglican code into
-Clojure before any inference is performed.
-
-\texttt{anglican.runtime} is the Anglican runtime library. For
-convenience, it exposes common mathematical functions
-(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
-\texttt{exp}, etc.), but most importantly, it provides
-definitions of common distributions. Each distribution object
-implements a distribution interface with the \texttt{sample*} and \texttt{observe*}
-methods; this interface is defined using Clojure's protocol mechanism
-(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
-The \texttt{sample*} \textit{method} returns a random sample and
-roughly corresponds to the \texttt{sample} checkpoint,
-the \texttt{observe*} \textit{method} returns the log probability
-of the value and roughly corresponds to the \texttt{observe}
-checkpoint. The methods can be, and sometimes are called from
-handlers of the corresponding checkpoints, but do not have to
-be. For example, in LMH either the \texttt{sample*} or
-the \texttt{observe*} \textit{method} is called for a
-\texttt{sample} checkpoint, depending on whether the value is
-drawn or reused.
-
-Macro \texttt{defdist} should be used to define distributions.
-\texttt{defdist} takes care of defining a separate type
-for every distribution so that Clojure multimethods (or overloaded
-methods) can be dispatched on
-distribution types when needed, e.g. for custom proposal distributions 
-used in an inference algorithm. The
-Bernoulli distribution could be defined as follows:
-\begin{lstlisting}[style=default]
-(defdist bernoulli
-  "Bernoulli distribution"
-  [p]
-  (sample* [this] (if (< (rand) p) 1 0))
-  (observe* [this value]
-    (Math/log (case value
-                1 p
-                0 (- 1. p)
-                0.))))
-\end{lstlisting}
-
-In addition to distributions, Anglican provides \textit{random processes},
-which define sequences of random variables that are not independent and
-identically distributed. Random processes are similar to the so called
-`exchangeable random procedures' in Church \cite{GMR+08} and
-Venture \cite{MSP14}. However, random sequences generated by Anglican random 
-processes are not required to be exchangeable. Random processes are
-defined using the \texttt{defproc} macro and implement the
-\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
-protocol has two methods
-\begin{itemize}
-\item \texttt{produce}, which returns the distribution on the next
-random variable in the sequence, and
-\item \texttt{absorb}, which incorporates the value for the next random
-variable and returns an updated random process.
-\end{itemize}
- As an example, here is a definition of a beta-Bernoulli process, in which
-each random variable is distributed according to a Bernoulli distribution with
-an unknown parameter that is drawn from a beta distribution:
-\begin{lstlisting}[style=default]
-(defproc beta-bernoulli
-  "beta-Bernoulli process"
-  [a b]
-  (produce [this] (bernoulli (/ a (+ a b))))
-  (absorb [this x]
-    (case x
-      0 (beta-bernoulli a (inc b))
-      1 (beta-bernoulli (inc a) b))))
-\end{lstlisting}
-Unlike typical implementations of exchangeable random processes, Anglican's random processes do not have mutable state.
-The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican.
-A sequence of random values can be generated using a recursive loop in which \texttt{absorb} returns the updated process for the next iteration.
-For example:
-\begin{lstlisting}[style=default]
-(defm sample-beta-binomial [n a b]
-  (loop [process (beta-bernoulli a b)
-         values []]
-    (if (= (count values) n)
-      values
-      (let [dist (produce process)
-            value (sample dist)]
-      (recur (absorb process value)
-             (conj values value))))))
-\end{lstlisting}
-Similarly, random processes can also be used to recursively observe a sequence of values:
-\begin{lstlisting}[style=default]
-(loop [process (beta-bernoulli 1.0 1.0)
-       [value0 & values0] values] 
-  (let [dist (produce process)]
-    (observe dist value0)
-    (recur (absorb process value0) values0)))
-\end{lstlisting}
-
-% \begin{itemize}
-%     \item Values produced by the process are sampled from the distribution
-%         returned by \texttt{produce}.
-%     \item When a value is observed from the distribution, a new
-%         process, with the observed value absorbed, is returned
-%         by \texttt{absorb}.
-% \end{itemize}
-
-% A common pattern when programming with random processes in 
-% Anglican is to generate a sequences of sampled values and 
-% constructing an updated process at each step
-
-\section{Case Study}
-\label{seq:study}
-
-The description of Anglican language and environment would be
-incomplete without a case study showing the process of building
-and executing a solution of an inference problem. This case
-study takes a problem for which a solution is not immediately
-obvious, \textit{the Deli dilemma}, and guides through
-writing an Anglican program for the problem, executing the
-program, and post-processing results.
-
-The program presented in this section is intentionally short and
-simple. Anglican is capable of compiling and running elaborated
-programs handling large amounts of data. Advanced examples of
-Anglican programs and inference can be found in literature on
-applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.
-
-\subsection{The Deli Dilemma}
-\label{seq:deli}
-
-Imagine that we are facing the following dilemma:
-\begin{quote}
-A customer wearing round sunglasses came to a deli at 1:13pm,
-and grabbed a sandwich and a coffee. Later on the same day, a
-customer wearing round sunglasses came at 6:09pm and ordered a
-dinner.  Was it the same customer?
-\end{quote}
-In addition to the story above, we know that:
-\begin{itemize}
-    \item There is an adjacent office quarter, and it takes
-        between 5 and 15 minutes on average to walk from an
-        office to the deli, where different average times are  for
-        different buildings in the quarter.
-    \item Depending on traffic lights, the walking time varies by about 2
-        minutes.
-    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
-    \item The waiter's odds that this is the same customer are 2 to 1.
-\end{itemize}
-
-\subsection{Anglican Query}
-\label{seq:deli-query}
-
-We want to formalize the dilemma in an Anglican query, based on
-the knowledge we have. Let us formalize the knowledge in
-Clojure (the times are in minutes). First, we encode our prior
-information which holds true independently of the customer's
-visit:
-
-\begin{lstlisting}[style=default]
-(def mean-time-to-arrive
-     "average time to arrive"
-     10.)
-(def sd-time-to-arrive
-     "standard deviation of arrival time"
-     3.)
-(def time-sd
-     "walking time deviation"
-     1.)
-\end{lstlisting}
-
-Then, we record our observations, based on which we want to
-solve the dilemma:
-
-\begin{lstlisting}[style=default]
-(def lunch-delay
-     "time between lunch break and lunch order"
-     13.)
-(def dinner-delay
-     "time between end of day and dinner order"
-     9.)
-(def p-same
-     "prior probability of the same customer"
-     (/ 2. 3.))
-\end{lstlisting}
-
-For inference, one often chooses a known distribution to
-represent uncertainty. We choose the normal
-distribution for representing uncertainty about average arrival
-time.
-
-\begin{lstlisting}[style=default]
-(def time-to-arrive-prior 
-     "prior distribution of average arrival time"
-     (normal mean-time-to-arrive
-             sd-time-to-arrive))
-\end{lstlisting}
-
-There are two possibilities: either the same customer visited
-the deli twice, or two different customers came to the deli,
-one for lunch and the other for dinner. We define an \textbf{Anglican} function
-for each case. Note that this is the first time we switch from
-Clojure to Anglican. The functions must be written in Anglican
-(and hence defined using \texttt{defm} instead of \texttt{defn})
-because they contain probabilistic forms \texttt{sample} and
-\texttt{observe}.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-  "observe the same customer twice"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive time-sd)
-             dinner-delay)
-    [time-to-arrive]))
-
-(defm different-customers
-  "observe different customers"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive-1
-          (sample time-to-arrive-prior)
-        time-to-arrive-2
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive-1 time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive-2 time-sd)
-             dinner-delay)
-    [time-to-arrive-1 time-to-arrive-2]))
-\end{lstlisting}
-
-Both functions have the same structure: we first `guess' the
-average arrival time and then observe the actual time from a
-distribution parameterized by the guessed time. However, in
-\texttt{same-customer} the average arrival time is the same for
-both the lunch and the dinner, while in
-\texttt{different-customers} two average arrival times are
-guessed independently.
-
-We are finally ready to define the query in \textbf{Anglican}.
-
-\begin{lstlisting}[style=default]
-(defquery deli [time-to-arrive-prior
-                lunch-delay
-                dinner-delay]
-  (let [is-same-customer (sample (flip p-same))
-        observe-customer (if is-same-customer
-                            same-customer
-                            different-customers)]
-    {:same-customer is-same-customer,
-     :times-to-arrive (observe-customer 
-                        time-to-arrive-prior
-                        lunch-delay
-                        dinner-delay)}))
-\end{lstlisting}
-
-Performing inference on query \texttt{deli} computes the
-probability that the same customer visited the deli twice, as
-well as probability distributions of average arrival times for
-both cases.
-
-\subsection{Inference}
-\label{seq:deli-inference}
-
-Having defined the query, we are now ready to run the query
-using an inference algorithm. Function \texttt{doquery} provided
-by the \texttt{anglican.core} namespace accepts the inference
-algorithm, the query, and optional parameters, and returns a
-lazy sequence of samples. We use here the inference algorithm
-called Lightweight Metropolis-Hastings (LMH). The algorithm is
-somewhat slow to converge but can be used with any Anglican
-query, and should be robust enough for our simple problem.
-
-We bind the results of \texttt{doquery} to variable
-\texttt{samples}, to analyse the results later. However, since
-the sequence is lazy, no inference is performed and no samples
-are generated until they are retrieved and processed.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :lmh deli nil))
-\end{lstlisting}
-
-To approximate the inferred distribution, we extract a finite
-subsequence of samples. Many algorithms use an initial
-subsequence of samples to converge to the target distribution.
-Hence, we drop initial $N$ samples ($N$ is 5000 in the code
-snippet below), and collect the next $N$ samples.
-
-\begin{lstlisting}[style=default]
-(def N 5000)
-(def results (map get-result
-                   (take N (drop N samples))))
-\end{lstlisting}
-
-Based on the collected samples we compute an approximation of
-the posterior probability \texttt{p-same+} that the same
-customer visited the deli twice.
-
-\begin{lstlisting}[style=default]
-(def p-same+ 
-     (/ (count (filter :same-customer results))
-        (double N)))
-\end{lstlisting}
-The \texttt{:same-customer} here represents a function
-that looks up the entry with the same name in a given map.
-The map stores the result of a single sampled execution,
-and the \texttt{:same-customer} entry in the map
-records whether the same customer visits the deli in the execution.
-The \texttt{filter} function in \texttt{p-same+} 
-goes through all the maps in \texttt{results}
-and picks only the ones whose executions involve just one customer.
-
-With the specified observations, \texttt{p-same+} is $\approx
-0.12$. The probability is much lower than the
-waiter's guess \texttt{p-same} ($\frac 2 3$).
-Of course, the results may vary from run to run, and, for a given
-algorithm, the accuracy depends on the number of samples we
-decide to retrieve.
-
-In addition to computing the posterior probability that the same
-customer visited the deli twice, we may want to know the average
-time (or times) to arrive. In Bayesian inference, it is common
-to report distributions instead of `most likely' values. We use
-query results in retrieved samples to approximate the
-distributions, and plot distribution histograms for the same
-customer visiting twice (Figure~\ref{fig:dist-same}) and
-two different customers (Figure~\ref{fig:dist-diff})
-
-\begin{lstlisting}[style=default]
-;; single customer                      
-(def time-to-arrive+ 
-    (map (fn [x] (first (:times-to-arrive x)))
-         (filter :same-customer results)))
-(def mean-to-arrive+ (mean time-to-arrive+))
-(def sd-to-arrive+ (std time-to-arrive+))
-
-;; two customers
-(def times-to-arrive+ 
-     (map :times-to-arrive 
-          (filter  
-            (fn [x] (not (:same-customer x)))
-            results)))
-(def mean-1-to-arrive+
-     (mean (map first times-to-arrive+)))
-(def sd-1-to-arrive+
-     (std (map first times-to-arrive+)))
-(def mean-2-to-arrive+
-     (mean (map second times-to-arrive+)))
-(def sd-2-to-arrive+
-     (std (map second times-to-arrive+)))
-\end{lstlisting}
-
-In addition to plotting the distribution histograms, we use
-functions \texttt{mean}, \texttt{std} provided
-in the \texttt{anglican.{\linebreak[0]}stat} namespace
-along with other
-useful statistical functions.
-This is another illustration of advantage of tight
-integration between Clojure and Anglican --- probabilistic
-models are expressed in Anglican. However, processing of data and
-results can rely on the full power of Clojure.
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
-    \caption{Arrival time distribution for a single customer.}
-    \label{fig:dist-same}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
-    \caption{Arrival time distributions for two customers.}
-    \label{fig:dist-diff}
-\end{figure}
-
-For inference, we chose to use  Lightweight
-Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
-probabilistic programming is that models are separated from
-inference. To switch to a different inference algorithm we just
-need to pass a different value to \texttt{doquery}. For example,
-we may decide to use Black-Box Variational Bayes (BBVB), which
-may not work equally well for all probabilistic programs, but is
-much faster to converge.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :bbvb deli nil))
-\end{lstlisting}
-
-We can still use samples and summary statistics with BBVB to
-approximate the posterior distribution. However, variational
-inference approximates the posterior by known distributions,
-and we can directly retrieve the distribution parameters.
-
-\begin{lstlisting}[style=default]
-(clojure.pprint/pprint
-  (anglican.bbvb/get-variational 
-    (nth samples N)))
-\end{lstlisting}
-
-\begin{lstlisting}[style=default]
-{S28209
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- S28217
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- S28215
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- S28219
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
-\end{lstlisting}
-
-We can guess that the variational distributions correspond to the
-prior distributions in the \texttt{sample} forms, in the order
-of appearance. However, it would help if we could use more
-informative labels instead of automatically generated symbols
-\texttt{S28209}, \texttt{S28217}, etc. Here the option to
-specify identifiers for probabilistic forms explicitly comes
-handy. If we modify the \texttt{sample} forms to use explicit identifiers
-(only the forms are shown for brevity), the output becomes much
-easier to analyse.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-      ...
-      (sample :arrival-time-same
-              time-to-arrive-prior)]
-      ...)
-
-(defm different-customers
-      ... 
-      (sample :arrival-time-first
-              time-to-arrive-prior)
-      ...
-      (sample :arrival-time-second
-              time-to-arrive-prior)]
-      ...)
-
-(defquery deli ...
-      ...
-      (sample :same-or-different
-              (flip p-same))
-      ...)
-\end{lstlisting}
-
-The output becomes much easier to interpret and analyse
-programmatically.
-
-\begin{lstlisting}[style=default]
-{:arrivate-time-same
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- :arrival-time-first
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- :arrival-time-second
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- :same-or-different
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
-\end{lstlisting}
-
-This completes the case study, where we showed a probabilistic
-programming solution of a problem, implemented and analysed in
-Anglican and Clojure, using two different inference algorithms.
-
-\section{Conclusion}
-\label{seq:summary}
-
-In this paper, we presented design and implementation internals
-of the \emph{probabilistic programming system} Anglican. Implementing
-a language is an interesting endeavour, in particular when the
-language implements a new paradigm, in this case probabilistic
-programming. Functional programming is a natural complement of
-probabilistic programming --- the latter allows both concise and
-expressive specification of probabilistic generative models and
-efficient implementation of inference algorithms.
-
-Implementing a probabilistic language on top of and in tight
-integration with a functional language, Clojure, both helped
-us to accomplish an ambitious goal in a short time span, and
-provided important insights on structure and semantics of
-probabilistic concepts incorporated in Anglican. Computational
-efficiency and expressive power of Anglican owe to adherence to
-the functional approach as much as to rich inference
-opportunities of the Anglican environment.
-
-\acks
-
-This work was partially supported under DARPA PPAML through the
-U.S. AFRL under Cooperative Agreement number FA8750- 14-2-0006,
-Sub Award number 61160290-111668.
-
-\bibliographystyle{abbrvnat}
-\bibliography{refs}
-
-\end{document}

### 2017-02-04 paper/paper.tex
many changes in progress

@@
 \newtoggle{full}
-\togglefalse{full}
+\input{toggle}
 
@@
 language is implemented `inside' another language of the same
-level of abstraction. 
-}{%
-One way to implement a programming language is to embed it within another language.
+level of abstraction. Different languages provide different
+means for this; }{%
+There are several ways to build a new language on top of an
+existing one;
 }%
-Different languages provide different
-means for this; Lisp is famous for the macro facility
+Lisp is famous for the macro facility
 that allows to extend the language almost without
@@
 
+\subsection*{Contributions}
+
 \section{Design Outline}
@@
 \texttt{anglican.emit} namespace), which call the CPS
-transformations and define Clojure values
-%Clojure objects
-suitable for passing
+transformations and define Clojure values suitable for passing
 as arguments to inference algorithms (\texttt{defquery},
@@
 \label{sec:related}
+
+{\it TODO}
+
+\section{Probabilistic Program}
+\label{sec:pp-def}
 
@@
 \iftoggle{full}{%
-\subsection{Expression Kinds}
-\label{sec:expr}
-
-There are three different kinds of inputs to CPS transformation:
-\begin{itemize}
-    \item Literals, which are constant expressions. They
-       are passed as an argument to the
-        continuation unmodified.
-    \item Value expressions such as the \texttt{fn} form
-        (called \textit{opaque}
-        expressions in the code). They must be transformed to
-        CPS, but the transformed object is passed to the
-        continuation as a whole, opaquely.
-    \item General expressions (which we call
-        \textit{transparent} expressions).
-        The continuation is threaded through such an expression
-        in an expression-specific way, and can be
-        called in multiple locations of the CPS-transformed
-        code, such as in all branches of an \texttt{if}
-        statement.
-\end{itemize}
-
-\subsubsection{Literals}
-\label{sec:literals}
-
-Literals are the same in Anglican and Clojure. They are
-left unmodified; literals are a subset of opaque expressions.
-However, the Clojure syntax has a peculiarity
-of using the syntax of compound literals (vectors, hash maps,
-and sets) for data constructors. Hence, compound literals must
-be traversed recursively, and if there is a nested non-literal
-component, transformed into a call to the corresponding data
-constructor. Functions \texttt{cps-of-vector},
-\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
-\texttt{cps-of-expression}, transform Clojure constructor syntax
-(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
-corresponding calls:
-\begin{lstlisting}[style=default]
-=> (cps-of-vector [0 1 2] 'cont)
-(cont (vector 0 1 2) $state)
-=> (cps-of-hash-map {:a 1, :b 2} 'cont)
-(cont (hash-map :a 1, :b 2) $state)
-=> (cps-of-set #{0 1} 'cont)
-(cont (set (list 0 1)) $state)
-\end{lstlisting}
-
-\subsubsection{Opaque Expressions}
-\label{sec:opaque}
-
-Opaque, or value, expressions, have a different shape in the
-original and the CPS form. However, their CPS form follows the
-pattern \texttt{(continuation transformed-expression \$state)}, and thus
-the transformation does not depend on the continuation parameter, and
-can be accomplished without passing the parameter as a
-transformation argument. Primitive (non-CPS) procedures used in
-Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
-forms are opaque and transformed by
-\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps},
-\texttt{fn-cps}, and \texttt{mem-cps}, correspondingly: a
-slightly simplified CPS form of expression
-\begin{lstlisting}[style=default]
-(fn [x y]
-  (+ x y))
-\end{lstlisting}
-would be
-\begin{lstlisting}[style=default]
-(fn [cont $state x y]
-  (cont (+ x y) $state))
-\end{lstlisting}
-In the actual code an automatically generated fresh symbol is
-used instead of \texttt{cont}.
-
-\subsubsection{General Expressions}
-\label{seq:general}
-
-The most general form of CPS transformation receives an
-expression and a continuation as parameters, and returns the expression
-in CPS form with the continuation parameter potentially called in multiple
-tail positions. General expressions can be somewhat voluntarily
-divided into several groups:
-\begin{itemize}
-    \item binding forms --- \texttt{let} and
-        \texttt{loop/recur};
-    \item flow control --- \texttt{if}, \texttt{when},
-        \texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
-    \item function applications and \texttt{apply};
-    \item probabilistic forms --- \texttt{observe},
-        \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
-\end{itemize}
-Functions that transform general expressions accept the
-expression and the continuation as parameters, and are
-consistently named \texttt{cps-of-}\textit{form}, for example,
-\texttt{cps-of-do}, \texttt{cps-of-store}.
-
+\input{expression-kinds.tex}
 }{%
-}
+}%
 
@@
 easily exhaust the stack. Anglican employs a workaround called
-\textit{trampolining} --- instead of inserting a continuation
-call directly, the transformer always wraps the call into a
-\textit{thunk}, or parameterless function. The thunk is returned
-and called by the trampoline (Clojure provides function
-\texttt{trampoline} for this purpose) --- this way the
+\textit{trampolining}~\cite{B95} --- instead of inserting a
+continuation call directly, the transformer always wraps the
+call into a \textit{thunk}, or parameterless function. The thunk
+is returned and called by the trampoline (Clojure provides
+function \texttt{trampoline} for this purpose) --- this way the
 computation continues, but the stack is collapsed on every
@@
 \iftoggle{full}{%
-\paragraph{Avoiding exponential code growth}
-
-To realize potential danger of code size explosion, consider
-CPS transformation of code
-\begin{lstlisting}[style=default]
-(if (adult? person)
-  (if (male? person)
-     (choose-beer)
-     (choose-wine))
-  (choose-juice))
-\end{lstlisting}
-with continuation
-\begin{lstlisting}[style=default]
-(fn [choice _]
-  (case (kind choice)
-    :beer (beer-jar choice)
-    :wine (wine-glass choice)
-    :juice (juice-bottle choice)))
-\end{lstlisting}
-During CPS transformation, if we substitute the code of this continuation 
-for all of its calls, the code will be repeated three times in 
-the CPS-transformed expression. In general, CPS code can
-grow extremely large if the code of continuations is substituted
-repeatedly.
-
-To circumvent this inefficiency, CPS transformers for
-expressions with multiple continuation points (\texttt{if} and
-derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
-the continuation to a fresh symbol if it is not yet a symbol.
-Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
-establishes the binding automatically:
-\begin{lstlisting}[style=default]
-=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
-(let [cont (fn [x] (* 2 x))]
-  (if c
-      (fn [] (cont t $state))
-      (fn [] (cont f $state))))
-\end{lstlisting}
-}{
-}
+\input{code-explosion.tex}
+}{%
+}%
 
@@
 
-To overcome the above problems, Anglican uses a different scheme
-which is almost as efficient as the scheme based on stack addresses for
-reuse of previously drawn values, while producing addresses of constant size,
-and allows manual computation of checkpoint addresses at runtime
-when the default automatic scheme is insufficient. According to
-the scheme:
+To overcome the above problems, Anglican introduces a new
+addressing scheme which is almost as efficient as the scheme
+based on stack addresses for reuse of previously drawn values,
+while producing addresses of constant size, and allows manual
+computation of checkpoint addresses at runtime when the default
+automatic scheme is insufficient. According to the scheme:
 \begin{itemize}
@@
 
-% \begin{itemize}
-%     \item Values produced by the process are sampled from the distribution
-%         returned by \texttt{produce}.
-%     \item When a value is observed from the distribution, a new
-%         process, with the observed value absorbed, is returned
-%         by \texttt{absorb}.
-% \end{itemize}
-
-% A common pattern when programming with random processes in 
-% Anglican is to generate a sequences of sampled values and 
-% construct an updated process at each step.
-
 \section{Performance}

### 2017-02-04 paper/paper.tex
definition of pp, related work before design outline

@@
 
+\section{Related Work}
+\label{sec:related}
+
+{\it TODO}
+
 \section{Design Outline}
@@
 
-\section{Related Work}
-\label{sec:related}
-
-{\it TODO}
-
 \section{Probabilistic Program}
@@
 
-{\it TODO}
+Anglican is a language for expressing probabilistic programs.
+A probabilistic program is a stateful \textit{deterministic} computation
+$\mathcal{P}$ with the following properties~\cite{TMP+15}:
+\begin{itemize}
+\item Initially, $\mathcal{P}$ expects no arguments.
+\item On every call, $\mathcal{P}$ returns either a distribution
+  and an address $(F,\alpha)$, a distribution and a value $(G, y)$, a
+  value $z$, or $\bot$.
+\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
+  the argument to the next call.
+\item Upon returning $(G, y)$ or $z$, $\mathcal{P}$ is invoked again
+  without arguments.
+\item Upon returning $\bot$, $\mathcal{P}$ terminates.
+\end{itemize}
+A program is run by calling $\mathcal{P}$ repeatedly until
+termination.
+
+A program need not generate the same sequence of random variables in
+every execution. For this reason we assume that each random variable
+$x$ is assigned a unique label $\alpha$, which we call an address,
+that induces a correspondence between variables in different
+executions. Every execution implicitly produces a sequence of triples
+of distributions, values of \textit{latent} random variables, and
+addresses, $(F_i, x_i, \alpha_i)$.  We call this sequence a
+\textit{trace} and denote it by $\pmb{x}$. A trace induces a sequence
+of pairs $(G_j, y_j)$ of distributions and values of
+\textit{observed} random variables. We call this
+sequence an \textit{image} and denote it by $\pmb{y}$. For notational simplicity we assume that the program always generates the same ordered set $\pmb{z}$ of \textit{output} values $z_k$.
+
+The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
+is defined in terms of the product of the probabilities of all random
+choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
+\begin{align}
+  \gamma(\pmb{x}) 
+  &:= 
+  \prod_{i=1}^{\left|\pmb{x}\right|}
+  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
+  \label{eqn:p-trace}
+\end{align}
+The objective of inference in probabilistic program $\mathcal{P}$
+is to discover the distribution of $\pmb{z}$.
 
@@
 \label{sec:language}
+
+\subsection{Syntax}
 

### 2017-02-05 paper/paper.tex
more compact illustrations
less detail, contributions upfront

@@
 
-\subsection*{Contributions}
+\subsection*{Contributions} This paper brings the following
+major contributions:
+\begin{itemize}
+	\item Design and implementation of a probabilistic
+		programming language Anglican involving tight bilateral
+		integration with a general-purpose programming language.
+	\item Definition of semantics of probabilistic forms in
+		Anglican which facilitates full separation of
+		probabilistic programs from inference algorithms.
+	\item Stateless representation of exchangeable
+		random procedures, or processes, seamlessly integrated
+		into pure functional computation.
+	\item Techniques for efficient and compact implementation of
+		inference algorithms, such as representation of
+		inference results as a lazy sequence of samples and 
+		a novel scheme for addressing of checkpoints.
+\end{itemize}
 
@@
 involved, and normal, linear execution of the program is
-disrupted. In Anglican and similar languages there are two types
+disrupted.
+\iftoggle{full}{%
+In Anglican and similar languages there are two types
 of such checkpoints:
@@
 
+}{%
+}%
 Anglican can be mostly implemented as a regular programming
 language, except for the handling of these checkpoints.
+\iftoggle{full}{%
 Depending on the \textit{inference algorithm}, \texttt{sample}
@@
 
+}{%
+}%
 Because of the checkpoints, Anglican programs must allow the
@@
 maintenance of program continuations at checkpoints. Anglican
-follows the latter option. Clojure is
+follows the latter option.
+\iftoggle{full}{%
+Clojure is
 a functional language, and continuation-passing style (CPS)
@@
 computation with access to continuations in probabilistic
-checkpoints. The Anglican `compiler', represented by a set of
-functions in the \texttt{anglican.trap} namespace, accepts a
+checkpoints. 
+}{%
+}%
+The Anglican `compiler'\iftoggle{full}{, represented by a set of
+functions in the \texttt{anglican.trap} namespace,}{} accepts a
 Clojure subset and transforms it into a variant of CPS
@@
 the source of a Clojure program. To facilitate this, Anglican
-programs, or queries, are wrapped by macros (defined in the
-\texttt{anglican.emit} namespace), which call the CPS
+programs, or queries, are wrapped by macros\iftoggle{full}{ (defined in the
+\texttt{anglican.emit} namespace)}{} which call the CPS
 transformations and define Clojure values suitable for passing
@@
 the definitions of \textit{probabilistic functions} using
-\texttt{defm}
-%and \texttt{fm}
-(Anglican counterparts of Clojure \texttt{defn}).
-%and \texttt{fn}).
+\texttt{defm} (Anglican counterpart of Clojure \texttt{defn}).
 Probabilistic functions are
@@
 of Clojure code, therefore queries must be called through
-inference algorithms, rather than `directly'.  The
-\texttt{anglican.inference} namespace declares the (ad-hoc) polymorphic
+inference algorithms, rather than `directly'. \iftoggle{full}{The
+\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
 function \texttt{infer} using Clojure's multimethod mechanism.
@@
 
-Finally, Anglican queries use `primitive', or commonly known
-and used, distributions, to draw random samples and condition
-observations. Many primitive distributions are provided by the
-\texttt{anglican.runtime} namespace, and an additional
-distribution can be defined by the user by implementing a
-particular set of functions for the distribution
-(via Clojure's protocol mechanism). The \texttt{defdist} macro
-provides a convenient syntax for defining primitive distributions.
+Finally, Anglican queries use `primitive', or commonly known and
+used, distributions, to draw random samples and condition
+observations. Many primitive distributions are provided by
+\iftoggle{full}{the \texttt{anglican.runtime} namespace}{the
+runtime}, and an additional distribution can be defined by the
+user by implementing a particular set of functions for the
+distribution (via Clojure's protocol mechanism). The
+\texttt{defdist} macro provides a convenient syntax for defining
+primitive distributions.
 
@@
 Having defined the query, we are now ready to run the query
-using an inference algorithm. Function \texttt{doquery} provided
-by the \texttt{anglican.core} namespace accepts the inference
+using an inference algorithm. Function
+\texttt{doquery}\iftoggle{full}{ provided
+by the \texttt{anglican.core} namespace}{} accepts the inference
 algorithm, the query, and optional parameters, and returns a
@@
 \begin{figure}
+	\centering
     \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
@@
 \begin{figure}
+	\centering
     \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
@@
 The compilation relies on the Clojure \textit{macro} facility,
-and implemented as a library of \textit{functions} in namespace
-\texttt{anglican.trap}, which are invoked by macros.  The CPS
+and implemented as a library of
+\textit{functions}\iftoggle{full}{ in namespace
+\texttt{anglican.trap}, which are}{} invoked by macros.  The CPS
 transformation is organized in top-down manner.  The top-level
@@
 ---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
-modify program state. These forms are translated into
+modify program state. \iftoggle{full}{These forms are translated into
 expressions involving calls of functions from the
-\texttt{anglican.state} namespace. The \texttt{mem} form, which
+\texttt{anglican.state} namespace. }{}The \texttt{mem} form, which
 implements memoization, deserves a more detailed explanation.
@@
 instantiations of the (ad-hoc) polymorphic function
-\texttt{infer} declared as Clojure's multimethod in the
-\texttt{anglican.{\linebreak[0]}inference} namespace. The function accepts an
-algorithm identifier, a query --- the probabilistic program in
-which to perform the inference, an initial value for the query,
-and optional algorithm parameters.
+\texttt{infer} declared as Clojure's multimethod\iftoggle{full}{
+in the \texttt{anglican.{\linebreak[0]}inference}
+namespace}{}. The function accepts an algorithm identifier,
+a query --- the probabilistic program in which to perform the
+inference, an initial value for the query, and optional
+algorithm parameters.
 
@@
 \begin{lstlisting}[style=default]
-    [C1 0]
-    [C2 0]
-    [C2 1]
-    [C1 16]
-    [C1 17]
-    [C1 18]
-    [C2 16]
-    [C3 0]
+  [C1 0][C2 0][C2 1][C1 16][C1 17][C1 18][C2 16][C3 0]
 \end{lstlisting}
@@
 \begin{lstlisting}[style=default]
-    [C1 0]
-    [C2 0]
-    [C1 16]
-    [C1 17]
-    [C2 16]
-    [C2 17]
-    [C3 0]
+  [C1 0][C2 0][C1 16][C1 17][C2 16][C2 17][C3 0]
 \end{lstlisting}
@@
 of 16 proved to be appropriate for all practical purposes.
+\iftoggle{full}{%
 
@@
 algorithm.
+}{%
+}%
 

### 2017-02-05 paper/paper.tex
related work

@@
 Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
-\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
+\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
 of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
@@
 		probabilistic programs from inference algorithms.
-	\item Stateless representation of exchangeable
-		random procedures, or processes, seamlessly integrated
+	\item Stateless representation of random processes, seamlessly integrated
 		into pure functional computation.
@@
 
-\section{Related Work}
-\label{sec:related}
-
-{\it TODO}
+\section{Related Work} \label{sec:related}
+
+Efficient implementation of expressive probabilistic programming languages has
+recently been an active idea of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
+SDT14,WVM14,GS15,GSX+16}. There is often a compromise between the
+expressiveness of the language and the efficiency of inference.  Some languages
+emphasize expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict the
+class of models which can be expressed by the language~\cite{MWG+10,SDT14,NN14}
+to facilitate application of efficient inference algorithms. Anglican allows
+unrestricted specification of probabilistic models in the spirit of
+Church~\cite{GMR+08}, while still supporting efficient scalable
+inference~\cite{WVM14,PWD+14,RNL+2016}.
+
+Probabilistic programming languages are implemented as
+interpreters~\cite{GMR+08,MSP14}, embedded languages~\cite{P09,SGG15,GSX+16},
+compiled languages~\cite{MMR+07,SDT14}, as through source-to-source
+transformation with augmentation~\cite{WSG11,GS15}. Each of these methods
+emphasizes different priorities in language design, such as computational
+efficiency, integration with an existing development environment, ease of
+implementation, or efficiency of inference. Anglican implemented using a
+combination of embedding and source-to-source transformational compilation to
+combine advantages of both approaches.  CPS transformation, employed by
+Anglican compiler, is also used in WebPPL~\cite{GS15} and facilitates clean
+separation between probabilistic programs and inference algorithms and
+diversity of applicable inference algorithms.
+
+Probabilistic programming languages are often implemented as pure functional
+languages~\cite{GMR+08,MSP14,SGG15}. This is because reasoning about
+probabilistic programsand distribution defined by the programs is easier in
+pure functional setting. Anglican is also implemented as a language without
+side-effects. However, there are also languages which follow imperative
+paradigm and support mutable state directly~\cite{SDT14,GSX+16}, or through
+probabilistic primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
+alternative to stateful random primitives in the form of random processes.
 
@@
 For inference, we chose to use  Lightweight
-Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
+Metropolis-Hastings, perhaps somewhat arbitrarily. A strength of
 probabilistic programming is that models are separated from
@@
 however, at runtime a checkpoint corresponding to a single code
-location may occur multiple time due to recurrent invocation of
+location may occur multiple times due to recurrent invocation of
 the function containing the checkpoint. Every unique occurrence
@@
 identically distributed. Random processes are similar to the so called
-`exchangeable random procedures' in Church \cite{GMR+08} and
+`exchangeable random primitives' in Church \cite{GMR+08} and
 Venture \cite{MSP14}. However, random sequences generated by Anglican random 

### 2017-02-05 paper/paper.tex
restructured compilation section

@@
 Efficient implementation of expressive probabilistic programming languages has
-recently been an active idea of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
+recently been an active area of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
 SDT14,WVM14,GS15,GSX+16}. There is often a compromise between the
@@
 languages~\cite{GMR+08,MSP14,SGG15}. This is because reasoning about
-probabilistic programsand distribution defined by the programs is easier in
-pure functional setting. Anglican is also implemented as a language without
+probabilistic programs and distributions defined by the programs, as well as
+implementation of certain inference algorithms~\cite{MYM+15}, is easier in pure
+functional setting. Anglican is also implemented as a language without
 side-effects.  However, there are also languages which follow imperative
@@
 
+\iftoggle{full}{%
 \subsection{Implementation Highlights}
@@
 
-\iftoggle{full}{%
 So far we introduced the basics of Anglican compilation to
@@
 for grasping the language implementation but relatively well-known.
-}{%
-Anglican compilation to Clojure is based on CPS transformation
-and compilation techniques, many of which are
-well known and widely used~\cite{AJ89,A07}.
-}%
 In the rest of the section we focus on challenges we met and
@@
 programming language.
-
-\subsubsection{Continuations}
+}{%
+}%
+
+\iftoggle{full}{\subsubsection}{\subsection}{Probabilistic Forms}
+\label{seq:forms}
+
+There are two proper probabilistic forms turning Anglican into a
+probabilistic programming language --- \texttt{sample} and
+\texttt{observe}. Their purpose is to interrupt deterministic
+computation and transfer control to the inference algorithm.
+Practically, this is achieved through returning
+\textbf{checkpoints} --- Clojure values of the corresponding
+types (\texttt{anglican.trap.sample} or
+\texttt{anglican.trap.observe}). The types contain fields
+specific to each form, as well as the continuation; calling the
+continuation resumes the computation. Checkpoints expose the
+program state to the inference algorithm, and the updated state
+is re-injected into the computation when the continuation is
+called:
+\begin{lstlisting}[style=default]
+=> (cps-of-expression '(sample dist) 'cont)
+(->sample dist cont $state)
+=> (cps-of-expression '(observe dist v) 'cont)
+(->observe dist v cont $state)
+\end{lstlisting}
+Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
+expressions are constructors for Clojure records, and they take 
+values of their fields as arguments.
+
+\iftoggle{full}{\subsubsection }{\subsection}{Memoization}
+
+\iftoggle{full}{%
+In addition to checkpoints, there are a few other special forms
+---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
+modify program state. These forms are translated into
+expressions involving calls of functions from the
+\texttt{anglican.state} namespace. The \texttt{mem} form, which
+implements memoization, deserves a more detailed explanation.
+}{%
+}%
+
+The author of a probabilistic model might want to 
+randomly draw a feature value for each entity in a collection,
+but to retain the same drawn
+value for the same entity during a single run of the probabilistic
+program. For example, a person may have brown or green eyes
+with some probability, but the \textit{same} person will
+always have the same eye color. This can be achieved
+through the use of memoized functions. In Anglican, one
+might write:
+\begin{lstlisting}[style=default]
+(let [eye-color (mem (fn [person]
+                       (sample
+                         (categorical
+                           ['brown 0.5]
+                           ['green 0.5]))))]
+  (if (not= (eye-color 'bill) (eye-color 'john))
+    (eye-color 'bill)
+    (eye-color 'john)))
+\end{lstlisting}
+The \texttt{mem} form converts a function to a memoized variant,
+which remembers past inputs and the corresponding outputs, and 
+returns the remembered output when given such a past input, 
+instead of calling the original function with it. As a result,
+for every input, random draws will be made only for the first time 
+that the memoized function is called with the input; all subsequent
+calls with the input will just reuse these draws and return the same output.
+
+Memoization is often implemented on top of a mutable
+dictionary, where the key is the argument list and the value is
+the returned value. However, there are no mutable data
+structures in a probabilistic program. Hence, \texttt{mem}'s
+memory is stored as a nested dictionary in the program state
+introduced during CPS transformation (function \texttt{mem-cps}).  
+Every memoized function gets a
+unique automatically generated identifier. Each time a memoized
+function is called, one of two continuations is chosen,
+depending on whether the same function (a function with the same
+identifier) was previously called in the same run of the
+probabilistic program with the same arguments. If the memoized
+result is available, the continuation of the memoized function
+call is immediately called with the stored result. Otherwise,
+the argument of \texttt{mem} is called with a continuation which
+first creates an updated state with the memoized result, and
+then calls the `outer' continuation with the result and the
+updated state:
+\begin{lstlisting}[style=default]
+=> (mem-cps '(foo))
+(let [M (gensym "M")]
+  (fn mem23623 [C $state & P]
+    (if (in-mem? $state M P)
+      ;; previously memoized result
+      (fn []
+        (C (get-mem $state M P) $state))
+      ;; new computation
+      (clojure.core/apply
+        foo
+        ;; memoize result in state
+        (fn [V $state]
+          (fn [] (C V (set-mem $state M P V))))
+        $state
+        P)))))
+\end{lstlisting}
+
+Memoized results are not shared among multiple runs of a
+probabilistic program, which is intended. Otherwise, it would be
+impossible to memoize functions with random results.
+
+\iftoggle{full}{\subsubsection{Continuations}}{\subsection{Managing stack size}}
 \label{seq:continuations}
@@
 
-\subsubsection{Primitive Procedures}
+\iftoggle{full}{\subsubsection}{\subsection}{Primitive Procedures}
 \label{seq:primitive}
@@
 \texttt{shading-primitive-procedures} automates the shading}{.}
-
-\subsubsection{Probabilistic Forms}
-\label{seq:forms}
-
-There are two proper probabilistic forms turning Anglican into a
-probabilistic programming language --- \texttt{sample} and
-\texttt{observe}. Their purpose is to interrupt deterministic
-computation and transfer control to the inference algorithm.
-Practically, this is achieved through returning
-\textbf{checkpoints} --- Clojure values of the corresponding
-types (\texttt{anglican.trap.sample} or
-\texttt{anglican.trap.observe}). The types contain fields
-specific to each form, as well as the continuation; calling the
-continuation resumes the computation. Checkpoints expose the
-program state to the inference algorithm, and the updated state
-is re-injected into the computation when the continuation is
-called:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression '(sample dist) 'cont)
-(->sample dist cont $state)
-=> (cps-of-expression '(observe dist v) 'cont)
-(->observe dist v cont $state)
-\end{lstlisting}
-Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
-expressions are constructors for Clojure records, and they take 
-values of their fields as arguments.
-
-In addition to checkpoints, there are a few other special forms
----  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
-modify program state. \iftoggle{full}{These forms are translated into
-expressions involving calls of functions from the
-\texttt{anglican.state} namespace. }{}The \texttt{mem} form, which
-implements memoization, deserves a more detailed explanation.
-
-\paragraph{Memoization}
-The author of a probabilistic model might want to 
-randomly draw a feature value for each entity in a collection,
-but to retain the same drawn
-value for the same entity during a single run of the probabilistic
-program. For example, a person may have brown or green eyes
-with some probability, but the \textit{same} person will
-always have the same eye color. This can be achieved
-through the use of memoized functions. In Anglican, one
-might write:
-\begin{lstlisting}[style=default]
-(let [eye-color (mem (fn [person]
-                       (sample
-                         (categorical
-                           ['brown 0.5]
-                           ['green 0.5]))))]
-  (if (not= (eye-color 'bill) (eye-color 'john))
-    (eye-color 'bill)
-    (eye-color 'john)))
-\end{lstlisting}
-The \texttt{mem} form converts a function to a memoized variant,
-which remembers past inputs and the corresponding outputs, and 
-returns the remembered output when given such a past input, 
-instead of calling the original function with it. As a result,
-for every input, random draws will be made only for the first time 
-that the memoized function is called with the input; all subsequent
-calls with the input will just reuse these draws and return the same output.
-
-Memoization is often implemented on top of a mutable
-dictionary, where the key is the argument list and the value is
-the returned value. However, there are no mutable data
-structures in a probabilistic program. Hence, \texttt{mem}'s
-memory is stored as a nested dictionary in the program state
-introduced during CPS transformation (function \texttt{mem-cps}).  
-Every memoized function gets a
-unique automatically generated identifier. Each time a memoized
-function is called, one of two continuations is chosen,
-depending on whether the same function (a function with the same
-identifier) was previously called in the same run of the
-probabilistic program with the same arguments. If the memoized
-result is available, the continuation of the memoized function
-call is immediately called with the stored result. Otherwise,
-the argument of \texttt{mem} is called with a continuation which
-first creates an updated state with the memoized result, and
-then calls the `outer' continuation with the result and the
-updated state:
-\begin{lstlisting}[style=default]
-=> (mem-cps '(foo))
-(let [M (gensym "M")]
-  (fn mem23623 [C $state & P]
-    (if (in-mem? $state M P)
-      ;; previously memoized result
-      (fn []
-        (C (get-mem $state M P) $state))
-      ;; new computation
-      (clojure.core/apply
-        foo
-        ;; memoize result in state
-        (fn [V $state]
-          (fn [] (C V (set-mem $state M P V))))
-        $state
-        P)))))
-\end{lstlisting}
-
-Memoized results are not shared among multiple runs of a
-probabilistic program, which is intended. Otherwise, it would be
-impossible to memoize functions with random results.
 

### 2017-02-05 paper/paper.tex
restructuring and spelling

@@
 
-A program need not generate the same sequence of random variables in
-every execution. For this reason we assume that each random variable
-$x$ is assigned a unique label $\alpha$, which we call an address,
-that induces a correspondence between variables in different
-executions. Every execution implicitly produces a sequence of triples
-of distributions, values of \textit{latent} random variables, and
-addresses, $(F_i, x_i, \alpha_i)$.  We call this sequence a
-\textit{trace} and denote it by $\pmb{x}$. A trace induces a sequence
-of pairs $(G_j, y_j)$ of distributions and values of
-\textit{observed} random variables. We call this
-sequence an \textit{image} and denote it by $\pmb{y}$. For notational simplicity we assume that the program always generates the same ordered set $\pmb{z}$ of \textit{output} values $z_k$.
+A program need not generate the same sequence of random variables in every
+execution. For this reason we assume that each random variable $x$ is assigned
+a unique label $\alpha$, which we call an address, that induces a
+correspondence between variables in different executions. Every execution
+implicitly produces a sequence of triples of distributions, values of
+\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
+call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
+a sequence of pairs $(G_j, y_j)$ of distributions and values of
+\textit{observed} random variables. We call this sequence an \textit{image} and
+denote it by $\pmb{y}$. For notational simplicity we assume that the program
+always generates the same ordered set $\pmb{z}$ of \textit{output} values
+$z_k$.
 

### 2017-02-06 paper/paper.tex
anglican vs clojure

@@
 \begin{itemize}
-	\item Design and implementation of a probabilistic
-		programming language Anglican involving tight bilateral
-		integration with a general-purpose programming language.
-	\item Definition of semantics of probabilistic forms in
-		Anglican which facilitates full separation of
-		probabilistic programs from inference algorithms.
+    \item Design and implementation of a probabilistic programming language
+        Anglican involving tight bilateral integration with a general-purpose
+        programming language.
+    \item Semantics of probabilistic forms in Anglican which facilitates full
+        separation of probabilistic programs from inference algorithms.
     \item Stateless representation of random processes, seamlessly integrated
         into pure functional computation.
-	\item Techniques for efficient and compact implementation of
-		inference algorithms, such as representation of
-		inference results as a lazy sequence of samples and 
-		a novel scheme for addressing of checkpoints.
+    \item Techniques for efficient and compact implementation of inference
+        algorithms, such as representation of inference results as a lazy
+        sequence of samples and a novel scheme for addressing of checkpoints.
 \end{itemize}
@@
 \texttt{exp}, etc.), but most importantly, it provides
-definitions of common distributions. Each distribution object
-implements a distribution interface with the \texttt{sample*} and \texttt{observe*}
-methods; this interface is defined using Clojure's protocol mechanism
+definitions of common distributions. Each distribution object implements a
+distribution interface with the \texttt{sample*} and \texttt{observe*} methods;
+this interface is defined using Clojure's protocol mechanism
 (\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
@@
 
-{\it TODO}
+To justify the claim that Anglican is an \textit{efficient}
+implementation, performance of Anglican programs must evaluated
+both against Clojure programs and against inference tasks in
+other probabilistic programming languages and environments.
+Anglican passes two additional parameters to every function, adds
+extra code to variable bindings, and passes every function
+call through the trampoline. Hence difference in performance
+between Anglican and Clojure should be most noticeable on
+programs which involve many function calls. One such example is
+the famous \textit{Towers of Hanoi}:
+
+\begin{lstlisting}[style=default]
+(defn towers-of-hanoi [n from to via]
+  (when (not= n 1)
+      (towers-of-hanoi (dec n) from via to)
+      (towers-of-hanoi (dec n) via to from)))
+\end{lstlisting}
+
+We run this program for n equal to 10, 15, 20, and 25. Clojure
+is consistently only twice as fast as Anglican
+(Table~\ref{tbl:hanoi}); considering the amount of overhead
+Anglican introduces, this is a good result.
+\begin{table}
+    \centering
+    \begin{tabular}{r|r|r|r|r}
+        {\bf n}  & {\bf 10} & {\bf 15} & {\bf 20} &{\bf 25} \\ \hline
+        Clojure &  0.32 & 1.45 & 38.1  &  1245.9 \\ \hline
+        Anglican & 0.63 & 3.00 & 75.9 & 2381.1 
+    \end{tabular}
+    \caption{Towers of Hanoi. Average running times, in
+    seconds, of Clojure and Anglican, averaged over 100 runs.
+    Clojure is only twice as fast as Anglican.}
+    \label{tbl:hanoi}
+\end{table}
 

### 2017-02-06 paper/paper.tex
performance

@@
 
+The other comparison must be made between Anglican and other
+probabilistic programming environment. \cite[pp. 32 -- 33]{P16} 
+provides a comparison of Anglican to an older, interpreted
+version of Anglican~\cite{WVM14}, and Probabilistic
+Scheme~\cite{PW14} on a rather complicated inference task of
+program synthesis. According to this comparison, Anglican is at
+least 10 times faster than the older interpreted version, and
+almost as fast as Probabilistic Scheme which uses inference
+engine implemented in C. In addition, \cite[p. 171]{SGG15}
+compares Anglican to a DSL for probabilistic programming in
+Haskell and to Probabilistic C~\cite{PW14} on two simple
+probabilistic programs. Anglican shows similar performance
+to the Haskell implementation of Sequential Monte
+Carlo~\cite{DFG+01} and scales better with the number of
+particles. Given the simplicity of the programs used in the
+evaluation, this comparison is a confirmation that the flexible
+checkpoint and trampoline based interaction between Anglican
+program and inference engine does not introduce any noticeable
+overhead compared to more rigid designs.
+
 \section{Conclusion}

### 2017-02-06 paper/paper.tex
spelling

@@
 evaluation, this comparison is a confirmation that the flexible
-checkpoint and trampoline based interaction between Anglican
-program and inference engine does not introduce any noticeable
+checkpoint and trampoline based interaction between an Anglican
+program and the inference engine does not introduce any noticeable
 overhead compared to more rigid designs.

### 2017-02-06 paper/paper.tex
more details on probabilistic forms

@@
         programming language.
-    \item Semantics of probabilistic forms in Anglican which facilitates full
+    \item Specification of probabilistic forms in Anglican which facilitates 
         separation of probabilistic programs from inference algorithms.
@@
 
-Efficient implementation of expressive probabilistic programming languages has
-recently been an active area of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
-SDT14,WVM14,GS15,GSX+16}. There is often a compromise between the
-expressiveness of the language and the efficiency of inference.  Some languages
-emphasize expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict the
-class of models which can be expressed by the language~\cite{MWG+10,SDT14,NN14}
-to facilitate application of efficient inference algorithms. Anglican allows
-unrestricted specification of probabilistic models in the spirit of
+Efficient implementation of expressive probabilistic programming
+languages has recently been an active area of
+research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
+SDT14,WVM14,GS15,GSX+16}. There is often a compromise between
+the expressiveness of the language and the efficiency of
+inference.  Some languages emphasize
+expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict
+the class of models which can be expressed by the
+language~\cite{MWG+10,SDT14,NN14} to facilitate application of
+efficient inference algorithms. Anglican allows unrestricted
+specification of probabilistic models in the spirit of
 Church~\cite{GMR+08}, while still supporting efficient scalable
@@
 Probabilistic programming languages are implemented as
-interpreters~\cite{GMR+08,MSP14}, embedded languages~\cite{P09,SGG15,GSX+16},
-compiled languages~\cite{MMR+07,SDT14}, as through source-to-source
-transformation with augmentation~\cite{WSG11,GS15}. Each of these methods
-emphasizes different priorities in language design, such as computational
-efficiency, integration with an existing development environment, ease of
-implementation, or efficiency of inference. Anglican implemented using a
-combination of embedding and source-to-source transformational compilation to
-combine advantages of both approaches.  CPS transformation, employed by
-Anglican compiler, is also used in WebPPL~\cite{GS15} and facilitates clean
-separation between probabilistic programs and inference algorithms and
-diversity of applicable inference algorithms.
-
-Probabilistic programming languages are often implemented as pure functional
-languages~\cite{GMR+08,MSP14,SGG15}. This is because reasoning about
-probabilistic programs and distributions defined by the programs, as well as
-implementation of certain inference algorithms~\cite{MYM+15}, is easier in pure
-functional setting. Anglican is also implemented as a language without
-side-effects.  However, there are also languages which follow imperative
-paradigm and support mutable state directly~\cite{SDT14,GSX+16}, or through
-probabilistic primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
-alternative to stateful random primitives in the form of random processes.
+interpreters~\cite{GMR+08,MSP14}, embedded
+languages~\cite{P09,SGG15,GSX+16}, compiled
+languages~\cite{MMR+07,SDT14}, as through source-to-source
+transformation with augmentation~\cite{WSG11,GS15}. Each of
+these methods emphasizes different priorities in language
+design, such as computational efficiency, integration with an
+existing development environment, ease of implementation, or
+efficiency of inference. Anglican implemented using a
+combination of embedding and source-to-source transformational
+compilation to combine advantages of both approaches.  CPS
+transformation, employed by Anglican compiler, is also used in
+WebPPL~\cite{GS15} and facilitates clean separation between
+probabilistic programs and inference algorithms and diversity of
+applicable inference algorithms.
+
+Probabilistic programming languages are often implemented as
+pure functional languages~\cite{GMR+08,MSP14,SGG15}. This is
+because reasoning about probabilistic programs and distributions
+defined by the programs, as well as implementation of certain
+inference algorithms~\cite{MYM+15}, is easier in pure functional
+setting. Anglican is also implemented as a language without
+side-effects.  However, there are also languages which follow
+imperative paradigm and support mutable state
+directly~\cite{SDT14,GSX+16}, or through probabilistic
+primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
+alternative to stateful random primitives in the form of random
+processes.
 
@@
 
-There are two proper probabilistic forms turning Anglican into a
-probabilistic programming language --- \texttt{sample} and
-\texttt{observe}. Their purpose is to interrupt deterministic
-computation and transfer control to the inference algorithm.
-Practically, this is achieved through returning
-\textbf{checkpoints} --- Clojure values of the corresponding
-types (\texttt{anglican.trap.sample} or
-\texttt{anglican.trap.observe}). The types contain fields
-specific to each form, as well as the continuation; calling the
-continuation resumes the computation. Checkpoints expose the
-program state to the inference algorithm, and the updated state
-is re-injected into the computation when the continuation is
-called:
+The purpose of probabilistic forms \texttt{sample} and
+\texttt{observe} is to interrupt deterministic computation and
+transfer control to the inference algorithm.  Practically, this
+is achieved through returning \textbf{checkpoints} --- Clojure
+values of the corresponding types
+(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
+\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe}).
+The types contain fields specific to each form, as well as the
+continuation; calling the continuation resumes the computation.
+Checkpoints expose the program state to the inference algorithm,
+and the updated state is re-injected into the computation when
+the continuation is called:
 \begin{lstlisting}[style=default]
@@
 expressions are constructors for Clojure records, and they take 
-values of their fields as arguments.
-
-\iftoggle{full}{\subsubsection }{\subsection}{Memoization}
+values of their fields as arguments. How the state is updated, 
+the number of times the continuation is called, and the value
+passed to the continuation of \texttt{sample} depend on the
+inference algorithm executing the program.
+Section~\ref{sec:inference} provides more detail on internals of
+inference algorithms and their interaction with probabilistic
+programs.
 
 \iftoggle{full}{%
-In addition to checkpoints, there are a few other special forms
----  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
-modify program state. These forms are translated into
+\smallskip
+\begin{center}
+*\qquad*\qquad*
+\end{center}
+
+In addition to the checkpoints, there are a few other special
+forms ---  \texttt{store}, \texttt{retrieve}, \texttt{mem} ---
+which modify program state. These forms are translated into
 expressions involving calls of functions from the
@@
 }%
+
+\iftoggle{full}{\subsubsection }{\subsection}{Memoization}
 
@@
 defined anonymously via the \texttt{query} form:
-%If needed, the query
-%can be defined anonymously, right in the argument list of a call
-%to \texttt{infer}:
 \begin{lstlisting}[style=default]
@@
 parameters of the query via the destructuring mechanism of Clojure.
-%The value is destructured via pattern match in case of a structured binding.
 For example,
@@
 passed to \texttt{infer}.
-%Finally, an indefinite number of auxiliary arguments can be
-%passed to \texttt{infer}.
 By convention, the arguments should be
@@
 \subsection{Internals of an Inference Algorithm}
-\label{seq:internals}
+\label{sec:internals}
 

### 2017-02-06 paper/paper.tex
Clojure-independent implementation of sample and observe

@@
 	\centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
+    \includegraphics[trim={12pt 0 0 0},scale=0.6]{same-customer-time.pdf}
     \caption{Arrival time distribution for a single customer.}
@@
 	\centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
+    \includegraphics[trim={12pt 0 0 0},scale=0.6]{different-customers-times.pdf}
     \caption{Arrival time distributions for two customers.}
@@
 transfer control to the inference algorithm.  Practically, this
-is achieved through returning \textbf{checkpoints} --- Clojure
-values of the corresponding types
+is achieved by  returning a \textbf{checkpoint} instead of
+invoking a continuation in a CPS-transformed program.  In
+Clojure, checkpoints are represented by values of the
+corresponding record types
 (\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
@@
 continuation; calling the continuation resumes the computation.
-Checkpoints expose the program state to the inference algorithm,
-and the updated state is re-injected into the computation when
-the continuation is called:
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
-expressions are constructors for Clojure records, and they take 
-values of their fields as arguments. How the state is updated, 
-the number of times the continuation is called, and the value
-passed to the continuation of \texttt{sample} depend on the
-inference algorithm executing the program.
-Section~\ref{sec:inference} provides more detail on internals of
-inference algorithms and their interaction with probabilistic
-programs.
+Here \texttt{->sample} and \texttt{->observe} in the
+CPS-transformed expressions are constructors for Clojure
+records, and they take values of their fields as arguments.
+
+Upon encountering a checkpoint, the inference algorithm computes
+the updated program state and the value to be passed to the
+continuation.  How the state is updated, the number of times the
+continuation is called, and the value passed to the continuation
+of \texttt{sample} depend on the inference algorithm executing
+the program.  Section~\ref{sec:inference} provides more detail
+on internals of inference algorithms and their interaction with
+probabilistic programs.
 

### 2017-02-06 paper/paper.tex
better language

@@
 
-In addition to the checkpoints, there are a few other special
+In addition to \texttt{sample} and \texttt{observe}, there are a few other special
 forms ---  \texttt{store}, \texttt{retrieve}, \texttt{mem} ---

### 2017-02-06 paper/paper.tex
spelling

@@
 When an Anglican function is transformed into a Clojure
-function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary parameters are added to the
-beginning of the parameter list --- continuation and state.
-Correspondingly, when a function \textit{call} is
-transformed\iftoggle{full}{ (by \texttt{cps-of-application} or
-\texttt{cps-of-apply})}{}, the
+function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary
+parameters are added to the beginning of the parameter list ---
+continuation and state.  Correspondingly, when a function
+\textit{call} is transformed\iftoggle{full}{ (by
+\texttt{cps-of-application} or \texttt{cps-of-apply})}{}, the
 current continuation and the state are passed to the called
@@
 
-Implementing an inference algorithm in Anglican amounts
-to defining an appropriate version of the \texttt{infer} function and 
-checkpoint handlers for \texttt{sample} and \texttt{observe}.
-The definitions may just reuse default implementations or override them
-with new algorithm-specific treatment of \texttt{sample} and \texttt{observe} forms
-and inference state.
+Implementing an inference algorithm in Anglican amounts to
+defining an appropriate version of the \texttt{infer} function
+and checkpoint handlers for \texttt{sample} and
+\texttt{observe}.  The definitions may just reuse default
+implementations or override them with new algorithm-specific
+treatment of \texttt{sample} and \texttt{observe} forms and
+inference state.
 
@@
 \end{lstlisting}
-If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
+If the program is run by a Metropolis-Hastings algorithm, then a
+small change usually takes place in the sequence of checkpoints
+with each invocation, and the new sequence may become
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-Unlike typical implementations of exchangeable random processes, Anglican's random processes do not have mutable state.
-The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican.
-A sequence of random values can be generated using a recursive loop in which \texttt{absorb} returns the updated process for the next iteration.
-For example:
+Unlike typical implementations of exchangeable random processes,
+Anglican's random processes do not have mutable state. The
+\texttt{produce} and \texttt{absorb} methods are deterministic
+and functional, and therefore do not have corresponding special
+forms in Anglican. A sequence of random values can be generated
+using a recursive loop in which \texttt{absorb} returns the
+updated process for the next iteration. For example:
 \begin{lstlisting}[style=default]
@@
 \end{lstlisting}
-Similarly, random processes can also be used to recursively observe a sequence of values:
+Similarly, random processes can also be used to recursively
+observe a sequence of values:
 \begin{lstlisting}[style=default]

### 2017-02-07 paper/anglican.tex
updated the colors

@@
 		counted?,create-ns,create-struct,cycle,dec,decimal?,declare,def,defdist,definline,defm,defmacro,defmethod,%
-		defmulti,defn,defn-,defonce,defquery,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%
+		defmulti,defn,defn-,defonce,defquery,defproc,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%
 		descendants,destructure,disj,disj!,dissoc,dissoc!,distinct,distinct?,do,do-template,doall,doc,%
@@
 		load-reader,load-string,loaded-libs,locking,long,long-array,longs,loop,macroexpand,macroexpand-1,%
-		make-array,make-hierarchy,map,map?,mapcat,max,max-key,memfn,memoize,merge,merge-with,meta,%
+		make-array,make-hierarchy,map,map?,mapcat,max,max-key,mem,memfn,memoize,merge,merge-with,meta,%
 		method-sig,methods,min,min-key,mod,monitor-enter,monitor-exit,name,namespace,neg?,new,newline,%
@@
 		release-pending-sends,rem,remove,remove-method,remove-ns,remove-watch,repeat,repeatedly,%
-		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,reverse,reversible?,%
+		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,retrieve,reverse,reversible?,%
 		rseq,rsubseq,sample,second,select-keys,send,send-off,seq,seq?,seque,sequence,sequential?,set,set!,%
@@
 		sorted-map-by,sorted-set,sorted-set-by,sorted?,special-form-anchor,special-symbol?,split-at,%
-		split-with,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%
+		split-with,store,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%
 		sync,syntax-symbol-anchor,take,take-last,take-nth,take-while,test,the-ns,throw,time,to-array,%

### 2017-02-07 paper/paper.tex
updated the colors

@@
 \documentclass[preprint]{sigplanconf}
-\linespread{1.00}
 
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican, basicstyle=\linespread{0.95}\ttfamily, columns=flexible, showstringspaces=false}
 
@@
 \begin{lstlisting}[style=default]
-(defquery model data
-  "chooses a distribution
-   which describes the data"
+(defquery model "model selection" data
   (let [;;; Guess a distribution.
@@
 \item On every call, $\mathcal{P}$ returns either a distribution
-  and an address $(F,\alpha)$, a distribution and a value $(G, y)$, a
-  value $z$, or $\bot$.
+  and an address $(F,\alpha)$, a distribution and a value $(G,
+        y)$, or a value $z$.
 \item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
-  the argument to the next call.
-\item Upon returning $(G, y)$ or $z$, $\mathcal{P}$ is invoked again
-  without arguments.
-\item Upon returning $\bot$, $\mathcal{P}$ terminates.
+	the argument to the next call (\texttt{sample} in Anglican).
+\item Upon returning $(G, y)$, $\mathcal{P}$ is invoked again
+	without arguments (\texttt{observe} in Anglican).
+\item Upon returning $z$, $\mathcal{P}$ terminates.
 \end{itemize}
@@
 denote it by $\pmb{y}$. For notational simplicity we assume that the program
-always generates the same ordered set $\pmb{z}$ of \textit{output} values
-$z_k$.
+always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
 
@@
 specified and implementation specific.}.  Anglican queries
-are defined within \texttt{defquery} as shown in the previous section,
-using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case},
-\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
-other forms of Clojure may be supported in the future but are not now.
-In \texttt{let} bindings and \texttt{fn} argument lists,
-Clojure's pattern matching mechanism for vector data type
-(called vector destructuring) is supported. Also, compound
-literals for vectors, hash maps, and
-sets are supported just like in Clojure.
-
-\subsection{Probabilistic Forms}
-
-{\it TODO}
+are defined within \texttt{defquery}. The value of the last
+expression in the query body is the result of the query.
+\begin{lstlisting}[style=default]
+(defquery name doc-string? param? expr*)
+\end{lstlisting}
+Anglican functions outside of a query are defined using
+\texttt{defm} with the same syntax as Clojure \texttt{defn},
+however \texttt{name} is bound to an Anglican function.
+\begin{lstlisting}[style=default]
+(defm name doc-string? [param*] expr*)
+\end{lstlisting}
+Within the body of \texttt{defquery} and \texttt{defm}, the
+following special forms are supported, with syntax and semantics 
+matching those of Clojure:
+\begin{lstlisting}[style=default]
+(if test then else?)
+(when test expr*)
+(cond clause*)
+(case expr clause* default?)
+(let [binding*] expr*)
+(and expr*)
+(or expr*)
+(fn name? [param*] expr*)
+(loop [binding*] expr*)
+(recur expr*)
+\end{lstlisting}
+In \texttt{defquery} parameter, \texttt{let} bindings, and
+\texttt{fn} argument lists, Clojure vector destructuring is
+supported. Literals for atomic types and compound literals for
+vectors, hash maps, and sets are supported just like in Clojure.
+
+Clojure provides special forms \texttt{loop} and \texttt{recur}
+for writing tail-recursive programs.  Anglican programs are
+CPS-converted and do not use the stack; recursive calls
+in Anglican cannot lead to stack overflow.  However,
+\texttt{loop}/\texttt{recur} is provided in Anglican for
+convenience as a way to express loops. Unlike in Clojure though,
+\texttt{recur} supported only inside a \texttt{loop}.
 
@@
 
-\subsection{Tail Call}
-\label{sec:tail}
-
-Clojure provides special forms \texttt{loop} and \texttt{recur}
-for writing tail-recursive programs. Such forms are not necessary
-in Anglican, because Anglican programs are CPS-converted and do
-not use the stack. For instance, no recursive call in Anglican
-can lead to stack overflow.  In fact, in Anglican, it is
-recommended to use recursive calls to functions instead of
-\texttt{recur}. However, \texttt{loop}/\texttt{recur} is
-provided in Anglican for convenience as a way to express loops.
-\texttt{recur} outside of \texttt{loop} will lead to
-unpredictable behaviour and hard-to-catch errors.
+\subsection{Probabilistic Forms}
+
+In addition to re-implementing a subset of Clojure, Anglican
+provides special forms for probabilistic inference:
+\begin{lstlisting}[style=default]
+(sample address? distribution)
+(observe address? distribution value)
+(mem function)
+(store tag value)
+(retrieve tag)
+\end{lstlisting}
+
+\texttt{sample} and \texttt{observe} forms correspond to
+handling $(F, \alpha)$ and $(G, y)$ correspondingly in the
+definition of probabilistic program (Section~\ref{sec:pp-def}).
+Their operational semantics can be defined in terms of
+importance sampling(Equation~\ref{eqn:p-trace}): 
+\begin{itemize}
+	\item \texttt{sample} returns a value drawn from
+		the distribution.
+	\item \texttt{observe} multiplies the unnormalized
+		probability of the program trace by the probability of
+		drawing the value from the distribution.
+\end{itemize}
+However, the actual computation involved in handling of
+\texttt{sample} and \texttt{observe} depends on the inference
+algorithm. Different inference algorithms may treat
+\texttt{sample} and \texttt{observe} differently as long as they
+compute the same distribution of program results.
+Denotational and operational semantics for an idealized version
+of Anglican is introduced in \cite{SYH+16}; in general though,
+specifying semantics of higher-order probabilistic programming
+languages is an open problem.
+
+\texttt{sample} and \texttt{observe} may appear anywhere in the
+code of an Anglican program. To support these forms, the
+inference engine must be able to intervene into execution of the
+program, perform computations related to inference, and control
+further execution of the program. This tight and complicated
+interaction between the program and the inference engine
+necessiates implementing Anglican as a language rather than a
+library.
+
+The remaining Anglican special forms support carrying values in
+the hidden program state. \texttt{mem} implements stochastic
+function  memoization (Section~\ref{sec:mem}). \texttt{store}
+stores a value for a tag in the program state, which can be
+later retrieved by the tag using \texttt{retrieve} in the same
+program execution.
 
@@
   {:p 0.11671977016875924,
-   :dist
-   {:min 0.0,
-    :max 1.0}}}}
+   :dist {:min 0.0, :max 1.0}}}}
 \end{lstlisting}
@@
 \iftoggle{full}{\subsubsection }{\subsection}{Memoization}
+\label{sec:mem}
 
@@
 (let [eye-color (mem (fn [person]
-                       (sample
-                         (categorical
+                       (sample (categorical
                                  ['brown 0.5]
@@
       ;; previously memoized result
-      (fn []
-        (C (get-mem $state M P) $state))
+      (fn [] (C (get-mem $state M P) $state))
       ;; new computation
-      (clojure.core/apply
-        foo
+      (clojure.core/apply foo
         ;; memoize result in state
@@
           (fn [] (C V (set-mem $state M P V))))
-        $state
-        P)))))
+        $state P)))))
 \end{lstlisting}

### 2017-02-07 paper/paper.tex
everything is in place, 12 pages

@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\linespread{0.95}\ttfamily, columns=flexible, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican, basicstyle=\linespread{0.92}\ttfamily, columns=flexible, showstringspaces=false}
 
@@
 stores a value for a tag in the program state, which can be
-later retrieved by the tag using \texttt{retrieve} in the same
+later retrieved by the tag using \texttt{retrieve} during the same
 program execution.

### 2017-02-07 paper/IFL_2016_reviews_RESPONSE.txt
added responses to reviews

1) Add a clear and concise statement of the contributions up front.

Added a bullet list of contributions to the end of introduction.

2) The treatment of related work must be significantly improved.

Added section 2. Related work.

3) A clear and concise overview of the constructs of the language has
to be added, along with a good characterisation of at least the
semantics of the core constructs "sample" and "observe", as formal
as possible.

Rewrote section 5. Language, with the syntax of Clojure forms
supported, core library support, and section 5.3 on
probabilistic forms, with details on their syntax and semantics, 
related to section 4, the definition of probabilistic program.

4) Clarify why the language constructs justify building a new language
instead of making a library.

Added a clarification at the end of subsection 5.3.

5) Clearly bring out the central ideas and techniques of implementing
the core constructs "sample" and "observe" as independently as
possible of the specifics of Clojure so as to facilitate using
these in other (embedded) DSL implementations or, if feasible,
as libraries.

Rewrote subsection 7.2 'Probabilistic Forms' of section 7.
Macro-based compilation to provide more details and techniques
and separate Clojure dependent and language independent issues.

6) Reduce the stress (detail) on the *general* CPS based compilation techniques
so as to allow an improved coverage of the core primitives or other 
aspects that need further explanation.

Removed description of general CPS based compilation techniques.

7) Make it clear that section addressing of checkpoints refers to a known
addressing scheme, and reduce amount of detail if necessary to save
space, referring to earlier work instead.

Upon discussing this point with the reviewers we discovered that
there was misunderstanding, and addressing of checkpoints is
indeed a contribution of the paper:

> The reviewers have re-read this part, and it is now clear there was
> a misunderstanding on their part.
> 
> Thus, ignore point 7 as it stands, and instead just make it clear
> both up-front and where the addressing scheme is discussed that this
> is novel and one of the central contributions of the paper.
> 
> Additionally, in general, try to bring out the central contributions
> and ideas in a way that does not get obscured by too many details
> specific to your setting. I believe many people who seek to implement
> similar languages can learn a lot from your work, and it would be
> great if that is made as easy as possible!

Added addressing of checkpoints to the list of contribution in
the beginning of the paper. In subsection 8.3, Addressing of
checkpoint, changed the language 'uses a different scheme' to
'introduces a new scheme'.

8) Bring the convincing example at page 9 significantly forward.

Case study is now Section 6, right after language definition and
before discussing any implementation issues.

9) Substantiate performance claims by at least some concrete results
from benchmarking.

Added section 10 Performance, with performance comparison
between Anglican and Clojure, as well as between Anglican and
other probabilistic programming environment, with numbers.

10) If more space needs to be saved, remove the two paragraphs starting
"There are several ways to build a programming language..."

Removed the aforementioned paragraphs as well as other content
which is either too specific or trivial for the audience of IFL.

### 2017-02-08 paper/paper.tex
reformatted a paragraph, fixed a typo
more performance evaluation, shorter case study

@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican, basicstyle=\linespread{0.92}\ttfamily, columns=flexible, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{1.0}\ttfamily, columns=flexible, showstringspaces=false}
 
@@
 \iftoggle{full}{%
-Clojure is
-a functional language, and continuation-passing style (CPS)
-transformation is a well-developed technique in the area of
-functional languages. Implementing a variant of CPS
+Clojure is a functional language, and continuation-passing style
+(CPS) transformation is a well-developed technique in the area
+of functional languages. Implementing a variant of CPS
 transformation seemed to be the most flexible and lightweight
-option --- any other form of concurrency would put a higher burden
-on the underlying runtime (JVM) and the operating system.
+option --- any other form of concurrency would put a higher
+burden on the underlying runtime (JVM) and the operating system.
 Consequently, Anglican has been implemented as a CPS-transformed
@@
 
-For inference, we chose to use  Lightweight
-Metropolis-Hastings, perhaps somewhat arbitrarily. A strength of
-probabilistic programming is that models are separated from
-inference. To switch to a different inference algorithm we just
-need to pass a different value to \texttt{doquery}. For example,
-we may decide to use Black-Box Variational Bayes (BBVB), which
-may not work equally well for all probabilistic programs, but is
-much faster to converge.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :bbvb deli nil))
-\end{lstlisting}
-
-We can still use samples and summary statistics with BBVB to
-approximate the posterior distribution. However, variational
-inference approximates the posterior by known distributions,
-and we can directly retrieve the distribution parameters.
-
-\begin{lstlisting}[style=default]
-(clojure.pprint/pprint
-  (anglican.bbvb/get-variational 
-    (nth samples N)))
-\end{lstlisting}
-
-\begin{lstlisting}[style=default]
-{S28209
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- S28217
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- S28215
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- S28219
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist {:min 0.0, :max 1.0}}}}
-\end{lstlisting}
-
-We can guess that the variational distributions correspond to the
-prior distributions in the \texttt{sample} forms, in the order
-of appearance. However, it would help if we could use more
-informative labels instead of automatically generated symbols
-\texttt{S28209}, \texttt{S28217}, etc. Here the option to
-specify identifiers for probabilistic forms explicitly comes
-handy. If we modify the \texttt{sample} forms to use explicit identifiers
-(only the forms are shown for brevity), the output becomes much
-easier to analyse.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-      ...
-      (sample :arrival-time-same
-              time-to-arrive-prior)]
-      ...)
-
-(defm different-customers
-      ... 
-      (sample :arrival-time-first
-              time-to-arrive-prior)
-      ...
-      (sample :arrival-time-second
-              time-to-arrive-prior)]
-      ...)
-
-(defquery deli ...
-      ...
-      (sample :same-or-different
-              (flip p-same))
-      ...)
-\end{lstlisting}
-
-The output becomes much easier to interpret and analyse
-programmatically.
-
-\begin{lstlisting}[style=default]
-{:arrivate-time-same
- {(0 anglican.runtime.normal-distribution)
-  {:mean 10.99753360180663,
-   :sd 0.7290976433082352}},
- :arrival-time-first
- {(0 anglican.runtime.normal-distribution)
-  {:mean 12.668050292254202,
-   :sd 0.9446695174790353}},
- :arrival-time-second
- {(0 anglican.runtime.normal-distribution)
-  {:mean 9.104132559955836,
-   :sd 0.9479290526821788}},
- :same-or-different
- {(0 anglican.runtime.flip-distribution)
-  {:p 0.11671977016875924,
-   :dist {:min 0.0, :max 1.0}}}}
-\end{lstlisting}
+\iftoggle{full}{%
+\input{case-study-bbvb}
+}{}%
 
@@
 programming solution of a problem, implemented and analysed in
-Anglican and Clojure, using two different inference algorithms.
+Anglican and Clojure\iftoggle{full}{, using two different
+inference algorithms}{}.
 
@@
         {\bf n}  & {\bf 10} & {\bf 15} & {\bf 20} &{\bf 25} \\ \hline
-        Clojure &  0.32 & 1.45 & 38.1  &  1245.9 \\ \hline
-        Anglican & 0.63 & 3.00 & 75.9 & 2381.1 
+        Clojure &  0.32 & 1.45 & 38.1  &  1245. \\ \hline
+        Anglican & 0.63 & 3.00 & 75.9 & 2381. 
     \end{tabular}
-    \caption{Towers of Hanoi. Average running times, in
+    \caption{Towers of Hanoi. Running times, in
     seconds, of Clojure and Anglican, averaged over 100 runs.
@@
 
-The other comparison must be made between Anglican and other
-probabilistic programming environment. \cite[pp. 32 -- 33]{P16} 
+The other comparison must be made between Anglican and another
+probabilistic programming environment. WebPPL~\cite{GS15} is
+similar to Anglican in expressivity.  WebPPL runs on top of
+\href{http://v8project.blogspot.com/}{V8}, a high performance
+JavaScript engine featuring a native compiler.
+Table~\ref{tbl:webppl} shows results for two models from WebPPL
+distribution, \textit{LDA} and
+\textit{Logistic regression}. For
+comparison, we re-implemented the models in Anglican, as close
+as possible to the WebPPL implementations, and run the same
+inference algorithm (called MCMC in WebPPL, LMH in Anglican)
+with the same number of iterations. Anglican outperforms WebPPL
+on both models. This can be attributed both to a better
+transformed representation of probabilistic programs and to more
+efficient architecture of the inference engine.
+\begin{table}
+    \centering
+    \begin{tabular}{r|c|c}
+        {\bf model}  & {\bf LDA} & {\bf Logistic regression} \\ \hline
+        Anglican & 22. & 4.8 \\ \hline
+        WebPPL & 31. & 9.4
+    \end{tabular}
+
+    \caption{Anglican vs. WebPPL. Running times, in seconds, of
+    Anglican and WebPPL, for $100\,000$ iterations of Markov Chain
+    Monte Carlo (Lighweight Metropolis-Hastings), averaged over
+    100 runs. Anglican outperforms WebPPL on both models.}
+    \label{tbl:webppl}
+\end{table}
+
+In addition, \cite[pp. 32 -- 33]{P16} 
 provides a comparison of Anglican to an older, interpreted

### 2017-02-08 paper/paper.tex
link to WebPPL distribution
added missing

@@
 
-We run this program for n equal to 10, 15, 20, and 25. Clojure
+We run this program for \texttt{n} equal to 10, 15, 20, and 25. Clojure
 is consistently only twice as fast as Anglican
@@
 JavaScript engine featuring a native compiler.
-Table~\ref{tbl:webppl} shows results for two models from WebPPL
-distribution, \textit{LDA} and
+Table~\ref{tbl:webppl} shows results for two models from
+\href{https://github.com/probmods/webppl}{WebPPL
+distribution}, \textit{LDA} and
 \textit{Logistic regression}. For

### 2017-02-08 paper/case-study-bbvb.tex
added missing

For inference, we chose to use  Lightweight
Metropolis-Hastings, perhaps somewhat arbitrarily. A strength of
probabilistic programming is that models are separated from
inference. To switch to a different inference algorithm we just
need to pass a different value to \texttt{doquery}. For example,
we may decide to use Black-Box Variational Bayes (BBVB), which
may not work equally well for all probabilistic programs, but is
much faster to converge.

\begin{lstlisting}[style=default]
(def samples (doquery :bbvb deli nil))
\end{lstlisting}

We can still use samples and summary statistics with BBVB to
approximate the posterior distribution. However, variational
inference approximates the posterior by known distributions,
and we can directly retrieve the distribution parameters.

\begin{lstlisting}[style=default]
(clojure.pprint/pprint
  (anglican.bbvb/get-variational 
    (nth samples N)))
\end{lstlisting}

\begin{lstlisting}[style=default]
{S28209
 {(0 anglican.runtime.normal-distribution)
  {:mean 10.99753360180663,
   :sd 0.7290976433082352}},
 S28217
 {(0 anglican.runtime.normal-distribution)
  {:mean 12.668050292254202,
   :sd 0.9446695174790353}},
 S28215
 {(0 anglican.runtime.normal-distribution)
  {:mean 9.104132559955836,
   :sd 0.9479290526821788}},
 S28219
 {(0 anglican.runtime.flip-distribution)
  {:p 0.11671977016875924,
   :dist {:min 0.0, :max 1.0}}}}
\end{lstlisting}

We can guess that the variational distributions correspond to the
prior distributions in the \texttt{sample} forms, in the order
of appearance. However, it would help if we could use more
informative labels instead of automatically generated symbols
\texttt{S28209}, \texttt{S28217}, etc. Here the option to
specify identifiers for probabilistic forms explicitly comes
handy. If we modify the \texttt{sample} forms to use explicit identifiers
(only the forms are shown for brevity), the output becomes much
easier to analyse.

\begin{lstlisting}[style=default]
(defm same-customer 
      ...
      (sample :arrival-time-same
              time-to-arrive-prior)]
      ...)

(defm different-customers
      ... 
      (sample :arrival-time-first
              time-to-arrive-prior)
      ...
      (sample :arrival-time-second
              time-to-arrive-prior)]
      ...)

(defquery deli ...
      ...
      (sample :same-or-different
              (flip p-same))
      ...)
\end{lstlisting}

The output becomes much easier to interpret and analyse
programmatically.

\begin{lstlisting}[style=default]
{:arrivate-time-same
 {(0 anglican.runtime.normal-distribution)
  {:mean 10.99753360180663,
   :sd 0.7290976433082352}},
 :arrival-time-first
 {(0 anglican.runtime.normal-distribution)
  {:mean 12.668050292254202,
   :sd 0.9446695174790353}},
 :arrival-time-second
 {(0 anglican.runtime.normal-distribution)
  {:mean 9.104132559955836,
   :sd 0.9479290526821788}},
 :same-or-different
 {(0 anglican.runtime.flip-distribution)
  {:p 0.11671977016875924,
   :dist {:min 0.0, :max 1.0}}}}
\end{lstlisting}

### 2017-02-08 paper/paper.tex
better? language

@@
 \end{quote}
-In addition to the story above, we know that:
+Additionally, we know that:
 \begin{itemize}
@@
 
-In addition to computing the posterior probability that the same
+Besides computing the posterior probability that the same
 customer visited the deli twice, we may want to know the average
@@
 in the \texttt{anglican.{\linebreak[0]}stat} namespace
-along with other
-useful statistical functions.
-This is another illustration of advantage of tight
-integration between Clojure and Anglican --- probabilistic
-models are expressed in Anglican. However, processing of data and
-results can rely on the full power of Clojure.
+along with other useful statistical functions to compute
+summary statistics. This is another illustration of advantage of
+tight integration between Clojure and Anglican --- probabilistic
+models are expressed in Anglican. However, processing of data
+and results can rely on the full power of Clojure.
 

### 2017-02-09 paper/paper.tex
link to source code
empirical evaluation in progress

@@
 
+\usepackage{siunitx}
 \usepackage[color]{changebar}
@@
     \begin{tabular}{r|r|r|r|r}
-        {\bf n}  & {\bf 10} & {\bf 15} & {\bf 20} &{\bf 25} \\ \hline
-        Clojure &  0.32 & 1.45 & 38.1  &  1245. \\ \hline
-        Anglican & 0.63 & 3.00 & 75.9 & 2381. 
+        {\bf n}  & 10 & 15 & 20 & 25 \\ \hline
+		{\bf Anglican} & 0.63 & 3.00 & 75.9 & 2381.  \\ \hline
+		{\bf Clojure} &  0.32 & 1.45 & 38.1  &  1245.
     \end{tabular}
@@
 JavaScript engine featuring a native compiler.
-Table~\ref{tbl:webppl} shows results for two models from
-\href{https://github.com/probmods/webppl}{WebPPL
-distribution}, \textit{LDA} and
-\textit{Logistic regression}. For
-comparison, we re-implemented the models in Anglican, as close
-as possible to the WebPPL implementations, and run the same
-inference algorithm (called MCMC in WebPPL, LMH in Anglican)
-with the same number of iterations. Anglican outperforms WebPPL
-on both models. This can be attributed both to a better
-transformed representation of probabilistic programs and to more
-efficient architecture of the inference engine.
+Table~\ref{tbl:webppl} shows results for a few models from
+\href{https://github.com/probmods/webppl}{WebPPL distribution},
+as well as from Anglican test suite. For the comparison, we
+re-implemented WebPPL models in Anglican, and Anglican models in
+WebPPL, following as close as possible the original
+implementations, and run the same inference algorithm (called
+MCMC in WebPPL, LMH in Anglican) with the same number of
+iterations. Anglican outperforms WebPPL on all models. This can
+be attributed both to a better transformed representation of
+probabilistic programs and to more efficient architecture of the
+inference engine.
 \begin{table}
     \centering
-    \begin{tabular}{r|c|c}
-        {\bf model}  & {\bf LDA} & {\bf Logistic regression} \\ \hline
-        Anglican & 22. & 4.8 \\ \hline
-        WebPPL & 31. & 9.4
+	\begin{tabular}{r|S[table-format=2.2]|S[table-format=2.2]}
+		{\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
+		LDA & 22. & 31. \\ 
+		\shortstack{Linear regression}  & 0. & 0. \\
+		\shortstack{Logistic regression} & 6.2 & 9.4 \\
+		\shortstack{Multivariate regression} & 0. & 0. \\
+		\shortstack{Simple branching} & 0. & 0.
     \end{tabular}
@@
     Monte Carlo (Lighweight Metropolis-Hastings), averaged over
-    100 runs. Anglican outperforms WebPPL on both models.}
+    100 runs. Anglican outperforms WebPPL on all models.}
     \label{tbl:webppl}
 \end{table}
+The source code of the examples is available in the
+\href{https://bitbucket.org/probprogs/anglican-white-paper}{paper
+repository}.
 

### 2017-02-09 paper/paper.tex
added hmm
expanded names of tests

@@
 \href{https://github.com/probmods/webppl}{WebPPL distribution},
-as well as from Anglican test suite. For the comparison, we
-re-implemented WebPPL models in Anglican, and Anglican models in
-WebPPL, following as close as possible the original
-implementations, and run the same inference algorithm (called
-MCMC in WebPPL, LMH in Anglican) with the same number of
-iterations. Anglican outperforms WebPPL on all models. This can
-be attributed both to a better transformed representation of
-probabilistic programs and to more efficient architecture of the
-inference engine.
+as well as from
+\href{https://bitbucket.org/probprog/anglican}{Anglican test
+suite}. For the comparison, we re-implemented WebPPL models in
+Anglican, and Anglican models in WebPPL, following as close as
+possible the original implementations, and run the same
+inference algorithm (called MCMC in WebPPL, LMH in Anglican)
+with the same number of iterations. Anglican outperforms WebPPL
+on all models. This can be attributed both to a better
+transformed representation of probabilistic programs and to more
+efficient architecture of the inference engine.
 \begin{table}
@@
 		{\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
-		LDA & 22. & 31. \\ 
-		\shortstack{Linear regression}  & 0. & 0. \\
-		\shortstack{Logistic regression} & 6.2 & 9.4 \\
-		\shortstack{Multivariate regression} & 0. & 0. \\
-		\shortstack{Simple branching} & 0. & 0.
+		Latent Dirichlet allocation & 22. & 31. \\ 
+		Linear regression  & 0. & 0. \\
+		Logistic regression & 6.2 & 9.4 \\
+		Multivariate regression & 0. & 0. \\
+		Simple branching & 0. & 0. \\
+		Hidden Markov model & 0. & 0.
     \end{tabular}

### 2017-02-09 paper/paper.tex
branching in webppl

@@
     \centering
-	\begin{tabular}{r|S[table-format=2.2]|S[table-format=2.2]}
+    \begin{tabular}{r|S[table-format=2.1]|S[table-format=2.1]}
         {\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
-		Latent Dirichlet allocation & 22. & 31. \\ 
-		Linear regression  & 0. & 0. \\
-		Logistic regression & 6.2 & 9.4 \\
-		Multivariate regression & 0. & 0. \\
-		Simple branching & 0. & 0. \\
-		Hidden Markov model & 0. & 0.
+        Latent Dirichlet allocation & 22.2 & 31.5 \\ 
+        Linear regression  & 6.1 & 8.5 \\
+        Logistic regression & 8.4 & 10.4 \\
+        Multivariate regression & 6.9 & 14.2 \\
+        Simple branching & 2.4 & 5.7 \\
+        Hidden Markov model & 8.7 & 0.
     \end{tabular}

### 2017-02-09 paper/paper.tex
typos

@@
 The other comparison must be made between Anglican and another
-probabilistic programming environment. WebPPL~\cite{GS15} is
-similar to Anglican in expressivity.  WebPPL runs on top of
-\href{http://v8project.blogspot.com/}{V8}, a high performance
-JavaScript engine featuring a native compiler.
+probabilistic programming environment. WebPPL~\cite{GS15} and
+Anglican have similar expressivity and capabilities.  WebPPL
+runs on top of \href{http://v8project.blogspot.com/}{V8}, a high
+performance JavaScript engine featuring a native compiler.
 Table~\ref{tbl:webppl} shows results for a few models from

### 2017-02-09 paper/paper.tex
minor edits of text
experiments done

--- 2017-02-09 paper/paper.tex
@@
     \centering
-	\begin{tabular}{r|S[table-format=2.2]|S[table-format=2.2]}
+    \begin{tabular}{r|S[table-format=2.1]|S[table-format=2.1]}
         {\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
-		Latent Dirichlet allocation & 22. & 31. \\ 
-		Linear regression  & 0. & 0. \\
-		Logistic regression & 6.2 & 9.4 \\
-		Multivariate regression & 0. & 0. \\
-		Simple branching & 0. & 0. \\
-		Hidden Markov model & 0. & 0.
+        Latent Dirichlet allocation & 22.2 & 31.5 \\ 
+        Linear regression  & 6.1 & 8.5 \\
+        Logistic regression & 8.4 & 10.4 \\
+        Multivariate regression & 6.9 & 14.2 \\
+        Simple branching & 2.4 & 5.7 \\
+        Hidden Markov model & 8.7 & 10.2
     \end{tabular}

### 2017-02-09 paper/paper.tex
typoe fixed

@@
 To justify the claim that Anglican is an \textit{efficient}
-implementation, performance of Anglican programs must evaluated
+implementation, performance of Anglican programs must be evaluated
 both against Clojure programs and against inference tasks in

### 2017-02-13 paper/paper.tex
typo fixed: this -> these
in the middle of revising the intro slightly.

@@
 probabilistic code fragments. Anglican claims its right to count
-as a separate language because of the ubiquitous probabilistic
+as a separate language, rather than being a Clojure library, 
+because of the ubiquitous probabilistic
 execution semantics rather than a different syntax,
@@
 syntax to write probabilistic code.
+
+\hy{I am here.}
+
+Anglican programs are executed by inference algorithms, which
+answer probabilistic queries on these programs approximately 
+often using a large number of samples. As already mentioned,
+this execution is significantly different from that of 
+usual programs. An Anglican program is executed multiple times, often hundreds
+of thousands or even millions of times for a single inference
+task. Random choices may affect which parts of the program code
+are executed and how often. Many inference algorithms require
+re-running the program multiple times partially, from a certain
+point on.  Different executions may employ different random
+choices. However, for efficient inference a correspondence
+between random choices in different executions should be
+maintained. These are just some of the challenges which were
+faced and solved during development of Anglican.
 
@@
 According to the addressing scheme, the addresses generated
-for this invocations are
+for these invocations are
 \begin{lstlisting}[style=default]

### 2017-02-13 paper/paper.tex
update

--- 2017-02-09 paper/paper.tex
@@
 probabilistic code fragments. Anglican claims its right to count
-as a separate language because of the ubiquitous probabilistic
+as a separate language
+because of the ubiquitous probabilistic
 execution semantics rather than a different syntax,
@@
 syntax to write probabilistic code.
+
+Anglican programs are executed by inference algorithms, which
+compute approximate answers to certain probabilistic queries, such as posterior
+inference and marginal likelihood. This execution 
+is significantly different from that of usual programs. For instance, 
+inference algorithms typically run Anglican programs multiple times, 
+often hundreds of thousands or even millions of times for a single inference
+task. Also, they make random choices, some of which do not even appear in the program
+explicitly, and these choices may affect which parts of the program code
+are executed and how often. Some inference algorithms 
+re-run the program multiple times partially, from a certain
+point on, while reusing random choices made in the previous runs as much as possible. 
+In a sense, each inference algorithm is nothing but a virtual machine for Anglican that
+executes Anglican programs according to a non-standard operational semantics. Supporting 
+a wide range of non-standard semantics of inference algorithms is the main reason that we
+developed Anglican as a language rather than as a library.
+
+\hy{I am here.}
 

### 2017-02-13 paper/paper.tex
minor

--- 2017-02-09 paper/paper.tex
@@
 
-An implementation of Anglican must therefore address three issues:
-\begin{itemize}
-    \item the Clojure syntax to introduce probabilistic Anglican
-        code inside Clojure modules;
-    \item source-to-source transformation of Anglican programs
-        into Clojure, so that probabilistic execution becomes
-        possible;
-    \item algorithms which run Clojure code, obtained by
-        transforming Anglican programs, according to the
-        probabilistic operational semantics.
-\end{itemize}
-Execution of probabilistic programs by inference algorithms is
-different from execution of deterministic programs.  A
-probabilistic program is executed multiple times, often hundreds
-of thousands or even millions of times for a single inference
-task. Random choices may affect which parts of the program code
-are executed and how often. Many inference algorithms require
-re-running the program multiple times partially, from a certain
-point on.  Different executions may employ different random
-choices. However, for efficient inference a correspondence
-between random choices in different executions should be
-maintained. These are just some of the challenges which were
-faced and solved during development of Anglican.
+Inference algorithms execute Anglican programs while trying to
+answer probabilistic queries on those programs. This execution is 
+significantly different from the one described in the standard 
+operational semantics. These algorithms typically run Anglican programs 
+multiple times, often hundreds of thousands or even millions of times 
+for a single inference task. The algorithms may make random choices 
+that do not correspond to any statements in the program, and decide which parts 
+of the program code are executed and how often. Some inference algorithms 
+re-run the program multiple times partially, from a certain
+point on, while reusing random choices made in the previous runs as much as possible. 
+A good high-level picture is that each inference algorithm specifies a 
+virtual machine that executes Anglican programs according to a 
+non-standard (often probabilistic) operational semantics. Supporting a wide range 
+of inference algorithms and their unusual operational semantics efficiently is the main reason 
+that we developed Anglican as a language rather than as a library.
+
+%An implementation of Anglican must therefore address three issues:
+%\begin{itemize}
+%    \item the Clojure syntax to introduce probabilistic Anglican
+%        code inside Clojure modules;
+%    \item source-to-source transformation of Anglican programs
+%        into Clojure, so that probabilistic execution becomes
+%        possible;
+%    \item algorithms which run Clojure code, obtained by
+%        transforming Anglican programs, according to the
+%        probabilistic operational semantics.
+%\end{itemize}
+%Execution of probabilistic programs by inference algorithms is
+%different from execution of deterministic programs.  A
+%probabilistic program is executed multiple times, often hundreds
+%of thousands or even millions of times for a single inference
+%task. Random choices may affect which parts of the program code
+%are executed and how often. Many inference algorithms require
+%re-running the program multiple times partially, from a certain
+%point on.  Different executions may employ different random
+%choices. However, for efficient inference a correspondence
+%between random choices in different executions should be
+%maintained. These are just some of the challenges which were
+%faced and solved during development of Anglican.
 

### 2017-02-14 paper/paper.tex
revised the intro slightly.

@@
 re-run the program multiple times partially, from a certain
-point on, while reusing random choices made in the previous runs as much as possible. 
-A good high-level picture is that each inference algorithm specifies a 
+point on, while reusing random choices made in the previous runs as much as possible. A good high-level picture is that each inference algorithm specifies a 
 virtual machine that executes Anglican programs according to a 
-non-standard (often probabilistic) operational semantics. Supporting a wide range 
-of inference algorithms and their unusual operational semantics efficiently is the main reason 
-that we developed Anglican as a language rather than as a library.
+non-standard (usually probabilistic) operational semantics. 
+Supporting a wide range of inference algorithms and their unusual semantics 
+the main reason that we developed Anglican as a language rather 
+than as a library.
 
@@
         programming language.
-    \item Specification of probabilistic forms in Anglican which facilitates 
-        separation of probabilistic programs from inference algorithms.
-    \item Stateless representation of random processes, seamlessly integrated
-        into pure functional computation.
     \item Techniques for efficient and compact implementation of inference
@@
         sequence of samples and a novel scheme for addressing of checkpoints.
+    \item Anglican's novel representation of random processes 
+            from statistics and machine learning, such as Beta-Binomial 
+            process. The representation is
+            \emph{stateless} and seamlessly integrates into pure functional 
+            computation. This contrasts with typical \emph{stateful}
+            implementations of these processes in other probabilistic 
+            programming languages.
+            These constructs allowrepresents these p
+%    \item Specification of probabilistic forms in Anglican which facilitates 
+%        separation of probabilistic programs from inference algorithms.
 \end{itemize}

### 2017-02-17 paper/paper.tex
update
minor

@@
             programming languages.
-            These constructs allowrepresents these p
 %    \item Specification of probabilistic forms in Anglican which facilitates 
@@
 interpreters~\cite{GMR+08,MSP14}, embedded
-languages~\cite{P09,SGG15,GSX+16}, compiled
-languages~\cite{MMR+07,SDT14}, as through source-to-source
+languages~\cite{P09,SGG15,GSX+16}, and compiled
+languages~\cite{MMR+07,SDT14}, and also through source-to-source
 transformation with augmentation~\cite{WSG11,GS15}. Each of
@@
 existing development environment, ease of implementation, or
-efficiency of inference. Anglican implemented using a
+efficiency of inference. Anglican is implemented using a
 combination of embedding and source-to-source transformational
@@
 Probabilistic programming languages are often implemented as
-pure functional languages~\cite{GMR+08,MSP14,SGG15}. This is
+pure functional languages~\cite{GMR+08,MSP14,SGG15} in
+the sense that they do not allow mutable states. This is
 because reasoning about probabilistic programs and distributions
 defined by the programs, as well as implementation of certain
-inference algorithms~\cite{MYM+15}, is easier in pure functional
+inference algorithms~\cite{MYM+15}, is easier in such pure functional
 setting. Anglican is also implemented as a language without
-side-effects.  However, there are also languages which follow
+mutable states.  However, there are also languages which follow
 imperative paradigm and support mutable state

### 2017-02-17 paper/paper.tex
minor

@@
 
-Operational semantics of Anglican queries is different from that
-of Clojure code, therefore queries must be called through
-inference algorithms, rather than `directly'. \iftoggle{full}{The
+%Operational semantics of Anglican queries is different from that
+%of Clojure code, therefore queries must be called through
+%inference algorithms, rather than `directly'. 
+Operational semantics of an Anglican query is different from that
+of Clojure code, and is determined by an inference algorithm. Thus,
+Anglican queries must be called through these inference algorithms, rather
+rather than `directly'. For this purpose,
+\iftoggle{full}{the
 \texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
 function \texttt{infer} using Clojure's multimethod mechanism.
-This function accepts an Anglican query and
-returns a lazy sequence of weighted samples from the
-distribution defined by the query.  When inference
-is performed on an Anglican query, the query is run by a
-particular inference algorithm. Inference algorithms must
-provide an implementation for \texttt{infer}, as well as
+The function accepts and runs an Anglican query, 
+and returns a lazy sequence of weighted samples from the
+distribution defined by the query. Providing an implementation of this function
+is a responsibilit of an inference algorithm, which should also
 override the polymorphic function \texttt{checkpoint} (defined
@@
 termination of a probabilistic program.
+
+%When inference
+%is performed on an Anglican query, the query is run by a
+%particular inference algorithm. Inference algorithms must
+%provide an implementation for \texttt{infer}, as well as
+%override the polymorphic function \texttt{checkpoint} (defined
+%as a multimethod) so as to handle \texttt{sample} and
+%\texttt{observe} in an algorithm-specific manner and
+%to construct an appropriate result on the
+%termination of a probabilistic program.
 

### 2017-02-20 paper/paper.tex
update
in the middle of revising the Probabilistic Program section

@@
 
-Anglican is a language for expressing probabilistic programs.
-A probabilistic program is a stateful \textit{deterministic} computation
-$\mathcal{P}$ with the following properties~\cite{TMP+15}:
+\hy{Need to be revised.}
+
+Programs in Anglican define probability distributions 
+over sequences of sampled values through their executions.
+
+Imagine an interpreter of Anglican that runs the deterministic parts of a Anglican program 
+according to its standard semantics, executes its \texttt{sample} form by
+generating a random sample, and treats its $\texttt{observe}$ form by skip.
+In addition to this execution, the interpreter keeps a log that records
+information about all the \texttt{sample} and \texttt{observe} forms encountered 
+during execution. The information kept for \texttt{sample} is a triple $(F,x,\alpha)$ of
+the following data:
 \begin{itemize}
-\item Initially, $\mathcal{P}$ expects no arguments.
-\item On every call, $\mathcal{P}$ returns either a distribution
-  and an address $(F,\alpha)$, a distribution and a value $(G,
-        y)$, or a value $z$.
-\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
-    the argument to the next call (\texttt{sample} in Anglican).
-\item Upon returning $(G, y)$, $\mathcal{P}$ is invoked again
-    without arguments (\texttt{observe} in Anglican).
-\item Upon returning $z$, $\mathcal{P}$ terminates.
+        \item $F$ is one of the built-in primitive probability distributions, 
+                such as the standard normal, for which we have
+                the probability density or probability mass function $p_F$.
+        \item $x$ is a value sampled from the distribution $F$.
+        \item $\alpha$ is an address that uniquely and systematically identifies the random 
+                choice made; this address is exploited in inference algorithms, as
+                explained later.
 \end{itemize}
-A program is run by calling $\mathcal{P}$ repeatedly until
-termination.
-
-A program need not generate the same sequence of random variables in every
-execution. For this reason we assume that each random variable $x$ is assigned
-a unique label $\alpha$, which we call an address, that induces a
-correspondence between variables in different executions. Every execution
-implicitly produces a sequence of triples of distributions, values of
-\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
-call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
+The information kept for \texttt{observe} is a pair $(G,y)$ where $G$
+is a primitive probability distribution as $F$ above and $y$ is an observed value.
+Thus, a log is just a sequence of triples $(F,x,\alpha)$ and pairs $(G,y)$.
+
+We call the triple part of a log \textit{trace}. This part records all the random 
+choices of an execution. Thus, if the trace of a log is fixed, the rest of the log 
+is decided determistically by the program. A trace induces
 a sequence of pairs $(G_j, y_j)$ of distributions and values of
@@
 is to discover the distribution of $\pmb{z}$.
+
+%\section{Probabilistic Program}
+%\label{sec:pp-def}
+%
+%Anglican is a language for expressing probabilistic programs.
+%A probabilistic program is a stateful \textit{deterministic} computation
+%$\mathcal{P}$ with the following properties~\cite{TMP+15}:
+%\begin{itemize}
+%\item Initially, $\mathcal{P}$ expects no arguments.
+%\item On every call, $\mathcal{P}$ returns either a distribution
+%  and an address $(F,\alpha)$, a distribution and a value $(G,
+%        y)$, or a value $z$.
+%\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
+%    the argument to the next call (\texttt{sample} in Anglican).
+%\item Upon returning $(G, y)$, $\mathcal{P}$ is invoked again
+%    without arguments (\texttt{observe} in Anglican).
+%\item Upon returning $z$, $\mathcal{P}$ terminates.
+%\end{itemize}
+%A program is run by calling $\mathcal{P}$ repeatedly until
+%termination.
+%
+%A program need not generate the same sequence of random variables in every
+%execution. For this reason we assume that each random variable $x$ is assigned
+%a unique label $\alpha$, which we call an address, that induces a
+%correspondence between variables in different executions. Every execution
+%implicitly produces a sequence of triples of distributions, values of
+%\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
+%call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
+%a sequence of pairs $(G_j, y_j)$ of distributions and values of
+%\textit{observed} random variables. We call this sequence an \textit{image} and
+%denote it by $\pmb{y}$. For notational simplicity we assume that the program
+%always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
+%
+%The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
+%is defined in terms of the product of the probabilities of all random
+%choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
+%\begin{align}
+%  \gamma(\pmb{x}) 
+%  &:= 
+%%  \prod_{i=1}^{\left|\pmb{x}\right|}
+%  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
+%  \label{eqn:p-trace}
+%\end{align}
+%The objective of inference in probabilistic program $\mathcal{P}$
+%is to discover the distribution of $\pmb{z}$.
 

### 2017-02-20 paper/paper.tex
update

@@
 
-\hy{Need to be revised.}
-
-Programs in Anglican define probability distributions 
-over sequences of sampled values through their executions.
-
-Imagine an interpreter of Anglican that runs the deterministic parts of a Anglican program 
-according to its standard semantics, executes its \texttt{sample} form by
-generating a random sample, and treats its $\texttt{observe}$ form by skip.
-In addition to this execution, the interpreter keeps a log that records
+Anglican programs define probability distributions 
+over sequences of annotated values, implicitly by means
+of program execution.
+Imagine an interpreter that runs the deterministic parts of a program 
+according to the standard semantics, executes the \texttt{sample} form by
+generating a random sample, and treats the $\texttt{observe}$ form by skip.
+In addition to this execution, the interpreter also keeps a log that records
 information about all the \texttt{sample} and \texttt{observe} forms encountered 
-during execution. The information kept for \texttt{sample} is a triple $(F,x,\alpha)$ of
-the following data:
-\begin{itemize}
-        \item $F$ is one of the built-in primitive probability distributions, 
+during execution. The information recorded for \texttt{sample} is 
+a triple $(F,x,\alpha)$ of (i) a primitive probability distribution $F$, 
 such as the standard normal, for which we have 
-                the probability density or probability mass function $p_F$.
-        \item $x$ is a value sampled from the distribution $F$.
-        \item $\alpha$ is an address that uniquely and systematically identifies the random 
-                choice made; this address is exploited in inference algorithms, as
-                explained later.
-\end{itemize}
-The information kept for \texttt{observe} is a pair $(G,y)$ where $G$
-is a primitive probability distribution as $F$ above and $y$ is an observed value.
-Thus, a log is just a sequence of triples $(F,x,\alpha)$ and pairs $(G,y)$.
-
-We call the triple part of a log \textit{trace}. This part records all the random 
-choices of an execution. Thus, if the trace of a log is fixed, the rest of the log 
-is decided determistically by the program. A trace induces
-a sequence of pairs $(G_j, y_j)$ of distributions and values of
-\textit{observed} random variables. We call this sequence an \textit{image} and
-denote it by $\pmb{y}$. For notational simplicity we assume that the program
-always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
-
-The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
-is defined in terms of the product of the probabilities of all random
-choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
+the probability density $p_F$; (ii) a value $x$ sampled from the distribution $F$; 
+and (iii) an address $\alpha$ that uniquely and systematically identifies the random 
+choice made. The information recorded for \texttt{observe} is a pair $(G,y)$ where $G$
+is a primitive probability distribution as $F$ from above and $y$ is an 
+observed value. Thus, a log is a sequence of triples $(F,x,\alpha)$ and 
+pairs $(G,y)$.
+
+A \textit{trace} $\pmb{x}$ is a sequence of triples $(F,x,\alpha)$. It is feasible if
+it is precisely the triple part of the log of some execution. 
+Such a feasible trace records all the random choices made during execution,
+and uniquely determines the rest of the execution and its log, because all the non-sample
+forms of Anglican programs are deterministic. In particular, each of these feasible traces
+decides a sequence of pairs $(G_j, y_j)$ for observed values.
+We call the sequence \textit{image of $\pmb{x}$} and denote it by~$\pmb{y}$.
+
+A (almost surely terminating) probabilistic program defines a probability distribution 
+over feasible traces $\pmb{x}$, which has the probability density 
+$\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
 \begin{align}
@@
 \end{align}
-The objective of inference in probabilistic program $\mathcal{P}$
-is to discover the distribution of $\pmb{z}$.
+for the image $\pmb{y}$ of $\pmb{x}$, and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
+of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of
+a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
+a countable sum of Lebesque and counting measures.}
 

### 2017-02-21 paper/paper.tex
update

@@
 Anglican programs define probability distributions 
-over sequences of annotated values, implicitly by means
-of program execution.
-Imagine an interpreter that runs the deterministic parts of a program 
+over sequences of values, implicitly by means of program execution.
+A good way to understand this is to imagine the following 
+interpreter of Anglican programs. Starting from a fixed initial state,
+the interpreter runs the deterministic parts of a program 
 according to the standard semantics, executes the \texttt{sample} form by
 generating a random sample, and treats the $\texttt{observe}$ form by skip.
-In addition to this execution, the interpreter also keeps a log that records
+More importantly, the interpreter keeps a log that records
 information about all the \texttt{sample} and \texttt{observe} forms encountered 
@@
 
-A \textit{trace} $\pmb{x}$ is a sequence of triples $(F,x,\alpha)$. It is feasible if
-it is precisely the triple part of the log of some execution. 
-Such a feasible trace records all the random choices made during execution,
-and uniquely determines the rest of the execution and its log, because all the non-sample
-forms of Anglican programs are deterministic. In particular, each of these feasible traces
-decides a sequence of pairs $(G_j, y_j)$ for observed values.
-We call the sequence \textit{image of $\pmb{x}$} and denote it by~$\pmb{y}$.
-
-A (almost surely terminating) probabilistic program defines a probability distribution 
-over feasible traces $\pmb{x}$, which has the probability density 
+One important property of logs is that they are determined by 
+their projections to triples: when two logs project to the same
+sequence of triples, they must be the same. This is because
+the triples in a log contain all the information about random choices
+made during execution and all the non-sample forms in Anglican programs
+are deterministic. We define a \textit{trace} $\pmb{x}$ to be a sequence
+of triples $(F,x,\alpha)$, and say that $\pmb{x}$ is feasible if
+the trace is precisely the triple part of the log of some execution. 
+Such a feasible trace uniquely determines the rest of the execution and its log.
+In particular, it decides the pair part of the log, namely, 
+a sequence of $(G_j, y_j)$ for observed values. We call this sequence \textit{image of $\pmb{x}$} 
+and denote it by~$\pmb{y}$.
+
+A (almost-surely terminating) probabilistic program defines a probability distribution 
+over finite feasible traces $\pmb{x}$. The distribution has the probability density 
 $\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
@@
   \prod_{i=1}^{\left|\pmb{x}\right|}
-  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
+  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})
+  \quad \mbox{for the image $\pmb{y}$ of $\pmb{x}$}, 
   \label{eqn:p-trace}
 \end{align}
-for the image $\pmb{y}$ of $\pmb{x}$, and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
+and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
 of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of

### 2017-02-21 paper/paper.tex
update

@@
 
-\texttt{sample} and \texttt{observe} forms correspond to
-handling $(F, \alpha)$ and $(G, y)$ correspondingly in the
-definition of probabilistic program (Section~\ref{sec:pp-def}).
-Their operational semantics can be defined in terms of
-importance sampling(Equation~\ref{eqn:p-trace}): 
-\begin{itemize}
-    \item \texttt{sample} returns a value drawn from
-        the distribution.
-    \item \texttt{observe} multiplies the unnormalized
-        probability of the program trace by the probability of
-        drawing the value from the distribution.
-\end{itemize}
-However, the actual computation involved in handling of
-\texttt{sample} and \texttt{observe} depends on the inference
-algorithm. Different inference algorithms may treat
+A good way to understand the \texttt{sample} and \texttt{observe} forms
+is to consider the imaginary interpreter in Section~\ref{sec:pp-def}.
+Under this interpreter, the \texttt{sample} form draws a value $x$ from 
+its parameter distribution $F$, and 
+generates a unique address $\alpha$ for this random draw.
+Then, it appends $(F, x, \alpha)$, the record of this random
+draw, to the log of the current execution. Finally, the
+form returns $x$. The computation of the \texttt{observe} form is
+simpler. It just appends its two parameters, distribution $G$
+and observed value $v$, to the log of the current execution,
+and returns \texttt{nil}.
+
+What we have just described is idealised semantics 
+of \texttt{sample} and \texttt{observe}. The actual computation 
+involved in handling \texttt{sample} and \texttt{observe} 
+depends on the inference algorithm. Different inference algorithms may treat
 \texttt{sample} and \texttt{observe} differently as long as they
-compute the same distribution of program results.
-Denotational and operational semantics for an idealized version
-of Anglican is introduced in \cite{SYH+16}; in general though,
+compute the same distribution on traces (i.e. sequences of sampled values), 
+namely, the one in Equation~\ref{eqn:p-trace}.
+Formal semantics for an idealized version
+of Anglican or other higher-order probabilistic programming languages are introduced 
+in \cite{SYH+16,BorgstromLGS16}; in general though,
 specifying semantics of higher-order probabilistic programming
 languages is an open problem.
+
+%Their operational semantics can be defined in terms of
+%importance sampling(Equation~\ref{eqn:p-trace}): 
+%\begin{itemize}
+%    \item \texttt{sample} returns a value drawn from
+%        the distribution.
+%    \item \texttt{observe} multiplies the unnormalized
+%        probability of the program trace by the probability of
+%        drawing the value from the distribution.
+%\end{itemize}
+%However, the actual computation involved in handling of
+%\texttt{sample} and \texttt{observe} depends on the inference
+%algorithm. Different inference algorithms may treat
+%\texttt{sample} and \texttt{observe} differently as long as they
+%compute the same distribution of program results.
+%Denotational and operational semantics for an idealized version
+%of Anglican is introduced in \cite{SYH+16}; in general though,
+%specifying semantics of higher-order probabilistic programming
+%languages is an open problem.
 

### 2017-02-22 paper/paper.tex
adding a reference to selective CPS. minor edit.

@@
   "initial program state"
-  {:log-weight 0.0,
-   :result nil,
-   ::mem {},
-   ::store nil,
+  {:log-weight 0.0, ;; map :log-weight to 0.0
+   :result nil,       ;; map :result to nil
+   ::store nil,       ;; map :store to nil
+   ::mem {},          ;; map ::mem to the empty map 
    ... })
@@
 is achieved by  returning a \textbf{checkpoint} instead of
-invoking a continuation in a CPS-transformed program.  In
-Clojure, checkpoints are represented by values of the
+invoking a continuation in a CPS-transformed program.  
+Checkpoints are values of the
 corresponding record types
 (\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
-\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe}).
-The types contain fields specific to each form, as well as the
+\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
+in Clojure. The types contain fields specific to each form, as well as the
 continuation; calling the continuation resumes the computation.
@@
 Memoization is often implemented on top of a mutable
-dictionary, where the key is the argument list and the value is
+dictionary, where the key is the argument  and the value is
 the returned value. However, there are no mutable data
@@
 identifier) was previously called in the same run of the
-probabilistic program with the same arguments. If the memoized
+probabilistic program with the same argument. If the memoized
 result is available, the continuation of the memoized function
@@
 To allow the mixing of Anglican (CPS-transformed) and Clojure
-function calls in Anglican code, the Anglican compiler must be
+function calls in Anglican code, the Anglican compiler must CPS-transform
+expressions selectively~\cite{Nielsen01}. It must be
 able to recognize `primitive' (that is, implemented in Clojure
-rather than in Anglican) functions.
+rather than in Anglican) functions, and invoke
+those functions in a direct, not CPS, style.
 

### 2017-02-22 paper/paper.tex
minor

@@
 namespace}{}. The function accepts an algorithm identifier,
-a query --- the probabilistic program in which to perform the
-inference, an initial value for the query, and optional
+a query (the probabilistic program in which to perform the
+inference), an initial value for the query, and optional
 algorithm parameters.
@@
 \begin{lstlisting}[style=default]
-(derive ::algorithm
-        :anglican.inference/algorithm)
+;; Make ::algorithm be converted to 
+;; :anglican.inference/algorithm whenever necessary
+(derive ::algorithm :anglican.inference/algorithm)
+
 ;; invoked when algo parameter is :importance

### 2017-02-22 paper/paper.tex
minor
update

@@
     \item Anglican's novel representation of random processes 
-            from statistics and machine learning, such as Beta-Binomial 
+            from statistics and machine learning, such as beta-Bernoulli
             process. The representation is
@@
 invoking a continuation in a CPS-transformed program.  
-Checkpoints are values of the
+Checkpoints are represented by values of the
 corresponding record types
@@
 The \texttt{sample*} \textit{method} returns a random sample and
-roughly corresponds to the \texttt{sample} checkpoint,
-the \texttt{observe*} \textit{method} returns the log probability
+roughly corresponds to the \texttt{sample} checkpoint.
+The \texttt{observe*} \textit{method} returns the log probability
 of the value and roughly corresponds to the \texttt{observe}
@@
        [value0 & values0] values] 
+  (if (= (count values0) 0)
+    nil
     (let [dist (produce process)]
       (observe dist value0)
-    (recur (absorb process value0) values0)))
+      (recur (absorb process value0) values0))))
 \end{lstlisting}
@@
 (defn towers-of-hanoi [n from to via]
-  (when (not= n 1)
+  (when (not= n 1) ;; return nil if n = 1.
       (towers-of-hanoi (dec n) from via to)
@@
 probabilistic programming environment. WebPPL~\cite{GS15} is
-similar to Anglican in expressivity.  WebPPL runs on top of
+similar to Anglican in expressiveness.  WebPPL runs on top of
 \href{http://v8project.blogspot.com/}{V8}, a high performance

### 2017-02-23 paper/paper.tex
ack to Hongseok's funding.
minor (mostly cosmetic) edits

--- 2017-02-22 paper/paper.tex
@@
 \newtoggle{full}
-\input{toggle}
+\input{toggle-ifl}
 
@@
     a random process, and an inference algorithm.
-
     We show that a probabilistic functional language can be
@@
 models making the inference inherently faster.
+\jw{How exactly does probabilistic programming make inference inherently faster?}
 
@@
 \begin{lstlisting}[style=default]
-(defquery model "model selection" data
+  (defquery model "model selection" [data]
     (let [;;; Guess a distribution.
@@
 %inference algorithms, rather than `directly'. 
-Operational semantics of an Anglican query is different from that
-of Clojure code, and is determined by an inference algorithm. Thus,
+The operational semantics of an Anglican query are different from those
+of Clojure code, and are determined by an inference algorithm. Thus,
 Anglican queries must be called through these inference algorithms, rather
@@
 function \texttt{infer} using Clojure's multimethod mechanism.
-The function accepts and runs an Anglican query, 
+This function accepts and runs an Anglican query, 
 and returns a lazy sequence of weighted samples from the
 distribution defined by the query. Providing an implementation of this function
-is a responsibilit of an inference algorithm, which should also
+is a responsibility of the inference algorithm, which should also
 override the polymorphic function \texttt{checkpoint} (defined
@@
 
-Anglican programs define probability distributions 
-over sequences of values, implicitly by means of program execution.
-A good way to understand this is to imagine the following 
-interpreter of Anglican programs. Starting from a fixed initial state,
-the interpreter runs the deterministic parts of a program 
-according to the standard semantics, executes the \texttt{sample} form by
-generating a random sample, and treats the $\texttt{observe}$ form by skip.
-More importantly, the interpreter keeps a log that records
+\hy{Need to be revised.}
+
+Programs in Anglican define probability distributions 
+over sequences of sampled values through their executions.
+
+Imagine an interpreter of Anglican that runs the deterministic parts of a Anglican program 
+according to its standard semantics, executes its \texttt{sample} form by
+generating a random sample, and treats its $\texttt{observe}$ form by skip.
+In addition to this execution, the interpreter keeps a log that records
 information about all the \texttt{sample} and \texttt{observe} forms encountered 
-during execution. The information recorded for \texttt{sample} is 
-a triple $(F,x,\alpha)$ of (i) a primitive probability distribution $F$, 
+during execution. The information kept for \texttt{sample} is a triple $(F,x,\alpha)$ of
+the following data:
+\begin{itemize}
+        \item $F$ is one of the built-in primitive probability distributions, 
                 such as the standard normal, for which we have
-the probability density $p_F$; (ii) a value $x$ sampled from the distribution $F$; 
-and (iii) an address $\alpha$ that uniquely and systematically identifies the random 
-choice made. The information recorded for \texttt{observe} is a pair $(G,y)$ where $G$
-is a primitive probability distribution as $F$ from above and $y$ is an 
-observed value. Thus, a log is a sequence of triples $(F,x,\alpha)$ and 
-pairs $(G,y)$.
-
-One important property of logs is that they are determined by 
-their projections to triples: when two logs project to the same
-sequence of triples, they must be the same. This is because
-the triples in a log contain all the information about random choices
-made during execution and all the non-sample forms in Anglican programs
-are deterministic. We define a \textit{trace} $\pmb{x}$ to be a sequence
-of triples $(F,x,\alpha)$, and say that $\pmb{x}$ is feasible if
-the trace is precisely the triple part of the log of some execution. 
-Such a feasible trace uniquely determines the rest of the execution and its log.
-In particular, it decides the pair part of the log, namely, 
-a sequence of $(G_j, y_j)$ for observed values. We call this sequence \textit{image of $\pmb{x}$} 
-and denote it by~$\pmb{y}$.
-
-A (almost-surely terminating) probabilistic program defines a probability distribution 
-over finite feasible traces $\pmb{x}$. The distribution has the probability density 
-$\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
+                the probability density or probability mass function $p_F$.
+        \item $x$ is a value sampled from the distribution $F$.
+        \item $\alpha$ is an address that uniquely and systematically identifies the random 
+                choice made; this address is exploited in inference algorithms, as
+                explained later.
+\end{itemize}
+The information kept for \texttt{observe} is a pair $(G,y)$ where $G$
+is a primitive probability distribution as $F$ above and $y$ is an observed value.
+Thus, a log is just a sequence of triples $(F,x,\alpha)$ and pairs $(G,y)$.
+
+We call the triple part of a log \textit{trace}. This part records all the random 
+choices of an execution. Thus, if the trace of a log is fixed, the rest of the log 
+is decided determistically by the program. A trace induces
+a sequence of pairs $(G_j, y_j)$ of distributions and values of
+\textit{observed} random variables. We call this sequence an \textit{image} and
+denote it by $\pmb{y}$. For notational simplicity we assume that the program
+always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
+
+The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
+is defined in terms of the product of the probabilities of all random
+choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
 \begin{align}
@@
   \prod_{i=1}^{\left|\pmb{x}\right|}
-  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})
-  \quad \mbox{for the image $\pmb{y}$ of $\pmb{x}$}, 
+  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
   \label{eqn:p-trace}
 \end{align}
-and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
-of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of
-a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
-a countable sum of Lebesque and counting measures.}
+The objective of inference in probabilistic program $\mathcal{P}$
+is to characterize the distribution of $\pmb{z}$.
 
@@
 
-A good way to understand the \texttt{sample} and \texttt{observe} forms
-is to consider the imaginary interpreter in Section~\ref{sec:pp-def}.
-Under this interpreter, the \texttt{sample} form draws a value $x$ from 
-its parameter distribution $F$, and 
-generates a unique address $\alpha$ for this random draw.
-Then, it appends $(F, x, \alpha)$, the record of this random
-draw, to the log of the current execution. Finally, the
-form returns $x$. The computation of the \texttt{observe} form is
-simpler. It just appends its two parameters, distribution $G$
-and observed value $v$, to the log of the current execution,
-and returns \texttt{nil}.
-
-What we have just described is idealised semantics 
-of \texttt{sample} and \texttt{observe}. The actual computation 
-involved in handling \texttt{sample} and \texttt{observe} 
-depends on the inference algorithm. Different inference algorithms may treat
+\texttt{sample} and \texttt{observe} forms correspond to
+handling $(F, \alpha)$ and $(G, y)$ correspondingly in the
+definition of probabilistic program (Section~\ref{sec:pp-def}).
+Their operational semantics can be defined in terms of
+importance sampling(Equation~\ref{eqn:p-trace}): 
+\begin{itemize}
+    \item \texttt{sample} returns a value drawn from
+        the distribution.
+    \item \texttt{observe} multiplies the unnormalized
+        probability of the program trace by the probability of
+        drawing the value from the distribution.
+\end{itemize}
+However, the actual computation involved in handling of
+\texttt{sample} and \texttt{observe} depends on the inference
+algorithm. Different inference algorithms may treat
 \texttt{sample} and \texttt{observe} differently as long as they
-compute the same distribution on traces (i.e. sequences of sampled values), 
-namely, the one in Equation~\ref{eqn:p-trace}.
-Formal semantics for an idealized version
-of Anglican or other higher-order probabilistic programming languages are introduced 
-in \cite{SYH+16,BorgstromLGS16}; in general though,
+compute the same distribution of program results.
+Denotational and operational semantics for an idealized version
+of Anglican is introduced in \cite{SYH+16}; in general though,
 specifying semantics of higher-order probabilistic programming
 languages is an open problem.
-
-%Their operational semantics can be defined in terms of
-%importance sampling(Equation~\ref{eqn:p-trace}): 
-%\begin{itemize}
-%    \item \texttt{sample} returns a value drawn from
-%        the distribution.
-%    \item \texttt{observe} multiplies the unnormalized
-%        probability of the program trace by the probability of
-%        drawing the value from the distribution.
-%\end{itemize}
-%However, the actual computation involved in handling of
-%\texttt{sample} and \texttt{observe} depends on the inference
-%algorithm. Different inference algorithms may treat
-%\texttt{sample} and \texttt{observe} differently as long as they
-%compute the same distribution of program results.
-%Denotational and operational semantics for an idealized version
-%of Anglican is introduced in \cite{SYH+16}; in general though,
-%specifying semantics of higher-order probabilistic programming
-%languages is an open problem.
 
@@
     "initial program state"
-  {:log-weight 0.0, ;; map :log-weight to 0.0
-   :result nil,       ;; map :result to nil
-   ::store nil,       ;; map :store to nil
-   ::mem {},          ;; map ::mem to the empty map 
+    {:log-weight 0.0,
+     :result nil,
+     ::mem {},
+     ::store nil,
      ... })
@@
 is achieved by  returning a \textbf{checkpoint} instead of
-invoking a continuation in a CPS-transformed program.  
-Checkpoints are values of the
+invoking a continuation in a CPS-transformed program.  In
+Clojure, checkpoints are represented by values of the
 corresponding record types
 (\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
-\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
-in Clojure. The types contain fields specific to each form, as well as the
+\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe}).
+The types contain fields specific to each form, as well as the
 continuation; calling the continuation resumes the computation.
@@
 Memoization is often implemented on top of a mutable
-dictionary, where the key is the argument  and the value is
+dictionary, where the key is the argument list and the value is
 the returned value. However, there are no mutable data
@@
 identifier) was previously called in the same run of the
-probabilistic program with the same argument. If the memoized
+probabilistic program with the same arguments. If the memoized
 result is available, the continuation of the memoized function
@@
 To allow the mixing of Anglican (CPS-transformed) and Clojure
-function calls in Anglican code, the Anglican compiler must CPS-transform
-expressions selectively~\cite{Nielsen01}. It must be
+function calls in Anglican code, the Anglican compiler must be
 able to recognize `primitive' (that is, implemented in Clojure
-rather than in Anglican) functions, and invoke
-those functions in a direct, not CPS, style.
+rather than in Anglican) functions.
 

### 2017-02-24 paper/paper.tex
comment
update
clarification

@@
 
-\hy{Need to be revised.}
-
 Programs in Anglican define probability distributions 
@@
 
-We call the triple part of a log \textit{trace}. This part records all the random 
+We call the triple part of a log \textit{trace}, and denote this sequence of $(F,x,\alpha)$'s 
+by $\pmb{x}$. This part records all the random 
 choices of an execution. Thus, if the trace of a log is fixed, the rest of the log 
-is decided determistically by the program. A trace induces
+is decided determistically by the program. A trace $\pmb{x}$ induces
 a sequence of pairs $(G_j, y_j)$ of distributions and values of
-\textit{observed} random variables. We call this sequence an \textit{image} and
-denote it by $\pmb{y}$. For notational simplicity we assume that the program
-always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
+\textit{observed} random variables. We call this sequence \textit{image of $\pmb{x}$} and
+denote it by $\pmb{y}$. A subset of random choices in $\pmb{x}$ forms the result of the program.
+We denote this result (i.e. the subsequence of $\pmb{x}$ with these choices) by $\pmb{z}$. 
+%For notational simplicity we assume that the program
+%always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
+%
+%\hy{What is $\pmb{z}$? It is not explained here. I found it easier to think
+%that probabilistic programs define distributions over traces. Also, $\vec{x}$ is not defined.}
 
@@
 is defined in terms of the product of the probabilities of all random
-choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
+choices $\pmb{x}$ and the likelihood of all observations in the image $\pmb{y}$ 
+of $\pmb{x}$:
 \begin{align}
@@
 \end{align}
-The objective of inference in probabilistic program $\mathcal{P}$
-is to characterize the distribution of $\pmb{z}$.
+This density induces the distribution of the results $\pmb{z}$ of the program
+by marginalization. The objective of probabilistic inference 
+is to characterize this distribution of $\pmb{z}$.
 

### 2017-02-25 paper/paper.tex
update
Parially reverted 0f995f
Edits in 0f995f were made relative to an out-of-date version of
the text (maybe due to the text editor not reloading after a pull?).

This commit keeps the intended changes from 0f995f, but reverts
the accidental reversions.

--- 2017-02-22 paper/paper.tex
@@
 U.S. AFRL under Cooperative Agreement number FA8750- 14-2-0006,
-Sub Award number 61160290-111668.
+Sub Award number 61160290-111668. Yang was supported by Institute for 
+Information \& communications Technology Promotion (IITP) grant funded 
+by the Korea government (MSIP) (No.R0190-16-2011, Development of 
+Vulnerability Discovery Technologies for IoT Software Security).
 

### 2017-02-25 paper/paper.tex
wip commit

@@
 \newtoggle{full}
-\input{toggle}
+\input{toggle-ifl }
 
@@
 models making the inference inherently faster.
+\jw{I'm not sure I understand how PP facilitates simpler implmenetations that make inference inherently faster.}
 
@@
       (when (not-empty observations)
-        ;; Retrieve the first observation as `o',
-        ;; and store the rest of observations in 
-        ;; 'observations*'.
-        (let [[o & observations*] observations]
+        ;; Retrieve the first observation as `o'
+        (let [o (first observations)]
           ;; Observe 'o' from the guessed 
@@
         ;; the rest of observations.
-        (recur observations*)))
+        (recur (rest observations))))
 
@@
 \end{lstlisting}
+
 %\end{minipage}
@@
 $\texttt{data}$ using the \texttt{observe} form.
-Samples from this adapted distribution (also called posterior distribution)
+Samples from this conditioned distribution (also called posterior distribution)
 can be obtained by running the query under one of Anglican's
@@
 
-Internally, an Anglican query is represented by a computation in
-\textit{continuation passing style} (CPS)~\cite{AJ89}, and inference algorithms
-exploit the CPS structure of the code to intercept probabilistic
-operations in an algorithm-specific way\footnote{\cite{GS15}
-also describe a CPS-based implementation of a probabilistic
-programming language.}. Among the available
-inference algorithms there are Particle Cascade~\cite{PWD+14},
-Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
-Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
-Inference on Anglican queries generates a lazy sequence of
-samples, which can be processed asynchronously in Clojure
-code for analysis, integration, and decision making.
-
-Clojure (and Anglican) runs on the JVM and gets access to a wide choice of Java
-libraries for data processing, networking, presentation, and imaging.
-Conversely, Anglican queries can be called from Java and other JVM languages.
-Programs involving Anglican queries can be deployed as JVM
-\textit{jars}, and run without modification on any platform for
-which JVM is available.
-
-A probabilistic program, or query, mostly runs deterministic
-code, except for certain checkpoints, in which probabilities are
-involved, and normal, linear execution of the program is
-disrupted.
-\iftoggle{full}{%
-In Anglican and similar languages there are two types
-of such checkpoints:
-
-\begin{itemize}
-    \item drawing a value from a random source (\texttt{sample});
-    \item conditioning a computed value on a random source
-        (\texttt{observe}).
-\end{itemize}
-
-}{%
-}%
-Anglican can be mostly implemented as a regular programming
-language, except for the handling of these checkpoints.
+Clojure (and Anglican) programs run on the JVM and are able to make
+use of a wide range of Java libraries for data processing,
+networking, presentation, and imaging. Conversely, Anglican queries
+can be called from Java and other JVM languages. Programs involving
+Anglican queries can be deployed as JVM \textit{jars}, and run without modification on any platform for which the JVM is available.
+
+A probabilistic program, or query, mostly runs deterministic code,
+except for \texttt{sample} and \texttt{observe} forms, which are
+probabilistic in nature.
+For this reason, Anglican can be mostly implemented as a regular programming language, except for the handling of these two special forms, which we refer to as checkpoints.
+Internally, an Anglican query is represented by a
+computation in \textit{continuation passing style} (CPS)~\cite{AJ89}.
+Inference algorithms exploit the CPS structure of the code to
+evaluate \texttt{sample} and \texttt{observe} checkpoints in an
+algorithm-specific manner\footnote{\cite{GS15} also describe a
+CPS-based implementation of a probabilistic programming language.}.
+The available inference algorithms include the Particle
+Cascade~\cite{PWD+14}, Lightweight Metropolis-Hastings~\cite{WSG11},
+Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
+\cite{WVM14}, and others. Inference on Anglican queries generates a
+lazy sequence of samples, which can be processed asynchronously in
+Clojure for analysis, integration, and decision making.
+
+% \iftoggle{full}{%
+% In Anglican and similar languages there are two types
+% of such checkpoints:
+
+% \begin{itemize}
+%     \item drawing a value from a random source (\texttt{sample});
+%     \item conditioning a computed value on a random source
+%         (\texttt{observe}).
+% \end{itemize}
+
+% }{%
+% }%
 \iftoggle{full}{%

### 2017-02-25 paper/paper.tex
edits to section 3

@@
 except for \texttt{sample} and \texttt{observe} forms, which are
-probabilistic in nature.
-For this reason, Anglican can be mostly implemented as a regular programming language, except for the handling of these two special forms, which we refer to as checkpoints.
-Internally, an Anglican query is represented by a
-computation in \textit{continuation passing style} (CPS)~\cite{AJ89}.
-Inference algorithms exploit the CPS structure of the code to
-evaluate \texttt{sample} and \texttt{observe} checkpoints in an
-algorithm-specific manner\footnote{\cite{GS15} also describe a
+probabilistic in nature. For this reason, Anglican can be implemented
+as a regular programming language, except for the handling of these
+two special forms, which we refer to as checkpoints. At these
+checkpoints, Anglican programs must allow the inference algorithm to
+step in, recording information and affecting control flow. This can
+be implemented through coroutines/cooperative multitasking, and
+parallel execution/preemptive multitasking, as well as through
+explicit maintenance of program continuations at checkpoints.
+Anglican follows the latter option.
+
+Internally, an Anglican query is represented by a computation in
+\textit{continuation passing style} (CPS) \cite{AJ89}. The Anglican 
+`compiler'\iftoggle{full}{, represented by a set of functions in the
+\texttt{anglican.trap} namespace,}{} accepts a Clojure subset and 
+transforms it into a variant of CPS representation, which allows 
+inference algorithms to intervene in the execution flow at 
+probabilistic checkpoints \footnote{\cite{GS15} also describe a
 CPS-based implementation of a probabilistic programming language.}.
-The available inference algorithms include the Particle
-Cascade~\cite{PWD+14}, Lightweight Metropolis-Hastings~\cite{WSG11},
-Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
-\cite{WVM14}, and others. Inference on Anglican queries generates a
-lazy sequence of samples, which can be processed asynchronously in
-Clojure for analysis, integration, and decision making.
+The available inference algorithms include the Particle Cascade
+\cite{PWD+14}, Lightweight Metropolis-Hastings \cite{WSG11}, Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
+\cite{WVM14}, and others. Inference on Anglican queries generates a lazy sequence of samples, which can be processed asynchronously in Clojure for analysis, integration, and decision making.
+
+% Inference algorithms exploit the CPS structure of the code to
+% evaluate \texttt{sample} and \texttt{observe} checkpoints in an algorithm-specific manner
 
@@
 }%
-Because of the checkpoints, Anglican programs must allow the
-inference algorithm to step in, recording information and
-affecting control flow. This can be implemented through
-coroutines/cooperative multitasking, and parallel execution/preemptive
-multitasking, as well as through explicit
-maintenance of program continuations at checkpoints. Anglican
-follows the latter option.
+% Because of the checkpoints, Anglican programs must allow the
+% inference algorithm to step in, recording information and
+% affecting control flow. This can be implemented through
+% coroutines/cooperative multitasking, and parallel execution/preemptive
+% multitasking, as well as through explicit
+% maintenance of program continuations at checkpoints. Anglican
+% follows the latter option.
 \iftoggle{full}{%
@@
 }%
-The Anglican `compiler'\iftoggle{full}{, represented by a set of
-functions in the \texttt{anglican.trap} namespace,}{} accepts a
-Clojure subset and transforms it into a variant of CPS
-representation, which allows inference algorithms to intervene
-in the execution flow at probabilistic checkpoints.
 
@@
 \texttt{query}). In addition to defining entire queries,
-Anglican promotes modularization of inference algorithms through
+Anglican promotes modularization of probabilistic models through
 the definitions of \textit{probabilistic functions} using

### 2017-02-25 paper/paper.tex
edits to section 4

@@
 \usepackage{graphicx}
+%\usepackage{microtype}
 
@@
 
-\section{Probabilistic Program}
+\section{Probability of a Program Execution}
 \label{sec:pp-def}
@@
 A (almost-surely terminating) probabilistic program defines a probability distribution 
-over finite feasible traces $\pmb{x}$. The distribution has the probability density 
+over finite feasible traces $\pmb{x}$ with probability density 
 $\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
@@
 \end{align}
-and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
-of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of
+and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$. The integral and the density use the default measure obtained by the standard extension of
 a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
-a countable sum of Lebesque and counting measures.}
+a countable sum of Lebesque and counting measures.
 

### 2017-02-25 paper/paper.tex
backend -> back end

@@
 performed for any probabilistic program using one or more generic
-inference techniques provided by the system backend, such as
+inference techniques provided by the system back end, such as
 Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
@@
 models making the inference inherently faster.
-\jw{I'm not sure I understand how PP facilitates simpler implmenetations that make inference inherently faster.}
+\jw{I'm not sure I understand how PP facilitates simpler implementations that make inference inherently faster.}
 

### 2017-02-25 paper/paper.tex
edits to section 5

@@
 \end{lstlisting}
-In \texttt{defquery} parameter, \texttt{let} bindings, and
+In \texttt{defquery} parameter lists, \texttt{let} bindings, and
 \texttt{fn} argument lists, Clojure vector destructuring is
@@
 in Anglican cannot lead to stack overflow.  However,
-\texttt{loop}/\texttt{recur} is provided in Anglican for
+\texttt{loop} and \texttt{recur} are provided in Anglican for
 convenience as a way to express loops. Unlike in Clojure though,
-\texttt{recur} supported only inside a \texttt{loop}.
+\texttt{recur} is supported only inside a \texttt{loop}.
 

### 2017-02-25 paper/paper.tex
change test -> pred (to avoid syntax highlighting of primitive test)

@@
 \begin{lstlisting}[style=default]
-(if test then else?)
-(when test expr*)
+(if pred then else?)
+(when pred expr*)
 (cond clause*)

### 2017-02-25 paper/paper.tex
edits to section 6

@@
 
-\subsection{Probabilistic Forms}
+\subsection{Special Forms}
 
@@
 stores a value for a tag in the program state, which can be
-later retrieved by the tag using \texttt{retrieve} during the same
+later retrieved using \texttt{retrieve} during the same
 program execution.
@@
 
-Before delving into implementation details, we provide a case
-study of using the Anglican language and environment. This case
-study takes a problem for which a solution is not immediately
-obvious, \textit{the Deli dilemma}, and guides through writing
-an Anglican program for the problem, executing the program, and
-post-processing results.
+Before delving into implementation details, we discuss a case study
+of the use of the Anglican language and environment. This case study
+takes an inference problem for which a solution is not immediately
+obvious, \textit{the Deli dilemma}, and discusses how this problem
+can be solved by writing an Anglican program, executing the program,
+and post-processing results.
 
 The program presented in this section is intentionally short and
-simple. Anglican is capable of compiling and running elaborated
-programs handling large amounts of data. Advanced examples of
+simple. Anglican is capable of compiling and running elaborate
+programs and handling large amounts of data. Advanced examples of
 Anglican programs and inference can be found in literature on
@@
     \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
-    \item The waiter's odds that this is the same customer are 2 to 1.
+    \item Based on the similarity of appearance, the waiter assesses the odds that this is the same customer as 2 to 1.
 \end{itemize}
@@
 
-We want to formalize the dilemma in an Anglican query, based on
+We want to formalize the dilemma as an Anglican query, based on
 the knowledge we have. Let us formalize the knowledge in
@@
 \begin{lstlisting}[style=default]
-(def mean-time-to-arrive
+(def mean-office-distance
      "average time to arrive"
@@
 
-Performing inference on query \texttt{deli} computes the
+Performing inference on the query \texttt{deli} computes the
 probability that the same customer visited the deli twice, as

### 2017-02-25 paper/paper.tex
reverted changed variable name in case study, wip edits in sec 7

@@
 \begin{lstlisting}[style=default]
-(def mean-office-distance
+(def mean-time-to-arrive
      "average time to arrive"
@@
 
-Compilation of Anglican into Clojure is built around a variant of CPS
+Anglican queries are compiled into Clojure using a variant of CPS
 transformation. In a basic CPS-transformed program, a
 continuation receives a single argument --- the computed value.
-In Anglican, there are two flows of computation going in
+In Anglican, two flows of computation are performed in
 parallel: values are computed by functional code, and,

### 2017-02-25 paper/paper.tex
edits to section 7

@@
    :result nil,       ;; map :result to nil
-   ::store nil,       ;; map :store to nil
+   ::store nil,       ;; map ::store to nil
    ::mem {},          ;; map ::mem to the empty map 
@@
 \end{lstlisting}
-which records inference-relevant information under various keys
-such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
-The full list of map entries depends on the inference
-algorithm. Except for transformation of
-the \texttt{mem} form (which converts a function to one with memoization), 
-CPS transformation routines
-are not aware of contents of \texttt{\$state}, do not access or
-modify it directly, but rather just thread the state unmodified
-through the computation. Algorithm-specific handlers of
-checkpoints corresponding to the probabilistic forms
-(\texttt{sample}, \texttt{observe}) modify the
-state and reinject a new state into the computation.
+which records inference-relevant information under various keys such
+as $\texttt{:log-weight}$ and $\texttt{::mem}$. The full list of map
+entries depends on the inference algorithm. CPS transformation
+routines are not aware of the contents of \texttt{\$state} and do not
+access or modify it directly. Rather, they just thread the state
+unmodified through the computation. The sole exception is the
+transformation of  the \texttt{mem} form (which converts a function
+to one with memoization). Algorithm-specific handlers of checkpoints
+corresponding to the probabilistic forms (\texttt{sample},
+\texttt{observe}) modify the state and insert an updated state into
+the computation.
 

### 2017-02-25 paper/paper.tex
small edits to section 8

@@
 a probabilistic program computes, exactly or approximately, the
-distribution from which execution results are drawn.  Computing
-the distribution is facilitated by an inference algorithm.
+distribution from which execution results are drawn.  Characterising
+this distribution is done by the inference algorithm.
 
@@
 
-Consider the following simplified Anglican query:
+\noindent Consider the following simplified Anglican query:
 
@@
 \end{lstlisting}
-
 An execution of the query may result in the following
@@
 
-This way, each occurrence of a checkpoint has unique address,
+This way, each occurrence of a checkpoint has a unique address,
 but small disturbances --- removal or addition of a single or just

### 2017-02-25 paper/paper.tex
edits to section 9

@@
 
-Macro \texttt{defdist} should be used to define distributions.
+The macro \texttt{defdist} is used to define distributions.
 \texttt{defdist} takes care of defining a separate type
@@
 \begin{lstlisting}[style=default]
-(loop [process (beta-bernoulli 1.0 1.0)
-       [value0 & values0] values] 
-  (if (= (count values0) 0)
-    nil
-    (let [dist (produce process)]
-      (observe dist value0)
-      (recur (absorb process value0) values0))))
+(defm observe-beta-binomial [values a b]
+  (loop [process (beta-bernoulli a b)
+         values values] 
+    (when (not-empty values)
+      (let [dist (produce process)
+            value (first values)]
+        (observe dist value)
+        (recur (absorb process value)
+               (rest values))))))
 \end{lstlisting}
@@
 Anglican introduces, this is a good result.
-\begin{table}
+\begin{table}[!b]
     \centering

### 2017-02-25 paper/paper.tex
observe-bete-binomial -> observe-beta-bernoulli

@@
 \begin{lstlisting}[style=default]
-(defm observe-beta-binomial [values a b]
+(defm observe-beta-bernoulli [values a b]
   (loop [process (beta-bernoulli a b)

### 2017-02-25 paper/paper.tex
fixed some typos

@@
 interaction between the program and the inference engine
-necessiates implementing Anglican as a language rather than a
+necessitates implementing Anglican as a language rather than a
 library.
@@
     Anglican and WebPPL, for $100\,000$ iterations of Markov Chain
-    Monte Carlo (Lighweight Metropolis-Hastings), averaged over
+    Monte Carlo (Lightweight Metropolis-Hastings), averaged over
     100 runs. Anglican outperforms WebPPL on all models.}

### 2017-02-26 paper/paper.tex
spellchecked with GB dictionary

@@
     computational overhead. We also demonstrate how advanced
-    probabilistic modeling concepts are mapped naturally
+    probabilistic modelling concepts are mapped naturally
     to the functional foundation.
@@
 a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
-a countable sum of Lebesque and counting measures.
+a countable sum of Lebesgue and counting measures.
 

### 2017-02-26 paper/paper.tex
made toggle parameterized again, it is 'make ifl or make full'

@@
 \newtoggle{full}
-\input{toggle-ifl }
+\input{toggle}
 
@@
 
-%An implementation of Anglican must therefore address three issues:
-%\begin{itemize}
-%    \item the Clojure syntax to introduce probabilistic Anglican
-%        code inside Clojure modules;
-%    \item source-to-source transformation of Anglican programs
-%        into Clojure, so that probabilistic execution becomes
-%        possible;
-%    \item algorithms which run Clojure code, obtained by
-%        transforming Anglican programs, according to the
-%        probabilistic operational semantics.
-%\end{itemize}
-%Execution of probabilistic programs by inference algorithms is
-%different from execution of deterministic programs.  A
-%probabilistic program is executed multiple times, often hundreds
-%of thousands or even millions of times for a single inference
-%task. Random choices may affect which parts of the program code
-%are executed and how often. Many inference algorithms require
-%re-running the program multiple times partially, from a certain
-%point on.  Different executions may employ different random
-%choices. However, for efficient inference a correspondence
-%between random choices in different executions should be
-%maintained. These are just some of the challenges which were
-%faced and solved during development of Anglican.
+\iftoggle{full}{%
+An implementation of Anglican must therefore address three issues:
+\begin{itemize}
+    \item the Clojure syntax to introduce probabilistic Anglican
+        code inside Clojure modules;
+    \item source-to-source transformation of Anglican programs
+        into Clojure, so that probabilistic execution becomes
+        possible;
+    \item algorithms which run Clojure code, obtained by
+        transforming Anglican programs, according to the
+        probabilistic operational semantics.
+\end{itemize}
+Execution of probabilistic programs by inference algorithms is
+different from execution of deterministic programs.  A
+probabilistic program is executed multiple times, often hundreds
+of thousands or even millions of times for a single inference
+task. Random choices may affect which parts of the program code
+are executed and how often. Many inference algorithms require
+re-running the program multiple times partially, from a certain
+point on.  Different executions may employ different random
+choices. However, for efficient inference a correspondence
+between random choices in different executions should be
+maintained. These are just some of the challenges which were
+faced and solved during development of Anglican.
+}{}%
 

### 2017-02-26 paper/paper.tex
put back the best size of the figures
removed commented out and rewritten fragments to clean up latex.
new style

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[sigconf]{acmart}
+
+\setcopyright{rightsretained}
 
@@
 \usepackage[lighttt]{lmodern}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
 \usepackage{graphicx}
-%\usepackage{microtype}
 
 \usepackage{listings}
-\usepackage{lmodern}
+
 \ttfamily
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{1.0}\ttfamily, columns=flexible, showstringspaces=false}
-
-\usepackage{hyperref}
+\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
+
 \usepackage{url}
@@
 
-\title{Design and Implementation of Probabilistic Programming Language Anglican}
-\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
-
 \newtoggle{full}
@@
 
+\acmConference[IFL'16]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
+\acmYear{2016}
+\copyrightyear{2017}
+
 \begin{document}
-
-\maketitle
+\title{Design and Implementation of Probabilistic Programming Language Anglican}
+
+\author{David Tolpin}
+\affiliation{\institution{University of Oxford}}
+\email{dtolpin@robots.ox.ac.uk}
+
+\author{Jan-Willem van de Meent}
+\affiliation{\institution{University of Oxford}}
+\email{jwvdm@robots.ox.ac.uk}
+
+\author{Hongseok Yang}
+\affiliation{\institution{University of Oxford}}
+\email{hongseok.yang@cs.ox.ac.uk}
+
+\author{Frank Wood}
+\affiliation{\institution{University of Oxford}}
+\email{fwood@cs.ox.ac.uk}
 
@@
 \end{abstract}
+
+\maketitle
 
@@
             programming languages.
-%    \item Specification of probabilistic forms in Anglican which facilitates 
-%        separation of probabilistic programs from inference algorithms.
 \end{itemize}
@@
 
-% Inference algorithms exploit the CPS structure of the code to
-% evaluate \texttt{sample} and \texttt{observe} checkpoints in an algorithm-specific manner
-
-% \iftoggle{full}{%
-% In Anglican and similar languages there are two types
-% of such checkpoints:
-
-% \begin{itemize}
-%     \item drawing a value from a random source (\texttt{sample});
-%     \item conditioning a computed value on a random source
-%         (\texttt{observe}).
-% \end{itemize}
-
-% }{%
-% }%
 \iftoggle{full}{%
@@
 rejection of a sample.
-
 }{%
 }%
-% Because of the checkpoints, Anglican programs must allow the
-% inference algorithm to step in, recording information and
-% affecting control flow. This can be implemented through
-% coroutines/cooperative multitasking, and parallel execution/preemptive
-% multitasking, as well as through explicit
-% maintenance of program continuations at checkpoints. Anglican
-% follows the latter option.
+
 \iftoggle{full}{%
@@
 
-%Operational semantics of Anglican queries is different from that
-%of Clojure code, therefore queries must be called through
-%inference algorithms, rather than `directly'. 
 The operational semantics of an Anglican query are different from those
@@
 termination of a probabilistic program.
-
-%When inference
-%is performed on an Anglican query, the query is run by a
-%particular inference algorithm. Inference algorithms must
-%provide an implementation for \texttt{infer}, as well as
-%override the polymorphic function \texttt{checkpoint} (defined
-%as a multimethod) so as to handle \texttt{sample} and
-%\texttt{observe} in an algorithm-specific manner and
-%to construct an appropriate result on the
-%termination of a probabilistic program.
 
@@
 
-%\section{Probabilistic Program}
-%\label{sec:pp-def}
-%
-%Anglican is a language for expressing probabilistic programs.
-%A probabilistic program is a stateful \textit{deterministic} computation
-%$\mathcal{P}$ with the following properties~\cite{TMP+15}:
-%\begin{itemize}
-%\item Initially, $\mathcal{P}$ expects no arguments.
-%\item On every call, $\mathcal{P}$ returns either a distribution
-%  and an address $(F,\alpha)$, a distribution and a value $(G,
-%        y)$, or a value $z$.
-%\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
-%    the argument to the next call (\texttt{sample} in Anglican).
-%\item Upon returning $(G, y)$, $\mathcal{P}$ is invoked again
-%    without arguments (\texttt{observe} in Anglican).
-%\item Upon returning $z$, $\mathcal{P}$ terminates.
-%\end{itemize}
-%A program is run by calling $\mathcal{P}$ repeatedly until
-%termination.
-%
-%A program need not generate the same sequence of random variables in every
-%execution. For this reason we assume that each random variable $x$ is assigned
-%a unique label $\alpha$, which we call an address, that induces a
-%correspondence between variables in different executions. Every execution
-%implicitly produces a sequence of triples of distributions, values of
-%\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
-%call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
-%a sequence of pairs $(G_j, y_j)$ of distributions and values of
-%\textit{observed} random variables. We call this sequence an \textit{image} and
-%denote it by $\pmb{y}$. For notational simplicity we assume that the program
-%always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
-%
-%The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
-%is defined in terms of the product of the probabilities of all random
-%choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
-%\begin{align}
-%  \gamma(\pmb{x}) 
-%  &:= 
-%%  \prod_{i=1}^{\left|\pmb{x}\right|}
-%  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
-%  \label{eqn:p-trace}
-%\end{align}
-%The objective of inference in probabilistic program $\mathcal{P}$
-%is to discover the distribution of $\pmb{z}$.
-
 \section{Language}
@@
 languages is an open problem.
-
-%Their operational semantics can be defined in terms of
-%importance sampling(Equation~\ref{eqn:p-trace}): 
-%\begin{itemize}
-%    \item \texttt{sample} returns a value drawn from
-%        the distribution.
-%    \item \texttt{observe} multiplies the unnormalized
-%        probability of the program trace by the probability of
-%        drawing the value from the distribution.
-%\end{itemize}
-%However, the actual computation involved in handling of
-%\texttt{sample} and \texttt{observe} depends on the inference
-%algorithm. Different inference algorithms may treat
-%\texttt{sample} and \texttt{observe} differently as long as they
-%compute the same distribution of program results.
-%Denotational and operational semantics for an idealized version
-%of Anglican is introduced in \cite{SYH+16}; in general though,
-%specifying semantics of higher-order probabilistic programming
-%languages is an open problem.
 
@@
 This implementation is dispatched 
-when \texttt{infer} receives \texttt{:importance} as its \texttt{algo} 
+when \texttt{infer} receives {\tt :im\-por\-tan\-ce} as its \texttt{algo} 
 parameter. (Anglican includes multiple implementations of \texttt{infer}
@@
 relies on their default implementations.
-%\texttt{anglican.importance/{\linebreak[0]}infer} just calls
-%\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
-%implementations of checkpoint handlers.
 
@@
 
-\acks
+\section*{Acknowledgements}
 
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{refs}

### 2017-02-26 paper/case-study-mean-std.tex
squeezed the paper into the new format

--- 2017-02-26 paper/paper.tex
@@
 \usepackage{graphicx}
-
 \usepackage{listings}
+\usepackage{url}
+\usepackage{etoolbox}
+
+\newtoggle{full}
+\input{toggle}
 
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
-
-\usepackage{url}
-\usepackage{etoolbox}
+\iftoggle{full}{
+\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}
+}{
+\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.8}\ttfamily\small, columns=flexible, showstringspaces=false}
+}
 
@@
 % \cbend
-
-\newtoggle{full}
-\input{toggle}
 
@@
 \end{abstract}
+
+\begin{CCSXML}
+<ccs2012>
+    <concept>
+        <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+        <concept_desc>Software and its engineering~Functional languages</concept_desc>
+        <concept_significance>300</concept_significance>
+    </concept>
+    <concept>
+        <concept_id>10011007.10011006.10011050.10011023</concept_id>
+        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
+        <concept_significance>300</concept_significance>
+    </concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[300]{Software and its engineering~Functional languages}
+\ccsdesc[300]{Software and its engineering~Specialized application languages}
 
@@
 of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
-These generic techniques may be less statistically efficient
-than techniques tailored to a specific model. However,
-probabilistic programming facilitates simpler implementations of
-models making the inference inherently faster.
-\jw{I'm not sure I understand how PP facilitates simpler implementations that make inference inherently faster.}
 
@@
 programming language that tightly integrates with
-Clojure~\cite{H08,Clojure}, a general-purpose programming language
+Clojure~\cite{H08}, a general-purpose programming language
 that runs on the Java Virtual Machine (JVM). Both languages share a common
@@
         d (dist a b)]
-
     ;;; Observe samples from the distribution.
@@
         (recur (rest observations))))
-
     ;;; Return the distribution and parameters.
@@
 Imagine that we are facing the following dilemma:
-\begin{quote}
+
 A customer wearing round sunglasses came to a deli at 1:13pm,
@@
 dinner.  Was it the same customer?
-\end{quote}
+
 Additionally, we know that:
@@
 \begin{lstlisting}[style=default]
-(def mean-time-to-arrive
-     "average time to arrive"
-     10.)
+(def mean-time-to-arrive "average time to arrive" 10.)
 (def sd-time-to-arrive
-     "standard deviation of arrival time"
-     3.)
-(def time-sd
-     "walking time deviation"
-     1.)
+     "standard deviation of arrival time" 3.)
+(def time-sd "walking time deviation" 1.)
 \end{lstlisting}
@@
 (def lunch-delay
-     "time between lunch break and lunch order"
-     13.)
+     "time between lunch break and lunch order" 13.)
 (def dinner-delay
-     "time between end of day and dinner order"
-     9.)
+     "time between end of day and dinner order" 9.)
 (def p-same
-     "prior probability of the same customer"
-     (/ 2. 3.))
+     "prior probability of the same customer" (/ 2. 3.))
 \end{lstlisting}
@@
      "prior distribution of average arrival time"
-     (normal mean-time-to-arrive
-             sd-time-to-arrive))
+     (normal mean-time-to-arrive sd-time-to-arrive))
 \end{lstlisting}
@@
   "observe the same customer twice"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive
-          (sample time-to-arrive-prior)]
+  [time-to-arrive-prior lunch-delay dinner-delay]
+  (let [time-to-arrive (sample time-to-arrive-prior)]
     (observe (normal time-to-arrive time-sd)
@@
   "observe different customers"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive-1
-          (sample time-to-arrive-prior)
-        time-to-arrive-2
-          (sample time-to-arrive-prior)]
+  [time-to-arrive-prior lunch-delay dinner-delay]
+  (let [time-to-arrive-1 (sample time-to-arrive-prior)
+        time-to-arrive-2 (sample time-to-arrive-prior)]
     (observe (normal time-to-arrive-1 time-sd)
@@
 (defquery deli [time-to-arrive-prior
-                lunch-delay
-                dinner-delay]
+                  lunch-delay dinner-delay]
   (let [is-same-customer (sample (flip p-same))
@@
                           time-to-arrive-prior
-                        lunch-delay
-                        dinner-delay)}))
+                          lunch-delay dinner-delay)}))
 \end{lstlisting}
@@
 (def N 5000)
-(def results (map get-result
-                   (take N (drop N samples))))
+(def results (map get-result (take N (drop N samples))))
 \end{lstlisting}
@@
 \begin{lstlisting}[style=default]
-(def p-same+ 
-     (/ (count (filter :same-customer results))
+(def p-same+ (/ (count (filter :same-customer results))
                   (double N)))
@@
 customer visiting twice (Figure~\ref{fig:dist-same}) and
-two different customers (Figure~\ref{fig:dist-diff})
+two different customers (Figure~\ref{fig:dist-diff}).
 
@@
   (map :times-to-arrive 
-          (filter  
-            (fn [x] (not (:same-customer x)))
-            results)))
-(def mean-1-to-arrive+
-     (mean (map first times-to-arrive+)))
-(def sd-1-to-arrive+
-     (std (map first times-to-arrive+)))
-(def mean-2-to-arrive+
-     (mean (map second times-to-arrive+)))
-(def sd-2-to-arrive+
-     (std (map second times-to-arrive+)))
+    (filter (fn [x] (not (:same-customer x))) results)))
+(def mean-1-to-arrive+ (mean (map first times-to-arrive+)))
+(def sd-1-to-arrive+ (std (map first times-to-arrive+)))
+(def mean-2-to-arrive+ (mean (map second times-to-arrive+)))
+(def sd-2-to-arrive+ (std (map second times-to-arrive+)))
 \end{lstlisting}
@@
 (let [eye-color (mem (fn [person]
-                       (sample (categorical
-                                 ['brown 0.5]
+                       (sample (categorical ['brown 0.5]
                                               ['green 0.5]))))]
@@
 \begin{lstlisting}[style=default]
-(fn [cont $state x y]
-  (fn []
-    (cont (+ x y) $state)))
+(fn [cont $state x y] (fn [] (cont (+ x y) $state)))
 \end{lstlisting}
@@
     ;; recursive function without any parameter
-    [sample-seq 
-      []
+    [sample-seq []
       ;; lazy infinite sequence 
-      (lazy-seq
-        (cons
+      (lazy-seq (cons
                    (:state (exec ::algorithm prog value
                                    initial-state))
-          (sample-seq))))]
+                   (sample-seq)))]
     (sample-seq)))
@@
 \begin{lstlisting}[style=default]
-(let [dist (if use-gamma
-               (gamma 2. 2.)
-               (normal 0. 1.))]
+(let [dist (if use-gamma (gamma 2. 2.) (normal 0. 1.))]
   (sample dist))
@@
 (defm foo []
-  (if (sample 'C1 (flip 0.5))
-    (foo)
-    (bar)))
+  (if (sample 'C1 (flip 0.5)) (foo) (bar)))
 
@@
 
-(defquery baz
-  (foo))
+(defquery baz (foo))
 \end{lstlisting}
@@
 \begin{lstlisting}[style=default]
-(defdist bernoulli
-  "Bernoulli distribution"
-  [p]
+(defdist bernoulli "Bernoulli distribution" [p]
   (sample* [this] (if (< (rand) p) 1 0))
@@
 \begin{lstlisting}[style=default]
-(defproc beta-bernoulli
-  "beta-Bernoulli process"
-  [a b]
+(defproc beta-bernoulli "beta-Bernoulli process" [a b]
   (produce [this] (bernoulli (/ a (+ a b))))

### 2017-02-27 paper/paper.tex
last changes forgot to commit

@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
-
-\usepackage{url}
-\usepackage{etoolbox}
+\iftoggle{full}{
+\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}
+}{
+\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.85}\ttfamily\small, columns=flexible, showstringspaces=false}
+}
 
@@
 lazy sequence of samples. We use here the inference algorithm
-called Lightweight Metropolis-Hastings (LMH). The algorithm is
+called Lightweight Metropolis-Hastings (LMH). \iftoggle{full}{The algorithm is
 somewhat slow to converge but can be used with any Anglican
-query, and should be robust enough for our simple problem.
-
+query, and should be robust enough for our simple problem.}{}
 We bind the results of \texttt{doquery} to variable
@@
 To approximate the inferred distribution, we extract a finite
-subsequence of samples. Many algorithms use an initial
+subsequence of samples. \iftoggle{full}{Many algorithms use an initial
 subsequence of samples to converge to the target distribution.
 Hence, we drop initial $N$ samples ($N$ is 5000 in the code
-snippet below), and collect the next $N$ samples.
+snippet below), and collect the next $N$ samples.}{}
 
@@
 customer visiting twice (Figure~\ref{fig:dist-same}) and
-two different customers (Figure~\ref{fig:dist-diff})
-
-\begin{lstlisting}[style=default]
-;; single customer                      
-(def time-to-arrive+ 
-    (map (fn [x] (first (:times-to-arrive x)))
-         (filter :same-customer results)))
-(def mean-to-arrive+ (mean time-to-arrive+))
-(def sd-to-arrive+ (std time-to-arrive+))
-
-;; two customers
-(def times-to-arrive+ 
-     (map :times-to-arrive 
-          (filter  
-            (fn [x] (not (:same-customer x)))
-            results)))
-(def mean-1-to-arrive+
-     (mean (map first times-to-arrive+)))
-(def sd-1-to-arrive+
-     (std (map first times-to-arrive+)))
-(def mean-2-to-arrive+
-     (mean (map second times-to-arrive+)))
-(def sd-2-to-arrive+
-     (std (map second times-to-arrive+)))
-\end{lstlisting}
-
-In addition to plotting the distribution histograms, we use
-functions \texttt{mean}, \texttt{std} provided
-in the \texttt{anglican.{\linebreak[0]}stat} namespace
-along with other useful statistical functions to compute
-summary statistics. This is another illustration of advantage of
-tight integration between Clojure and Anglican --- probabilistic
-models are expressed in Anglican. However, processing of data
-and results can rely on the full power of Clojure.
+two different customers (Figure~\ref{fig:dist-diff}).
+
+\iftoggle{full}{%
+\input{case-study-mean-std}
+}{}%
 

### 2017-02-27 paper/paper.tex
merged final in old style

--- 2017-02-26 paper/paper.tex
@@
-\documentclass[sigconf]{acmart}
-
-\setcopyright{rightsretained}
+\documentclass[preprint]{sigplanconf}
 
@@
 \usepackage[lighttt]{lmodern}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsfonts}
 \usepackage{graphicx}
+%\usepackage{microtype}
 
 \usepackage{listings}
-
+\usepackage{lmodern}
 \ttfamily
@@
      {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
-
+\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{1.0}\ttfamily, columns=flexible, showstringspaces=false}
+
+\usepackage{hyperref}
 \usepackage{url}
@@
 
+\title{Design and Implementation of Probabilistic Programming Language Anglican}
+\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}
+
 \newtoggle{full}
@@
 
-\acmConference[IFL'16]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
-\acmYear{2016}
-\copyrightyear{2017}
-
 \begin{document}
-\title{Design and Implementation of Probabilistic Programming Language Anglican}
-
-\author{David Tolpin}
-\affiliation{\institution{University of Oxford}}
-\email{dtolpin@robots.ox.ac.uk}
-
-\author{Jan-Willem van de Meent}
-\affiliation{\institution{University of Oxford}}
-\email{jwvdm@robots.ox.ac.uk}
-
-\author{Hongseok Yang}
-\affiliation{\institution{University of Oxford}}
-\email{hongseok.yang@cs.ox.ac.uk}
-
-\author{Frank Wood}
-\affiliation{\institution{University of Oxford}}
-\email{fwood@cs.ox.ac.uk}
+
+\maketitle
 
@@
 \end{abstract}
-
-\maketitle
 
@@
 
+%Their operational semantics can be defined in terms of
+%importance sampling(Equation~\ref{eqn:p-trace}): 
+%\begin{itemize}
+%    \item \texttt{sample} returns a value drawn from
+%        the distribution.
+%    \item \texttt{observe} multiplies the unnormalized
+%        probability of the program trace by the probability of
+%        drawing the value from the distribution.
+%\end{itemize}
+%However, the actual computation involved in handling of
+%\texttt{sample} and \texttt{observe} depends on the inference
+%algorithm. Different inference algorithms may treat
+%\texttt{sample} and \texttt{observe} differently as long as they
+%compute the same distribution of program results.
+%Denotational and operational semantics for an idealized version
+%of Anglican is introduced in \cite{SYH+16}; in general though,
+%specifying semantics of higher-order probabilistic programming
+%languages is an open problem.
+
 \texttt{sample} and \texttt{observe} may appear anywhere in the
@@
     \centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.6]{same-customer-time.pdf}
+    \includegraphics[trim={12pt 0 0 0},scale=0.65]{same-customer-time.pdf}
     \caption{Arrival time distribution for a single customer.}
@@
     \centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.6]{different-customers-times.pdf}
+    \includegraphics[trim={12pt 0 0 0},scale=0.65]{different-customers-times.pdf}
     \caption{Arrival time distributions for two customers.}
@@
 This implementation is dispatched 
-when \texttt{infer} receives {\tt :im\-por\-tan\-ce} as its \texttt{algo} 
+when \texttt{infer} receives \texttt{:importance} as its \texttt{algo} 
 parameter. (Anglican includes multiple implementations of \texttt{infer}
@@
 
-\section*{Acknowledgements}
+\acks
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{abbrvnat}
 \bibliography{refs}

### 2017-02-28 paper/IFL_2016_reviews_RESPONSE.txt
updated review response

@@
+We thank the reviews for in-depth reviews. For convenience, here is a
+brief description of how we addressed the reviewers' recommendations:
+
 1) Add a clear and concise statement of the contributions up front.
@@
 
-Added section 2. Related work.
+Added section 2: Related work.
 
@@
 probabilistic forms, with details on their syntax and semantics, 
-related to section 4, the definition of probabilistic program.
+related to (new) section 4: Probability of a problem execution.
 
@@
 
-Added a clarification at the end of subsection 5.3.
+Added a general clarification at the end of section 3, design outline, 
+and an additional explanation at the end of subsection 5.3.
 
@@
 
-Added addressing of checkpoints to the list of contribution in
+Added addressing of checkpoints to the list of contributions in
 the beginning of the paper. In subsection 8.3, Addressing of
@@
 between Anglican and Clojure, as well as between Anglican and
-other probabilistic programming environment, with numbers.
+other probabilistic programming environments, with quantitative evaluation 
+on a set of probabilistic models.
 

### 2017-03-05 paper/paper.tex
edits to clarify the different meanings of checkpoint

--- 2017-02-27 paper/paper.tex
@@
 
-A probabilistic program, or query, mostly runs deterministic code,
-except for \texttt{sample} and \texttt{observe} forms, which are
-probabilistic in nature. For this reason, Anglican can be implemented
-as a regular programming language, except for the handling of these
-two special forms, which we refer to as checkpoints. At these
-checkpoints, Anglican programs must allow the inference algorithm to
-step in, recording information and affecting control flow. This can
-be implemented through coroutines/cooperative multitasking, and
-parallel execution/preemptive multitasking, as well as through
-explicit maintenance of program continuations at checkpoints.
+A probabilistic program, or query, mostly runs deterministic code.
+Aside from the special forms \texttt{sample} and \texttt{observe},
+which are probabilistic in nature, Anglican can be implemented as a
+regular programming language. At \texttt{sample} and
+\texttt{observe} forms, normal deterministic execution is 
+interrupted and Anglican programs must allow the inference algorithm
+to step in, recording information and affecting control flow. We refer these points in the execution as checkpoints. Handling of 
+checkpoints can be implemented through coroutines/cooperative multitasking, and parallel execution/preemptive multitasking, as well as through explicit maintenance of program continuations.
 Anglican follows the latter option.
@@
 transfer control to the inference algorithm.  Practically, this
-is achieved by  returning a \textbf{checkpoint} instead of
-invoking a continuation in a CPS-transformed program.  
-Checkpoints are represented by values of the
-corresponding record types
-(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
-\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
-in Clojure. The types contain fields specific to each form, as well as the
-continuation; calling the continuation resumes the computation.
+is achieved by returning a record that represents the state of program execution at the checkpoint, rather than invoking a continuation. The Clojure record used to interrupt execution at 
+\texttt{sample} and \texttt{observe} forms are \texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} and \texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe} respectively.
+These records contain a unique identifier (which we will discuss in detail in section \ref{sec:addressing}), the arguments to the special form, the continuation, and the state.
 \begin{lstlisting}[style=default]
 => (cps-of-expression '(sample dist) 'cont)
-(->sample dist cont $state)
+(->sample 'S1 dist cont $state)
 => (cps-of-expression '(observe dist v) 'cont)
-(->observe dist v cont $state)
+(->observe 'O1 dist v cont $state)
 \end{lstlisting}
@@
 records, and they take values of their fields as arguments.
+Calling the continuation resumes the computation.
 

### 2017-03-05 paper/paper.tex
edits to clarify the different meanings of checkpoint
small further edits for consistency of use of terminology

@@
 
-Upon encountering a checkpoint, the inference algorithm computes
-the updated program state and the value to be passed to the
-continuation.  How the state is updated, the number of times the
-continuation is called, and the value passed to the continuation
-of \texttt{sample} depend on the inference algorithm executing
-the program.  Section~\ref{sec:inference} provides more detail
-on internals of inference algorithms and their interaction with
-probabilistic programs.
+Upon encountering a \texttt{sample} or \texttt{observe} record, the
+inference algorithm computes the updated program state and the value
+to be passed to the continuation.  How the state is updated, the
+number of times the continuation is called, and the value passed to
+the continuation of \texttt{sample} depend on the inference algorithm
+executing the program.  Section~\ref{sec:inference} provides more
+detail on internals of inference algorithms and their interaction
+with probabilistic programs.
 
@@
 The \texttt{sample*} \textit{method} returns a random sample and
-roughly corresponds to the \texttt{sample} checkpoint.
+roughly corresponds to the default implementation of the \texttt{sample} checkpoint.
 The \texttt{observe*} \textit{method} returns the log probability
-of the value and roughly corresponds to the \texttt{observe}
+of the value, which roughly corresponds to the default implementation of the \texttt{observe}
 checkpoint. The methods can be, and sometimes are called from

### 2017-03-14 paper/paper.tex
small further edits for consistency of use of terminology
JW in NEU

--- 2017-03-05 paper/paper.tex
@@
 \newtoggle{full}
-\input{toggle-ifl}
+\input{toggle}
 
@@
 \author{Jan-Willem van de Meent}
-\affiliation{\institution{University of Oxford}}
-\email{jwvdm@robots.ox.ac.uk}
+\affiliation{\institution{Northeastern University}}
+\email{j.vandemeent@northeastern.edu}
 
@@
 
-A probabilistic program, or query, mostly runs deterministic code.
-Aside from the special forms \texttt{sample} and \texttt{observe},
-which are probabilistic in nature, Anglican can be implemented as a
-regular programming language. At \texttt{sample} and
-\texttt{observe} forms, normal deterministic execution is 
-interrupted and Anglican programs must allow the inference algorithm
-to step in, recording information and affecting control flow. We refer these points in the execution as checkpoints. Handling of 
-checkpoints can be implemented through coroutines/cooperative multitasking, and parallel execution/preemptive multitasking, as well as through explicit maintenance of program continuations.
+A probabilistic program, or query, mostly runs deterministic code,
+except for \texttt{sample} and \texttt{observe} forms, which are
+probabilistic in nature. For this reason, Anglican can be implemented
+as a regular programming language, except for the handling of these
+two special forms, which we refer to as checkpoints. At these
+checkpoints, Anglican programs must allow the inference algorithm to
+step in, recording information and affecting control flow. This can
+be implemented through coroutines/cooperative multitasking, and
+parallel execution/preemptive multitasking, as well as through
+explicit maintenance of program continuations at checkpoints.
 Anglican follows the latter option.
@@
 transfer control to the inference algorithm.  Practically, this
-is achieved by returning a record that represents the state of program execution at the checkpoint, rather than invoking a continuation. The Clojure record used to interrupt execution at 
-\texttt{sample} and \texttt{observe} forms are \texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} and \texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe} respectively.
-These records contain a unique identifier (which we will discuss in detail in section \ref{sec:addressing}), the arguments to the special form, the continuation, and the state.
+is achieved by  returning a \textbf{checkpoint} instead of
+invoking a continuation in a CPS-transformed program.  
+Checkpoints are represented by values of the
+corresponding record types
+(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
+\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
+in Clojure. The types contain fields specific to each form, as well as the
+continuation; calling the continuation resumes the computation.
 \begin{lstlisting}[style=default]
 => (cps-of-expression '(sample dist) 'cont)
-(->sample 'S1 dist cont $state)
+(->sample dist cont $state)
 => (cps-of-expression '(observe dist v) 'cont)
-(->observe 'O1 dist v cont $state)
+(->observe dist v cont $state)
 \end{lstlisting}
@@
 records, and they take values of their fields as arguments.
-Calling the continuation resumes the computation.
 

### 2017-03-14 paper/paper.tex

--- 2017-03-05 paper/paper.tex
@@
 \texttt{observe} forms, normal deterministic execution is 
-interrupted and Anglican programs must allow the inference algorithm
+interrupted, and Anglican programs must allow the inference algorithm
 to step in, recording information and affecting control flow. We refer these points in the execution as checkpoints. Handling of 
@@
 \begin{lstlisting}[style=default]
-=> (cps-of-expression '(sample dist) 'cont)
+=> (cps-of-expression '(sample dist) cont)
 (->sample 'S1 dist cont $state)
-=> (cps-of-expression '(observe dist v) 'cont)
+=> (cps-of-expression '(observe dist v) cont)
 (->observe 'O1 dist v cont $state)

### 2017-03-14 paper/paper.tex
input{toggle}, use make full or make ifl to switch
squeezed into 12 pages again

@@
 }{
-\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.85}\ttfamily\small, columns=flexible, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.842}\ttfamily\small, columns=flexible, showstringspaces=false}
 }
@@
     \centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.6]{same-customer-time.pdf}
+    \includegraphics[trim={6pt 12pt 0 12pt},scale=0.6]{same-customer-time.pdf}
     \caption{Arrival time distribution for a single customer.}
@@
     \centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.6]{different-customers-times.pdf}
+    \includegraphics[trim={6pt 12pt 0 12pt},clip,scale=0.6]{different-customers-times.pdf}
     \caption{Arrival time distributions for two customers.}
@@
 ;; invoked when algo parameter is :importance
-(defmethod infer :importance 
-  [algo prog value & {}]
+(defmethod infer :importance [algo prog value & {}]
   (letfn
@@
                    1 p
-                0 (- 1. p)
-                0.))))
+                   0 (- 1. p)))))
 \end{lstlisting}

### 2017-03-14 paper/case-study-mean-std.tex
better placed figures

--- 2017-02-26 paper/paper.tex
@@
-\documentclass[sigconf]{acmart}
-
-\setcopyright{rightsretained}
-
-\usepackage{siunitx}
-\usepackage[color]{changebar}
-\usepackage{enumitem}
-\usepackage[lighttt]{lmodern}
-\usepackage{graphicx}
-
-\usepackage{listings}
-
-\ttfamily
-\DeclareFontShape{OT1}{lmtt}{m}{it}
-     {<->sub*lmtt/m/sl}{}
-\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}
-
-\usepackage{url}
-\usepackage{etoolbox}
-
-\input{anglican-color}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-% Authors: comment using these commands
-
-\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:}
-        {\rm #1}\end{minipage}}}
-\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:}
-        {\rm #1}\end{minipage}}}
-
-\cbcolor{red}
-% Authors: use changebars to mark changed regions
-% \cbstart
-% ...
-% \cbend
-
-\newtoggle{full}
-\input{toggle}
-
-\acmConference[IFL'16]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
-\acmYear{2016}
-\copyrightyear{2017}
-
-\begin{document}
-\title{Design and Implementation of Probabilistic Programming Language Anglican}
-
-\author{David Tolpin}
-\affiliation{\institution{University of Oxford}}
-\email{dtolpin@robots.ox.ac.uk}
-
-\author{Jan-Willem van de Meent}
-\affiliation{\institution{University of Oxford}}
-\email{jwvdm@robots.ox.ac.uk}
-
-\author{Hongseok Yang}
-\affiliation{\institution{University of Oxford}}
-\email{hongseok.yang@cs.ox.ac.uk}
-
-\author{Frank Wood}
-\affiliation{\institution{University of Oxford}}
-\email{fwood@cs.ox.ac.uk}
-
-\begin{abstract}
-    Anglican is a probabilistic programming system designed to interoperate
-    with Clojure and other JVM languages. We introduce the
-    programming language Anglican, outline our design choices,
-    and discuss in depth the implementation of the Anglican language
-    and runtime, including macro-based compilation, extended
-    CPS-based evaluation model, and functional representations for
-    probabilistic paradigms, such as a distribution,
-    a random process, and an inference algorithm.
-
-    We show that a probabilistic functional language can be
-    implemented efficiently and integrated tightly with a
-    conventional functional language with only moderate
-    computational overhead. We also demonstrate how advanced
-    probabilistic modelling concepts are mapped naturally
-    to the functional foundation.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}
-\label{sec:intro}
-
-For data science practitioners, statistical inference is typically
-just one step in a more elaborate analysis workflow. The first stage
-of this work involves data acquisition, pre-processing and cleaning.
-This is often followed by several iterations of exploratory model
-design and testing of inference algorithms. Once a sufficiently
-robust statistical model and a corresponding inference algorithm have
-been identified, analysis results must be post-processed, visualized,
-and in some cases integrated into a wider production system.
-
-Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
-represent generative models as programs written in a specialized
-language that provides syntax for the definition and conditioning of
-random variables. The code for such models is generally concise,
-modular, and easy to modify or extend. Typically inference can be
-performed for any probabilistic program using one or more generic
-inference techniques provided by the system back end, such as
-Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
-\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
-of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
-These generic techniques may be less statistically efficient
-than techniques tailored to a specific model. However,
-probabilistic programming facilitates simpler implementations of
-models making the inference inherently faster.
-\jw{I'm not sure I understand how PP facilitates simpler implementations that make inference inherently faster.}
-
-While probabilistic programming systems shorten the iteration cycle in
-exploratory model design, they typically lack basic functionality needed for
-data I/O, pre-processing, and analysis and visualization of inference results.
-In this paper, we describe the implementation of
-Anglican~\cite{TMW15, Anglican}, a probabilistic
-programming language that tightly integrates with
-Clojure~\cite{H08,Clojure}, a general-purpose programming language
-that runs on the Java Virtual Machine (JVM). Both languages share a common
-syntax, and can be invoked from each other. This allows Anglican programs to
-make use of a rich set of libraries written in both Clojure and Java.
-Conversely, Anglican allows intuitive and compact specification of models for
-which inference may be performed as part of a larger Clojure project.
-
-\iftoggle{full}{%
-There are several ways to build a programming language
-on top of or besides another language.  The easiest is an
-interpreter --- a program that reads a program, in its entirety
-or line-by-line, and executes it by applying operational
-semantics of a certain kind to the language. \textsc{Basic} is
-famous for line-by-line interpreted implementations.
-
-Another approach is to write a compiler, either to a virtual
-architecture, so called p-code or byte-code, or to real
-hardware. Here, the whole program is translated from the
-`higher-level' source language to a `lower-level' object
-language, which can be directly executed, either by hardware or
-by an interpreter --- but the latter interpreter can be made
-simpler and more efficient  than an interpreter for the source
-language.
-
-On top of these two approaches are methods in which a new
-language is implemented `inside' another language of the same
-level of abstraction. Different languages provide different
-means for this; }{%
-There are several ways to build a new language on top of an
-existing one;
-}%
-Lisp is famous for the macro facility
-that allows to extend the language almost without
-restriction --- by writing \textit{macros}, one adds new
-constructs to the existing language. There are several uses of
-macros --- one is to extend the language \textit{syntax}, for
-example, by adding new control structures; another is to keep
-the existing syntax but alter the \textit{operational semantics}
---- the way programs are executed and compute their outputs.
-
-Anglican is implemented in just this way --- a macro facility
-provided by Clojure, a Lisp dialect, is used both to extend
-Clojure with constructs that delimit probabilistic code, and to
-alter the operational semantics of Clojure expressions inside
-probabilistic code fragments. Anglican claims its right to count
-as a separate language
-because of the ubiquitous probabilistic
-execution semantics rather than a different syntax,
-which is actually an advantage rather than a drawback ---
-Clojure programmers only need to know how to specify the
-boundaries of Anglican programs, but can use familiar Clojure
-syntax to write probabilistic code.
-
-Inference algorithms execute Anglican programs while trying to
-answer probabilistic queries on those programs. This execution is
-significantly different from the one described in the standard 
-operational semantics. These algorithms typically run Anglican programs 
-multiple times, often hundreds of thousands or even millions of times 
-for a single inference task. The algorithms may make random choices 
-that do not correspond to any statements in the program, and decide which parts 
-of the program code are executed and how often. Some inference algorithms 
-re-run the program multiple times partially, from a certain
-point on, while reusing random choices made in the previous runs as much as possible. A good high-level picture is that each inference algorithm specifies a 
-virtual machine that executes Anglican programs according to a 
-non-standard (usually probabilistic) operational semantics. 
-Supporting a wide range of inference algorithms and their unusual semantics 
-the main reason that we developed Anglican as a language rather 
-than as a library.
-
-\iftoggle{full}{%
-An implementation of Anglican must therefore address three issues:
-\begin{itemize}
-    \item the Clojure syntax to introduce probabilistic Anglican
-        code inside Clojure modules;
-    \item source-to-source transformation of Anglican programs
-        into Clojure, so that probabilistic execution becomes
-        possible;
-    \item algorithms which run Clojure code, obtained by
-        transforming Anglican programs, according to the
-        probabilistic operational semantics.
-\end{itemize}
-Execution of probabilistic programs by inference algorithms is
-different from execution of deterministic programs.  A
-probabilistic program is executed multiple times, often hundreds
-of thousands or even millions of times for a single inference
-task. Random choices may affect which parts of the program code
-are executed and how often. Many inference algorithms require
-re-running the program multiple times partially, from a certain
-point on.  Different executions may employ different random
-choices. However, for efficient inference a correspondence
-between random choices in different executions should be
-maintained. These are just some of the challenges which were
-faced and solved during development of Anglican.
-}{}%
-
-Comparisons of Anglican with other implementations of
-probabilistic programming languages~\cite{SGG15}\cite[pp.
-32--33]{P16} demonstrate that Anglican achieves state-of-the-art
-computational efficiency without sacrificing expressiveness.
-Anglican language syntax, compilation, invocation, and runtime
-support of Anglican queries are discussed in detail in further
-sections.
-
-\subsection*{Contributions} This paper brings the following
-major contributions:
-\begin{itemize}
-    \item Design and implementation of a probabilistic programming language
-        Anglican involving tight bilateral integration with a general-purpose
-        programming language.
-    \item Techniques for efficient and compact implementation of inference
-        algorithms, such as representation of inference results as a lazy
-        sequence of samples and a novel scheme for addressing of checkpoints.
-    \item Anglican's novel representation of random processes 
-            from statistics and machine learning, such as beta-Bernoulli
-            process. The representation is
-            \emph{stateless} and seamlessly integrates into pure functional 
-            computation. This contrasts with typical \emph{stateful}
-            implementations of these processes in other probabilistic 
-            programming languages.
-\end{itemize}
-
-\section{Related Work} \label{sec:related}
-
-Efficient implementation of expressive probabilistic programming
-languages has recently been an active area of
-research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
-SDT14,WVM14,GS15,GSX+16}. There is often a compromise between
-the expressiveness of the language and the efficiency of
-inference.  Some languages emphasize
-expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict
-the class of models which can be expressed by the
-language~\cite{MWG+10,SDT14,NN14} to facilitate application of
-efficient inference algorithms. Anglican allows unrestricted
-specification of probabilistic models in the spirit of
-Church~\cite{GMR+08}, while still supporting efficient scalable
-inference~\cite{WVM14,PWD+14,RNL+2016}.
-
-Probabilistic programming languages are implemented as
-interpreters~\cite{GMR+08,MSP14}, embedded
-languages~\cite{P09,SGG15,GSX+16}, and compiled
-languages~\cite{MMR+07,SDT14}, and also through source-to-source
-transformation with augmentation~\cite{WSG11,GS15}. Each of
-these methods emphasizes different priorities in language
-design, such as computational efficiency, integration with an
-existing development environment, ease of implementation, or
-efficiency of inference. Anglican is implemented using a
-combination of embedding and source-to-source transformational
-compilation to combine advantages of both approaches.  CPS
-transformation, employed by Anglican compiler, is also used in
-WebPPL~\cite{GS15} and facilitates clean separation between
-probabilistic programs and inference algorithms and diversity of
-applicable inference algorithms.
-
-Probabilistic programming languages are often implemented as
-pure functional languages~\cite{GMR+08,MSP14,SGG15} in
-the sense that they do not allow mutable states. This is
-because reasoning about probabilistic programs and distributions
-defined by the programs, as well as implementation of certain
-inference algorithms~\cite{MYM+15}, is easier in such pure functional
-setting. Anglican is also implemented as a language without
-mutable states.  However, there are also languages which follow
-imperative paradigm and support mutable state
-directly~\cite{SDT14,GSX+16}, or through probabilistic
-primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
-alternative to stateful random primitives in the form of random
-processes.
-
-\section{Design Outline}
-\label{sec:design}
-
-An Anglican program, or \textit{query}, is compiled into a
-Clojure function. When inference is performed with a provided
-algorithm, this produces a sequence of samples. Anglican shares
-a common syntax with Clojure; Clojure functions can be called
-from Anglican code and vice versa. A simple program in Anglican
-can look like the following code:
-%\begin{minipage}{\linewidth}
-\begin{lstlisting}[style=default]
-(defquery model "model selection" data
-  (let [;;; Guess a distribution.
-        dist (sample (categorical
-                       [[normal 0.5]
-                        [gamma 0.5]]))
-        a (sample (gamma 1 1))
-        b (sample (gamma 1 1))
-        d (dist a b)]
-
-    ;;; Observe samples from the distribution.
-    (loop [observations data]
-      (when (not-empty observations)
-        ;; Retrieve the first observation as `o'
-        (let [o (first observations)]
-          ;; Observe 'o' from the guessed 
-          ;; distribution 'd'.
-          (observe d o))
-        ;; Proceed to the next iteration with
-        ;; the rest of observations.
-        (recur (rest observations))))
-
-    ;;; Return the distribution and parameters.
-    [d a b]))
-\end{lstlisting}
-
-%\end{minipage}
-The query builds a model for the input $\texttt{data}$, 
-a sequence of data points.
-It defines a probability distribution on three variables, 
-$\texttt{d} \in \{\texttt{normal},\texttt{gamma}\}$
-for a distribution type, and $\texttt{a}$ and $\texttt{b}$ for
-positive parameters for the type. Concretely,
-using the $\texttt{sample}$ forms, the query
-first defines a so called prior distribution on these three variables,
-and then it adjusts this prior distribution based on observations in
-$\texttt{data}$ using the \texttt{observe} form.
-Samples from this conditioned distribution (also called posterior distribution)
-can be obtained by running the query under one of Anglican's
-inference algorithms.
-
-Clojure (and Anglican) programs run on the JVM and are able to make
-use of a wide range of Java libraries for data processing,
-networking, presentation, and imaging. Conversely, Anglican queries
-can be called from Java and other JVM languages. Programs involving
-Anglican queries can be deployed as JVM \textit{jars}, and run without modification on any platform for which the JVM is available.
-
-A probabilistic program, or query, mostly runs deterministic code,
-except for \texttt{sample} and \texttt{observe} forms, which are
-probabilistic in nature. For this reason, Anglican can be implemented
-as a regular programming language, except for the handling of these
-two special forms, which we refer to as checkpoints. At these
-checkpoints, Anglican programs must allow the inference algorithm to
-step in, recording information and affecting control flow. This can
-be implemented through coroutines/cooperative multitasking, and
-parallel execution/preemptive multitasking, as well as through
-explicit maintenance of program continuations at checkpoints.
-Anglican follows the latter option.
-
-Internally, an Anglican query is represented by a computation in
-\textit{continuation passing style} (CPS) \cite{AJ89}. The Anglican 
-`compiler'\iftoggle{full}{, represented by a set of functions in the
-\texttt{anglican.trap} namespace,}{} accepts a Clojure subset and 
-transforms it into a variant of CPS representation, which allows 
-inference algorithms to intervene in the execution flow at 
-probabilistic checkpoints \footnote{\cite{GS15} also describe a
-CPS-based implementation of a probabilistic programming language.}.
-The available inference algorithms include the Particle Cascade
-\cite{PWD+14}, Lightweight Metropolis-Hastings \cite{WSG11}, Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
-\cite{WVM14}, and others. Inference on Anglican queries generates a lazy sequence of samples, which can be processed asynchronously in Clojure for analysis, integration, and decision making.
-
-\iftoggle{full}{%
-Depending on the \textit{inference algorithm}, \texttt{sample}
-and \texttt{observe} may result in implicit input/output
-operations and control changes. For example, \texttt{observe} in
-particle filtering inference algorithms \cite{WVM14} is a non-deterministic
-control statement at which a particle (corresponding to
-a user-level thread executing a program)
-can be either replicated or terminated. Similarly, in 
-Metropolis-Hastings \cite{WSG11},
-\texttt{sample} is both an input
-statement which `reads' values from a random source,
-and a non-deterministic control statement
-(with delayed effect), eventually affecting acceptance or
-rejection of a sample.
-}{%
-}%
-
-\iftoggle{full}{%
-Clojure is a functional language, and continuation-passing style
-(CPS) transformation is a well-developed technique in the area
-of functional languages. Implementing a variant of CPS
-transformation seemed to be the most flexible and lightweight
-option --- any other form of concurrency would put a higher
-burden on the underlying runtime (JVM) and the operating system.
-Consequently, Anglican has been implemented as a CPS-transformed
-computation with access to continuations in probabilistic
-checkpoints. 
-}{%
-}%
-
-Anglican is intended to co-exist with Clojure and be a part of
-the source of a Clojure program. To facilitate this, Anglican
-programs, or queries, are wrapped by macros\iftoggle{full}{ (defined in the
-\texttt{anglican.emit} namespace)}{} which call the CPS
-transformations and define Clojure values suitable for passing
-as arguments to inference algorithms (\texttt{defquery},
-\texttt{query}). In addition to defining entire queries,
-Anglican promotes modularization of probabilistic models through
-the definitions of \textit{probabilistic functions} using
-\texttt{defm} (Anglican counterpart of Clojure \texttt{defn}).
-Probabilistic functions are
-written in Anglican, may include probabilistic forms
-\texttt{sample} and \texttt{observe}, and can be seamlessly
-called from inside Anglican queries, just like functions locally
-defined within the same query.
-
-The operational semantics of an Anglican query are different from those
-of Clojure code, and are determined by an inference algorithm. Thus,
-Anglican queries must be called through these inference algorithms, rather
-rather than `directly'. For this purpose,
-\iftoggle{full}{the
-\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
-function \texttt{infer} using Clojure's multimethod mechanism.
-This function accepts and runs an Anglican query, 
-and returns a lazy sequence of weighted samples from the
-distribution defined by the query. Providing an implementation of this function
-is a responsibility of the inference algorithm, which should also
-override the polymorphic function \texttt{checkpoint} (defined
-as a multimethod) so as to handle \texttt{sample} and
-\texttt{observe} in an algorithm-specific manner and
-to construct an appropriate result on the
-termination of a probabilistic program.
-
-Finally, Anglican queries use `primitive', or commonly known and
-used, distributions, to draw random samples and condition
-observations. Many primitive distributions are provided by
-\iftoggle{full}{the \texttt{anglican.runtime} namespace}{the
-runtime}, and an additional distribution can be defined by the
-user by implementing a particular set of functions for the
-distribution (via Clojure's protocol mechanism). The
-\texttt{defdist} macro provides a convenient syntax for defining
-primitive distributions.
-
-\section{Probability of a Program Execution}
-\label{sec:pp-def}
-
-Anglican programs define probability distributions 
-over sequences of values, implicitly by means of program execution.
-A good way to understand this is to imagine the following 
-interpreter of Anglican programs. Starting from a fixed initial state,
-the interpreter runs the deterministic parts of a program 
-according to the standard semantics, executes the \texttt{sample} form by
-generating a random sample, and treats the $\texttt{observe}$ form by skip.
-More importantly, the interpreter keeps a log that records
-information about all the \texttt{sample} and \texttt{observe} forms encountered 
-during execution. The information recorded for \texttt{sample} is 
-a triple $(F,x,\alpha)$ of (i) a primitive probability distribution $F$, 
-such as the standard normal, for which we have 
-the probability density $p_F$; (ii) a value $x$ sampled from the distribution $F$; 
-and (iii) an address $\alpha$ that uniquely and systematically identifies the random 
-choice made. The information recorded for \texttt{observe} is a pair $(G,y)$ where $G$
-is a primitive probability distribution as $F$ from above and $y$ is an 
-observed value. Thus, a log is a sequence of triples $(F,x,\alpha)$ and 
-pairs $(G,y)$.
-
-One important property of logs is that they are determined by 
-their projections to triples: when two logs project to the same
-sequence of triples, they must be the same. This is because
-the triples in a log contain all the information about random choices
-made during execution and all the non-sample forms in Anglican programs
-are deterministic. We define a \textit{trace} $\pmb{x}$ to be a sequence
-of triples $(F,x,\alpha)$, and say that $\pmb{x}$ is feasible if
-the trace is precisely the triple part of the log of some execution. 
-Such a feasible trace uniquely determines the rest of the execution and its log.
-In particular, it decides the pair part of the log, namely, 
-a sequence of $(G_j, y_j)$ for observed values. We call this sequence \textit{image of $\pmb{x}$} 
-and denote it by~$\pmb{y}$.
-
-A (almost-surely terminating) probabilistic program defines a probability distribution 
-over finite feasible traces $\pmb{x}$ with probability density 
-$\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
-\begin{align}
-  \gamma(\pmb{x}) 
-  &:= 
-  \prod_{i=1}^{\left|\pmb{x}\right|}
-  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})
-  \quad \mbox{for the image $\pmb{y}$ of $\pmb{x}$}, 
-  \label{eqn:p-trace}
-\end{align}
-and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$. The integral and the density use the default measure obtained by the standard extension of
-a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
-a countable sum of Lebesgue and counting measures.
-
-\section{Language}
-\label{sec:language}
-
-\subsection{Syntax}
-
-The Anglican language is a subset of Clojure\footnote{It would be
-possible to support almost full Clojure by expanding all macros
-in the Anglican source code. However, in Clojure, unlike in
-Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
-of macro-expansion of derived special forms is not well
-specified and implementation specific.}.  Anglican queries
-are defined within \texttt{defquery}. The value of the last
-expression in the query body is the result of the query.
-\begin{lstlisting}[style=default]
-(defquery name doc-string? param? expr*)
-\end{lstlisting}
-Anglican functions outside of a query are defined using
-\texttt{defm} with the same syntax as Clojure \texttt{defn},
-however \texttt{name} is bound to an Anglican function.
-\begin{lstlisting}[style=default]
-(defm name doc-string? [param*] expr*)
-\end{lstlisting}
-Within the body of \texttt{defquery} and \texttt{defm}, the
-following special forms are supported, with syntax and semantics 
-matching those of Clojure:
-\begin{lstlisting}[style=default]
-(if pred then else?)
-(when pred expr*)
-(cond clause*)
-(case expr clause* default?)
-(let [binding*] expr*)
-(and expr*)
-(or expr*)
-(fn name? [param*] expr*)
-(loop [binding*] expr*)
-(recur expr*)
-\end{lstlisting}
-In \texttt{defquery} parameter lists, \texttt{let} bindings, and
-\texttt{fn} argument lists, Clojure vector destructuring is
-supported. Literals for atomic types and compound literals for
-vectors, hash maps, and sets are supported just like in Clojure.
-
-Clojure provides special forms \texttt{loop} and \texttt{recur}
-for writing tail-recursive programs.  Anglican programs are
-CPS-converted and do not use the stack; recursive calls
-in Anglican cannot lead to stack overflow.  However,
-\texttt{loop} and \texttt{recur} are provided in Anglican for
-convenience as a way to express loops. Unlike in Clojure though,
-\texttt{recur} is supported only inside a \texttt{loop}.
-
-\subsection{Core Library}
-\label{sec:core}
-
-All of Clojure's core library except for higher-order functions
-(functions that accept other functions as arguments) is
-available in Anglican. Higher-order functions cannot be reused
-from Clojure, as they have to be re-implemented to accept
-functional arguments in CPS form. The following higher-order
-functions are implemented: \texttt{map}, \texttt{reduce},
-\texttt{filter}, \texttt{some}, \texttt{repeatedly},
-\texttt{comp}, \texttt{partial}.
-
-\subsection{Special Forms}
-
- In addition to re-implementing a subset of Clojure, Anglican
-provides special forms for probabilistic inference:
-\begin{lstlisting}[style=default]
-(sample address? distribution)
-(observe address? distribution value)
-(mem function)
-(store tag value)
-(retrieve tag)
-\end{lstlisting}
-
-A good way to understand the \texttt{sample} and \texttt{observe} forms
-is to consider the imaginary interpreter in Section~\ref{sec:pp-def}.
-Under this interpreter, the \texttt{sample} form draws a value $x$ from 
-its parameter distribution $F$, and 
-generates a unique address $\alpha$ for this random draw.
-Then, it appends $(F, x, \alpha)$, the record of this random
-draw, to the log of the current execution. Finally, the
-form returns $x$. The computation of the \texttt{observe} form is
-simpler. It just appends its two parameters, distribution $G$
-and observed value $v$, to the log of the current execution,
-and returns \texttt{nil}.
-
-What we have just described is idealised semantics 
-of \texttt{sample} and \texttt{observe}. The actual computation 
-involved in handling \texttt{sample} and \texttt{observe} 
-depends on the inference algorithm. Different inference algorithms may treat
-\texttt{sample} and \texttt{observe} differently as long as they
-compute the same distribution on traces (i.e. sequences of sampled values), 
-namely, the one in Equation~\ref{eqn:p-trace}.
-Formal semantics for an idealized version
-of Anglican or other higher-order probabilistic programming languages are introduced 
-in \cite{SYH+16,BorgstromLGS16}; in general though,
-specifying semantics of higher-order probabilistic programming
-languages is an open problem.
-
-\texttt{sample} and \texttt{observe} may appear anywhere in the
-code of an Anglican program. To support these forms, the
-inference engine must be able to intervene into execution of the
-program, perform computations related to inference, and control
-further execution of the program. This tight and complicated
-interaction between the program and the inference engine
-necessitates implementing Anglican as a language rather than a
-library.
-
-The remaining Anglican special forms support carrying values in
-the hidden program state. \texttt{mem} implements stochastic
-function memoization (Section~\ref{sec:mem}). \texttt{store}
-stores a value for a tag in the program state, which can be
-later retrieved using \texttt{retrieve} during the same
-program execution.
-
-\section{Case Study}
-\label{seq:study}
-
-Before delving into implementation details, we discuss a case study
-of the use of the Anglican language and environment. This case study
-takes an inference problem for which a solution is not immediately
-obvious, \textit{the Deli dilemma}, and discusses how this problem
-can be solved by writing an Anglican program, executing the program,
-and post-processing results.
-
-The program presented in this section is intentionally short and
-simple. Anglican is capable of compiling and running elaborate
-programs and handling large amounts of data. Advanced examples of
-Anglican programs and inference can be found in literature on
-applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.
-
-\subsection{The Deli Dilemma}
-\label{seq:deli}
-
-Imagine that we are facing the following dilemma:
-\begin{quote}
-A customer wearing round sunglasses came to a deli at 1:13pm,
-and grabbed a sandwich and a coffee. Later on the same day, a
-customer wearing round sunglasses came at 6:09pm and ordered a
-dinner.  Was it the same customer?
-\end{quote}
-Additionally, we know that:
-\begin{itemize}
-    \item There is an adjacent office quarter, and it takes
-        between 5 and 15 minutes on average to walk from an
-        office to the deli, where different average times are  for
-        different buildings in the quarter.
-    \item Depending on traffic lights, the walking time varies by about 2
-        minutes.
-    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
-    \item Based on the similarity of appearance, the waiter assesses the odds that this is the same customer as 2 to 1.
-\end{itemize}
-
-\subsection{Anglican Query}
-\label{seq:deli-query}
-
-We want to formalize the dilemma as an Anglican query, based on
-the knowledge we have. Let us formalize the knowledge in
-Clojure (the times are in minutes). First, we encode our prior
-information which holds true independently of the customer's
-visit:
-
-\begin{lstlisting}[style=default]
-(def mean-time-to-arrive
-     "average time to arrive"
-     10.)
-(def sd-time-to-arrive
-     "standard deviation of arrival time"
-     3.)
-(def time-sd
-     "walking time deviation"
-     1.)
-\end{lstlisting}
-
-Then, we record our observations, based on which we want to
-solve the dilemma:
-
-\begin{lstlisting}[style=default]
-(def lunch-delay
-     "time between lunch break and lunch order"
-     13.)
-(def dinner-delay
-     "time between end of day and dinner order"
-     9.)
-(def p-same
-     "prior probability of the same customer"
-     (/ 2. 3.))
-\end{lstlisting}
-
-For inference, one often chooses a known distribution to
-represent uncertainty. We choose the normal
-distribution for representing uncertainty about average arrival
-time.
-
-\begin{lstlisting}[style=default]
-(def time-to-arrive-prior 
-     "prior distribution of average arrival time"
-     (normal mean-time-to-arrive
-             sd-time-to-arrive))
-\end{lstlisting}
-
-There are two possibilities: either the same customer visited
-the deli twice, or two different customers came to the deli,
-one for lunch and the other for dinner. We define an \textbf{Anglican} function
-for each case. Note that this is the first time we switch from
-Clojure to Anglican. The functions must be written in Anglican
-(and hence defined using \texttt{defm} instead of \texttt{defn})
-because they contain probabilistic forms \texttt{sample} and
-\texttt{observe}.
-
-\begin{lstlisting}[style=default]
-(defm same-customer 
-  "observe the same customer twice"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive time-sd)
-             dinner-delay)
-    [time-to-arrive]))
-
-(defm different-customers
-  "observe different customers"
-  [time-to-arrive-prior
-   lunch-delay
-   dinner-delay]
-  (let [time-to-arrive-1
-          (sample time-to-arrive-prior)
-        time-to-arrive-2
-          (sample time-to-arrive-prior)]
-    (observe (normal time-to-arrive-1 time-sd) 
-             lunch-delay)
-    (observe (normal time-to-arrive-2 time-sd)
-             dinner-delay)
-    [time-to-arrive-1 time-to-arrive-2]))
-\end{lstlisting}
-
-Both functions have the same structure: we first `guess' the
-average arrival time and then observe the actual time from a
-distribution parameterized by the guessed time. However, in
-\texttt{same-customer} the average arrival time is the same for
-both the lunch and the dinner, while in
-\texttt{different-customers} two average arrival times are
-guessed independently.
-
-We are finally ready to define the query in \textbf{Anglican}.
-
-\begin{lstlisting}[style=default]
-(defquery deli [time-to-arrive-prior
-                lunch-delay
-                dinner-delay]
-  (let [is-same-customer (sample (flip p-same))
-        observe-customer (if is-same-customer
-                            same-customer
-                            different-customers)]
-    {:same-customer is-same-customer,
-     :times-to-arrive (observe-customer 
-                        time-to-arrive-prior
-                        lunch-delay
-                        dinner-delay)}))
-\end{lstlisting}
-
-Performing inference on the query \texttt{deli} computes the
-probability that the same customer visited the deli twice, as
-well as probability distributions of average arrival times for
-both cases.
-
-\subsection{Inference}
-\label{seq:deli-inference}
-
-Having defined the query, we are now ready to run the query
-using an inference algorithm. Function
-\texttt{doquery}\iftoggle{full}{ provided
-by the \texttt{anglican.core} namespace}{} accepts the inference
-algorithm, the query, and optional parameters, and returns a
-lazy sequence of samples. We use here the inference algorithm
-called Lightweight Metropolis-Hastings (LMH). The algorithm is
-somewhat slow to converge but can be used with any Anglican
-query, and should be robust enough for our simple problem.
-
-We bind the results of \texttt{doquery} to variable
-\texttt{samples}, to analyse the results later. However, since
-the sequence is lazy, no inference is performed and no samples
-are generated until they are retrieved and processed.
-
-\begin{lstlisting}[style=default]
-(def samples (doquery :lmh deli nil))
-\end{lstlisting}
-
-To approximate the inferred distribution, we extract a finite
-subsequence of samples. Many algorithms use an initial
-subsequence of samples to converge to the target distribution.
-Hence, we drop initial $N$ samples ($N$ is 5000 in the code
-snippet below), and collect the next $N$ samples.
-
-\begin{lstlisting}[style=default]
-(def N 5000)
-(def results (map get-result
-                   (take N (drop N samples))))
-\end{lstlisting}
-
-Based on the collected samples we compute an approximation of
-the posterior probability \texttt{p-same+} that the same
-customer visited the deli twice.
-
-\begin{lstlisting}[style=default]
-(def p-same+ 
-     (/ (count (filter :same-customer results))
-        (double N)))
-\end{lstlisting}
-The \texttt{:same-customer} here represents a function
-that looks up the entry with the same name in a given map.
-The map stores the result of a single sampled execution,
-and the \texttt{:same-customer} entry in the map
-records whether the same customer visits the deli in the execution.
-The \texttt{filter} function in \texttt{p-same+} 
-goes through all the maps in \texttt{results}
-and picks only the ones whose executions involve just one customer.
-
-With the specified observations, \texttt{p-same+} is $\approx
-0.12$. The probability is much lower than the
-waiter's guess \texttt{p-same} ($\frac 2 3$).
-Of course, the results may vary from run to run, and, for a given
-algorithm, the accuracy depends on the number of samples we
-decide to retrieve.
-
-Besides computing the posterior probability that the same
-customer visited the deli twice, we may want to know the average
-time (or times) to arrive. In Bayesian inference, it is common
-to report distributions instead of `most likely' values. We use
-query results in retrieved samples to approximate the
-distributions, and plot distribution histograms for the same
-customer visiting twice (Figure~\ref{fig:dist-same}) and
-two different customers (Figure~\ref{fig:dist-diff})
-
 \begin{lstlisting}[style=default]
@@
 and results can rely on the full power of Clojure.
-
-\begin{figure}
-    \centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.6]{same-customer-time.pdf}
-    \caption{Arrival time distribution for a single customer.}
-    \label{fig:dist-same}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[trim={12pt 0 0 0},scale=0.6]{different-customers-times.pdf}
-    \caption{Arrival time distributions for two customers.}
-    \label{fig:dist-diff}
-\end{figure}
-
-\iftoggle{full}{%
-\input{case-study-bbvb}
-}{}%
-
-This completes the case study, where we showed a probabilistic
-programming solution of a problem, implemented and analysed in
-Anglican and Clojure\iftoggle{full}{, using two different
-inference algorithms}{}.
-
-\section{Macro-based Compilation}
-\label{sec:compilation}
-
-Anglican queries are compiled into Clojure using a variant of CPS
-transformation. In a basic CPS-transformed program, a
-continuation receives a single argument --- the computed value.
-In Anglican, two flows of computation are performed in
-parallel: values are computed by functional code, and,
-at the same time, the state of the probabilistic program,
-used by inference algorithms, is updated by
-probabilistic forms. Because of that, in Anglican a continuation
-accepts  \textit{two} arguments:
-\begin{itemize}
-    \item the computed value;
-    \item the internal state, bound to the local variable
-        \texttt{\$state} in every lexical scope.
-\end{itemize}
-
-The compilation relies on the Clojure \textit{macro} facility,
-and implemented as a library of
-\textit{functions}\iftoggle{full}{ in namespace
-\texttt{anglican.trap}, which are}{} invoked by macros.  The CPS
-transformation is organized in top-down manner.  The top-level
-function is  \texttt{cps-of-expression}, which receives an
-expression and a continuation, and returns the expression in the
-CPS form. For example, the CPS transformation of constant
-\texttt{1} with continuation \texttt{cont} thus takes the
-following form:
-\begin{lstlisting}[style=default]
-=> (cps-of-expression 1 'cont)
-(cont 1 $state)
-\end{lstlisting}
-
-\subsection {The State}
-\label{sec:state}
-
-The state (\texttt{\$state}) is threaded through the computation
-and contains data used by inference algorithms. \texttt{\$state}
-is a Clojure hash map:
-\begin{lstlisting}[style=default]
-(def initial-state
-  "initial program state"
-  {:log-weight 0.0, ;; map :log-weight to 0.0
-   :result nil,       ;; map :result to nil
-   ::store nil,       ;; map ::store to nil
-   ::mem {},          ;; map ::mem to the empty map 
-   ... })
-\end{lstlisting}
-which records inference-relevant information under various keys such
-as $\texttt{:log-weight}$ and $\texttt{::mem}$. The full list of map
-entries depends on the inference algorithm. CPS transformation
-routines are not aware of the contents of \texttt{\$state} and do not
-access or modify it directly. Rather, they just thread the state
-unmodified through the computation. The sole exception is the
-transformation of  the \texttt{mem} form (which converts a function
-to one with memoization). Algorithm-specific handlers of checkpoints
-corresponding to the probabilistic forms (\texttt{sample},
-\texttt{observe}) modify the state and insert an updated state into
-the computation.
-
-\iftoggle{full}{%
-\input{expression-kinds.tex}
-}{%
-}%
-
-\iftoggle{full}{%
-\subsection{Implementation Highlights}
-\label{seq:highlights}
-
-So far we introduced the basics of Anglican compilation to
-Clojure. The described approaches and techniques are important
-for grasping the language implementation but relatively well-known.
-In the rest of the section we focus on challenges we met and
-resolved while implementing Anglican, as well as on
-implementation of unique features of Anglican as a probabilistic
-programming language.
-}{%
-}%
-
-\iftoggle{full}{\subsubsection}{\subsection}{Probabilistic Forms}
-\label{seq:forms}
-
-The purpose of probabilistic forms \texttt{sample} and
-\texttt{observe} is to interrupt deterministic computation and
-transfer control to the inference algorithm.  Practically, this
-is achieved by  returning a \textbf{checkpoint} instead of
-invoking a continuation in a CPS-transformed program.  
-Checkpoints are represented by values of the
-corresponding record types
-(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
-\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
-in Clojure. The types contain fields specific to each form, as well as the
-continuation; calling the continuation resumes the computation.
-\begin{lstlisting}[style=default]
-=> (cps-of-expression '(sample dist) 'cont)
-(->sample dist cont $state)
-=> (cps-of-expression '(observe dist v) 'cont)
-(->observe dist v cont $state)
-\end{lstlisting}
-Here \texttt{->sample} and \texttt{->observe} in the
-CPS-transformed expressions are constructors for Clojure
-records, and they take values of their fields as arguments.
-
-Upon encountering a checkpoint, the inference algorithm computes
-the updated program state and the value to be passed to the
-continuation.  How the state is updated, the number of times the
-continuation is called, and the value passed to the continuation
-of \texttt{sample} depend on the inference algorithm executing
-the program.  Section~\ref{sec:inference} provides more detail
-on internals of inference algorithms and their interaction with
-probabilistic programs.
-
-\iftoggle{full}{%
-\smallskip
-\begin{center}
-*\qquad*\qquad*
-\end{center}
-
-In addition to \texttt{sample} and \texttt{observe}, there are a few other special
-forms ---  \texttt{store}, \texttt{retrieve}, \texttt{mem} ---
-which modify program state. These forms are translated into
-expressions involving calls of functions from the
-\texttt{anglican.state} namespace. The \texttt{mem} form, which
-implements memoization, deserves a more detailed explanation.
-}{%
-}%
-
-\iftoggle{full}{\subsubsection }{\subsection}{Memoization}
-\label{sec:mem}
-
-The author of a probabilistic model might want to 
-randomly draw a feature value for each entity in a collection,
-but to retain the same drawn
-value for the same entity during a single run of the probabilistic
-program. For example, a person may have brown or green eyes
-with some probability, but the \textit{same} person will
-always have the same eye color. This can be achieved
-through the use of memoized functions. In Anglican, one
-might write:
-\begin{lstlisting}[style=default]
-(let [eye-color (mem (fn [person]
-                       (sample (categorical
-                                 ['brown 0.5]
-                                 ['green 0.5]))))]
-  (if (not= (eye-color 'bill) (eye-color 'john))
-    (eye-color 'bill)
-    (eye-color 'john)))
-\end{lstlisting}
-The \texttt{mem} form converts a function to a memoized variant,
-which remembers past inputs and the corresponding outputs, and 
-returns the remembered output when given such a past input, 
-instead of calling the original function with it. As a result,
-for every input, random draws will be made only for the first time 
-that the memoized function is called with the input; all subsequent
-calls with the input will just reuse these draws and return the same output.
-
-Memoization is often implemented on top of a mutable
-dictionary, where the key is the argument  and the value is
-the returned value. However, there are no mutable data
-structures in a probabilistic program. Hence, \texttt{mem}'s
-memory is stored as a nested dictionary in the program state
-introduced during CPS transformation (function \texttt{mem-cps}).  
-Every memoized function gets a
-unique automatically generated identifier. Each time a memoized
-function is called, one of two continuations is chosen,
-depending on whether the same function (a function with the same
-identifier) was previously called in the same run of the
-probabilistic program with the same argument. If the memoized
-result is available, the continuation of the memoized function
-call is immediately called with the stored result. Otherwise,
-the argument of \texttt{mem} is called with a continuation which
-first creates an updated state with the memoized result, and
-then calls the `outer' continuation with the result and the
-updated state:
-\begin{lstlisting}[style=default]
-=> (mem-cps '(foo))
-(let [M (gensym "M")]
-  (fn mem23623 [C $state & P]
-    (if (in-mem? $state M P)
-      ;; previously memoized result
-      (fn [] (C (get-mem $state M P) $state))
-      ;; new computation
-      (clojure.core/apply foo
-        ;; memoize result in state
-        (fn [V $state]
-          (fn [] (C V (set-mem $state M P V))))
-        $state P)))))
-\end{lstlisting}
-
-Memoized results are not shared among multiple runs of a
-probabilistic program, which is intended. Otherwise, it would be
-impossible to memoize functions with random results.
-
-\iftoggle{full}{\subsubsection{Continuations}}{\subsection{Managing stack size}}
-\label{seq:continuations}
-
-\iftoggle{full}{
-Continuations are functions that are called in tail positions
-with the computed value and state as their arguments --- in CPS
-there is always a function call in every tail position and never
-a value. Continuations are passed to CPS transformers, and when
-transformers are called recursively, the continuations are
-generated on the fly.
-
-There are two critical issues related to generation of
-continuations:
-\begin{itemize}
-    \item unbounded \textit{stack growth} in recursive code;
-    \item code size \textit{explosion} when a non-atomic
-        continuation is symbolically substituted in multiple
-        locations.
-\end{itemize}
-
-\paragraph{Managing stack size}
-}{%
-Continuations may lead to unbounded stack growth in
-recursive code.
-}%
-In implementations of functional programming languages stack
-growth is avoided through \textit{tail call optimization} (TCO).
-However, Clojure does not support a general form of TCO, and
-CPS-transformed code that creates deeply nested calls will
-easily exhaust the stack. Anglican employs a workaround called
-\textit{trampolining}~\cite{B95} --- instead of inserting a
-continuation call directly, the transformer always wraps the
-call into a \textit{thunk}, or parameterless function. The thunk
-is returned and called by the trampoline (Clojure provides
-function \texttt{trampoline} for this purpose) --- this way the
-computation continues, but the stack is collapsed on every
-continuation call. Function \texttt{continue} implements the
-wrapping and is invoked on every continuation call:
-\begin{lstlisting}[style=default]
-=> (continue 'cont 'value 'state)
-(fn [] (cont value state))
-\end{lstlisting}
-Correspondingly, the full, wrapped CPS form of
-\begin{lstlisting}[style=default]
-(fn [x y] (+ x y))
-\end{lstlisting}
-becomes
-\begin{lstlisting}[style=default]
-(fn [cont $state x y]
-  (fn []
-    (cont (+ x y) $state)))
-\end{lstlisting}
-When the CPS-transformed function is called, it returns a
-\textit{thunk} (a parameterless function) which is then
-re-invoked through the trampoline, with the stack collapsed.
-
-\iftoggle{full}{%
-\input{code-explosion.tex}
-}{%
-}%
-
-\iftoggle{full}{\subsubsection}{\subsection}{Primitive Procedures}
-\label{seq:primitive}
-
-When an Anglican function is transformed into a Clojure
-function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary
-parameters are added to the beginning of the parameter list ---
-continuation and state.  Correspondingly, when a function
-\textit{call} is transformed\iftoggle{full}{ (by
-\texttt{cps-of-application} or \texttt{cps-of-apply})}{}, the
-current continuation and the state are passed to the called
-function. Anglican can also call Clojure functions; however,
-Clojure functions do not expect these auxiliary parameters.
-To allow the mixing of Anglican (CPS-transformed) and Clojure
-function calls in Anglican code, the Anglican compiler must CPS-transform
-expressions selectively~\cite{Nielsen01}. It must be
-able to recognize `primitive' (that is, implemented in Clojure
-rather than in Anglican) functions, and invoke
-those functions in a direct, not CPS, style.
-
-Providing an explicit syntax for differentiating between
-Anglican and Clojure function calls would be cumbersome. Another
-option would be to use meta-data to identify Anglican function
-calls at runtime. However, this would impact performance, and
-good runtime performance is critical for probabilistic
-programs. The approach taken by Anglican is to maintain a list
-of unqualified names of primitive functions, as well of
-namespaces in which all functions are primitive, and recognize
-primitive functions by name --- if a function name is not in the
-list, the function is an Anglican function. Global dynamically-bound 
-variables \texttt{*primitive-procedures*} and
-\texttt{*primitive-namespaces*} contain the initial lists of
-names and namespaces, correspondingly. Of course, local bindings
-can shade global primitive function names. For example,
-\texttt{first} is a Clojure primitive function that takes the first
-element from an ordered collection such as list and vector,
-but inside the let block in the
-following example, \texttt{first} is an Anglican function:
-\begin{lstlisting}[style=default]
-(let [first (fn [[x & y]] x)]
-  (first '[1 2 3]))
-\end{lstlisting}
-The Anglican compiler takes care of the shading by rebinding
-\texttt{*primitive-procedures*} in every lexical
-scope\iftoggle{full}{ (\texttt{fn-cps}, \texttt{cps-of-let}). Macro
-\texttt{shading-primitive-procedures} automates the shading}{.}
-
-\section{Inference Algorithms}
-\label{sec:inference}
-
-A probabilistic program in Anglican may look almost like a
-Clojure program. However, the purpose of executing a
-probabilistic program is different from that of a `regular'
-program: instead of producing the result of a single execution,
-a probabilistic program computes, exactly or approximately, the
-distribution from which execution results are drawn.  Characterising
-this distribution is done by the inference algorithm.
-
-Probabilistic programming system Anglican provides a variety of
-approximate inference algorithms. Ideally, Anglican should
-automatically choose the most appropriate algorithm for each
-probabilistic program. In practice, selecting an inference
-algorithm, or a combination thereof, is still a challenging
-task, and program structure, intended use of the computation
-results, performance, approximation accuracy, and other factors
-must be taken into consideration. New algorithms are being
-developed and added to Anglican~\cite{TMP+15,MPT+16,RNL+2016},
-as a part of ongoing research.
-
-In the implementation of Anglican, inference algorithms are
-instantiations of the (ad-hoc) polymorphic function
-\texttt{infer} declared as Clojure's multimethod\iftoggle{full}{
-in the \texttt{anglican.{\linebreak[0]}inference}
-namespace}{}. The function accepts an algorithm identifier,
-a query (the probabilistic program in which to perform the
-inference), an initial value for the query, and optional
-algorithm parameters.
-
-\subsection{The \texttt{infer} Function}
-\label{sec:infer}
-
-The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
-an appropriate overloading or implementation of the function --- conventionally,
-the identifier is a Clojure keyword (a symbolic constant starting with colon)
-related to the algorithm name, such as \texttt{:lmh} for Lightweight
-Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
-The second parameter is a query as defined by
-the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
-the following Clojure code invokes \texttt{infer} on an Anglican query
-defined anonymously via the \texttt{query} form:
-\begin{lstlisting}[style=default]
-(let [x 1]
-  (infer :pgibbs (query x) nil))
-\end{lstlisting}
-A query is executed by calling the initial continuation of the
-query, which accepts a value and a state. The state is supplied
-by the inference algorithm, while the value is provided as a
-parameter of \texttt{infer}. A query does not have to have any
-parameters, in which case the value can be simply \texttt{nil}.
-When a query is defined with a binding for the initial value,
-the value becomes available inside the query. A query may
-accept multiple parameters using Clojure's structured binding.
-For instance, it may take multiple parameters as components
-of an input vector. In this case, the initial value is given
-as a structured value, such as a vector, and the components
-of this value are matched to the corresponding
-parameters of the query via the destructuring mechanism of Clojure.
-For example,
-\begin{lstlisting}[style=default]
-(defquery my-query [mean sd]
-  (sample (normal mean sd)))
-
-(def samples (infer :lmh my-query [1.0 3.0]))
-\end{lstlisting}
-Finally, any number of auxiliary arguments can be
-passed to \texttt{infer}.
-By convention, the arguments should be
-keyword arguments, and are interpreted in the algorithm-specific
-manner.
-
-\subsection{Internals of an Inference Algorithm}
-\label{sec:internals}
-
-Implementing an inference algorithm in Anglican amounts to
-defining an appropriate version of the \texttt{infer} function
-and checkpoint handlers for \texttt{sample} and
-\texttt{observe}.  The definitions may just reuse default
-implementations or override them with new algorithm-specific
-treatment of \texttt{sample} and \texttt{observe} forms and
-inference state.
-
-Let us illustrate this implementation step with
-\textit{importance sampling},
-the simplest inference algorithm.
-Here is an implementation of the \texttt{infer} function for the algorithm:
-\begin{lstlisting}[style=default]
-;; Make ::algorithm be converted to 
-;; :anglican.inference/algorithm whenever necessary
-(derive ::algorithm :anglican.inference/algorithm)
-
-;; invoked when algo parameter is :importance
-(defmethod infer :importance 
-  [algo prog value & {}]
-  (letfn
-    ;; recursive function without any parameter
-    [sample-seq 
-      []
-      ;; lazy infinite sequence 
-      (lazy-seq
-        (cons
-          (:state (exec ::algorithm prog value
-                        initial-state))
-          (sample-seq))))]
-    (sample-seq)))
-\end{lstlisting}
-This implementation is dispatched 
-when \texttt{infer} receives {\tt :im\-por\-tan\-ce} as its \texttt{algo} 
-parameter. (Anglican includes multiple implementations of \texttt{infer}
-for different inference algorithms.) Once dispatched, it lazily 
-constructs an infinite sequence of inference states by repeatedly 
-executing the program \texttt{prog} using \texttt{exec}
-and retrieving the final inference state of the execution 
-using \texttt{:state}. For checkpoint handlers, importance sampling simply
-relies on their default implementations.
-
-Typically, an inference
-algorithm has its own implementations of
-\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
-both, as well as invokes \texttt{exec} from an elaborated
-conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
-(\texttt{anglican.smc}) are examples of inference algorithms
-where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
-handler is overridden. In addition, SMC runs multiple particles
-(corresponding to user-level threads) simultaneously, while LMH re-runs
-programs from an intermediate continuation rather than from the
-beginning.
-
-\subsection{Addressing of Checkpoints}
-\label{sec:addressing}
-
-Many inference algorithms memoize and reuse earlier computations
-at checkpoints. Variants of Metropolis-Hastings reuse previously
-drawn values~\cite{WSG11}, as well as additional information
-used for adaptive sampling~\cite{TMP+15} at \texttt{sample}
-checkpoints. Asynchronous SMC (Particle Cascade) computes
-average particle weights at \texttt{observe}
-checkpoints~\cite{PWD+14}. Implementations of black-box
-variational inference~\cite{WW13,MPT+16} associate with random
-variables the learned parameters of variational posterior
-distribution.
-
-Checkpoints can be uniquely named at compilation time;
-however, at runtime a checkpoint corresponding to a single code
-location may occur multiple times due to recurrent invocation of
-the function containing the checkpoint. Every unique occurrence
-of a checkpoint must receive a different address. An addressing
-scheme based on computing of stack addresses of checkpoints was
-described in the context of Lightweight
-Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
-over the naive scheme where dynamic occurrences of checkpoints
-are numbered sequentially. However, it impacts the
-performance of probabilistic programs because of the computation
-cost of computing the stack addresses:
-\begin{enumerate}
-    \item On every function call, a component is added to the
-        address. Hence, the address size is linear in stack depth.
-    \item Every function call must be augmented by symbolic
-        information required to compute stack addresses.
-\end{enumerate}
-In addition, the above scheme can still lead to inferior
-correspondence between checkpoints in different traces: in
-Anglican and other probabilistic languages where distributions
-are first-class objects checkpoints with incompatible arguments
-can correspond to the same stack address. Consider the following
-program fragment:
-\begin{lstlisting}[style=default]
-(let [dist (if use-gamma
-               (gamma 2. 2.)
-               (normal 0. 1.))]
-  (sample dist))
-\end{lstlisting}
-The \texttt{sample} checkpoint has the same stack address in
-different traces. However, the random values should not be
-reused between different distributions. Further on, in some
-algorithms, such as black-box variational inference, the role of
-checkpoint addresses is semantic rather than heuristic ---
-appropriate correspondence must be established
-between checkpoints in different traces for the algorithm to
-work.
-
-To overcome the above problems, Anglican introduces a new
-addressing scheme which is almost as efficient as the scheme
-based on stack addresses for reuse of previously drawn values,
-while producing addresses of constant size, and allows manual
-computation of checkpoint addresses at runtime when the default
-automatic scheme is insufficient. According to the scheme:
-\begin{itemize}
-    \item A checkpoint may accept an auxiliary argument --- the
-        checkpoint identifier. If specified, the identifier is the first
-        argument of a checkpoint. For example \texttt{(sample
-        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
-        with identifier \texttt{x1}.
-    \item If the identifier is omitted, a unique identifier is
-        generated automatically as a fresh symbol.
-    \item At runtime, the address of a checkpoint invocation has
-        the form
-        \texttt{[}\textit{checkpoint-identifier
-        number-of-previous-occurrences}\texttt{]}, where the
-        occurrences are of a checkpoint with the same checkpoint
-        identifier.
-    \item If a sequence of invocations of the same checkpoint is
-        interrupted by a different checkpoint, the number of
-        previous occurrences is rounded up to a multiple of
-        an integer. For efficiency, a small power of 2 is used,
-        such as $2^4 = 16$.
-\end{itemize}
-
-\noindent Consider the following simplified Anglican query:
-
-\begin{lstlisting}[style=default]
-(defm foo []
-  (if (sample 'C1 (flip 0.5))
-    (foo)
-    (bar)))
-
-(defm bar []
-  (case (sample 'C2 (uniform-discrete 0 3))
-    0 (bar)
-    1 (foo)
-    2 (sample 'C3 (normal 0. 1.))))
-
-(defquery baz
-  (foo))
-\end{lstlisting}
-An execution of the query may result in the following
-sequence of checkpoint invocations:
-\begin{lstlisting}[style=default]
-  (sample 'C1 ...)
-  (sample 'C2 ...)
-  (sample 'C2 ...)
-  (sample 'C1 ...)
-  (sample 'C1 ...)
-  (sample 'C1 ...)
-  (sample 'C2 ...)
-  (sample 'C3 ...)
-\end{lstlisting}
-According to the addressing scheme, the addresses generated
-for these invocations are
-\begin{lstlisting}[style=default]
-[C1 0][C2 0][C2 1][C1 16][C1 17][C1 18][C2 16][C3 0]
-\end{lstlisting}
-If the program is run by a Metropolis-Hastings algorithm, then a
-small change usually takes place in the sequence of checkpoints
-with each invocation, and the new sequence may become
-\begin{lstlisting}[style=default]
-  (sample 'C1 ...)
-  (sample 'C2 ...)
-  (sample 'C1 ...)
-  (sample 'C1 ...)
-  (sample 'C2 ...)
-  (sample 'C2 ...)
-  (sample 'C3 ...)
-\end{lstlisting}
-The addresses for the new sequence are
-\begin{lstlisting}[style=default]
-  [C1 0][C2 0][C1 16][C1 17][C2 16][C2 17][C3 0]
-\end{lstlisting}
-Despite the change, the correspondence between checkpoints of
-similar types occurring in similar positions (relative positions
-in contiguous subsequences of a certain type) is preserved, and
-drawn values can be reused efficiently:
-
-\vspace{\baselineskip}
-{
-\begin{tabular}{l l}
-    \textbf{old} & \textbf{new} \\ \hline
-    {}\texttt{[C1 0]} & \texttt{[C1 0]} \\
-    {}\texttt{[C2 0]} & \texttt{[C2 0]} \\
-    {}\texttt{[C2 1]} & \textit{unused} \\
-    {}\texttt{[C1 16]} & \texttt{[C1 16]} \\
-    {}\texttt{[C1 17]} & \texttt{[C1 17]} \\
-    {}\texttt{[C1 18]} & \textit{unused} \\
-    {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
-    \textit{missing} & \texttt{[C2 17]} \textit{drawn}\\
-    {}\texttt{[C3 0]} & \texttt{[C3 0]}
-\end{tabular}
-\vspace{\baselineskip}}
-
-Note that correspondence between checkpoints in different
-traces plays the role of an heuristic in Metropolis-Hastings
-family of algorithms. Any correspondence (or no correspondence,
-meaning all values must be re-drawn from their proposal
-distributions) is valid, and reused values from the previous
-invocation which are not in support or have zero probability in
-the new invocation are simply ignored and new values are
-re-drawn.
-
-This way, each occurrence of a checkpoint has a unique address,
-but small disturbances --- removal or addition of a single or just
-a few checkpoints --- are unlikely to derail the entire sequence.
-The probability of derailment depends on the padding. The
-padding can be safely, and without any impact on performance,
-set to rather large numbers. However, rounding up to a multiple
-of 16 proved to be appropriate for all practical purposes.
-\iftoggle{full}{%
-
-Function \texttt{checkpoint-id} in the
-\texttt{anglican.inference} namespace automates generation of
-checkpoint addresses and can be used from any inference
-algorithm.
-}{%
-}%
-
-\section{Definitions and Runtime Library}
-\label{seq:runtime}
-
-A Clojure namespace that includes a definition of an Anglican
-program imports (`requires') two essential namespaces:
-\texttt{anglican.emit} and \texttt{anglican.runtime}. The former
-provides macros for defining Anglican programs
-(\texttt{defquery}, \texttt{query}) and functions
-(\texttt{defm}, \texttt{fm}, \texttt{mem}), as well as Anglican
-bootstrap definitions that must be included with every program
---- first of all, CPS implementations of higher-order functions.
-\texttt{anglican.emit} can be viewed as the Anglican
-\textit{compiler tool}, which helps transform Anglican code into
-Clojure before any inference is performed.
-
-\texttt{anglican.runtime} is the Anglican runtime library. For
-convenience, it exposes common mathematical functions
-(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
-\texttt{exp}, etc.), but most importantly, it provides
-definitions of common distributions. Each distribution object implements a
-distribution interface with the \texttt{sample*} and \texttt{observe*} methods;
-this interface is defined using Clojure's protocol mechanism
-(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
-The \texttt{sample*} \textit{method} returns a random sample and
-roughly corresponds to the \texttt{sample} checkpoint.
-The \texttt{observe*} \textit{method} returns the log probability
-of the value and roughly corresponds to the \texttt{observe}
-checkpoint. The methods can be, and sometimes are called from
-handlers of the corresponding checkpoints, but do not have to
-be. For example, in LMH either the \texttt{sample*} or
-the \texttt{observe*} \textit{method} is called for a
-\texttt{sample} checkpoint, depending on whether the value is
-drawn or reused.
-
-The macro \texttt{defdist} is used to define distributions.
-\texttt{defdist} takes care of defining a separate type
-for every distribution so that Clojure multimethods (or overloaded
-methods) can be dispatched on
-distribution types when needed, e.g. for custom proposal distributions 
-used in an inference algorithm. The
-Bernoulli distribution could be defined as follows:
-\begin{lstlisting}[style=default]
-(defdist bernoulli
-  "Bernoulli distribution"
-  [p]
-  (sample* [this] (if (< (rand) p) 1 0))
-  (observe* [this value]
-    (Math/log (case value
-                1 p
-                0 (- 1. p)
-                0.))))
-\end{lstlisting}
-
-In addition to distributions, Anglican provides \textit{random processes},
-which define sequences of random variables that are not independent and
-identically distributed. Random processes are similar to the so called
-`exchangeable random primitives' in Church \cite{GMR+08} and
-Venture \cite{MSP14}. However, random sequences generated by Anglican random 
-processes are not required to be exchangeable. Random processes are
-defined using the \texttt{defproc} macro and implement the
-\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
-protocol has two methods
-\begin{itemize}
-\item \texttt{produce}, which returns the distribution on the next
-random variable in the sequence, and
-\item \texttt{absorb}, which incorporates the value for the next random
-variable and returns an updated random process.
-\end{itemize}
- As an example, here is a definition of a beta-Bernoulli process, in which
-each random variable is distributed according to a Bernoulli distribution with
-an unknown parameter that is drawn from a beta distribution:
-\begin{lstlisting}[style=default]
-(defproc beta-bernoulli
-  "beta-Bernoulli process"
-  [a b]
-  (produce [this] (bernoulli (/ a (+ a b))))
-  (absorb [this x]
-    (case x
-      0 (beta-bernoulli a (inc b))
-      1 (beta-bernoulli (inc a) b))))
-\end{lstlisting}
-Unlike typical implementations of exchangeable random processes,
-Anglican's random processes do not have mutable state. The
-\texttt{produce} and \texttt{absorb} methods are deterministic
-and functional, and therefore do not have corresponding special
-forms in Anglican. A sequence of random values can be generated
-using a recursive loop in which \texttt{absorb} returns the
-updated process for the next iteration. For example:
-\begin{lstlisting}[style=default]
-(defm sample-beta-binomial [n a b]
-  (loop [process (beta-bernoulli a b)
-         values []]
-    (if (= (count values) n)
-      values
-      (let [dist (produce process)
-            value (sample dist)]
-        (recur (absorb process value)
-               (conj values value))))))
-\end{lstlisting}
-Similarly, random processes can also be used to recursively
-observe a sequence of values:
-\begin{lstlisting}[style=default]
-(defm observe-beta-bernoulli [values a b]
-  (loop [process (beta-bernoulli a b)
-         values values] 
-    (when (not-empty values)
-      (let [dist (produce process)
-            value (first values)]
-        (observe dist value)
-        (recur (absorb process value)
-               (rest values))))))
-\end{lstlisting}
-
-\section{Performance}
-
-To justify the claim that Anglican is an \textit{efficient}
-implementation, performance of Anglican programs must be evaluated
-both against Clojure programs and against inference tasks in
-other probabilistic programming languages and environments.
-Anglican passes two additional parameters to every function, adds
-extra code to variable bindings, and passes every function
-call through the trampoline. Hence difference in performance
-between Anglican and Clojure should be most noticeable on
-programs which involve many function calls. One such example is
-the famous \textit{Towers of Hanoi}:
-
-\begin{lstlisting}[style=default]
-(defn towers-of-hanoi [n from to via]
-  (when (not= n 1) ;; return nil if n = 1.
-      (towers-of-hanoi (dec n) from via to)
-      (towers-of-hanoi (dec n) via to from)))
-\end{lstlisting}
-
-We run this program for \texttt{n} equal to 10, 15, 20, and 25. Clojure
-is consistently only twice as fast as Anglican
-(Table~\ref{tbl:hanoi}); considering the amount of overhead
-Anglican introduces, this is a good result.
-\begin{table}[!b]
-    \centering
-    \begin{tabular}{r|r|r|r|r}
-        {\bf n}  & 10 & 15 & 20 & 25 \\ \hline
-        {\bf Anglican} & 0.63 & 3.00 & 75.9 & 2381.  \\ \hline
-        {\bf Clojure} &  0.32 & 1.45 & 38.1  &  1245.
-    \end{tabular}
-    \caption{Towers of Hanoi. Running times, in
-    seconds, of Clojure and Anglican, averaged over 100 runs.
-    Clojure is only twice as fast as Anglican.}
-    \label{tbl:hanoi}
-\end{table}
-
-The other comparison must be made between Anglican and another
-probabilistic programming environment. WebPPL~\cite{GS15} is
-similar to Anglican in expressiveness.  WebPPL runs on top of
-\href{http://v8project.blogspot.com/}{V8}, a high performance
-JavaScript engine featuring a native compiler.
-Table~\ref{tbl:webppl} shows results for a few models from
-\href{https://github.com/probmods/webppl}{WebPPL distribution},
-as well as from
-\href{https://bitbucket.org/probprog/anglican}{Anglican test
-suite}. For the comparison, we re-implemented WebPPL models in
-Anglican, and Anglican models in WebPPL, following as close as
-possible the original implementations, and run the same
-inference algorithm (called MCMC in WebPPL, LMH in Anglican)
-with the same number of iterations. Anglican outperforms WebPPL
-on all models. This can be attributed both to a better
-transformed representation of probabilistic programs and to more
-efficient architecture of the inference engine.
-\begin{table}
-    \centering
-    \begin{tabular}{r|S[table-format=2.1]|S[table-format=2.1]}
-        {\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
-        Latent Dirichlet allocation & 22.2 & 31.5 \\ 
-        Linear regression  & 6.1 & 8.5 \\
-        Logistic regression & 8.4 & 10.4 \\
-        Multivariate regression & 6.9 & 14.2 \\
-        Simple branching & 2.4 & 5.7 \\
-        Hidden Markov model & 8.7 & 10.2
-    \end{tabular}
-
-    \caption{Anglican vs. WebPPL. Running times, in seconds, of
-    Anglican and WebPPL, for $100\,000$ iterations of Markov Chain
-    Monte Carlo (Lightweight Metropolis-Hastings), averaged over
-    100 runs. Anglican outperforms WebPPL on all models.}
-    \label{tbl:webppl}
-\end{table}
-The source code of the examples is available in the
-\href{https://bitbucket.org/probprogs/anglican-white-paper}{paper
-repository}.
-
-In addition, \cite[pp. 32 -- 33]{P16} 
-provides a comparison of Anglican to an older, interpreted
-version of Anglican~\cite{WVM14}, and Probabilistic
-Scheme~\cite{PW14} on a rather complicated inference task of
-program synthesis. According to this comparison, Anglican is at
-least 10 times faster than the older interpreted version, and
-almost as fast as Probabilistic Scheme which uses inference
-engine implemented in C. In addition, \cite[p. 171]{SGG15}
-compares Anglican to a DSL for probabilistic programming in
-Haskell and to Probabilistic C~\cite{PW14} on two simple
-probabilistic programs. Anglican shows similar performance
-to the Haskell implementation of Sequential Monte
-Carlo~\cite{DFG+01} and scales better with the number of
-particles. Given the simplicity of the programs used in the
-evaluation, this comparison is a confirmation that the flexible
-checkpoint and trampoline based interaction between an Anglican
-program and the inference engine does not introduce any noticeable
-overhead compared to more rigid designs.
-
-\section{Conclusion}
-\label{seq:summary}
-
-In this paper, we presented design and implementation internals
-of the \emph{probabilistic programming system} Anglican. Implementing
-a language is an interesting endeavour, in particular when the
-language implements a new paradigm, in this case probabilistic
-programming. Functional programming is a natural complement of
-probabilistic programming --- the latter allows both concise and
-expressive specification of probabilistic generative models and
-efficient implementation of inference algorithms.
-
-Implementing a probabilistic language on top of and in tight
-integration with a functional language, Clojure, both helped
-us to accomplish an ambitious goal in a short time span, and
-provided important insights on structure and semantics of
-probabilistic concepts incorporated in Anglican. Computational
-efficiency and expressive power of Anglican owe to adherence to
-the functional approach as much as to rich inference
-opportunities of the Anglican environment.
-
-\section*{Acknowledgements}
-
-This work was partially supported under DARPA PPAML through the
-U.S. AFRL under Cooperative Agreement number FA8750- 14-2-0006,
-Sub Award number 61160290-111668. Yang was supported by Institute for 
-Information \& communications Technology Promotion (IITP) grant funded 
-by the Korea government (MSIP) (No.R0190-16-2011, Development of 
-Vulnerability Discovery Technologies for IoT Software Security).
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{refs}
-
-\end{document}

### 2017-03-14 paper/paper.tex
better placed figures

@@
     \centering
-    \includegraphics[trim={6pt 12pt 0 12pt},scale=0.6]{same-customer-time.pdf}
+    \includegraphics[trim={0pt 12pt 0 12pt},scale=0.6]{same-customer-time.pdf}
     \caption{Arrival time distribution for a single customer.}
@@
     \centering
-    \includegraphics[trim={6pt 12pt 0 12pt},clip,scale=0.6]{different-customers-times.pdf}
+    \includegraphics[trim={0pt 12pt 0 12pt},clip,scale=0.6]{different-customers-times.pdf}
     \caption{Arrival time distributions for two customers.}

### 2017-03-16 paper/paper.tex
fixed Franks email
forgot to commit changes

@@
 \affiliation{\institution{University of Oxford}}
-\email{fwood@cs.ox.ac.uk}
+\email{fwood@robots.ox.ac.uk}
 
@@
 \begin{lstlisting}[style=default]
-=> (cps-of-expression '(sample dist) cont)
-(->sample 'S1 dist cont $state)
-=> (cps-of-expression '(observe dist v) cont)
-(->observe 'O1 dist v cont $state)
+=> (cps-of-expression '(sample dist) 'cont)
+(->sample S1 dist cont $state)
+=> (cps-of-expression '(observe dist v) 'cont)
+(->observe O1 dist v cont $state)
 \end{lstlisting}

### 2017-03-16 paper/paper.tex
finally got quoting right

@@
 }{
-\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.842}\ttfamily\small, columns=flexible, showstringspaces=false}
+\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.839}\ttfamily\small, columns=flexible, showstringspaces=false}
 }
@@
 
-\acmConference[IFL'16]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
+\acmConference[IFL 2016]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
 \acmYear{2016}
@@
 \ccsdesc[300]{Software and its engineering~Specialized application languages}
+
+\copyrightyear{2016}
+\setcopyright{acmlicensed}
+\acmConference[IFL 2016]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
+\acmISBN{978-1-4503-4767-9/16/08}
+\acmPrice{15.00}
+\acmDOI{http://dx.doi.org/10.1145/3064899.3064910}
 
@@
 => (cps-of-expression '(sample dist) 'cont)
-(->sample S1 dist cont $state)
+(->sample 'S1 dist cont $state)
 => (cps-of-expression '(observe dist v) 'cont)
-(->observe O1 dist v cont $state)
+(->observe 'O1 dist v cont $state)
 \end{lstlisting}

### 2017-03-16 paper/paper.tex
removed unused packages
which font assignment is right? T1 or OT1

@@
 \documentclass[sigconf]{acmart}
-
-\setcopyright{rightsretained}
 
@@
 \usepackage[color]{changebar}
-\usepackage{enumitem}
 \usepackage[lighttt]{lmodern}
@@
 \usepackage{listings}
-\usepackage{url}
 \usepackage{etoolbox}
@@
 
+\setcopyright{acmlicensed}
+
 \ttfamily
-\DeclareFontShape{OT1}{lmtt}{m}{it}
+\DeclareFontShape{T1}{lmtt}{m}{it}
      {<->sub*lmtt/m/sl}{}
-\iftoggle{full}{
+\iftoggle{full}{%
 \lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}
-}{
+}{%
 \lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.839}\ttfamily\small, columns=flexible, showstringspaces=false}
-}
+}%
 
@@
 % \cbend
-
-\acmConference[IFL 2016]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
-\acmYear{2016}
-\copyrightyear{2017}
 
@@
 \copyrightyear{2016}
-\setcopyright{acmlicensed}
 \acmConference[IFL 2016]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
